### 2014-05-29 tex/manuscript.tex
files moved

\documentclass{bio}
%\documentclass[oupdraft]{bio}
\usepackage[colorlinks=true, urlcolor=citecolor, linkcolor=citecolor, citecolor=citecolor]{hyperref}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}

% Add history information for the article if required
%\history{}

%% NOTES TO SELF:
% Cite using:
%    \citet{...}  in text
%    \citealp{...}   within parantheses
%
% Temporary notifications and reminders:
\newcommand\NB[1]{{\color{red}#1}}
\newcommand\NOTE[1]{{\color{red}[#1]}}
\newcommand\CITE[1]{{\color{magenta}[#1]}}

\begin{document}

% Title of paper
\title{Methods that remove batch effects while retaining group differences may lead
to exaggerated confidence in downstream analyses
}

% List of authors, with corresponding author marked by asterisk
\author{VEGARD NYGAARD, EINAR ANDREAS RØDLAND, EIVIND HOVIG$^\ast$\\[4pt]
% Author addresses
\textit{Department of Tumor Biology,
Institute for Cancer Research,
Oslo University Hospital HF - Radiumhospitalet,
Montebello,
0310 Oslo,
Norway}
\\[2pt]
% E-mail address for correspondence
{ehovig@radium.uio.no}}

% Running headers of paper:
\markboth%
% First field is the short list of authors
{V. Nygaard, E. A. Rødland, E. Hovig}
% Second field is the short title of the paper
{Exaggerated effects after batch correction}

\maketitle

% Add a footnote for the corresponding author if one has been
% identified in the author list
\footnotetext{To whom correspondence should be addressed.}

\begin{abstract}
{Removal of, or adjustment for, batch effects or centre differences is generally required when such effects are present in data. In particular, when preparing microarray gene expression data from multiple cohorts, array platforms, or batches for later analyses, batch effects can have confounding effects. Many methods and tools exist for this purpose. One method, ComBat which is part of the R package sva, is particularly popular due to its ability to remove batch differences even when batches are small and heterogeneous. It also has the option of preserving the difference between study groups, estimated from a two-way ANOVA model, to avoid conflating batch effects and group differences during batch adjustments. Unfortunately, this frequently used and recommended approach may systematically induce incorrect group differences in downstream analyses when groups are distributed between the batches in an unbalanced manner. The scientific community seems to be largely unaware of this problem, which most likely has contributed to false discoveries being presented in the published literature.
}
\end{abstract}

\section{Introduction}\label{intro}

Extraneous variables, if left unaccounted for, have the potential to lead an investigator into drawing wrong conclusions. In molecular biology, extraneous variables are often called "batch effects", probably due to the fact that reagents and other equipment, for instance microarray chips, are made in batches, and this is frequently observed as an effect in the measurements. See \citet{Luo2010} for more examples.

For a typical experiment comparing group differences, the presence of batch effects will decrease the statistical power since it adds variation to the data. If the batch--group design is unbalanced, i.e if the study groups are not equally represented in all batches, batch effects may act as a confounder and induce false differences between groups (\citealp{Leek2010}).

The standard way to handle an extraneous variable is to include it in the statistical model employed in the inquiry. However, many analysis tools for high throughput data do not cater for this option, and when available it could still be outside the competence of the investigator. Therefore, an alternative two step procedure has emerged. First the batch effects are estimated and removed, creating a ``batch effect free'' data set. In the next step, the statistical analyses are performed on the adjusted data without further consideration of batch effects. This appealing compartmentalization is also convenient for practical purposes, for example when data-processing and statistical analyses are performed by different personnel.

The first step can be achieved by subtracting the mean of the measurements in one batch from all measurements in that batch, i.e mean adjustment or one-way ANOVA adjustment as implemented in the method \texttt{pamr.batchadjust} from the \texttt{pamr} package in R. When the batch--group design is balanced, mean-adjustment will remove most, but not necessarily all, variance attributed to batch and leave the between group variance, thus increasing the statistical power. However, when the batch--group design is unbalanced, batch differences will in part be influences by group differences, and thus batch correction will reduce group differences and thereby reduce the statistical power. In very uneven group--batch designs with multiple groups, spurious group differences may even be induced in this way. Figure~\ref{fig:boxplots} illustrates both these effects.

To mitigate the above problems, one may simultaneously estimate batch effects and group differences, e.g. using a two-way ANOVA, and only remove the batch differences from the data. Effectively, this means group differences are estimated based on within batch comparison, and applied to the batch adjusted data. In a balanced group--batch design, group differences and batch effects are independent, and this approach becomes identical to the above described zero-centering per batch. If the group--batch design is heavily unbalanced, estimation of group differences and batch effects are interdependent. Unfortunately, when group differences estimated from an unbalanced group--batch design are applied to the entire data set, if this batch adjusted data set is later analysed for group differences, the confidence will be exaggerated as the estimation of the group differences in the unbalanced design is less accurate than in the corresponding balanced design. Figure~\ref{fig:boxplots} illustrates how statistical uncertainties are deflated by this batch adjustment method by comparing them to the uncertainties from the original ANOVA.

The ComBat method described in \citet{Johnson2007}, and included in the sva package (\citealp{Leek2012}), can use either of the two above described approaches to estimate batch differences, but uses an empirical Bayes approach to avoid over-correction for batch effects for small batches. It has popularised the two-way ANOVA procedure for retaining group differences when adjusting for batch effects, and the inclusion of group difference as a covariate when removing batch effects has been recommended in the sva tutorial and user fora. Based on actual use of ComBat by the authors and others, ourselves included, we suspect thus adjusted data are commonly treated as "batch effect free" in subsequent analyses. And as a consequence, confidence in group effects has been overestimated and false results reported.

\NB{Kort oppsummere behovet for batch-korrigering: generelt og i microarraydata.}

\NB{Kort beskrive metoder som gjoer dette (batch-sentrering, ANOVA, ComBat).}

\NB{Forklare tilsiktet effekt disse paa batch+gruppe-forskjeller (balansert versus ikke-balansert).}

\NB{Forklare kort hva faktisk effekt er: ANOVA gir redusert gruppe-forskjeller, mens for ComBat overestimeres den statistiske styrken/sikkerheten av forskjellen.}

\section{Methods for batch effect correction}

\subsection{Model for data with batch effects}

We will base our discussion of data with batch effects on a simple model:
\begin{equation}
Y_{ijr}=\alpha+\beta_j+\gamma_i+\epsilon_{ijr}
\end{equation}
where $i=1,\ldots,m$ are the different batches, $j=1,\ldots,M$ are different study groups that we wish to compare, and $r=1,\ldots,n_{ij}$ are the different samples within batch $i$ and group $j$.

When combining data from more diverse data sources, e.g. microarray data from different platforms, a more general model is required. One such model, used by \citet{Johnson2007}, is
\begin{equation}
Y_{ijgr}=\alpha_g+X_r\beta_{g}+\gamma_{ig}+\delta_{ig}\epsilon_{ijgr}
\end{equation}
where $g=1,\ldots,G$ are different measurements, e.g. genes, performed for each sample, and $X$ is the design matrix which in our case will indicate the study group. This permits independent rescaling of data from different batches. In addition, \citet{Johnson2007} uses an empirical Bayes approach to estimate $\gamma_{ig}$ and $\delta_{ig}$ to stabilise estimates, which is critical for use with small batches.

For simplicity, we consider the case with a single gene and constant scale, i.e. $\delta_{ig}=1$. We will discuss the effect of empirical Bayes estimation of $\gamma_{ig}$ later, but out main argument is more easily made in the simpler context.

\subsection{Standard batch correction methods}

The main ambition of batch effect adjustments is to be able to remove batch differences in such a way that downstream analyses of the adjusted data may be done without further batch adjustments.

The most common method for removing batch effects is to zero-centre each batch:
\begin{equation}
\Delta\widetilde Y_{ijr}=Y_{ijr}-\bar{Y}_{i}
\quad\text{where}\quad
\bar{Y}_i=\frac{\sum_{j=1}^M\sum_{r=1}^{n_{ij}} Y_{ijr}}{\sum_{j=1}^M n_{ij}}.
\end{equation}
An alternative is to centre each batch to the common average by adding the average value $\bar{Y}$ across the entire data set: i.e. $\widetilde Y_{ijr}=\Delta\widetilde Y_{ijr}+\bar{Y}$. When comparing groups, the common value $\bar{Y}$ has no effect, and so this is equilvalent to zero-centring each batch. However, if the different groups are unevenly represented in the different batches, the batch average $\bar{Y}_i$ will tend to capture group differences as well as batch effects. Thus, batch centering may reduce group differences, and thus reduce the power of downstream analyses.

Removing batch effects while retaining group differences can be done through an ANOVA analysis in which the group effects, $\beta_j$, and the batch effects, $\gamma_i$, are estimated simultaneously. Batch adjusted values may then be obtained by subtracting the estimated batch effects, $\hat\beta_j$:
\begin{equation}
\widetilde Y_{ijr}=Y_{ijr}-\hat\beta_j=\alpha+(\beta_j-\hat\beta_j)+\gamma_i+\epsilon_{ijr}.
\end{equation}
This will yield batch adjusted values where any systematic bias induced by the batch differences has been removed, while the group differences are retained.

Unfortunately, the estimation error $\hat\beta_j-\beta_j$ affects all values within the same batch in the same manner, inducing a dependency between the values. If the study groups are evenly represented in all batches, this will not influence estimated group differences as all groups are equally affected. However, if groups are unevenly represented, this induced dependency can have severe impact on downstream analyses.

\section{Results}

\subsection{A simple sanity check}

The undesired consequences of preserving group effects when correcting for batch effect is readily illustrated with a sanity check using random numbers . The documentation accompanying the sva library has a runnable example demonstrating how to adjust a data set with ComBat followed by a F-test. Swapping the real data with random numbers from a normal distribution (mean=0, sd=1), but otherwise following the instructions, will generate the p-value distribution shown in Figure 1. The skewed distribution is a indication that this approach may have a unintentional adverse effect. \NOTE{Jeg har ikke beskrevet permutatsjon checken fordi jeg er usikker paa om vi skal bruke den.}

%% FIGURE 1: Original position

\subsection{Explanation for the simple two-group comparison}

To explain more clearly what is happening, and quantify the size of the problem, we may consider the simple case of estimating the difference $\Delta\beta=\beta_A-\beta_B$ between two groups, $A$ and $B$, when there are $m$ batches with batch $i$ containing $n_{iA}$ and $n_{iB}$ samples from each of the two groups for $i=1,\ldots,m$.

If we estimate the group difference within batch $i$, we get
\begin{equation}
\widehat{\Delta\beta}_i=\bar{Y}_{iA}-\bar{Y}_{iB}\sim\text{N}\left(\Delta\beta,\frac{\sigma^2}{\nu_i}\right)
\quad\text{where}\quad
\bar{Y}_{ij}=\frac{\sum_{r=1}^{n_{ij}}Y_{ijr}}{n_{ij}},\,
\nu_i=\frac{1}{\frac{1}{n_{iA}}+\frac{1}{n_{iB}}},
\end{equation}
from which we may express the overall estimate of $\Delta\beta$
\begin{equation}
\widehat{\Delta\beta}=\frac{\sum_{i=1}^m \nu_i\,\widehat{\Delta\beta}_i}{\nu}
\sim\text{N}\left(\Delta\beta,\frac{\sigma^2}{\nu}\right)
\quad\text{where}\quad
\nu=\sum_{i=1}^m\nu_i.
\end{equation}

If batch and group effects are estimated using a two-way ANOVA, the estimate $\widehat{\Delta\beta}$ will be as stated above, and so the estimated group difference is unaffected. The batch effects are then removed, and the estimated group differences retained, leaving the estimated $\widehat{\Delta\beta}$ unchanged by the batch adjustment. However, if this batch adjusted data set is analysed without considering batch effects, the variance of $\widehat{\Delta\beta}$ will be computed under the assumption that it is derived from a comparison of $n_A=\sum_{i=1}^m n_{iA}$ versus $n_B=\sum_{i=1}^m n_{iB}$ samples, and thus satisfy
\begin{equation}
\widehat{\Delta\beta}
\sim\text{N}\left(\Delta\beta,\sigma^2\cdot\left(\frac{1}{n_A}+\frac{1}{n_B}\right)\right).
\end{equation}
Using Jensen's inequality, we may then derive
\begin{equation}
\frac{1}{\frac{1}{n_A}+\frac{1}{n_B}}=n\cdot\frac{n_A}{n}\cdot\frac{n_B}{n}
\le
\nu
=\sum_{i=1}^m \frac{1}{\frac{1}{n_{iA}}+\frac{1}{n_{iB}}}
=\sum_{i=1}^m n_i\cdot\frac{n_{iA}}{n_i}\cdot\frac{n_{iB}}{n_i}
\end{equation}
with equality if and only if the ratios $n_{iA}:n_{iB}=n_A:n_B$ for all batches $i=1,\ldots,m$.

\subsection{Examples of undesired consequences}

As the amount of false positive results when trying to retain group differences depends on the batch/group balance, we will show two examples with varying degree of unbalancedness.

1) In the first experiment(\citealp{Towfic2014}), cells were treated with glatiramer acetate (a medicine for multiple sclerosis) or a generic and mRNA was measured using microarrays alongside control samples. A batch effect correlating to the chip (Illumina WG-6\_V2, six samples per chip) was observed and adjusted for with ComBat, whereafter the data was tested for differetially expressed genes, yielding a list of 1000 genes (Table S5, \citealp{Towfic2014}). Unfortunately the batch/treatment design was unbalanced with several batches having only one of the main treatments of interest. When we re-analyzed their data without using ComBat, but instead blocked for batch effect in limma, only 9 genes were found ($\text{FDR}<0.05$). Additional sanity checks with random numbers or perumted labels were also carried out and the distribution of p-values for differents settings are shown in Fig. 2.  Our conclusion is that most of the genes reported as differtially expresssed in (\citealp{Towfic2014}) are false positives. This example is a sort of "worst case" scenario for applying ComBat, since it both has a very unbalanced batch/group design and a a priori assumpotion of no difference. The R-code for our analysis and a more extensive report can be downloaded from github (\href{https://github.com/vegardny/combat_tests.git}).

%% FIGURE 2: Original position

2) The second example is taken from the supporting information for the original ComBat article (\citealp{Johnson2007}) where it is denoted "Data Set 2". Cells inhibited for the expression of the TAL1 gene were compared to controls on a microarray platform. The experiment was conducted on three different time points (used as batches) with a total of 30 samples and a fairly balanced batch/treatment set up (6-2, 3-4 and 9-6). ComBat was applied followed by a t-test in order to identify differentially expressed genes. First, we reproduced their analysis including the adjustment by ComBat, but using limma instead of the t-test, resulting in 1003 probes ($q<0.05$).  Then, we analysed their data without batch adjustment in ComBat, but blocking for batch in limma, resulting in 377 probes ($q<0.05$). In addition, the two sanity checks outlined above were performed. In contrast to the above results obtained for(\citealp{Towfic2014}), the P-value distributions for the alternative analyses (Fig.3 a,b) does not indicate a huge difference. Nevertheless, we belive that  P-values are deflated for the ComBat adjusted analysis. The R-code for our analysis and a more extensive report can be downloaded from github (\href{https://github.com/vegardny/combat_tests.git}).

%% FIGURE 3: Original position

%% FIGURE 4: Original position

3) \NOTE{Her kan vi fylle paa med flere eksempler etter samme mal hvis det trengs. Feks en fersk artikkel fra Nature Genetic (\citealp{Yoo2014}). Der har de kombinert sekvenseringsdata  (deres data) fra 9 lymfom pasienter med et (eksternt) microarray data set som har 6 lymfom proever og 5 kontroller, dvs kontroll gruppen er helt fravaerende hos dem. Dvs veldig ubalansert (9/0, 6/5), MEN skille mellom phenotypene er saa stor at de sikkert hadde funnet omtrent de samme genene signifikante med aa bare bruke de balanserte dataene (men det var jo ikke deres.) Dette er ikke en viktig del av paperet, men det er veldig feil selvom konsekvensense ikke blir store. Dog gir det et falskt inntrykk av at dataene deres ligner tidligere data, se heatmap i suppl.. Dessuten lager de GO analyser av dette som egentlig er 7 aar gamle andres data og ikke deres}

\section{Discussion}

\subsection{Increased emphasis on preserving group difference}

In the original ComBat article (\citealp{Johnson2007}) it is clear that the primary motivation behind ComBat was to employ an Emprical Bayes method for batch-effect removal. The feature of retaining group differences for unbalanced designs is optional and seems to be subordinate, only exemplified in the supplementary information. However, over the years this feature became more important judging from advice given by the author to users
(
[1](\href{https://groups.google.com/d/msg/combat-user-forum/eSVSKwGtuyE/ZIWV2juYmmAJ})
,[2](\href{https://groups.google.com/d/msg/combat-user-forum/Vkb9p7wekd4/h5Etie7FTVQJ})
,[3](\href{https://groups.google.com/d/msg/combat-user-forum/fpBTcgDjiR8/uo4QIZL4sZgJ})
,[4](\href{https://groups.google.com/d/msg/combat-user-forum/26FZlgU2LFQ/W6U_Lhh_64EJ})
). \NOTE{Aa linke direkte er kanskje litt ufint, alternativt kan det linkes til forumet uten konkrete innlegg.}  
And when ComBat was incorporated in the sva package (\citealp{Leek2012}), inclusion of group labels was made almost mandatory (an undocumented option of passing a NULL value exists). Thus, this problematic use has likely been common.

\subsection{Motivation for this warning}

Our knowledge of ComBat came through a typical use case when trying to salvage unbalanced data which had batch effects. Upon realizing that the confidence on our group differences were exaggerated, the literature was searched for a better understanding of correct use and potential overseen limitations of ComBat. But the authors of ComBat and the sva package recommended our usage (\citealp{Johnson2007}, \citealp{Leek2012}). In addition, other works looking into the problem of batch effects were mostly recommending ComBat without much concern (\citealp{Kupfer2012}, \citealp{Kitchen2011}). A brief inquiry into  some of the articles citing ComBat (574 Google Scholar) revealed few problems, and their method descriptions regarding ComBat were mostly sparse, limited to one or two sentences (\CITE{24391845}, \CITE{18414638}, \CITE{21731603}, \CITE{23630272}, \CITE{24584070}). A further indication of their carefree use of this potentially devastating procedure was the frequent neglect to state the program parameters, i.e batch labels (\CITE{18414638}, \CITE{21731603}) or group labels (\CITE{24391845}, \CITE{18414638}, \CITE{21731603}, \CITE{23630272}, \CITE{23482648}, \CITE{24584070}). Often no effort was done in order to substantiate the existence of batch effects in their data, except for stating the presence of batches (\CITE{24391845}?, \CITE{18414638}, \CITE{21731603}, \CITE{23630272}). In one instance ComBat was even applied on data were effects due to batch were investigated, but not found \NOTE{husker ikke ref}. The incorporation of the method into analysis pipelines (\CITE{16642009}, TCGA) and other packages (\CITE{23452776}, \CITE{21937664}) could make its usage even more trivial and parameters setting harder to perceive.
Taken together we fear that many published results from data adjusted by ComBat are completely or partially false. And knowing that scientists don't give up where one analysis fails, a method that almost ensures a result given a sufficiently unbalanced design will continue to be used.

\subsection{Practical advice}

We have shown that adjusting for batch effects while preserve the group difference may lead to varying degree of false results. Knowing this, to what degree can an investigator trust a result from a work applying such a method? Essentially, when the batch/group configuration is balanced, or group difference is ignored  (i.e. no group labels not given as parameters to ComBat), problems related to preserving group differences will not occur. For other cases, a re-analysis without using this approach is the most rigorous path. However, this thoroughness is not feasible if the downstream  analysis can not adjust for batch effects by it self. To reach a reliable result, batch effects need to be handled in some way or another. To make matters worse, a re-analysis relies on the availability of the raw data and a description of processing and analysis steps taken in the original work. Even when this is available, the necessary bioinformatic skills and work hours could still be in short supply. For such situations, a superficial assessment can be performed, taking special note of batches were groups of interest are near missing and how likely a group difference is. In essence asking if the balanced parts (effective sample size?) of the data has enough power to detect the presumed effects and if this is the case (\citealp{Johnson2007}, \CITE{24584070}), treat the results more like an ordered list with the most likely true positives on top while de-emphasizing the somewhat deflated p-values . In contrast, if biological knowledge suggest that a group effect is unlikely  (\CITE{24391845}, \CITE{18414638}?, \CITE{21731603}, \CITE{23630272}), an intermediate lack of batch/group balance  could lead to a mostly false result.

\section{Supplementary Material}
\label{supp}

Supplementary material is available online at
\href{http://biostatistics.oxfordjournals.org}%
{http://biostatistics.oxfordjournals.org}.

\section*{Reproducible research}
\NOTE{vi boer klare aa tilfredstille kravene i http://biostatistics.oxfordjournals.org/content/10/3/405.full}

\section*{Acknowledgments}

[...Acknowledgements...]
{\it Conflict of Interest}: None declared.

\bibliographystyle{biorefs}
\bibliography{D:/Projects/_Appl_/_TeX_/library}
%\bibliography{refs}

%\begin{figure}[!p]
%\centering\includegraphics{fig1.eps}
%\caption{...}
%\label{Fig1}
%\end{figure}

\begin{figure}[!p]
\centering\includegraphics[width=13cm]{Fig/boxplots.png}
\caption{Simulated data from four study groups was generated where groups 1 and 2 have lower means than groups 3 and 4. These were placed in three different batches with batch effect added. Values and boxes showing mean and two standard errors of the mean are displayed for data without batch effects, after adding batch effects, after batch centering, and after ANOVA based batch centering. The last frame shows the least squares estimates of the group means from a two-way ANOVA analysis with 2 standard errors. This case, design and effects, was selected to illustrate the spurious effects that may arise.}
\label{fig:boxplots}
\end{figure}

\begin{figure}[!p]
\centering\includegraphics[width=13cm]{Fig/pvaluesleeksva.png}
\caption{
Two sanity checks were the recommended use of ComBat fails. Adapted from the user guide in the sva package. 
a) Real data is substituted with random numbers from a normal distribution (mean=0, sd=1), but the batch/group design is retained, followed by batch adjustment in ComBat and a F-test.
b) 10 runs of real data with the "cancer" labels permuted within batches, followed by ComBat adjustment and a F-test. All permutations produces a skewed p-value distribution.
}
\label{fig:p-sva}
\end{figure}

\begin{figure}[!p]
\centering\includegraphics[width=13cm]{Fig/pvaluestowfic.png}
\caption{
P-value distribution for the main comparison in (data set 2 24421904) glatiramer acetate vs. generic.  
a) Real data adjusted by ComBat as described in the article followed by a significance test in limma. 
b) Real data not adjusted by ComBat, tested for significance in limma using batch as a blocking factor. 
c) Real data is substituted with random numbers from a normal distribution (mean=0, sd=1), but the batch/group design is retained, followed by batch adjustment in ComBat and a significance test in limma.
d) 10 runs of real data with the glatiramer acetate("DP") and generic("N") labels permuted within batches, followed by ComBat adjustment and significance test in limma.  
\NOTE{jeg tok med begge sanitycheckene her. Er d) slik einar foreslo?. Merk jeg har ikke permutert alle labeler, bare de mellom DP og N. Det er en del andre.}
}
\label{p-towfic}
\end{figure}

\begin{figure}[!p]
\centering\includegraphics[width=13cm]{Fig/pvaluesjohnson.png}
\caption{
P-value distribution for the main comparison in (data set 2, \citealp{Johnson2007}) for  
a) Real data adjusted by ComBat as described in the article followed by a significance test in limma. 
b) Real data not adjusted by ComBat, tested for significance in limma using batch as a blocking factor. 
c) Real data is substituted with random numbers from a normal distribution (mean=0, sd=1), but the batch/group design is retained, followed by batch adjustment in ComBat and a significance test in limma.
d) 10 runs of real data with the group labels permuted within batches, followed by ComBat adjustment and significance test in limma.   
\NOTE{Problemet med permutatsjons sanitychechen slik jeg har utfoert og plottet den her er at den neppe kan sies aa feile. Dessuten kan de 3 foerst plottene bli plottet i ett som vist under i en alternativ version jeg synes er finere, men dette gaar ikke for permutasjonstesten tror jeg}
}
\label{fig:p-johnson}
\end{figure}

\begin{figure}[!p]
\centering\includegraphics[width=13cm]{Fig/pvaluesjohnson2.png}
\caption{
\NOTE{Alternativ illustrasjon av de tre p-value distribusjonene}
}
\label{fig:pdist-johnson}
\end{figure}

%\begin{table}[!p]
%\tblcaption{... \protect\citet{...} ... \label{Table1}}
%{\tabcolsep=4.25pt
%\begin{tabular}{@{}cccccccccc@{}}
%\tblhead{Dis & Chr & Marker & MLE-OR & MLE & MLE & EB-HM-OR & EB-HM & EB-HM & ES \\
%&&& (SE) & \textit{P}-value & rank & (SE) & \textit{P}-value & rank & rank}
%CAD & 2 & rs2943634 & 1.22(0.04) & $1.23 \times 10^{-5}$ & 32 & 1.02(0.01) & 0.19 & 28 & 37 \lastline
%\end{tabular}}
%\end{table}

\end{document}

### 2014-05-29 boxplotscript.r

@@ -1,0 +1,141 @@
+
+
+
+
+
+library(sva)
+
+ 
+adhocboxplot = function(y, grouplabels, batchlabels, addlegend=FALSE, ylim=NULL, xlab="x")
+{
+  #x = batchlabels + grouplabels/8
+  colpalette = c("blue", "red", "darkgreen", "magenta")
+  pchselection = c(1, 2, 3)
+  x = batchlabels
+  lwd=2
+  
+  # add extra x for better visual separation
+  for(b in unique(batchlabels))
+  {
+    
+    separation = 0.15
+    xoffset=0    
+    for(g in unique(grouplabels[batchlabels==b]))
+    {
+      x[batchlabels==b & grouplabels==g] = x[batchlabels==b & grouplabels==g] + xoffset
+      xoffset = xoffset+separation
+    }
+  }
+  
+  xlim = c(0, length(unique(batchlabels))+2)
+  plot(x, y, col=colpalette[grouplabels], pch=pchselection[batchlabels], 
+       xlim=xlim, lwd=lwd, cex=2, ylim=ylim, xlab=xlab, ylab="", axes=FALSE, cex.lab =2)
+  
+  #add boxplot
+  groupnames = unique(grouplabels)
+  groupnames = groupnames[order(groupnames)]
+  xoffset=length(unique(batchlabels))+1
+  boxwidth=0.35
+  boxseparation = 0.25
+  for(g in groupnames)
+  { 
+    #boxplot(y[grouplabels==g], notch=TRUE, border=colpalette[g], 
+    #        add=TRUE, at=xoffset, col=rgb(1,0.5,0.3, alpha=0.0),
+    #        outline=FALSE, boxwex=0.5)
+    m = mean(y[grouplabels==g])
+    ci = t.test(y[grouplabels==g])[["conf.int"]]
+    xleft = xoffset
+    ybottom = ci[1]
+    xright = xleft + boxwidth
+    ytop = ci[2]
+    rect(xleft, ybottom, xright, ytop, border=colpalette[g], lwd=lwd)
+    lines(c(xoffset,xright), c(m,m), col=colpalette[g], lwd=lwd)
+    xoffset = xoffset+boxseparation
+  }
+  
+  if(addlegend)
+  {
+    legendcex=2
+    legend("topright", legend=paste("Group ", groupnames, sep=""), 
+           text.col=colpalette[groupnames], bty="n", cex=legendcex)
+    legend("bottomright", legend=paste("Batch ", unique(batchlabels), sep=""),
[…]

### 2014-05-29 figurescript.r

@@ -1,0 +1,32 @@
+
+
+
+
+
+startwd = getwd()
+rseed=3
+for(i in 6:6)
+{
+  rseed=i
+  source("boxplotscript.r")
+}
+
+
+setwd("reanalysis/Leek2012")
+source("figurescript_leek.r")
+setwd(startwd)
+
+setwd("reanalysis/Towfic2014")
+source("figurescript_towfic.r")
+setwd(startwd)
+
+setwd("reanalysis/Johnson2007")
+source("figurescript_johnson.r")
+setwd(startwd)
+
+
+
+
+
+
+

### 2014-05-29 reanalysis/Johnson2007/figurescript_johnson.r

@@ -1,0 +1,75 @@
+
+# source("figurescriptjohnson.r")
+starttime = Sys.time()
+
+library("sva")
+library("limma")
+source("../../commonscripts/helperfunctions.r")
+source("helperfunctions_johnson.r")
+
+tmp = loadjohnsondata()
+sampleannotation = tmp[["sampleannotation"]]
+normdata = tmp[["normdata"]]
+
+combatdata = as.matrix( ComBat(
+  dat=normdata, 
+  batch=sampleannotation$Batch, 
+  mod=model.matrix(~as.factor(sampleannotation$"Type")), 
+  numCovs=NULL, par.prior=TRUE, prior.plots=FALSE))
+
+# DE pvalues ComBat adjusted real data
+Type = as.factor(sampleannotation$Type)
+design = model.matrix(~0 + Type)
+fit = lmFit(combatdata, design)
+cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design)  
+fit2 = contrasts.fit(fit, cont.matrix)
+combatp = eBayes(fit2)$p.value[,1]
+rm(Type, design, fit, cont.matrix, fit2)
+print(paste("ComBat adjusted real data, significant probes: ",  sum(p.adjust(combatp, "fdr")<0.05)))
+
+# DE pvalues not ComBat adjusted, 
+# batch blocked in limma
+Type = as.factor(sampleannotation$Type)
+Block = as.factor(sampleannotation$Batch)
+design = model.matrix(~0+Type+Block)
+fit = lmFit(normdata, design)
+cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design)  
+fit2 = contrasts.fit(fit, cont.matrix)
+limmap = eBayes(fit2)$p.value[,1]
+rm(Type, Block, design, fit, cont.matrix, fit2)
+print(paste("Limma batch adjusted real data, significant probes: ",  sum(p.adjust(limmap, "fdr")<0.05)))
+
+#Random data
+set.seed(100)
+randdata = normdata
+randdata[,] =rnorm(length(randdata), mean=0, sd=1)
+randcombatdata = as.matrix(ComBat(
+  dat=randdata, 
+  batch=sampleannotation$Batch, 
+  mod=model.matrix(~as.factor(sampleannotation$Type)), 
+  numCovs=NULL, par.prior=TRUE,  prior.plots=FALSE))
+
+Type = as.factor(sampleannotation$Type)
[…]

### 2014-05-29 reanalysis/Leek2012/figurescript_leek.r

@@ -1,0 +1,33 @@
+
+
+library(sva)
+library(bladderbatch)
+data(bladderdata)
+library(limma)
+
+pheno = pData(bladderEset)
+edata = exprs(bladderEset)
+
+# Substitute the data with random numbers from a normal distribution.
+set.seed(100)
+edata[,] = rnorm(length(edata), mean=0, sd=1)
+
+
+mod0 = model.matrix(~1,data=pheno)
+
+batch = pheno$batch
+
+mod = model.matrix(~as.factor(cancer), data=pheno)
+
+combat_edata = ComBat(dat=edata, batch=batch, 
+                      mod=mod, numCovs=NULL, 
+                      par.prior=TRUE, prior.plots=FALSE)
+
+pValuesComBat = f.pvalue(combat_edata,mod,mod0)
+qValuesComBat = p.adjust(pValuesComBat,method="BH")
+
+figfile = paste(getwd(), "/leekrandomdatapvalues.pdf", sep="")
+pdf(file=figfile)
+hist(pValuesComBat,  main="P-values, Random numbers", breaks=100, xlab="p-value")
+dev.off()
+print( paste("Figure created; ",figfile ))

### 2014-05-29 reanalysis/Towfic2014/figurescript_towfic.r

@@ -1,0 +1,80 @@
+
+# source("figuretowfic.r")
+starttime = Sys.time()
+debug = FALSE
+downloaddata=TRUE
+set.seed(100)
+
+
+includelibs = c("Biobase", "GEOquery", "sva", "limma")
+lapply(includelibs, require, character.only=T)
+source("../../commonscripts/helperfunctions.r")
+source("helperfunctions_towfic.r")
+
+ret = loadtowfic(downloaddata)
+sampleannotation = ret[["sampleannotation"]]
+rawdata = ret[["data"]]
+
+if(debug)
+  rawdata = rawdata[1:1000,]
+
+qnormdata = normalizeBetweenArrays(rawdata, method="quantile") 
+
+# combat adjust
+combatdata= as.matrix(ComBat(dat=qnormdata,
+                             batch=sampleannotation$chip,
+                             mod=model.matrix(~as.factor(sampleannotation$covariate)),
+                             numCovs=NULL, par.prior=TRUE, prior.plots=FALSE))
+
+# Significance test DP vs N
+group = factor(sampleannotation$covariate)
+design = model.matrix(~0 + group)
+fit = lmFit(combatdata, design)
+cont.matrix = makeContrasts ( contrasts="groupDP-groupN", levels=design)  
+fit2 = contrasts.fit(fit, cont.matrix)
+limma_p_alt = eBayes(fit2)$p.value[,1]
+print(paste("ComBat adjusted real data, significant probes: ",
+            sum(p.adjust(limma_p_alt, method="fdr")<0.05)))
+
+#Limma blocked batch and significance test
+group = factor(sampleannotation$covariate)
+block = factor(sampleannotation$chip)
+design = model.matrix(~0+group+block)
+fit = lmFit(qnormdata, design)
+cont.matrix = makeContrasts ( contrasts="groupDP-groupN", levels=design)  
+fit2 = contrasts.fit(fit, cont.matrix)
+limma_ret_woc = eBayes(fit2)
+limma_p_woc = limma_ret_woc$p.value[,1]
+print(paste("Limma adjusted real data, significant probes: ",  
+            sum(p.adjust(limma_p_woc, method="fdr")<0.05)))
+
+# radnom, ComBat adjusted
+set.seed(100)
+randdata = rawdata
+randdata[,] = matrix(rnorm(length(randdata), mean=0, sd=1))
+randcombatdata= as.matrix(ComBat(dat=randdata,
+                                 batch=sampleannotation$chip,
[…]

### 2014-06-05 figurescript.r
figurescript as r

@@ -1,16 +1,13 @@
 
 
 
-
-
 startwd = getwd()
 rseed=3
 for(i in 6:6)
 {
   rseed=i
-  source("boxplotscript.r")
+  source("overviewplot.r")
 }
-
 
 setwd("reanalysis/Leek2012")
 source("figurescript_leek.r")
@@ -24,7 +21,9 @@ setwd("reanalysis/Johnson2007")
 source("figurescript_johnson.r")
 setwd(startwd)
 
-
+sink("sessionInfo.txt")
+sessionInfo()
+sink()
 
 
 

### 2014-06-11 commonscripts/helperfunctions.r
fiksing

@@ -1,16 +1,16 @@
 
+# Creates a sampleannotation data.frame based on a list of batches with a vector of treatmentcounts.
+# Used when creating artificial data.
 createsampleannotation = function( treatment_in_batches)
 {
 	batches = vector()
 	treatments = vector()
 	for(i in 1:length(treatment_in_batches))
 	{
-		#thisbatch = paste("batch", i, sep="")
 		thisbatch = i
-		for(s in 1:length(treatment_in_batches[[i]]))
+		for(s in 1:length(treatment_in_batches[[i]]))  
 		{
 			batches = c(batches, rep(thisbatch, treatment_in_batches[[i]][s]))
-			#treatments = c(treatments, rep(paste("treatment", s, sep=""), treatment_in_batches[[i]][s]))
 			treatments = c(treatments, rep( s, treatment_in_batches[[i]][s]))
 		}
 	}
@@ -19,279 +19,9 @@ createsampleannotation = function( treatment_in_batches)
 }
 
 
-
-
-createrandomdata = function(ngenes, sa, mean=0, sd=1)
-{
-	ret = matrix( rnorm(dim(sa)[1]*ngenes, mean=mean, sd=sd), ncol=dim(sa)[1], nrow=ngenes)
-	dimnames(ret) = list( paste("gene", 1:ngenes, sep=""),  sa$id)
-	return(ret)
-}
-
-addbatcheffect = function(df, batches, thismean=1, thissd=1)
-{
-	ret = df
-	for(i in 1:length(unique(batches)))
-	{
-		for(s in 1:dim(df)[1])
-		{
-			thisgenesbatcheffect = rnorm(1,mean=thismean, sd=thissd)
-			a=batches==i
-			#ret[s, a] = ret[s, a] + rnorm(length(ret[s, a]), mean=thisgenesbatcheffect)	
-			ret[s, a] = ret[s, a] + thisgenesbatcheffect
-		}
-	}
-	return(ret)
-}
-
-# 
-addconditioneffect = function(df, labels, ndiffgenes, thismean=1, betweensd=1, insidesd=1, affectedconditions)
-{
-	ret = df
-	if(ndiffgenes>0)
-	{
-		for(i in 1:length(affectedconditions))
-		{
-			for(s in 1:ndiffgenes)
-			{
-				thisgeneseffect = rnorm(1, thismean, sd=betweensd)
-				a=labels==affectedconditions[i]
[…]

### 2014-06-11 figurescript.r

@@ -2,12 +2,7 @@
 
 
 startwd = getwd()
-rseed=3
-for(i in 6:6)
-{
-  rseed=i
-  source("overviewplot.r")
-}
+source("overviewplot.r")
 
 setwd("reanalysis/Leek2012")
 source("figurescript_leek.r")

### 2014-06-11 overviewplot.r

@@ -74,7 +74,7 @@ table(sampleannotation[,2:3])
 
 ngenes=1000
 #matrix_random = createrandomdata(ngenes, sampleannotation, mean=0, sd=1)
-set.seed(rseed)
+set.seed(6)
 matrix_random = matrix(rnorm(ngenes * nrow(sampleannotation), mean=0, sd=1), nrow=ngenes, ncol=nrow(sampleannotation))
 
 matrix_condition = matrix_random
@@ -93,7 +93,8 @@ matrix_combat = ComBat(dat=matrix_conditionbatch, batch=sampleannotation$batch,
 
 
 index=1
-figfile = paste( getwd(), "/", "figure1_", rseed, ".pdf", sep="")
+#figfile = paste( getwd(), "/", "figure1_", rseed, ".pdf", sep="")
+figfile = paste( getwd(), "/", "figure1",  ".pdf", sep="")
 pdf(file =figfile, width=16, height=8)
 #png(file = paste("figure/", "figure1_", rseed, ".png", sep=""), width=1600, height=800)
 par(mfrow=c(1, 4))

### 2014-06-11 commonscripts/helperfunctions.r
no color

@@ -22,7 +22,7 @@ createsampleannotation = function( treatment_in_batches)
 # plots 3 histograms from p-value distibutions on top on each other.
 # draws a line instead of bars for clarity.
 # binsize is 0.01
-adhocpvalueplot = function(realcombatp, reallimmap, randomp, main="P-values", xrange=1:25)
+oldadhocpvalueplot = function(realcombatp, reallimmap, randomp, main="P-values", xrange=1:25)
 {
   thiscolors = c("red", "blue", "black")
   
@@ -49,4 +49,31 @@ adhocpvalueplot = function(realcombatp, reallimmap, randomp, main="P-values", xr
                   "Random data, ComBat adjusted")) 
 }
 
+
+# plots 3 histograms from p-value distibutions on top on each other.
+# draws a line instead of bars for clarity.
+# binsize is 0.01
+adhocpvalueplot = function(realcombatp, reallimmap, randomp, main="P-values", xrange=1:25)
+{
+
+	a = hist(realcombatp, breaks=100, plot=F)$counts[xrange]
+	b = hist(reallimmap, breaks=100, plot=F)$counts[xrange]
+	c = hist(randomp, breaks=100, plot=F)$counts[xrange]
+	ylim=c(0, max(c(a,b,c)))
+	# reproduced ComBat + limma
+	plot((xrange)/100, a , ylim=ylim,
+			 main=main, xlab="p-value", ylab="frequency",
+			 type="l", lwd=2, lty=1)
+	
+	# batch handled in limma
+	lines((xrange)/100, b, lwd=2, lty=2)
+	
+	# random ComBat + limma
+	lines((xrange)/100, c, lwd=2, lty=3)
+	
+	legend("topright", lwd=2, lty=1:3,
+				 legend=c("Real data, ComBat adjusted",
+				 				 "Real data, batch handled by Limma",
+				 				 "Random data, ComBat adjusted")) 
+}
 

### 2014-06-12 tex/manuscript_vegardedit.tex
mer manuscript

--- 2014-05-29 tex/manuscript.tex
@@
 1) In the first experiment(\citealp{Towfic2014}), cells were treated with glatiramer acetate (a medicine for multiple sclerosis) or a generic and mRNA was measured using microarrays alongside control samples.
-A batch effect correlating to the chip (Illumina WG-6\_V2, six samples per chip) was observed and adjusted for with ComBat, whereafter the data was tested for differetially expressed genes, yielding a list of 1000 genes (Table S5, \citealp{Towfic2014}).
+A batch effect correlating to the chip (Illumina WG-6\_V2, six samples per chip) was observed and adjusted for with ComBat, whereafter the data was tested for differentially expressed genes, yielding a list of 1000 genes (Table S5, \citealp{Towfic2014}).
 Unfortunately the batch/treatment design was unbalanced with several batches having only one of the main treatments of interest.
 When we re-analyzed their data without using ComBat, but instead blocked for batch effect in limma, only 9 genes were found ($\text{FDR}<0.05$).
-Additional sanity checks with random numbers or perumted labels were also carried out and the distribution of p-values for differents settings are shown in Fig.
-2.
-Our conclusion is that most of the genes reported as differtially expresssed in (\citealp{Towfic2014}) are false positives.
-This example is a sort of "worst case" scenario for applying ComBat, since it both has a very unbalanced batch/group design and a a priori assumpotion of no difference.
-The R-code for our analysis and a more extensive report can be downloaded from github (\href{https://github.com/vegardny/combat_tests.git}).
+The above outlined sanity check with random numbers were also performed.
+The distributions of p-values for different settings are shown in Fig.
+3a.
+Our conclusion is that most of the genes reported as differentially expressed in (\citealp{Towfic2014}) are false positives.
+This example is a sort of "worst case" scenario for applying ComBat, since it both has a very unbalanced batch/group design and a a priori assumption of no difference.
+The R-code for our analysis can be downloaded from Github (\href{https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-figures.git}).
 
@@
 Then, we analysed their data without batch adjustment in ComBat, but blocking for batch in limma, resulting in 377 probes ($q<0.05$).
-In addition, the two sanity checks outlined above were performed.
-In contrast to the above results obtained for(\citealp{Towfic2014}), the P-value distributions for the alternative analyses (Fig.3 a,b) does not indicate a huge difference.
+In addition, the sanity check outlined above were performed.
+The distributions of p-values for different settings are shown in Fig.
+3b.
+In contrast to the results obtained for \citealp{Towfic2014} (Fig.3b) , the P-value distributions for the alternative analyses  does not indicate a huge difference.
 Nevertheless, we belive that  P-values are deflated for the ComBat adjusted analysis.
-The R-code for our analysis and a more extensive report can be downloaded from github (\href{https://github.com/vegardny/combat_tests.git}).
+The R-code for our analysis can be downloaded from Github (\href{https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-figures.git}).
 
@@
 
-We have shown that adjusting for batch effects while preserve the group difference may lead to varying degree of false results.
-Knowing this, to what degree can an investigator trust a result from a work applying such a method?
+For an investigator facing an unbalanced data set with batch effects, our advice is to adjust for batch inside the statistical test and avoid the two step procedure outlined above.
+If this is not possible, then applying the method described in this article should only be done with great caution and downstream confidence estimates must be treated with suspicion.
+
+Knowing that adjusting for batch effects while preserving the group difference may lead to varying degree of false results, to what extent can an investigator trust a result from a work applying such a method?
 Essentially, when the batch/group configuration is balanced, or group difference is ignored  (i.e.
@@
 
-\section*{Reproducible research}
+\subsection{Reproducible research}
+
+The data and scripts used to generate the results in this work, i.e figures, are made available at https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-figures.git.
+Additional analyses were also performed, but considered excessive for the current  article.
+However, some readers might find this very interesting and we provide it as an extra resource at https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-reports.git
+
 \NOTE{vi boer klare aa tilfredstille kravene i http://biostatistics.oxfordjournals.org/content/10/3/405.full}
@@
 \begin{figure}[!p]
-\centering\includegraphics[width=13cm]{Fig/boxplots.png}
+\centering\includegraphics[width=13cm]{../figure1.pdf}
 \caption{Simulated data from four study groups was generated where groups 1 and 2 have lower means than groups 3 and 4.
@@
 \begin{figure}[!p]
-\centering\includegraphics[width=13cm]{Fig/pvaluesleeksva.png}
+\centering\includegraphics[width=13cm]{../reanalysis/Leek2012/leekrandomdatapvalues.pdf}
 \caption{
-Two sanity checks were the recommended use of ComBat fails. Adapted from the user guide in the sva package. 
-a) Real data is substituted with random numbers from a normal distribution (mean=0, sd=1), but the batch/group design is retained, followed by batch adjustment in ComBat and a F-test.
-b) 10 runs of real data with the "cancer" labels permuted within batches, followed by ComBat adjustment and a F-test. All permutations produces a skewed p-value distribution.
+A sanity checks were the recommended use of ComBat fails.
+Adapted from the user guide in the sva package.
+Real data are substituted with random numbers from a normal distribution, but the batch/group design is retained.
+ComBat is applied followed by a F-test.
+a) P-value distribution
+b) QQ plot of the F-statistics
 }
@@
 \begin{figure}[!p]
-\centering\includegraphics[width=13cm]{Fig/pvaluestowfic.png}
-\caption{
-P-value distribution for the main comparison in (data set 2 24421904) glatiramer acetate vs. generic.  
-a) Real data adjusted by ComBat as described in the article followed by a significance test in limma. 
-b) Real data not adjusted by ComBat, tested for significance in limma using batch as a blocking factor. 
-c) Real data is substituted with random numbers from a normal distribution (mean=0, sd=1), but the batch/group design is retained, followed by batch adjustment in ComBat and a significance test in limma.
-d) 10 runs of real data with the glatiramer acetate("DP") and generic("N") labels permuted within batches, followed by ComBat adjustment and significance test in limma.  
-\NOTE{jeg tok med begge sanitycheckene her. Er d) slik einar foreslo?. Merk jeg har ikke permutert alle labeler, bare de mellom DP og N. Det er en del andre.}
+\begin{center}$
+\begin{array}{cc}
+\includegraphics[width=6.5cm]{../reanalysis/Towfic2014/towficpvalues.pdf} &
+\includegraphics[width=6.5cm]{../reanalysis/Johnson2007/dataset2pvalues.pdf}
+\end{array}$
+\end{center}
+\caption
+{
+Three analyses of two works where batch effects were adjusted for with ComBat.
+First as described in the original works with ComBat on real data, secondly with ComBat but with random numbers instead of real data, thirdly without ComBat and instead blocking batch with limma on real data
+a) Re-analysis of  \citealp{Towfic2014}, glatiramer acetate vs. generic 
+b) Re-analysis of "Data set 2" \citealp{Johnson2007},  TAL1 inhibition vs. control
 }
-\label{p-towfic}
-\end{figure}
-
-\begin{figure}[!p]
-\centering\includegraphics[width=13cm]{Fig/pvaluesjohnson.png}
-\caption{
-P-value distribution for the main comparison in (data set 2, \citealp{Johnson2007}) for  
-a) Real data adjusted by ComBat as described in the article followed by a significance test in limma. 
-b) Real data not adjusted by ComBat, tested for significance in limma using batch as a blocking factor. 
-c) Real data is substituted with random numbers from a normal distribution (mean=0, sd=1), but the batch/group design is retained, followed by batch adjustment in ComBat and a significance test in limma.
-d) 10 runs of real data with the group labels permuted within batches, followed by ComBat adjustment and significance test in limma.   
-\NOTE{Problemet med permutatsjons sanitychechen slik jeg har utfoert og plottet den her er at den neppe kan sies aa feile.
-Dessuten kan de 3 foerst plottene bli plottet i ett som vist under i en alternativ version jeg synes er finere, men dette gaar ikke for permutasjonstesten tror jeg}
-}
-\label{fig:p-johnson}
-\end{figure}
-
-\begin{figure}[!p]
-\centering\includegraphics[width=13cm]{Fig/pvaluesjohnson2.png}
-\caption{
-\NOTE{Alternativ illustrasjon av de tre p-value distribusjonene}
-}
-\label{fig:pdist-johnson}
 \end{figure}

### 2014-06-16 tex/manuscript_vegardedit.tex
Mer manus og figurer

@@
 \maketitle
+
+\NOTE{Er det denne emailen du vil gi fra deg? Var det ikke snakk om aa pensjonere radium emailer for noen aar siden? Vil denne virke om to aar?}
 
@@
 And as a consequence, confidence in group effects has been overestimated and false results reported.
-
-\NB{Kort oppsummere behovet for batch-korrigering: generelt og i microarraydata.}
-
-\NB{Kort beskrive metoder som gjoer dette (batch-sentrering, ANOVA, ComBat).}
-
-\NB{Forklare tilsiktet effekt disse paa batch+gruppe-forskjeller (balansert versus ikke-balansert).}
-
-\NB{Forklare kort hva faktisk effekt er: ANOVA gir redusert gruppe-forskjeller, mens for ComBat overestimeres den statistiske styrken/sikkerheten av forskjellen.}
 
@@
 The skewed distribution is a indication that this approach may have a unintentional adverse effect.
-\NOTE{Jeg har ikke beskrevet permutatsjon checken fordi jeg er usikker paa om vi skal bruke den.}
 
@@
 
-As the amount of false positive results when trying to retain group differences depends on the batch/group balance, we will show two examples with varying degree of unbalancedness.
+As the amount of false positive results when trying to retain group differences depends on the batch-group balance, we will show two examples with varying degree of unbalancedness.
 
@@
 A batch effect correlating to the chip (Illumina WG-6\_V2, six samples per chip) was observed and adjusted for with ComBat, whereafter the data was tested for differentially expressed genes, yielding a list of 1000 genes (Table S5, \citealp{Towfic2014}).
-Unfortunately the batch/treatment design was unbalanced with several batches having only one of the main treatments of interest.
+Unfortunately the batch-treatment design was unbalanced with several batches having only one of the main treatments of interest.
 When we re-analyzed their data without using ComBat, but instead blocked for batch effect in limma, only 9 genes were found ($\text{FDR}<0.05$).
@@
 Our conclusion is that most of the genes reported as differentially expressed in (\citealp{Towfic2014}) are false positives.
-This example is a sort of "worst case" scenario for applying ComBat, since it both has a very unbalanced batch/group design and a a priori assumption of no difference.
+This example is a sort of "worst case" scenario for applying ComBat, since it both has a very unbalanced batch-group design and a a priori assumption of no difference.
 The R-code for our analysis can be downloaded from Github (\href{https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-figures.git}).
@@
 Cells inhibited for the expression of the TAL1 gene were compared to controls on a microarray platform.
-The experiment was conducted on three different time points (used as batches) with a total of 30 samples and a fairly balanced batch/treatment set up (6-2, 3-4 and 9-6).
+The experiment was conducted on three different time points (used as batches) with a total of 30 samples and a fairly balanced batch-treatment set up (6-2, 3-4 and 9-6).
 ComBat was applied followed by a t-test in order to identify differentially expressed genes.
@@
 
-3) \NOTE{Her kan vi fylle paa med flere eksempler etter samme mal hvis det trengs.
-Feks en fersk artikkel fra Nature Genetic (\citealp{Yoo2014}).
-Der har de kombinert sekvenseringsdata  (deres data) fra 9 lymfom pasienter med et (eksternt) microarray data set som har 6 lymfom proever og 5 kontroller, dvs kontroll gruppen er helt fravaerende hos dem.
-Dvs veldig ubalansert (9/0, 6/5), MEN skille mellom phenotypene er saa stor at de sikkert hadde funnet omtrent de samme genene signifikante med aa bare bruke de balanserte dataene (men det var jo ikke deres.) Dette er ikke en viktig del av paperet, men det er veldig feil selvom konsekvensense ikke blir store.
-Dog gir det et falskt inntrykk av at dataene deres ligner tidligere data, se heatmap i suppl..
-Dessuten lager de GO analyser av dette som egentlig er 7 aar gamle andres data og ikke deres}
-
 \section{Discussion}
@@
 The feature of retaining group differences for unbalanced designs is optional and seems to be subordinate, only exemplified in the supplementary information.
-However, over the years this feature became more important judging from advice given by the author to users
-(
-[1](\href{https://groups.google.com/d/msg/combat-user-forum/eSVSKwGtuyE/ZIWV2juYmmAJ})
-,[2](\href{https://groups.google.com/d/msg/combat-user-forum/Vkb9p7wekd4/h5Etie7FTVQJ})
-,[3](\href{https://groups.google.com/d/msg/combat-user-forum/fpBTcgDjiR8/uo4QIZL4sZgJ})
-,[4](\href{https://groups.google.com/d/msg/combat-user-forum/26FZlgU2LFQ/W6U_Lhh_64EJ})
-). \NOTE{Aa linke direkte er kanskje litt ufint, alternativt kan det linkes til forumet uten konkrete innlegg.}  
+However, over the years this feature became more important judging from advice given on user fora.
 And when ComBat was incorporated in the sva package (\citealp{Leek2012}), inclusion of group labels was made almost mandatory (an undocumented option of passing a NULL value exists).
@@
 
-Our knowledge of ComBat came through a typical use case when trying to salvage unbalanced data which had batch effects.
+Our knowledge of the problem discussed in this article came through a typical use case when trying to adjust for batch effects in an unbalanced data set using ComBat.
 Upon realizing that the confidence on our group differences were exaggerated, the literature was searched for a better understanding of correct use and potential overseen limitations of ComBat.
@@
 A brief inquiry into  some of the articles citing ComBat (574 Google Scholar) revealed few problems, and their method descriptions regarding ComBat were mostly sparse, limited to one or two sentences (\CITE{24391845}, \CITE{18414638}, \CITE{21731603}, \CITE{23630272}, \CITE{24584070}).
-A further indication of their carefree use of this potentially devastating procedure was the frequent neglect to state the program parameters, i.e batch labels (\CITE{18414638}, \CITE{21731603}) or group labels (\CITE{24391845}, \CITE{18414638}, \CITE{21731603}, \CITE{23630272}, \CITE{23482648}, \CITE{24584070}).
+A further indication of their carefree use of this potentially devastating procedure was the frequent omittance of program parameters that were used, i.e batch labels (\CITE{18414638}, \CITE{21731603}) or whether group labels were supplied as covariates(\CITE{24391845}, \CITE{18414638}, \CITE{21731603}, \CITE{23630272}, \CITE{23482648}, \CITE{24584070}).
 Often no effort was done in order to substantiate the existence of batch effects in their data, except for stating the presence of batches (\CITE{24391845}?, \CITE{18414638}, \CITE{21731603}, \CITE{23630272}).
-In one instance ComBat was even applied on data were effects due to batch were investigated, but not found \NOTE{husker ikke ref}.
 The incorporation of the method into analysis pipelines (\CITE{16642009}, TCGA) and other packages (\CITE{23452776}, \CITE{21937664}) could make its usage even more trivial and parameters setting harder to perceive.
 Taken together we fear that many published results from data adjusted by ComBat are completely or partially false.
-And knowing that scientists don't give up where one analysis fails, a method that almost ensures a result given a sufficiently unbalanced design will continue to be used.
+We hope that our warning will enlighten the community and  reduce this unfortunate combination of methods.
 
@@
 Knowing that adjusting for batch effects while preserving the group difference may lead to varying degree of false results, to what extent can an investigator trust a result from a work applying such a method?
-Essentially, when the batch/group configuration is balanced, or group difference is ignored  (i.e.
+Essentially, when the batch-group configuration is balanced, or group difference is ignored  (i.e.
 no group labels not given as parameters to ComBat), problems related to preserving group differences will not occur.
@@
 In essence asking if the balanced parts (effective sample size?) of the data has enough power to detect the presumed effects and if this is the case (\citealp{Johnson2007}, \CITE{24584070}), treat the results more like an ordered list with the most likely true positives on top while de-emphasizing the somewhat deflated p-values .
-In contrast, if biological knowledge suggest that a group effect is unlikely  (\CITE{24391845}, \CITE{18414638}?, \CITE{21731603}, \CITE{23630272}), an intermediate lack of batch/group balance  could lead to a mostly false result.
+In contrast, if biological knowledge suggest that a group effect is unlikely  (\CITE{24391845}, \CITE{18414638}?, \CITE{21731603}, \CITE{23630272}), an intermediate lack of batch-group balance  could lead to a mostly false result.
 
@@
 \NOTE{vi boer klare aa tilfredstille kravene i http://biostatistics.oxfordjournals.org/content/10/3/405.full}
+\NOTE{For aa tilfredstille kravene skal data og script vare available.
+Jeg mener vi klarer dette.
+paa githuben er ikke data fra towfic, men det blir lastet av scriptet fra GEO.
+Ikke beskrevet her men i scriptet.
+Videre i forklaringen til biostatistics staar det at script og data skal bli lagt paa deres sider, som vi kan hvis de vil}
 
@@
 \begin{figure}[!p]
-\centering\includegraphics[width=13cm]{../reanalysis/Leek2012/leekrandomdatapvalues.pdf}
-\caption{
-A sanity checks were the recommended use of ComBat fails.
+\begin{center}$
+\begin{array}{cc}
+\includegraphics[width=6.5cm]{../reanalysis/Leek2012/leekrandomdatapvalues.pdf} &
+\includegraphics[width=6.5cm]{../reanalysis/Leek2012/leekqqplot.png}
+\end{array}$
+\end{center}
+\caption
+{
+A sanity checks where the recommended use of ComBat fails.
 Adapted from the user guide in the sva package.
-Real data are substituted with random numbers from a normal distribution, but the batch/group design is retained.
+Real data are substituted with random numbers from a normal distribution, but the batch-group design is retained.
 ComBat is applied followed by a F-test.
@@
 }
-\label{fig:p-sva}
 \end{figure}
@@
 Three analyses of two works where batch effects were adjusted for with ComBat.
-First as described in the original works with ComBat on real data, secondly with ComBat but with random numbers instead of real data, thirdly without ComBat and instead blocking batch with limma on real data
+First as described in the original works with ComBat on real data.
+Secondly, with ComBat but with random numbers instead of real data.
+Thirdly without ComBat and instead blocking batch with limma on real data.
 a) Re-analysis of  \citealp{Towfic2014}, glatiramer acetate vs. generic 

### 2014-06-16 commonscripts/helperfunctions.r

@@ -77,3 +77,40 @@ adhocpvalueplot = function(realcombatp, reallimmap, randomp, main="P-values", xr
 				 				 "Random data, ComBat adjusted")) 
 }
 
+
+
+
+# Run F test and store results in an object
+Ftest = function (dat, mod, mod0,adjust=1)
+{
+	obj=new.env()
+	class(obj)="Ftest"
+	n <- dim(dat)[2]
+	m <- dim(dat)[1]
+	df1 <- dim(mod)[2]
+	df0 <- dim(mod0)[2]
+	p <- rep(0, m)
+	Id <- diag(n)
+	resid <- dat %*% (Id - mod %*% solve(t(mod) %*% mod) %*% t(mod))
+	rss1 <- rowSums(resid * resid)
+	rm(resid)
+	resid0 <- dat %*% (Id - mod0 %*% solve(t(mod0) %*% mod0) %*%  t(mod0))
+	rss0 <- rowSums(resid0 * resid0)
+	rm(resid0)
+	fstats <- ((rss0 - rss1)/(df1 - df0))/(rss1/(n - df1))
+	if (is.null(adjust)) {
+		adjust=mean(rss1/(n - df1))/mean((rss0 - rss1)/(df1 - df0))
+	}
+	fadj=fstats*adjust
+	p <- 1 - pf(fadj, df1 = (df1 - df0), df2 = (n - df1))
+	obj$n1=n; obj$n0=m; obj$df1=df1-df0; obj$df0=n-df1;
+	obj$F0=fstats; obj$F=fadj
+	obj$MS1=(rss0-rss1)/(df1-df0)
+	obj$MS0=rss1/(n-df1)
+	obj$P=p
+	obj$adjust=adjust
+	return(obj)
+}
+
+
+

### 2014-06-16 overviewplot.r

@@ -6,7 +6,7 @@
 library(sva)
 
  
-adhocboxplot = function(y, grouplabels, batchlabels, addlegend=FALSE, ylim=NULL, xlab="x")
+adhocboxplot = function(y, grouplabels, batchlabels, addlegend=FALSE, ylim=NULL, xlab="x", figureletter=NA)
 {
   #x = batchlabels + grouplabels/8
   colpalette = c("blue", "red", "darkgreen", "magenta")
@@ -52,14 +52,19 @@ adhocboxplot = function(y, grouplabels, batchlabels, addlegend=FALSE, ylim=NULL,
     lines(c(xoffset,xright), c(m,m), col=colpalette[g], lwd=lwd)
     xoffset = xoffset+boxseparation
   }
-  
+  legendcex=2
   if(addlegend)
   {
-    legendcex=2
+    
     legend("topright", legend=paste("Group ", groupnames, sep=""), 
            text.col=colpalette[groupnames], bty="n", cex=legendcex)
     legend("bottomright", legend=paste("Batch ", unique(batchlabels), sep=""),
            pch=pchselection[unique(batchlabels)], bty="n", cex=legendcex)
+  }
+  if(!is.na(figureletter))
+  {
+  	figureletter=paste("(", figureletter, ")", sep="")
+  	text(x=0, y=ylim[2], labels=figureletter, cex=legendcex)
   }
 }
 
@@ -104,10 +109,10 @@ alldata = c(matrix_condition[index,],
             matrix_meancenter[index,],
             matrix_combat[index,])
 ylim = c(min(alldata), max(alldata))
-adhocboxplot(matrix_condition[index,], sampleannotation$treatment, sampleannotation$batch, addlegend=FALSE, ylim=ylim, xlab="Without batch effects")
-adhocboxplot(matrix_conditionbatch[index,], sampleannotation$treatment, sampleannotation$batch, addlegend=FALSE, ylim=ylim, xlab="With batch effects added")
-adhocboxplot(matrix_meancenter[index,], sampleannotation$treatment, sampleannotation$batch, addlegend=FALSE, ylim=ylim, xlab="Zero-centered per batch")
-adhocboxplot(matrix_combat[index,], sampleannotation$treatment, sampleannotation$batch, addlegend=TRUE, ylim=ylim, xlab="ANOVA centered values (for tiden combat)")
[…]

### 2014-06-16 reanalysis/Johnson2007/figurescript_johnson.r

@@ -60,7 +60,7 @@ print(paste("ComBat adjusted random data, significant probes: ",  sum(p.adjust(r
 # create pvalue plot
 figfile = paste(getwd(), "/dataset2pvalues.pdf", sep="")
 pdf(figfile)
-adhocpvalueplot( combatp, limmap, randp)
+adhocpvalueplot( combatp, limmap, randp,  main="(b) P-values")
 dev.off()
 print( paste("Figure created; ",figfile ))
 

### 2014-06-16 reanalysis/Leek2012/figurescript_leek.r

@@ -1,5 +1,11 @@
 
 
+# Script to create a Figure 2.
+
+# This is a modified version of the ComBat use example from the sva tutorial:
+# http://www.bioconductor.org/packages/2.13/bioc/vignettes/sva/inst/doc/sva.pdf
+# The modifications consist of swapping the data with random numbers a plot of the p-value distibution and qq-plot of the F-statisitcs
+
 library(sva)
 library(bladderbatch)
 data(bladderdata)
@@ -28,6 +34,17 @@ qValuesComBat = p.adjust(pValuesComBat,method="BH")
 
 figfile = paste(getwd(), "/leekrandomdatapvalues.pdf", sep="")
 pdf(file=figfile)
-hist(pValuesComBat,  main="P-values, Random numbers", breaks=100, xlab="p-value")
+hist(pValuesComBat,  main="(a) P-values, Random numbers", breaks=100, xlab="p-value")
+dev.off()
+print( paste("Figure created; ",figfile ))
+
+
+figfile = paste(getwd(), "/leekqqplot.png", sep="")
+png(file=figfile)
+source("../../commonscripts/helperfunctions.r")
+test = Ftest(combat_edata,mod,mod0)
+Fquant=qf(ppoints(nrow(combat_edata)),test$df1,test$df0)
+qqplot(Fquant,test$F, ylim=c(0,max(test$F)), xlim=c(0,max(test$F)), main="(b) QQ-plot of F-statistics")
+abline(0,1,lwd=2, lty=2)
 dev.off()
 print( paste("Figure created; ",figfile ))

### 2014-06-16 reanalysis/Towfic2014/figurescript_towfic.r

@@ -70,7 +70,7 @@ print(paste("ComBat adjusted random data, significant probes: ",
 # create pvalue plot
 figfile = paste(getwd(), "/towficpvalues.pdf", sep="")
 pdf(file=figfile)
-adhocpvalueplot(limma_p_alt,limma_p_woc,limma_p_rand_combat)
+adhocpvalueplot(limma_p_alt,limma_p_woc,limma_p_rand_combat, main="(a) P-values")
 dev.off()
 print( paste("Figure created; ",figfile ))
 

### 2014-06-18 tex/manuscript.tex
Updated manuscript (main tex and pdf files)
Have merged (manually) the two different manuscript versions, and placed
as manuscript.tex, adding PDF and placing the figure files in the Fig
subfolder (renaming two of the files). Note that I do not now do the
LaTeX typesetting inside the Git folder, so the auxhiliary files are not
included including the tex-aux folder.

@@
 % Temporary notifications and reminders:
-\newcommand\NB[1]{{\color{red}#1}}
-\newcommand\NOTE[1]{{\color{red}[#1]}}
-\newcommand\CITE[1]{{\color{magenta}[#1]}}
+\usepackage{color}
+\definecolor{NBcol}{rgb}{0.7,0,1}
+\definecolor{CITEcol}{rgb}{1,0,0.7}
+\definecolor{COMMENTcol}{rgb}{0.7,0.2,0}
+\newcommand\COMMENT[2]{\textcolor{COMMENTcol}{#1}\NOTE{#2}}
+\newcommand\NB[1]{\textcolor{NBcol}{\textit{#1}}}
+\newcommand\NOTE[1]{\NB{[#1]}}
+\newcommand\CITE[1]{\textcolor{CITEcol}{[#1]}}
 
 \begin{document}
+
+\footnotetext{\color{blue}Manuscript version \today.}
 
@@
 % E-mail address for correspondence
-{ehovig@radium.uio.no}}
+{ehovig@ifi.uio.no}}
 
@@
 This appealing compartmentalization is also convenient for practical purposes, for example when data-processing and statistical analyses are performed by different personnel.
+Unfortunately, as we demonstrate in this paper, when the batch--group design is unbalanced, this approach may be unreliable.
 
-The first step can be achieved by subtracting the mean of the measurements in one batch from all measurements in that batch, i.e mean adjustment or one-way ANOVA adjustment as implemented in the method \texttt{pamr.batchadjust} from the \texttt{pamr} package in R.
+A simple removal of batch effects can be achieved by subtracting the mean of the measurements in one batch from all measurements in that batch, i.e mean adjustment or one-way ANOVA adjustment as implemented in the method \texttt{pamr.batchadjust} from the \texttt{pamr} package in R.
 When the batch--group design is balanced, mean-adjustment will remove most, but not necessarily all, variance attributed to batch and leave the between group variance, thus increasing the statistical power.
-However, when the batch--group design is unbalanced, batch differences will in part be influences by group differences, and thus batch correction will reduce group differences and thereby reduce the statistical power.
+However, when the batch--group design is unbalanced, batch differences will in part be influenced by group differences, and thus batch correction will reduce group differences and thereby reduce the statistical power.
 In very uneven group--batch designs with multiple groups, spurious group differences may even be induced in this way.
@@
 using a two-way ANOVA, and only remove the batch differences from the data.
-Effectively, this means group differences are estimated based on within batch comparison, and applied to the batch adjusted data.
+Effectively, group differences are estimated based on within batch comparisons, and applied to the batch adjusted data.
 In a balanced group--batch design, group differences and batch effects are independent, and this approach becomes identical to the above described zero-centering per batch.
 If the group--batch design is heavily unbalanced, estimation of group differences and batch effects are interdependent.
-Unfortunately, when group differences estimated from an unbalanced group--batch design are applied to the entire data set, if this batch adjusted data set is later analysed for group differences, the confidence will be exaggerated as the estimation of the group differences in the unbalanced design is less accurate than in the corresponding balanced design.
+However, in applying the estimated group differences across the entire data set, the uncertainties of these estimates are ignored.
+If the group--batch design is unbalanced and batch compositions ignored after batch adjustments have been made, later analyses will systematically underestimate the statistical uncertainties and exaggerate the confidence of group differences.
 Figure~\ref{fig:boxplots} illustrates how statistical uncertainties are deflated by this batch adjustment method by comparing them to the uncertainties from the original ANOVA.
@@
 The ComBat method described in \citet{Johnson2007}, and included in the sva package (\citealp{Leek2012}), can use either of the two above described approaches to estimate batch differences, but uses an empirical Bayes approach to avoid over-correction for batch effects for small batches.
-It has popularised the two-way ANOVA procedure for retaining group differences when adjusting for batch effects, and the inclusion of group difference as a covariate when removing batch effects has been recommended in the sva tutorial and user fora.
-Based on actual use of ComBat by the authors and others, ourselves included, we suspect thus adjusted data are commonly treated as "batch effect free" in subsequent analyses.
+It has thus improved and popularised the two-way ANOVA procedure for retaining group differences when adjusting for batch effects, and the inclusion of group difference as a covariate when removing batch effects has been recommended  in the sva tutorial and user fora.
+Based on actual use of ComBat by the authors and others, ourselves included, we suspect thus adjusted data are commonly treated as ``batch effect free'' in subsequent analyses.
 And as a consequence, confidence in group effects has been overestimated and false results reported.
 
-\NB{Kort oppsummere behovet for batch-korrigering: generelt og i microarraydata.}
-
-\NB{Kort beskrive metoder som gjoer dette (batch-sentrering, ANOVA, ComBat).}
-
-\NB{Forklare tilsiktet effekt disse paa batch+gruppe-forskjeller (balansert versus ikke-balansert).}
-
-\NB{Forklare kort hva faktisk effekt er: ANOVA gir redusert gruppe-forskjeller, mens for ComBat overestimeres den statistiske styrken/sikkerheten av forskjellen.}
+Statisticians would most likely take extra precautions should the sample or batch sizes be very small.
+However, the effects of batch adjustment using two-way ANOVA or ComBat on unbalanced data sets remain even as sample and batch sizes increase.
+For example, group comparisons using one-way ANOVA on the batch adjusted data will essentially result in $F$ statistics that are inflated by a fixed factor which depends on the unevenness of the design rather than the size of the sample or batches.
+The effect of this may be further exacerbated by running these analyses a large number of times, e.g.
+on thousands of genes, and use false discovery rate to determine significant cases: an approach that is particularly sensitive to inflated false positive rates.
 
@@
 
-We will base our discussion of data with batch effects on a simple model:
+We will base our discussion on a simple model for data with batch effects:
 \begin{equation}
@@
 $\delta_{ig}=1$.
-We will discuss the effect of empirical Bayes estimation of $\gamma_{ig}$ later, but out main argument is more easily made in the simpler context.
+We will discuss the effect of empirical Bayes estimation of $\gamma_{ig}$ later, but our main argument is more easily made in the simpler context.
+Empirical Bayes may reduce the amount of batch adjustment in cases where batch effects are small or cannot be accurately estimated, e.g.
+for small batches.
+However, in cases with large batches or substantial batch effects, it should differ little from the two-way ANOVA approach.
 
@@
 
-The main ambition of batch effect adjustments is to be able to remove batch differences in such a way that downstream analyses of the adjusted data may be done without further batch adjustments.
+A common ambition of batch effect adjustments is to remove batch differences in such a way that downstream analyses of the adjusted data may be done without further corrections for batches.
 
@@
 When comparing groups, the common value $\bar{Y}$ has no effect, and so this is equilvalent to zero-centring each batch.
-However, if the different groups are unevenly represented in the different batches, the batch average $\bar{Y}_i$ will tend to capture group differences as well as batch effects.
-Thus, batch centering may reduce group differences, and thus reduce the power of downstream analyses.
+If the groups are unevenly represented in the different batches, the batch average $\bar{Y}_i$ will tend to capture group differences as well as batch effects:
+\begin{equation}
+\Delta\widetilde{Y}_{ijr}=\beta_j-\bar{\beta}_i+\epsilon_{ijr}-\bar{\epsilon}_i
+\quad\text{where}\quad
+\bar{\beta}_i=\frac{\sum_{j=1}^M n_{ij}\beta_j}{\sum_{j=1}^M n_{ij}},\;
+\bar{\epsilon}_i=\frac{\sum_{j=1}^M\sum_{r=1}^{n_{ij}} \epsilon_{ijr}}{\sum_{j=1}^M n_{ij}}.
+\end{equation}
+Thus, batch centering will tend to reduce group differences in an unbalanced design, and thus reduce the power of downstream analyses.
+By reducing the differences between some groups, i.e.
+those found together in the same batches, it may also induce false differences between other groups.
 
-Removing batch effects while retaining group differences can be done through an ANOVA analysis in which the group effects, $\beta_j$, and the batch effects, $\gamma_i$, are estimated simultaneously.
-Batch adjusted values may then be obtained by subtracting the estimated batch effects, $\hat\beta_j$:
+Removing batch effects while retaining group differences can be done through a two-way ANOVA in which group effects, $\beta_j$, and batch effects, $\gamma_i$, are estimated simultaneously.
+Batch adjusted values may then be obtained by subtracting the estimated batch effects, $\hat\gamma_i$:
 \begin{equation}
-\widetilde Y_{ijr}=Y_{ijr}-\hat\beta_j=\alpha+(\beta_j-\hat\beta_j)+\gamma_i+\epsilon_{ijr}.
+\widetilde Y_{ijr}=Y_{ijr}-\hat\beta_j=\alpha+\beta_j+(\gamma_i-\hat\gamma_i)+\epsilon_{ijr}.
 \end{equation}
@@
 
-Unfortunately, the estimation error $\hat\beta_j-\beta_j$ affects all values within the same batch in the same manner, inducing a dependency between the values.
-If the study groups are evenly represented in all batches, this will not influence estimated group differences as all groups are equally affected.
-However, if groups are unevenly represented, this induced dependency can have severe impact on downstream analyses.
+The estimation error $\hat\gamma_i-\gamma_i$ affects all values within the same batch in the same manner.
+Thus, while the aim is to remove spurious dependencies within batches, it may also induce new dependencies.
+In a balanced group--batch design, this has the same effect on all groups and thus does not influence group comparisons.
+In an unbalanced design, however, the estimated batch effects $\hat\gamma_i$ will correlate with the estimated group effects $\hat\beta_j$.
+\COMMENT{Assuming the batch adjusted values are without further batch differences corresponds to assuming that $\gamma_i=\hat\gamma_i$, and ignoring the estimation uncertainty and its impact on the correlated $\hat\beta_j$.
+It results in over-confidence in the accuracy of the group differences, which can have severe impact on downstream analyses.}{Denne forklaringen må jeg gjøre mye klarere.}
 
@@
 The documentation accompanying the sva library has a runnable example demonstrating how to adjust a data set with ComBat followed by a F-test.
-Swapping the real data with random numbers from a normal distribution (mean=0, sd=1), but otherwise following the instructions, will generate the p-value distribution shown in Figure 1.
-The skewed distribution is a indication that this approach may have a unintentional adverse effect.
-\NOTE{Jeg har ikke beskrevet permutatsjon checken fordi jeg er usikker paa om vi skal bruke den.}
+Swapping the real data with random numbers from a normal distribution (mean=0, sd=1), but otherwise following the instructions, will generate the p-value distribution shown in Figure~\ref{fig:p-sva}.
+\NOTE{Perhaps show Q-Q plot of $F$-statistics.} The skewed distribution is a indication that this approach may have a unintentional adverse effect.
+\NOTE{Tror vi dropper permutasjonstesten.}
 
-%% FIGURE 1: Original position
+If the number of samples is increased, the empirical Bayes estimates used by ComBat will shrink the batch effect estimates and thus moderate the batch adjustments.
+However, if random data with different means in each batch, but no difference betwene groups, are drawn, the problem remains.
+\NOTE{Bør dette komme her eller senere?
+Evt.
+understøttes?}
+
+%% FIGURE 1=fig:p-sva: Original position
 
@@
 
-As the amount of false positive results when trying to retain group differences depends on the batch/group balance, we will show two examples with varying degree of unbalancedness.
+As the amount of false positive results when trying to retain group differences depends on the batch--group balance, we will show two examples with varying degree of unbalancedness.
 
-1) In the first experiment(\citealp{Towfic2014}), cells were treated with glatiramer acetate (a medicine for multiple sclerosis) or a generic and mRNA was measured using microarrays alongside control samples.
-A batch effect correlating to the chip (Illumina WG-6\_V2, six samples per chip) was observed and adjusted for with ComBat, whereafter the data was tested for differetially expressed genes, yielding a list of 1000 genes (Table S5, \citealp{Towfic2014}).
-Unfortunately the batch/treatment design was unbalanced with several batches having only one of the main treatments of interest.
+\subsubsection{Experiment 1}
+
+In the first experiment(\citealp{Towfic2014}), cells were treated with glatiramer acetate (a medicine for multiple sclerosis) or a generic and mRNA was measured using microarrays alongside control samples.
+A batch effect correlating to the chip (Illumina WG-6\_V2, six samples per chip) was observed and adjusted for with ComBat, whereafter the data was tested for differentially expressed genes, yielding a list of 1000 genes (Table S5, \citealp{Towfic2014}).
+Unfortunately the batch--treatment design was unbalanced with several batches having only one of the main treatments of interest.
 When we re-analyzed their data without using ComBat, but instead blocked for batch effect in limma, only 9 genes were found ($\text{FDR}<0.05$).
-Additional sanity checks with random numbers or perumted labels were also carried out and the distribution of p-values for differents settings are shown in Fig.
-2.
-Our conclusion is that most of the genes reported as differtially expresssed in (\citealp{Towfic2014}) are false positives.
-This example is a sort of "worst case" scenario for applying ComBat, since it both has a very unbalanced batch/group design and a a priori assumpotion of no difference.
-The R-code for our analysis and a more extensive report can be downloaded from github (\href{https://github.com/vegardny/combat_tests.git}).
+\NOTE{Also run analysis using a mixed random effects model which assumes random batch effects much like in the empirical Bayes model.} The above outlined sanity check with random numbers was also carried out.
+The distribution of p-values for different settings are shown in Figure~\ref{fig:p-towfic}.
+Our conclusion is that most of the genes reported as differentially expressed in (\citealp{Towfic2014}) are false positives.
+This example is a sort of ``worst case'' scenario for applying ComBat, since it both has a very unbalanced batch--group design and a a priori assumption of no difference.
+The R-code for our analysis and a more extensive report can be downloaded from \href{https://github.com/vegardny/combat_tests.git}{Github} \NOTE{Update reference}.
 
-%% FIGURE 2: Original position
+%% FIGURE 2=fig:p-towfic: Original position
 
-2) The second example is taken from the supporting information for the original ComBat article (\citealp{Johnson2007}) where it is denoted "Data Set 2".
+\subsubsection{Experiment 2}
+
+The second example is taken from the supporting information for the original ComBat article (\citealp{Johnson2007}) where it is denoted ``Data set 2''.
 Cells inhibited for the expression of the TAL1 gene were compared to controls on a microarray platform.
-The experiment was conducted on three different time points (used as batches) with a total of 30 samples and a fairly balanced batch/treatment set up (6-2, 3-4 and 9-6).
+The experiment was conducted on three different time points (used as batches) with a total of 30 samples and a fairly balanced batch--treatment set up (6-2, 3-4 and 9-6).
 ComBat was applied followed by a t-test in order to identify differentially expressed genes.
@@
 Then, we analysed their data without batch adjustment in ComBat, but blocking for batch in limma, resulting in 377 probes ($q<0.05$).
-In addition, the two sanity checks outlined above were performed.
-In contrast to the above results obtained for(\citealp{Towfic2014}), the P-value distributions for the alternative analyses (Fig.3 a,b) does not indicate a huge difference.
+In addition, the sanity checks outlined above were performed.
+The distribution of P-values for different settings are shown in Figure~\ref{fig:p-values}b.
+In contrast to the results obtained for\citealp{Towfic2014} (Figure~\ref{fig:p-values}a), the P-value distributions for the alternative analysis does not indicate a huge difference.
 Nevertheless, we belive that  P-values are deflated for the ComBat adjusted analysis.
-The R-code for our analysis and a more extensive report can be downloaded from github (\href{https://github.com/vegardny/combat_tests.git}).
+The R-code for our analysis can be downloaded from \href{https://github.com/vegardny/combat_tests.git}{Github} \NOTE{Update reference}.
 
-%% FIGURE 3: Original position
+%% FIGURE 3=fig:p-johnson: Original position
 
-%% FIGURE 4: Original position
+%% FIGURE 4=fig:pdist-johnson: Original position
 
-3) \NOTE{Her kan vi fylle paa med flere eksempler etter samme mal hvis det trengs.
+\subsubsection{More?}
+
+\NOTE{Her kan vi fylle paa med flere eksempler etter samme mal hvis det trengs.
 Feks en fersk artikkel fra Nature Genetic (\citealp{Yoo2014}).
@@
 
+\NB{Systematic underestimation of $P$-values combined with FDR is a particularly bad combination as FDR lends strength across tests.}
+
 \subsection{Increased emphasis on preserving group difference}
 
-In the original ComBat article (\citealp{Johnson2007}) it is clear that the primary motivation behind ComBat was to employ an Emprical Bayes method for batch-effect removal.
+In the original ComBat article (\citealp{Johnson2007}) it is clear that the primary motivation behind ComBat was to employ an Emprical Bayes method for batch-effect removal to allow better handling of small batches.
 The feature of retaining group differences for unbalanced designs is optional and seems to be subordinate, only exemplified in the supplementary information.
-However, over the years this feature became more important judging from advice given by the author to users
-(
-[1](\href{https://groups.google.com/d/msg/combat-user-forum/eSVSKwGtuyE/ZIWV2juYmmAJ})
-,[2](\href{https://groups.google.com/d/msg/combat-user-forum/Vkb9p7wekd4/h5Etie7FTVQJ})
-,[3](\href{https://groups.google.com/d/msg/combat-user-forum/fpBTcgDjiR8/uo4QIZL4sZgJ})
-,[4](\href{https://groups.google.com/d/msg/combat-user-forum/26FZlgU2LFQ/W6U_Lhh_64EJ})
-). \NOTE{Aa linke direkte er kanskje litt ufint, alternativt kan det linkes til forumet uten konkrete innlegg.}  
-And when ComBat was incorporated in the sva package (\citealp{Leek2012}), inclusion of group labels was made almost mandatory (an undocumented option of passing a NULL value exists).
-Thus, this problematic use has likely been common.
+However, over the years this feature became more important judging from advice given on user fora.
+When ComBat was later incorporated in the sva package (\citealp{Leek2012}), \COMMENT{batch adjustment without covariates was no longer a default as specification of covariates became a requirement, although specification of a null model was still possible although not clearly documented in the sva documentation}{Is this correct?}.
+Thus, we expect this usage to have grown more common.
 
@@
 In addition, other works looking into the problem of batch effects were mostly recommending ComBat without much concern (\citealp{Kupfer2012}, \citealp{Kitchen2011}).
-A brief inquiry into  some of the articles citing ComBat (574 Google Scholar) revealed few problems, and their method descriptions regarding ComBat were mostly sparse, limited to one or two sentences (\CITE{24391845}, \CITE{18414638}, \CITE{21731603}, \CITE{23630272}, \CITE{24584070}).
-A further indication of their carefree use of this potentially devastating procedure was the frequent neglect to state the program parameters, i.e batch labels (\CITE{18414638}, \CITE{21731603}) or group labels (\CITE{24391845}, \CITE{18414638}, \CITE{21731603}, \CITE{23630272}, \CITE{23482648}, \CITE{24584070}).
-Often no effort was done in order to substantiate the existence of batch effects in their data, except for stating the presence of batches (\CITE{24391845}?, \CITE{18414638}, \CITE{21731603}, \CITE{23630272}).
-In one instance ComBat was even applied on data were effects due to batch were investigated, but not found \NOTE{husker ikke ref}.
+A brief inquiry into  some of the articles citing ComBat (574 Google Scholar) revealed few problems, and their method descriptions regarding ComBat were mostly sparse, limited to one or two sentences.
+A further indication of their carefree use of this potentially devastating procedure was the frequent omittance of program parameters that were used, i.e batch labels or whether group labels were supplied as covariates.
+Often no effort was done in order to substantiate the existence of batch effects in their data, except for stating the presence of batches.
 The incorporation of the method into analysis pipelines (\CITE{16642009}, TCGA) and other packages (\CITE{23452776}, \CITE{21937664}) could make its usage even more trivial and parameters setting harder to perceive.
 Taken together we fear that many published results from data adjusted by ComBat are completely or partially false.
-And knowing that scientists don't give up where one analysis fails, a method that almost ensures a result given a sufficiently unbalanced design will continue to be used.
+We hope that our warning will enlighten the community and  reduce this unfortunate combination of methods.
 
@@
 
-We have shown that adjusting for batch effects while preserve the group difference may lead to varying degree of false results.
-Knowing this, to what degree can an investigator trust a result from a work applying such a method?
-Essentially, when the batch/group configuration is balanced, or group difference is ignored  (i.e.
-no group labels not given as parameters to ComBat), problems related to preserving group differences will not occur.
+For an investigator facing an unbalanced data set with batch effects, our advice is to adjust for batch inside the statistical test and avoid the two step procedure outlined above.
+If this is not possible, then applying the method described in this article should only be done with great caution and downstream confidence estimates must be treated with suspicion.
+
+Knowing that adjusting for batch effects while preserving the group difference may lead to varying degree of false results, to what extent can an investigator trust a result from a work applying such a method?
+Essentially, when the batch--group configuration is balanced, or group difference is ignored  (i.e.
+group labels not given as parameters to ComBat), problems related to preserving group differences will not occur.
 For other cases, a re-analysis without using this approach is the most rigorous path.
@@
 
-\section*{Reproducible research}
+\subsection{Reproducible research}
+
+The data and scripts used to generate the results in this work, i.e figures, are made available at https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-figures.git.
+Additional analyses were also performed, but considered excessive for the current  article.
+However, some readers might find this very interesting and we provide it as an extra resource at https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-reports.git
+
 \NOTE{vi boer klare aa tilfredstille kravene i http://biostatistics.oxfordjournals.org/content/10/3/405.full}
+\NOTE{For aa tilfredstille kravene skal data og script vare available.
+Jeg mener vi klarer dette.
+paa githuben er ikke data fra towfic, men det blir lastet av scriptet fra GEO.
+Ikke beskrevet her men i scriptet.
+Videre i forklaringen til biostatistics staar det at script og data skal bli lagt paa deres sider, som vi kan hvis de vil}
 
@@
 \begin{figure}[!p]
-\centering\includegraphics[width=13cm]{Fig/boxplots.png}
+\centering\includegraphics[width=13cm]{Fig/boxplots.pdf}
+%\centering\includegraphics[width=13cm]{Fig/boxplots.png}
 \caption{Simulated data from four study groups was generated where groups 1 and 2 have lower means than groups 3 and 4.
@@
 \begin{figure}[!p]
-\centering\includegraphics[width=13cm]{Fig/pvaluesleeksva.png}
-\caption{
-Two sanity checks were the recommended use of ComBat fails. Adapted from the user guide in the sva package. 
-a) Real data is substituted with random numbers from a normal distribution (mean=0, sd=1), but the batch/group design is retained, followed by batch adjustment in ComBat and a F-test.
-b) 10 runs of real data with the "cancer" labels permuted within batches, followed by ComBat adjustment and a F-test. All permutations produces a skewed p-value distribution.
+\begin{center}$
+\begin{array}{cc}
+\includegraphics[width=6.5cm]{Fig/leekrandomdatapvalues.pdf} &
+\includegraphics[width=6.5cm]{Fig/leekqqplot.png}
+\end{array}$
+\end{center}
+\caption
+{
+A sanity checks where the recommended use of ComBat fails.
+Adapted from the user guide in the sva package.
+Real data are substituted with random numbers from a normal distribution, but the batch-group design is retained.
+ComBat is applied followed by a F-test.
+a) P-value distribution
+b) QQ plot of the F-statistics
 }
-\label{fig:p-sva}
 \end{figure}
@@
 \begin{figure}[!p]
-\centering\includegraphics[width=13cm]{Fig/pvaluestowfic.png}
-\caption{
-P-value distribution for the main comparison in (data set 2 24421904) glatiramer acetate vs. generic.  
-a) Real data adjusted by ComBat as described in the article followed by a significance test in limma. 
-b) Real data not adjusted by ComBat, tested for significance in limma using batch as a blocking factor. 
-c) Real data is substituted with random numbers from a normal distribution (mean=0, sd=1), but the batch/group design is retained, followed by batch adjustment in ComBat and a significance test in limma.
-d) 10 runs of real data with the glatiramer acetate("DP") and generic("N") labels permuted within batches, followed by ComBat adjustment and significance test in limma.  
-\NOTE{jeg tok med begge sanitycheckene her. Er d) slik einar foreslo?. Merk jeg har ikke permutert alle labeler, bare de mellom DP og N. Det er en del andre.}
+\begin{center}$
+\begin{array}{cc}
+\includegraphics[width=6.5cm]{Fig/towficpvalues.pdf} &
+\includegraphics[width=6.5cm]{Fig/johnson2pvalues.pdf}
+\end{array}$
+\end{center}
+\caption
+{
+Three analyses of two works where batch effects were adjusted for with ComBat.
+First as described in the original works with ComBat on real data.
+Secondly, with ComBat but with random numbers instead of real data.
+Thirdly without ComBat and instead blocking batch with limma on real data.
+a) Re-analysis of  \citealp{Towfic2014}, glatiramer acetate vs. generic 
+b) Re-analysis of "Data set 2" \citealp{Johnson2007},  TAL1 inhibition vs. control
 }
-\label{p-towfic}
 \end{figure}
 
-\begin{figure}[!p]
-\centering\includegraphics[width=13cm]{Fig/pvaluesjohnson.png}
-\caption{
-P-value distribution for the main comparison in (data set 2, \citealp{Johnson2007}) for  
-a) Real data adjusted by ComBat as described in the article followed by a significance test in limma. 
-b) Real data not adjusted by ComBat, tested for significance in limma using batch as a blocking factor. 
-c) Real data is substituted with random numbers from a normal distribution (mean=0, sd=1), but the batch/group design is retained, followed by batch adjustment in ComBat and a significance test in limma.
-d) 10 runs of real data with the group labels permuted within batches, followed by ComBat adjustment and significance test in limma.   
-\NOTE{Problemet med permutatsjons sanitychechen slik jeg har utfoert og plottet den her er at den neppe kan sies aa feile.
-Dessuten kan de 3 foerst plottene bli plottet i ett som vist under i en alternativ version jeg synes er finere, men dette gaar ikke for permutasjonstesten tror jeg}
-}
-\label{fig:p-johnson}
-\end{figure}
-
-\begin{figure}[!p]
-\centering\includegraphics[width=13cm]{Fig/pvaluesjohnson2.png}
-\caption{
-\NOTE{Alternativ illustrasjon av de tre p-value distribusjonene}
-}
-\label{fig:pdist-johnson}
-\end{figure}
-
-%\begin{table}[!p]
-%\tblcaption{... \protect\citet{...} ... \label{Table1}}
-%{\tabcolsep=4.25pt
-%\begin{tabular}{@{}cccccccccc@{}}
-%\tblhead{Dis & Chr & Marker & MLE-OR & MLE & MLE & EB-HM-OR & EB-HM & EB-HM & ES \\
-%&&& (SE) & \textit{P}-value & rank & (SE) & \textit{P}-value & rank & rank}
-%CAD & 2 & rs2943634 & 1.22(0.04) & $1.23 \times 10^{-5}$ & 32 & 1.02(0.01) & 0.19 & 28 & 37 \lastline
-%\end{tabular}}
-%\end{table}
-
 \end{document}

### 2014-06-18 figurescript.r
rydding

@@ -1,4 +1,6 @@
 
+# Main script.
+# Calls the individual scripts that creates figures 1,2,3a,3b.
 
 
 startwd = getwd()

### 2014-06-18 overviewplot.r

@@ -1,6 +1,6 @@
 
 
-
+# Script to create Figure 1
 
 
 library(sva)

### 2014-06-18 reanalysis/Johnson2007/figurescript_johnson.r

@@ -1,11 +1,15 @@
 
-# source("figurescriptjohnson.r")
+
+# script to create Figure 3b
+
 starttime = Sys.time()
 
 library("sva")
 library("limma")
 source("../../commonscripts/helperfunctions.r")
 source("helperfunctions_johnson.r")
+
+print("Re-analysis and figure generation for example taken from Johnson et al. 2007, Data set 2")
 
 tmp = loadjohnsondata()
 sampleannotation = tmp[["sampleannotation"]]
@@ -25,7 +29,7 @@ cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design)
 fit2 = contrasts.fit(fit, cont.matrix)
 combatp = eBayes(fit2)$p.value[,1]
 rm(Type, design, fit, cont.matrix, fit2)
-print(paste("ComBat adjusted real data, significant probes: ",  sum(p.adjust(combatp, "fdr")<0.05)))
+print(paste("ComBat adjusted real data, significant probes (fdr<0.05): ",  sum(p.adjust(combatp, "fdr")<0.05)))
 
 # DE pvalues not ComBat adjusted, 
 # batch blocked in limma
@@ -37,7 +41,7 @@ cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design)
 fit2 = contrasts.fit(fit, cont.matrix)
 limmap = eBayes(fit2)$p.value[,1]
 rm(Type, Block, design, fit, cont.matrix, fit2)
-print(paste("Limma batch adjusted real data, significant probes: ",  sum(p.adjust(limmap, "fdr")<0.05)))
+print(paste("Limma batch adjusted real data, significant probes (fdr<0.05): ",  sum(p.adjust(limmap, "fdr")<0.05)))
 
 #Random data
 set.seed(100)
@@ -55,7 +59,7 @@ cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design)
 fit = lmFit(randcombatdata, design)
 fit2 = contrasts.fit(fit, cont.matrix)
 randp = eBayes(fit2)$p.value[,1]
-print(paste("ComBat adjusted random data, significant probes: ",  sum(p.adjust(randp, "fdr")<0.05)))
+print(paste("ComBat adjusted random data, significant probes (fdr<0.05): ",  sum(p.adjust(randp, "fdr")<0.05)))
 
 # create pvalue plot
 figfile = paste(getwd(), "/dataset2pvalues.pdf", sep="")

### 2014-06-18 reanalysis/Towfic2014/figurescript_towfic.r

@@ -1,5 +1,6 @@
 
-# source("figuretowfic.r")
+# script to create Figure 3a
+
 starttime = Sys.time()
 debug = FALSE
 downloaddata=TRUE
@@ -10,6 +11,8 @@ includelibs = c("Biobase", "GEOquery", "sva", "limma")
 lapply(includelibs, require, character.only=T)
 source("../../commonscripts/helperfunctions.r")
 source("helperfunctions_towfic.r")
+
+print("Re-analysis and figure generation for example taken from Towfic et al. 2014")
 
 ret = loadtowfic(downloaddata)
 sampleannotation = ret[["sampleannotation"]]
@@ -33,7 +36,7 @@ fit = lmFit(combatdata, design)
 cont.matrix = makeContrasts ( contrasts="groupDP-groupN", levels=design)  
 fit2 = contrasts.fit(fit, cont.matrix)
 limma_p_alt = eBayes(fit2)$p.value[,1]
-print(paste("ComBat adjusted real data, significant probes: ",
+print(paste("ComBat adjusted real data, significant probes (fdr<0.05): ",
             sum(p.adjust(limma_p_alt, method="fdr")<0.05)))
 
 #Limma blocked batch and significance test
@@ -45,7 +48,7 @@ cont.matrix = makeContrasts ( contrasts="groupDP-groupN", levels=design)
 fit2 = contrasts.fit(fit, cont.matrix)
 limma_ret_woc = eBayes(fit2)
 limma_p_woc = limma_ret_woc$p.value[,1]
-print(paste("Limma adjusted real data, significant probes: ",  
+print(paste("Limma adjusted real data, significant probes (fdr<0.05): ",  
             sum(p.adjust(limma_p_woc, method="fdr")<0.05)))
 
 # radnom, ComBat adjusted
@@ -63,7 +66,7 @@ fit = lmFit(randcombatdata, design)
 cont.matrix = makeContrasts ( contrasts="groupDP-groupN", levels=design)
 fit2 = contrasts.fit(fit, cont.matrix)
 limma_p_rand_combat = eBayes(fit2)$p.value[,1]
-print(paste("ComBat adjusted random data, significant probes: ",
+print(paste("ComBat adjusted random data, significant probes (fdr<0.05): ",
             sum(p.adjust(limma_p_rand_combat, method="fdr")<0.05)))
 
 

### 2014-06-18 overviewplot.r
figure1 fix

@@ -4,7 +4,7 @@
 
 
 library(sva)
-
+library(lsmeans)
  
 adhocboxplot = function(y, grouplabels, batchlabels, addlegend=FALSE, ylim=NULL, xlab="x", figureletter=NA)
 {
@@ -39,9 +39,7 @@ adhocboxplot = function(y, grouplabels, batchlabels, addlegend=FALSE, ylim=NULL,
   boxseparation = 0.25
   for(g in groupnames)
   { 
-    #boxplot(y[grouplabels==g], notch=TRUE, border=colpalette[g], 
-    #        add=TRUE, at=xoffset, col=rgb(1,0.5,0.3, alpha=0.0),
-    #        outline=FALSE, boxwex=0.5)
+
     m = mean(y[grouplabels==g])
     ci = t.test(y[grouplabels==g])[["conf.int"]]
     xleft = xoffset
@@ -67,6 +65,61 @@ adhocboxplot = function(y, grouplabels, batchlabels, addlegend=FALSE, ylim=NULL,
   	text(x=0, y=ylim[2], labels=figureletter, cex=legendcex)
   }
 }
+
+
+
+adhocboxplot2 = function(y, grouplabels, batchlabels, addlegend=FALSE, ylim=NULL, xlab="x", figureletter=NA)
+{
+	#x = batchlabels + grouplabels/8
+	colpalette = c("blue", "red", "darkgreen", "magenta")
+	pchselection = c(1, 2, 3)
+	x = batchlabels
+	lwd=2
+	
+	
+	xlim = c(0, length(unique(batchlabels))+2)
+	plot(1, 1, col=colpalette[grouplabels], pch=pchselection[batchlabels], 
+			 xlim=xlim, lwd=lwd, cex=2, ylim=ylim, xlab=xlab, ylab="", axes=FALSE, cex.lab =2, type="n")
+	
+	expression =  matrix_conditionbatch[1,]
+	group= factor(grouplabels)
+	batch= factor(batchlabels)
+	fit = lm(expression ~ group + batch )
+	anovaest = summary(lsmeans(fit,  ~group))
+	
+	
+	groupnames = unique(grouplabels)
+	groupnames = groupnames[order(groupnames)]
+	xoffset=2
+	boxwidth=0.35
+	boxseparation = 0.25
+	for(g in groupnames)
+	{ 
+		m =anovaest[g , "lsmean"]		
+		xleft = xoffset
[…]

### 2014-06-18 tex/manuscript.tex
fix

@@
 
-\subsubsection{More?}
-
-\NOTE{Her kan vi fylle paa med flere eksempler etter samme mal hvis det trengs.
-Feks en fersk artikkel fra Nature Genetic (\citealp{Yoo2014}).
-Der har de kombinert sekvenseringsdata  (deres data) fra 9 lymfom pasienter med et (eksternt) microarray data set som har 6 lymfom proever og 5 kontroller, dvs kontroll gruppen er helt fravaerende hos dem.
-Dvs veldig ubalansert (9/0, 6/5), MEN skille mellom phenotypene er saa stor at de sikkert hadde funnet omtrent de samme genene signifikante med aa bare bruke de balanserte dataene (men det var jo ikke deres.) Dette er ikke en viktig del av paperet, men det er veldig feil selvom konsekvensense ikke blir store.
-Dog gir det et falskt inntrykk av at dataene deres ligner tidligere data, se heatmap i suppl..
-Dessuten lager de GO analyser av dette som egentlig er 7 aar gamle andres data og ikke deres}
-
 \section{Discussion}
@@
 However, over the years this feature became more important judging from advice given on user fora.
-When ComBat was later incorporated in the sva package (\citealp{Leek2012}), \COMMENT{batch adjustment without covariates was no longer a default as specification of covariates became a requirement, although specification of a null model was still possible although not clearly documented in the sva documentation}{Is this correct?}.
+When ComBat was later incorporated in the sva package (\citealp{Leek2012}), batch adjustment without covariates was no longer a default as specification of covariates became a requirement.
+Although specification of a null model was still possible, the help file specifies that this should contain ``outcome of interest and other covariates besides batch''.
 Thus, we expect this usage to have grown more common.

### 2014-06-26 tex/manuscript.tex
Manuscript updated
Not entirely finished yet...

@@
 % Temporary notifications and reminders:
+%    \COMMENT{text}{comment to text}
+%    \NB{comment (as separate paragraph)}
+%    \NOTE{comment inside paragraph}
+%    \CITE{missing/unspecified citation}
 \usepackage{color}
@@
 Extraneous variables, if left unaccounted for, have the potential to lead an investigator into drawing wrong conclusions.
-In molecular biology, extraneous variables are often called "batch effects", probably due to the fact that reagents and other equipment, for instance microarray chips, are made in batches, and this is frequently observed as an effect in the measurements.
+A common example is ``batch effects'' caused by reagents, microarray chips, and other equipment are made in batches, often have systematic effects on the measurements.
+A similar example is ``centre effects'' when samples or data come from multiple sources.
 See \citet{Luo2010} for more examples.
 
-For a typical experiment comparing group differences, the presence of batch effects will decrease the statistical power since it adds variation to the data.
-If the batch--group design is unbalanced, i.e if the study groups are not equally represented in all batches, batch effects may act as a confounder and induce false differences between groups (\citealp{Leek2010}).
+For a typical experiment comparing differences between study groups, the presence of batch effects will decrease the statistical power since it adds variation to the data.
+If the batch--group design is unbalanced, i.e if the study groups are not equally represented in all batches, batch effects may also act as a confounder and induce false differences between groups (\citealp{Leek2010}).
 
@@
 First the batch effects are estimated and removed, creating a ``batch effect free'' data set.
-In the next step, the statistical analyses are performed on the adjusted data without further consideration of batch effects.
+Then, the statistical analyses are performed on the adjusted data without further consideration of batch effects.
 This appealing compartmentalization is also convenient for practical purposes, for example when data-processing and statistical analyses are performed by different personnel.
 Unfortunately, as we demonstrate in this paper, when the batch--group design is unbalanced, this approach may be unreliable.
+
+\NB{Er de neste tre avsnittene nødvendige eller kan de forkortes drastisk siden dette forklares grundigere siden?}
 
@@
 Effectively, group differences are estimated based on within batch comparisons, and applied to the batch adjusted data.
-In a balanced group--batch design, group differences and batch effects are independent, and this approach becomes identical to the above described zero-centering per batch.
-If the group--batch design is heavily unbalanced, estimation of group differences and batch effects are interdependent.
-However, in applying the estimated group differences across the entire data set, the uncertainties of these estimates are ignored.
-If the group--batch design is unbalanced and batch compositions ignored after batch adjustments have been made, later analyses will systematically underestimate the statistical uncertainties and exaggerate the confidence of group differences.
+In a balanced group--batch design, estimates of group differences and batch effects are independent, and this approach becomes identical to the above described zero-centering per batch.
+However, if the group--batch design is heavily unbalanced, estimation of group differences and batch effects are interdependent, and when applying the estimated group differences across the entire data set, the uncertainties of these estimates are ignored.
+Subsequent analyses will then systematically underestimate the statistical uncertainties and exaggerate the confidence of group differences.
 Figure~\ref{fig:boxplots} illustrates how statistical uncertainties are deflated by this batch adjustment method by comparing them to the uncertainties from the original ANOVA.
 
-The ComBat method described in \citet{Johnson2007}, and included in the sva package (\citealp{Leek2012}), can use either of the two above described approaches to estimate batch differences, but uses an empirical Bayes approach to avoid over-correction for batch effects for small batches.
-It has thus improved and popularised the two-way ANOVA procedure for retaining group differences when adjusting for batch effects, and the inclusion of group difference as a covariate when removing batch effects has been recommended  in the sva tutorial and user fora.
+The ComBat method described in \citet{Johnson2007}, and included in the sva package (\citealp{Leek2012}), can use either of the two above described approaches to estimate batch differences.
+In addition, it uses an empirical Bayes approach to avoid over-correction for batch effects, which is critical for small batches.
+Increasingly, the inclusion of group difference as a covariate when removing batch effects has been recommended, both in the sva tutorial and user fora, and ComBat has thus help popularise this approach.
 Based on actual use of ComBat by the authors and others, ourselves included, we suspect thus adjusted data are commonly treated as ``batch effect free'' in subsequent analyses.
@@
 \end{equation}
-where $i=1,\ldots,m$ are the different batches, $j=1,\ldots,M$ are different study groups that we wish to compare, and $r=1,\ldots,n_{ij}$ are the different samples within batch $i$ and group $j$.
+where $i=1,\ldots,m$ are the different batches, $j=1,\ldots,M$ are different study groups that we wish to compare, and $\epsilon_{ijr}\sim N(0,\sigma^2)$ are the error terms for samples $r=1,\ldots,n_{ij}$ within batch $i$ and group $j$.
 
@@
 $\delta_{ig}=1$.
-We will discuss the effect of empirical Bayes estimation of $\gamma_{ig}$ later, but our main argument is more easily made in the simpler context.
-Empirical Bayes may reduce the amount of batch adjustment in cases where batch effects are small or cannot be accurately estimated, e.g.
+Empirical Bayes will tend to shrink the estimates of  $\gamma_{ig}$, and thus the amount of batch adjustment, in cases where batch effects are small or cannot be accurately estimated, e.g.
 for small batches.
 However, in cases with large batches or substantial batch effects, it should differ little from the two-way ANOVA approach.
+We will therefore explain the problem in the simpler case where no empirical Bayes is applied, and later demonstrate that it also affects the ComBat approach.
 
@@
 A common ambition of batch effect adjustments is to remove batch differences in such a way that downstream analyses of the adjusted data may be done without further corrections for batches.
+We illustrate this using Figure~\ref{fig:boxplots}, where the first frame contains the ``true'' values with no batch differences, and the remaining frames show values with various levels of batch effects and batch effect corrections.
 
@@
 When comparing groups, the common value $\bar{Y}$ has no effect, and so this is equilvalent to zero-centring each batch.
-If the groups are unevenly represented in the different batches, the batch average $\bar{Y}_i$ will tend to capture group differences as well as batch effects:
+If the groups are unevenly represented in the different batches, the batch average $\bar{Y}_i$ will tend to capture, through $\bar\beta_i$, group differences as well as batch effects:
 \begin{equation}
@@
 \end{equation}
-Thus, batch centering will tend to reduce group differences in an unbalanced design, and thus reduce the power of downstream analyses.
+Thus, batch centering will tend to reduce group differences in an unbalanced design, and reduce the power of downstream analyses.
 By reducing the differences between some groups, i.e.
@@
 \begin{equation}
-\widetilde Y_{ijr}=Y_{ijr}-\hat\beta_j=\alpha+\beta_j+(\gamma_i-\hat\gamma_i)+\epsilon_{ijr}.
+\widetilde Y_{ijr}=Y_{ijr}-\hat\gamma_i=\alpha+\beta_j+(\gamma_i-\hat\gamma_i)+\epsilon_{ijr}.
 \end{equation}
@@
 Thus, while the aim is to remove spurious dependencies within batches, it may also induce new dependencies.
-In a balanced group--batch design, this has the same effect on all groups and thus does not influence group comparisons.
-In an unbalanced design, however, the estimated batch effects $\hat\gamma_i$ will correlate with the estimated group effects $\hat\beta_j$.
-\COMMENT{Assuming the batch adjusted values are without further batch differences corresponds to assuming that $\gamma_i=\hat\gamma_i$, and ignoring the estimation uncertainty and its impact on the correlated $\hat\beta_j$.
-It results in over-confidence in the accuracy of the group differences, which can have severe impact on downstream analyses.}{Denne forklaringen må jeg gjøre mye klarere.}
+The batch effect estimation errors will influence group effects in proportion to how well the group is represented in each batch:
+\begin{equation}
+\widetilde{Y}_{\textrm{-}j}
+=\frac{\sum_{i=1}^m\sum_{r=1}^{n_{ij}}\widetilde{Y}_{ijr}}{n_{\textrm{-}j}}
+=\alpha+\beta_j+\bar{\epsilon}_{\textrm{-}j}
+  -\sum_{i=1}^m \frac{n_{ij}}{n_{\textrm{-}j}}(\hat{\gamma}_i-\gamma_i),
+\quad
+n_{\textrm{-}j}=\sum_{i=1}^m n_{ij}.
+\end{equation}
+In a balanced group--batch design, the effect is the same all groups and thus does not influence group comparisons.
+In an unbalanced design, however, the estimation errors $\hat\gamma_i-\gamma_i$ will induce systematic differences between groups which, when ignored in downstream analyses, may lead to over-confidence in estimated group differences.
 
@@
 
-The undesired consequences of preserving group effects when correcting for batch effect is readily illustrated with a sanity check using random numbers .
-The documentation accompanying the sva library has a runnable example demonstrating how to adjust a data set with ComBat followed by a F-test.
-Swapping the real data with random numbers from a normal distribution (mean=0, sd=1), but otherwise following the instructions, will generate the p-value distribution shown in Figure~\ref{fig:p-sva}.
-\NOTE{Perhaps show Q-Q plot of $F$-statistics.} The skewed distribution is a indication that this approach may have a unintentional adverse effect.
-\NOTE{Tror vi dropper permutasjonstesten.}
+The undesired consequences of preserving group effects when correcting for batch effect is readily illustrated with a sanity check using random numbers.
+The documentation accompanying the sva library has a runnable example demonstrating how to adjust a data set with ComBat followed by an F-test.
+Replacing the real data with random numbers from a standard normal distribution (mean=0, sd=1), but otherwise following the instructions, will generate the p-value distribution shown in Figure~\ref{fig:sanity}.
 
-If the number of samples is increased, the empirical Bayes estimates used by ComBat will shrink the batch effect estimates and thus moderate the batch adjustments.
-However, if random data with different means in each batch, but no difference betwene groups, are drawn, the problem remains.
-\NOTE{Bør dette komme her eller senere?
-Evt.
-understøttes?}
+As can be seen from the Q-Q plot, the main effect of the procedure is to inflate the F-statistic by a factor.
+The size of this factor depends on how unbalanced the group--batch design is rather on the sample size in itself.
+Thus, increasing the sample size will not reduce the problem.
 
-%% FIGURE 1=fig:p-sva: Original position
+If the number of samples is increased and there are no actual batch effects present, the empirical Bayes estimates used by ComBat will shrink the batch effect estimates and thus moderate the batch adjustments.
+However, if batch differences are added which are not constant across all genes, the problem remains even as the samples size increases.
 
@@
 
-In the first experiment(\citealp{Towfic2014}), cells were treated with glatiramer acetate (a medicine for multiple sclerosis) or a generic and mRNA was measured using microarrays alongside control samples.
+An experiment by \citealp{Towfic2014} treated cells with either glatiramer acetate (a medicine for multiple sclerosis) or a generic drug, and mRNA was measured using microarrays alongside control samples.
 A batch effect correlating to the chip (Illumina WG-6\_V2, six samples per chip) was observed and adjusted for with ComBat, whereafter the data was tested for differentially expressed genes, yielding a list of 1000 genes (Table S5, \citealp{Towfic2014}).
@@
 When we re-analyzed their data without using ComBat, but instead blocked for batch effect in limma, only 9 genes were found ($\text{FDR}<0.05$).
-\NOTE{Also run analysis using a mixed random effects model which assumes random batch effects much like in the empirical Bayes model.} The above outlined sanity check with random numbers was also carried out.
-The distribution of p-values for different settings are shown in Figure~\ref{fig:p-towfic}.
+The above outlined sanity check with random numbers was also carried out.
+The distribution of p-values for different settings are shown in Figure~\ref{fig:comparison}a.
 Our conclusion is that most of the genes reported as differentially expressed in (\citealp{Towfic2014}) are false positives.
@@
 
-%% FIGURE 2=fig:p-towfic: Original position
-
 \subsubsection{Experiment 2}
 
-The second example is taken from the supporting information for the original ComBat article (\citealp{Johnson2007}) where it is denoted ``Data set 2''.
-Cells inhibited for the expression of the TAL1 gene were compared to controls on a microarray platform.
+The supporting information of the original ComBat article (\citealp{Johnson2007}) demonstrates the method on cells inhibited for the expression of the TAL1 gene were compared to controls on a microarray platform (denoted ``Data set 2'').
 The experiment was conducted on three different time points (used as batches) with a total of 30 samples and a fairly balanced batch--treatment set up (6-2, 3-4 and 9-6).
@@
 In addition, the sanity checks outlined above were performed.
-The distribution of P-values for different settings are shown in Figure~\ref{fig:p-values}b.
-In contrast to the results obtained for\citealp{Towfic2014} (Figure~\ref{fig:p-values}a), the P-value distributions for the alternative analysis does not indicate a huge difference.
-Nevertheless, we belive that  P-values are deflated for the ComBat adjusted analysis.
+The distribution of P-values for different settings are shown in Figure~\ref{fig:comparison}b.
+In contrast to the results obtained for \citealp{Towfic2014} (Figure~\ref{fig:comparison}a), the P-value distributions for the alternative analysis does not indicate a huge difference.
+Nevertheless, the P-values may still be somewhat deflated in the ComBat adjusted analysis.
 The R-code for our analysis can be downloaded from \href{https://github.com/vegardny/combat_tests.git}{Github} \NOTE{Update reference}.
-
-%% FIGURE 3=fig:p-johnson: Original position
-
-%% FIGURE 4=fig:pdist-johnson: Original position
 
@@
 
-\NB{Systematic underestimation of $P$-values combined with FDR is a particularly bad combination as FDR lends strength across tests.}
+The use of study group, or other form of outcome, as a covariate when estimating and removing batch effects is problematic if the data is treated as ``batch effect free'' in subsequent analyses.
+When the group--batch distribution is unbalanced, i.e.
+where batches do not have the same composition of groups, this will lead to deflated uncertainty estimates and over-confidence in the results.
+The problem is essentially independent of sample size, rather than being particular to small samples.
+
+The size and impact of the problem will depend greatly on how unbalanced the group--batch distribution is: if it is only moderately unbalanced, it need not be a concern, whereas in heavily unbalanced cases it may have a huge influence.
+The impact is also more likely to be notable when used to analyse a large number of cases, e.g.
+a large set of gene, followed by multiple testing corrections such as false discovery rate.
 
@@
 
-In the original ComBat article (\citealp{Johnson2007}) it is clear that the primary motivation behind ComBat was to employ an Emprical Bayes method for batch-effect removal to allow better handling of small batches.
-The feature of retaining group differences for unbalanced designs is optional and seems to be subordinate, only exemplified in the supplementary information.
+In the original ComBat article (\citealp{Johnson2007}), the main objective was to employ an emprical Bayes method to allow better handling of small batches.
+The inclusion of covariates, e.g.
+for retaining group differences for unbalanced designs, was optional and appeared subordinate, only exemplified in the supplementary information.
 However, over the years this feature became more important judging from advice given on user fora.
-When ComBat was later incorporated in the sva package (\citealp{Leek2012}), batch adjustment without covariates was no longer a default as specification of covariates became a requirement.
-Although specification of a null model was still possible, the help file specifies that this should contain ``outcome of interest and other covariates besides batch''.
+Later, when ComBat was incorporated in the sva package (\citealp{Leek2012}), specification of a covariate model became required: batch adjustment without covariates was no longer a default behaviour.
+Although specification of a null model was still possible, the help file specifies that covariates should contain ``outcome of interest and other covariates besides batch''.
 Thus, we expect this usage to have grown more common.
@@
 Our knowledge of the problem discussed in this article came through a typical use case when trying to adjust for batch effects in an unbalanced data set using ComBat.
-Upon realizing that the confidence on our group differences were exaggerated, the literature was searched for a better understanding of correct use and potential overseen limitations of ComBat.
-But the authors of ComBat and the sva package recommended our usage (\citealp{Johnson2007}, \citealp{Leek2012}).
-In addition, other works looking into the problem of batch effects were mostly recommending ComBat without much concern (\citealp{Kupfer2012}, \citealp{Kitchen2011}).
-A brief inquiry into  some of the articles citing ComBat (574 Google Scholar) revealed few problems, and their method descriptions regarding ComBat were mostly sparse, limited to one or two sentences.
-A further indication of their carefree use of this potentially devastating procedure was the frequent omittance of program parameters that were used, i.e batch labels or whether group labels were supplied as covariates.
-Often no effort was done in order to substantiate the existence of batch effects in their data, except for stating the presence of batches.
-The incorporation of the method into analysis pipelines (\CITE{16642009}, TCGA) and other packages (\CITE{23452776}, \CITE{21937664}) could make its usage even more trivial and parameters setting harder to perceive.
-Taken together we fear that many published results from data adjusted by ComBat are completely or partially false.
-We hope that our warning will enlighten the community and  reduce this unfortunate combination of methods.
+Upon realizing that the confidence on our group differences were exaggerated, we searched the literature for a better understanding of correct use and potential limitations of ComBat.
+However, the authors of ComBat and the sva package recommended including study group as a covariate as we had done.
+In addition, other studies investigating batch effects were mostly recommending ComBat without much concern (\citealp{Kupfer2012}, \citealp{Kitchen2011}).
+A brief inquiry into  some of the articles citing ComBat (574 on Google Scholar) revealed few reported problems, although their method descriptions regarding ComBat were mostly sparse, limited to one or two sentences.
+A further indication of the carefree use of the procedure was the frequent omittance of program parameters that were used, i.e batch labels or whether group labels were supplied as covariates.
+Often no effort was done in order to substantiate the existence of batch effects beyond stating the presence of batches.
+The incorporation of the method into analysis pipelines (\CITE{16642009}, TCGA) and other packages (\CITE{23452776}, \CITE{21937664}) could make its even more accessible, but its usage less transparent.
+We are concerned that many published results from data adjusted by ComBat with study group as covariate may be unreliable in light of our findings, and furthermore that the frequent lack of proper method description accompanying published results make it hard to judge they are affected or not.
+We hope that our warning will help caution the scientific community against this particular approach.
 
@@
 }
+\label{fig:sanity}
 \end{figure}
@@
 }
+\label{fig:comparison}
 \end{figure}

### 2014-06-27 tex/manuscript.tex
Updated manuscript, ready for proof-reading
Minor details left to fix: Figure 1 needs fixing, axis labels on Figure
2, references in "Motivation..." section.

@@
 % List of authors, with corresponding author marked by asterisk
-\author{VEGARD NYGAARD, EINAR ANDREAS RØDLAND, EIVIND HOVIG$^\ast$\\[4pt]
+\author{VEGARD NYGAARD, EINAR ANDREAS RØDLAND\\[4pt]
 % Author addresses
@@
 Norway}
+\\[8pt]
+EIVIND HOVIG$^\ast$\\[4pt]
+% Author addresses
+\textit{Department of Tumor Biology,
+Institute for Cancer Research,
+Oslo University Hospital HF - Radiumhospitalet,
+Montebello,
+0310 Oslo,
+Norway}
+\\[2pt]
+\textit{\NOTE{Andre adresser}}
 \\[2pt]
@@
 
-\NB{Er de neste tre avsnittene nødvendige eller kan de forkortes drastisk siden dette forklares grundigere siden?}
-
 A simple removal of batch effects can be achieved by subtracting the mean of the measurements in one batch from all measurements in that batch, i.e mean adjustment or one-way ANOVA adjustment as implemented in the method \texttt{pamr.batchadjust} from the \texttt{pamr} package in R.
@@
 
-The ComBat method described in \citet{Johnson2007}, and included in the sva package (\citealp{Leek2012}), can use either of the two above described approaches to estimate batch differences.
+The ComBat method described in \citet{Johnson2007}, and included in the \texttt{sva} package (\citealp{Leek2012}), can use either of the two above described approaches to estimate batch differences.
 In addition, it uses an empirical Bayes approach to avoid over-correction for batch effects, which is critical for small batches.
-Increasingly, the inclusion of group difference as a covariate when removing batch effects has been recommended, both in the sva tutorial and user fora, and ComBat has thus help popularise this approach.
+Increasingly, the inclusion of group difference as a covariate when removing batch effects has been recommended, both in the \texttt{sva} tutorial and user fora, and ComBat has thus help popularise this approach.
 Based on actual use of ComBat by the authors and others, ourselves included, we suspect thus adjusted data are commonly treated as ``batch effect free'' in subsequent analyses.
@@
 By reducing the differences between some groups, i.e.
-those found together in the same batches, it may also induce false differences between other groups.
+those found together in the same batches, it may also induce false differences between other groups, as is demonstrated in Figure~\ref{fig:boxplots}.
 
@@
 In an unbalanced design, however, the estimation errors $\hat\gamma_i-\gamma_i$ will induce systematic differences between groups which, when ignored in downstream analyses, may lead to over-confidence in estimated group differences.
+In Figure~\ref{fig:boxplots}, the effect of batch correction using group as covariate is shown with confidence intervals of the corrected data.
+For comparison, least square means estimates (R package \texttt{lsmeans}) are used to illustrate more appropriate confidence intervals that incorporate the uncertainties of the estimation errors.
 
@@
 The undesired consequences of preserving group effects when correcting for batch effect is readily illustrated with a sanity check using random numbers.
-The documentation accompanying the sva library has a runnable example demonstrating how to adjust a data set with ComBat followed by an F-test.
+The documentation accompanying the \texttt{sva} library has a runnable example demonstrating how to adjust a data set with ComBat followed by an F-test.
 Replacing the real data with random numbers from a standard normal distribution (mean=0, sd=1), but otherwise following the instructions, will generate the p-value distribution shown in Figure~\ref{fig:sanity}.
 
-As can be seen from the Q-Q plot, the main effect of the procedure is to inflate the F-statistic by a factor.
-The size of this factor depends on how unbalanced the group--batch design is rather on the sample size in itself.
+As can be seen from the QQ plot, the main effect of the procedure is to inflate the F-statistic by a factor.
+The size of this factor depends on how unbalanced the group--batch design is rather than on the sample size in itself.
 Thus, increasing the sample size will not reduce the problem.
@@
 \subsection{Explanation for the simple two-group comparison}
+\label{sec:twogroups}
 
@@
 \widehat{\Delta\beta}
-\sim\text{N}\left(\Delta\beta,\sigma^2\cdot\left(\frac{1}{n_A}+\frac{1}{n_B}\right)\right).
+\sim\text{N}\left(\Delta\beta,\frac{\sigma^2}{\nu_0}\right)
+\quad\text{where}\quad
+\nu_0=\frac{1}{\frac{1}{n_A}+\frac{1}{n_B}}.
 \end{equation}
@@
 \begin{equation}
-\frac{1}{\frac{1}{n_A}+\frac{1}{n_B}}=n\cdot\frac{n_A}{n}\cdot\frac{n_B}{n}
+\nu_0
+=\frac{1}{\frac{1}{n_A}+\frac{1}{n_B}}=n\cdot\frac{n_A}{n}\cdot\frac{n_B}{n}
 \le
@@
 
+In effect, $\nu$ represents the effective sample size in the unbalanced group--batch design, while $\nu_0$ represents the nominal sample size when batches are ignored.
+The ratio $\nu/\nu_0\ge1$ indicates to what extent the uncertainty of the estimated group difference is deflated by the two-step procedure.
+
 \subsection{Examples of undesired consequences}
@@
 
-An experiment by \citealp{Towfic2014} treated cells with either glatiramer acetate (a medicine for multiple sclerosis) or a generic drug, and mRNA was measured using microarrays alongside control samples.
+In an experiment by \citet{Towfic2014}, cells were treated with either glatiramer acetate (a medicine for multiple sclerosis) or a generic drug, and mRNA was measured using microarrays alongside control samples.
 A batch effect correlating to the chip (Illumina WG-6\_V2, six samples per chip) was observed and adjusted for with ComBat, whereafter the data was tested for differentially expressed genes, yielding a list of 1000 genes (Table S5, \citealp{Towfic2014}).
@@
 The distribution of p-values for different settings are shown in Figure~\ref{fig:comparison}a.
-Our conclusion is that most of the genes reported as differentially expressed in (\citealp{Towfic2014}) are false positives.
+Our conclusion is that most of the genes reported as differentially expressed in \citet{Towfic2014} are false positives.
 This example is a sort of ``worst case'' scenario for applying ComBat, since it both has a very unbalanced batch--group design and a a priori assumption of no difference.
@@
 The distribution of P-values for different settings are shown in Figure~\ref{fig:comparison}b.
-In contrast to the results obtained for \citealp{Towfic2014} (Figure~\ref{fig:comparison}a), the P-value distributions for the alternative analysis does not indicate a huge difference.
+In contrast to the results obtained for \citet{Towfic2014} (Figure~\ref{fig:comparison}a), the P-value distributions for the alternative analysis does not indicate a huge difference.
 Nevertheless, the P-values may still be somewhat deflated in the ComBat adjusted analysis.
@@
 where batches do not have the same composition of groups, this will lead to deflated uncertainty estimates and over-confidence in the results.
-The problem is essentially independent of sample size, rather than being particular to small samples.
+The problem is essentially independent of sample size.
 
 The size and impact of the problem will depend greatly on how unbalanced the group--batch distribution is: if it is only moderately unbalanced, it need not be a concern, whereas in heavily unbalanced cases it may have a huge influence.
-The impact is also more likely to be notable when used to analyse a large number of cases, e.g.
-a large set of gene, followed by multiple testing corrections such as false discovery rate.
+The impact is also more likely to be notable when used to analyse a large number of features, e.g.
+a large set of gene, followed by multiple testing corrections such as false discovery rate as the effect is more pronounced for more extreme values.
 
@@
 However, over the years this feature became more important judging from advice given on user fora.
-Later, when ComBat was incorporated in the sva package (\citealp{Leek2012}), specification of a covariate model became required: batch adjustment without covariates was no longer a default behaviour.
+Later, when ComBat was incorporated into the \texttt{sva} package (\citealp{Leek2012}), specification of a covariate model became required: batch adjustment without covariates was no longer a default behaviour.
 Although specification of a null model was still possible, the help file specifies that covariates should contain ``outcome of interest and other covariates besides batch''.
@@
 Our knowledge of the problem discussed in this article came through a typical use case when trying to adjust for batch effects in an unbalanced data set using ComBat.
-Upon realizing that the confidence on our group differences were exaggerated, we searched the literature for a better understanding of correct use and potential limitations of ComBat.
-However, the authors of ComBat and the sva package recommended including study group as a covariate as we had done.
+Upon realizing that the confidence in the estimated group differences were exaggerated, we searched the literature for a better understanding of correct use and potential limitations of ComBat.
+However, the authors of ComBat and the \texttt{sva} package recommended including study group as a covariate as we had done.
 In addition, other studies investigating batch effects were mostly recommending ComBat without much concern (\citealp{Kupfer2012}, \citealp{Kitchen2011}).
@@
 The incorporation of the method into analysis pipelines (\CITE{16642009}, TCGA) and other packages (\CITE{23452776}, \CITE{21937664}) could make its even more accessible, but its usage less transparent.
-We are concerned that many published results from data adjusted by ComBat with study group as covariate may be unreliable in light of our findings, and furthermore that the frequent lack of proper method description accompanying published results make it hard to judge they are affected or not.
+We are concerned, in light of our findings, that many published results from data adjusted by ComBat with study group as covariate may be unreliable.
+Furthermore, the frequent lack of proper method description accompanying published results makes it hard to judge if they are affected or not.
 We hope that our warning will help caution the scientific community against this particular approach.
@@
 
-For an investigator facing an unbalanced data set with batch effects, our advice is to adjust for batch inside the statistical test and avoid the two step procedure outlined above.
-If this is not possible, then applying the method described in this article should only be done with great caution and downstream confidence estimates must be treated with suspicion.
+For an investigator facing an unbalanced data set with batch effects, our primary advice would be to adjust for batch inside the statistical test and avoid the two step procedure outlined above.
+If this is not possible, batch correction using outcome as covariate should only be done with great caution and downstream confidence estimates must be treated with suspicion.
 
-Knowing that adjusting for batch effects while preserving the group difference may lead to varying degree of false results, to what extent can an investigator trust a result from a work applying such a method?
+Knowing that adjusting for batch effects while preserving the group difference may lead to varying degree of false results, to what extent can an investigator trust published results which has applied such a method?
 Essentially, when the batch--group configuration is balanced, or group difference is ignored  (i.e.
@@
 For other cases, a re-analysis without using this approach is the most rigorous path.
-However, this thoroughness is not feasible if the downstream  analysis can not adjust for batch effects by it self.
+However, such a re-analysis is not feasible if the downstream  analysis can not adjust for batch effects by itself.
 To reach a reliable result, batch effects need to be handled in some way or another.
 To make matters worse, a re-analysis relies on the availability of the raw data and a description of processing and analysis steps taken in the original work.
-Even when this is available, the necessary bioinformatic skills and work hours could still be in short supply.
-For such situations, a superficial assessment can be performed, taking special note of batches were groups of interest are near missing and how likely a group difference is.
-In essence asking if the balanced parts (effective sample size?) of the data has enough power to detect the presumed effects and if this is the case (\citealp{Johnson2007}, \CITE{24584070}), treat the results more like an ordered list with the most likely true positives on top while de-emphasizing the somewhat deflated p-values .
-In contrast, if biological knowledge suggest that a group effect is unlikely  (\CITE{24391845}, \CITE{18414638}?, \CITE{21731603}, \CITE{23630272}), an intermediate lack of batch-group balance  could lead to a mostly false result.
+Even when this is available, the necessary statistics and bioinformatics skills and work hours could still be in short supply.
+
+A superficial assessment of the reliability of results based on batch corrected data with group differences retain can often be made.
+The effective sample size, $\nu$, may be compared against the nominal sample size, $\nu_0$, as in section~\label{sec:twogroups} for the case of two groups: similar computations, although more complicated, can be made for multiple groups, but the two-group assessment should still give a fair idea of the reliability of pairwise group comparisons.
+A major concern would be batches were either of the groups of interest are missing or strongly under-represented, which would contribute to the nominal sample size, but not to the effective sample size.
+
+Alternatively, one may treat the results more like an ordered list of candidates, with the most likely true positives on top, de-emphasizing the somewhat deflated p-values.
+This would, however, be hypothesis generating rather than hypothesis testing.
+While investigators should always assess the extent to which findings make biological or clinical sense, this is particularly true when the statistical assessment may be unreliable.
+
+Finally, we would like to emphasise the importance of proper description of how data has been prepared for analysis, and what corrections and adjustments have been made.
+In cases where data preparation is done prior to and separate from the data analysis, as is now often the case, this is of particular importance as artifacts may be introduced in the data preparation which could influence the reliability of downstream analyses.
 
@@
 
-The data and scripts used to generate the results in this work, i.e figures, are made available at https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-figures.git.
-Additional analyses were also performed, but considered excessive for the current  article.
-However, some readers might find this very interesting and we provide it as an extra resource at https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-reports.git
+The data and scripts used to generate the results in this work, i.e figures, are available at
+\href{https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-figures.git}{GitHub}.
+Additional analyses, performed but not included in the article, may be found at
+\href{ https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-reports.git}{GitHub}.
 
-\NOTE{vi boer klare aa tilfredstille kravene i http://biostatistics.oxfordjournals.org/content/10/3/405.full}
-\NOTE{For aa tilfredstille kravene skal data og script vare available.
+\NOTE{Vi bør klare aa tilfredstille kravene i \href{http://biostatistics.oxfordjournals.org/content/10/3/405.full}{Biostatistics}.
+For å tilfredstille kravene skal data og script vare available.
 Jeg mener vi klarer dette.
-paa githuben er ikke data fra towfic, men det blir lastet av scriptet fra GEO.
+På githuben er ikke data fra Towfic, men det blir lastet av scriptet fra GEO.
 Ikke beskrevet her men i scriptet.
-Videre i forklaringen til biostatistics staar det at script og data skal bli lagt paa deres sider, som vi kan hvis de vil}
+Videre i forklaringen til Biostatistics står det at script og data skal bli lagt på deres sider, som vi kan hvis de vil.}
 
@@
 \caption{Simulated data from four study groups was generated where groups 1 and 2 have lower means than groups 3 and 4.
-These were placed in three different batches with batch effect added.
-Values and boxes showing mean and two standard errors of the mean are displayed for data without batch effects, after adding batch effects, after batch centering, and after ANOVA based batch centering.
-The last frame shows the least squares estimates of the group means from a two-way ANOVA analysis with 2 standard errors.
+These were placed in three different batches with batch effects added.
+Values and boxes showing group mean with 95\%~confidence intervals are displayed for data without batch effects, after adding batch effects, after batch centering, and after two-way ANOVA based batch centering.
+The last frame shows the least squares estimates of the group means from a two-way ANOVA.
 This case, design and effects, was selected to illustrate the spurious effects that may arise.}
@@
 {
-A sanity checks where the recommended use of ComBat fails.
-Adapted from the user guide in the sva package.
-Real data are substituted with random numbers from a normal distribution, but the batch-group design is retained.
-ComBat is applied followed by a F-test.
+This is a sanity check where the recommended use of ComBat fails, adapted from the user guide in the \texttt{sva} package.
+Real data are substituted with random numbers from a normal distribution, but the batch--group design is retained.
+ComBat is applied followed by an F-test.
 a) P-value distribution
@@
 {
-Three analyses of two works where batch effects were adjusted for with ComBat.
-First as described in the original works with ComBat on real data.
+Three analyses of two published data sets where batch effects were adjusted for with ComBat.
+First, analysed as described on the real data using ComBat.
 Secondly, with ComBat but with random numbers instead of real data.
-Thirdly without ComBat and instead blocking batch with limma on real data.
-a) Re-analysis of  \citealp{Towfic2014}, glatiramer acetate vs. generic 
-b) Re-analysis of "Data set 2" \citealp{Johnson2007},  TAL1 inhibition vs. control
+Thirdly, instead of ComBat, analyses of the real data applied \texttt{limma} blocking by batch.
+a) Re-analysis of \citet{Towfic2014}, glatiramer acetate vs. generic 
+b) Re-analysis of "Data set 2" \citet{Johnson2007},  TAL1 inhibition vs. control
 }

### 2014-07-16 tex/manuscript.tex
Updated manuscript, figures, R scripts
The R script helperfunctions.r was updated. Need to check if this may
have affected compatibility with reanalysis plots.

@@
 Often no effort was done in order to substantiate the existence of batch effects beyond stating the presence of batches.
-The incorporation of the method into analysis pipelines (\CITE{16642009}, TCGA) and other packages (\CITE{23452776}, \CITE{21937664}) could make its even more accessible, but its usage less transparent.
+The incorporation of the method into analysis tools and pipelines, like \COMMENT{TCGA}{does not seem to offer ComBat with covariates} and GenePattern (\citealp{Reich2006}), and other packages like \texttt{sva} (\citealp{Leek2012}) and \COMMENT{inSilicoDb (\citealp{Taminau2011})}{doesn't seem to offer ComBat with covariates} could make it even more accessible, but its use less transparent.
 We are concerned, in light of our findings, that many published results from data adjusted by ComBat with study group as covariate may be unreliable.
@@
 The last frame shows the least squares estimates of the group means from a two-way ANOVA.
-This case, design and effects, was selected to illustrate the spurious effects that may arise.}
+This case, design and effects, was selected to illustrate the spurious effects that may arise.
+\NOTE{Replace ComBat with ANOVA in plot.}}
 \label{fig:boxplots}

### 2014-07-16 commonscripts/boxplot_function.r

@@ -1,0 +1,97 @@
+# Functions for producing a boxplot
+
+adhoc.palette =  c("blue", "red", "darkgreen", "darkorange")
+adhoc.pch = c(1, 2, 3)
+adhoc.cex=3
+adhoc.legendcex=adhoc.cex
+adhoc.labcex=adhoc.cex
+
+# Boxplot component: values and intervals
+adhocboxplot = function(y, grouplabels, batchlabels, addlegend=FALSE, ylim=NULL, xlab="x",figureletter=NA,cex=adhoc.cex,labelcex=adhoc.labcex, legendcex=adhoc.legendcex)
+{
+  colpalette = adhoc.palette
+  pchselection = adhoc.pch
+  x = as.numeric(batchlabels)
+  lwd=2
+  
+  # add extra x for better visual separation
+  for(b in unique(batchlabels)) {
+    separation = 0.15
+    xoffset=0
+    for(g in unique(grouplabels[batchlabels==b])) {
+      x[batchlabels==b & grouplabels==g] = x[batchlabels==b & grouplabels==g] + xoffset
+      xoffset = xoffset+separation
+    }
+  }
+  
+  xlim = c(0, length(unique(batchlabels))+2)
+  plot(x, y, col=colpalette[grouplabels], pch=pchselection[batchlabels], 
+       xlim=xlim, lwd=lwd, cex=cex, ylim=ylim, xlab=xlab, ylab="", axes=FALSE, cex.lab =labelcex)
+
+  #add boxplot
+  groupnames = sort(unique(grouplabels))
+  xoffset=length(unique(batchlabels))+1
+  boxwidth=0.35
+  boxseparation = 0.25
+  for(g in groupnames) {
+    g=factor(g,levels=levels(groupnames))
+    m = mean(y[grouplabels==g])
+    ci = t.test(y[grouplabels==g])[["conf.int"]]
+    xleft = xoffset
+    ybottom = ci[1]
+    xright = xleft + boxwidth
+    ytop = ci[2]
+    rect(xleft, ybottom, xright, ytop, border=colpalette[g], lwd=lwd)
+    lines(c(xoffset,xright), c(m,m), col=colpalette[g], lwd=lwd)
+    xoffset = xoffset+boxseparation
+  }
+	if(!is.na(figureletter)) {
+		figureletter=paste("(", figureletter, ")", sep="")
[…]

### 2014-07-16 commonscripts/helperfunctions-vegard.r

@@ -1,0 +1,116 @@
+
+# Creates a sampleannotation data.frame based on a list of batches with a vector of treatmentcounts.
+# Used when creating artificial data.
+createsampleannotation = function( treatment_in_batches)
+{
+	batches = vector()
+	treatments = vector()
+	for(i in 1:length(treatment_in_batches))
+	{
+		thisbatch = i
+		for(s in 1:length(treatment_in_batches[[i]]))  
+		{
+			batches = c(batches, rep(thisbatch, treatment_in_batches[[i]][s]))
+			treatments = c(treatments, rep( s, treatment_in_batches[[i]][s]))
+		}
+	}
+	sampleannotation = data.frame(id=paste("sample", 1:length(treatments), sep=""),    treatment = treatments, batch=batches)
+	return(sampleannotation)
+}
+
+
+# plots 3 histograms from p-value distibutions on top on each other.
+# draws a line instead of bars for clarity.
+# binsize is 0.01
+oldadhocpvalueplot = function(realcombatp, reallimmap, randomp, main="P-values", xrange=1:25)
+{
+  thiscolors = c("red", "blue", "black")
+  
+  a = hist(realcombatp, breaks=100, plot=F)$counts[xrange]
+  b = hist(reallimmap, breaks=100, plot=F)$counts[xrange]
+  c = hist(randomp, breaks=100, plot=F)$counts[xrange]
+  ylim=c(0, max(c(a,b,c)))
+  # reproduced ComBat + limma
+  plot((xrange)/100, a , ylim=ylim,
+       main=main, xlab="p-value", ylab="frequency",
+       type="l", lwd=2, col=thiscolors[1])
+  
+  # batch handled in limma
+  lines((xrange)/100, b,
+        col=thiscolors[2], lwd=2)
+  
+  # random ComBat + limma
+  lines((xrange)/100, c,
+        col=thiscolors[3], lwd=2)
+  
+  legend("topright", text.col=thiscolors,
+         legend=c("Real data, ComBat adjusted",
+                  "Real data, batch handled by Limma",
+                  "Random data, ComBat adjusted")) 
+}
+
+
[…]

### 2014-07-16 commonscripts/helperfunctions.r

@@ -1,16 +1,16 @@
 
-# Creates a sampleannotation data.frame based on a list of batches with a vector of treatmentcounts.
-# Used when creating artificial data.
 createsampleannotation = function( treatment_in_batches)
 {
 	batches = vector()
 	treatments = vector()
 	for(i in 1:length(treatment_in_batches))
 	{
+		#thisbatch = paste("batch", i, sep="")
 		thisbatch = i
-		for(s in 1:length(treatment_in_batches[[i]]))  
+		for(s in 1:length(treatment_in_batches[[i]]))
 		{
 			batches = c(batches, rep(thisbatch, treatment_in_batches[[i]][s]))
+			#treatments = c(treatments, rep(paste("treatment", s, sep=""), treatment_in_batches[[i]][s]))
 			treatments = c(treatments, rep( s, treatment_in_batches[[i]][s]))
 		}
 	}
@@ -19,10 +19,280 @@ createsampleannotation = function( treatment_in_batches)
 }
 
 
-# plots 3 histograms from p-value distibutions on top on each other.
-# draws a line instead of bars for clarity.
-# binsize is 0.01
-oldadhocpvalueplot = function(realcombatp, reallimmap, randomp, main="P-values", xrange=1:25)
+
+
+createrandomdata = function(ngenes, sa, mean=0, sd=1)
+{
+	ret = matrix( rnorm(dim(sa)[1]*ngenes, mean=mean, sd=sd), ncol=dim(sa)[1], nrow=ngenes)
+	dimnames(ret) = list( paste("gene", 1:ngenes, sep=""),  sa$id)
+	return(ret)
+}
+
+addbatcheffect = function(df, batches, thismean=1, thissd=1)
+{
+	ret = df
+	for(i in 1:length(unique(batches)))
+	{
+		for(s in 1:dim(df)[1])
+		{
+			thisgenesbatcheffect = rnorm(1,mean=thismean, sd=thissd)
+			a=batches==i
+			#ret[s, a] = ret[s, a] + rnorm(length(ret[s, a]), mean=thisgenesbatcheffect)	
+			ret[s, a] = ret[s, a] + thisgenesbatcheffect
+		}
+	}
+	return(ret)
+}
+
+# 
+addconditioneffect = function(df, labels, ndiffgenes, thismean=1, betweensd=1, insidesd=1, affectedconditions)
+{
+	ret = df
[…]

### 2014-07-16 overviewplot-vegard.r

@@ -1,0 +1,172 @@
+
+
+# Script to create Figure 1
+
+library(sva)
+library(lsmeans)
+
+adhocboxplot = function(y, grouplabels, batchlabels, addlegend=FALSE, ylim=NULL, xlab="x", figureletter=NA)
+{
+  #x = batchlabels + grouplabels/8
+  colpalette = c("blue", "red", "darkgreen", "magenta")
+  pchselection = c(1, 2, 3)
+  x = batchlabels
+  lwd=2
+  
+  # add extra x for better visual separation
+  for(b in unique(batchlabels))
+  {
+    
+    separation = 0.15
+    xoffset=0    
+    for(g in unique(grouplabels[batchlabels==b]))
+    {
+      x[batchlabels==b & grouplabels==g] = x[batchlabels==b & grouplabels==g] + xoffset
+      xoffset = xoffset+separation
+    }
+  }
+  
+  xlim = c(0, length(unique(batchlabels))+2)
+  plot(x, y, col=colpalette[grouplabels], pch=pchselection[batchlabels], 
+       xlim=xlim, lwd=lwd, cex=2, ylim=ylim, xlab=xlab, ylab="", axes=FALSE, cex.lab =2)
+  
+  #add boxplot
+  groupnames = unique(grouplabels)
+  groupnames = groupnames[order(groupnames)]
+  xoffset=length(unique(batchlabels))+1
+  boxwidth=0.35
+  boxseparation = 0.25
+  for(g in groupnames)
+  { 
+
+    m = mean(y[grouplabels==g])
+    ci = t.test(y[grouplabels==g])[["conf.int"]]
+    xleft = xoffset
+    ybottom = ci[1]
+    xright = xleft + boxwidth
+    ytop = ci[2]
+    rect(xleft, ybottom, xright, ytop, border=colpalette[g], lwd=lwd)
+    lines(c(xoffset,xright), c(m,m), col=colpalette[g], lwd=lwd)
+    xoffset = xoffset+boxseparation
+  }
+  legendcex=2
+  if(addlegend)
+  {
+    
+    legend("topright", legend=paste("Group ", groupnames, sep=""), 
+           text.col=colpalette[groupnames], bty="n", cex=legendcex)
+    legend("bottomright", legend=paste("Batch ", unique(batchlabels), sep=""),
+           pch=pchselection[unique(batchlabels)], bty="n", cex=legendcex)
+  }
+  if(!is.na(figureletter))
+  {
[…]

### 2014-07-16 overviewplot.r

@@ -1,126 +1,11 @@
 
-
-# Script to create Figure 1
-
+### Figure of box-plots and confidence intervals
 
 library(sva)
+library(limma)
 library(lsmeans)
- 
-adhocboxplot = function(y, grouplabels, batchlabels, addlegend=FALSE, ylim=NULL, xlab="x", figureletter=NA)
-{
-  #x = batchlabels + grouplabels/8
-  colpalette = c("blue", "red", "darkgreen", "magenta")
-  pchselection = c(1, 2, 3)
-  x = batchlabels
-  lwd=2
-  
-  # add extra x for better visual separation
-  for(b in unique(batchlabels))
-  {
-    
-    separation = 0.15
-    xoffset=0    
-    for(g in unique(grouplabels[batchlabels==b]))
-    {
-      x[batchlabels==b & grouplabels==g] = x[batchlabels==b & grouplabels==g] + xoffset
-      xoffset = xoffset+separation
-    }
-  }
-  
-  xlim = c(0, length(unique(batchlabels))+2)
-  plot(x, y, col=colpalette[grouplabels], pch=pchselection[batchlabels], 
-       xlim=xlim, lwd=lwd, cex=2, ylim=ylim, xlab=xlab, ylab="", axes=FALSE, cex.lab =2)
-  
-  #add boxplot
-  groupnames = unique(grouplabels)
-  groupnames = groupnames[order(groupnames)]
-  xoffset=length(unique(batchlabels))+1
-  boxwidth=0.35
-  boxseparation = 0.25
-  for(g in groupnames)
-  { 
 
-    m = mean(y[grouplabels==g])
-    ci = t.test(y[grouplabels==g])[["conf.int"]]
-    xleft = xoffset
-    ybottom = ci[1]
-    xright = xleft + boxwidth
-    ytop = ci[2]
-    rect(xleft, ybottom, xright, ytop, border=colpalette[g], lwd=lwd)
-    lines(c(xoffset,xright), c(m,m), col=colpalette[g], lwd=lwd)
-    xoffset = xoffset+boxseparation
-  }
-  legendcex=2
-  if(addlegend)
-  {
-    
-    legend("topright", legend=paste("Group ", groupnames, sep=""), 
-           text.col=colpalette[groupnames], bty="n", cex=legendcex)
-    legend("bottomright", legend=paste("Batch ", unique(batchlabels), sep=""),
[…]

### 2014-07-16 reanalysis/Leek2012/figurescript_leek.r

@@ -39,12 +39,15 @@ dev.off()
 print( paste("Figure created; ",figfile ))
 
 
+source("../../commonscripts/helperfunctions.r")
 figfile = paste(getwd(), "/leekqqplot.png", sep="")
 png(file=figfile)
-source("../../commonscripts/helperfunctions.r")
 test = Ftest(combat_edata,mod,mod0)
 Fquant=qf(ppoints(nrow(combat_edata)),test$df1,test$df0)
-qqplot(Fquant,test$F, ylim=c(0,max(test$F)), xlim=c(0,max(test$F)), main="(b) QQ-plot of F-statistics")
+qqplot(Fquant,test$F, ylim=c(0,max(test$F)), xlim=c(0,max(test$F)),
+       xlab='Quantiles of F distribution',
+       ylab='F statistics from analyses',
+       main="(b) QQ-plot of F-statistics")
 abline(0,1,lwd=2, lty=2)
 dev.off()
 print( paste("Figure created; ",figfile ))

### 2014-07-17 tex/manuscript.tex
Updated boxplot with ANOVA
Replaced ComBat centering with two-way ANOVA (using limma) in Figure 1.
Updated manuscript text accordingly.

@@
 These were placed in three different batches with batch effects added.
-Values and boxes showing group mean with 95\%~confidence intervals are displayed for data without batch effects, after adding batch effects, after batch centering, and after two-way ANOVA based batch centering.
+Values and boxes showing group mean with 95\%~confidence intervals are displayed for data without batch effects, after adding batch effects, after batch centering, and after two-way ANOVA based batch centering (using \texttt{limma}).
 The last frame shows the least squares estimates of the group means from a two-way ANOVA.
 This case, design and effects, was selected to illustrate the spurious effects that may arise.
-\NOTE{Replace ComBat with ANOVA in plot.}}
+}
 \label{fig:boxplots}

### 2014-07-17 overviewplot.r

@@ -34,6 +34,7 @@ matrix_meancenter[,batch ==2] =  matrix_meancenter[,batch ==2] - rowMeans( matri
 matrix_meancenter[,batch ==3] =  matrix_meancenter[,batch ==3] - rowMeans( matrix_meancenter[,batch ==3])
 mod = model.matrix(~group)
 matrix_combat = ComBat(dat=matrix_conditionbatch, batch=batch,mod=mod, par.prior=TRUE, prior.plots=FALSE)
+matrix_limma = removeBatchEffect(matrix_conditionbatch, batch=batch, design=mod)
 
 #fit_anova = lmFit(matrix_conditionbatch,model.matrix(~group+batch))
 #fit_eb = eBayes(fit_anova)
@@ -45,7 +46,6 @@ means_lm=lsmeans(fit_lm,~group)
 summary(means_lm)
 contrast(means_lm,"pairwise")
 
-source("commonscripts/boxplot_function.r")
 figfilename = file.path( getwd(), "boxplots")
 #figfile = paste( figfilename, ".png", sep=""); png(file = figfile, width=1600, height=800)
 figfile = paste( figfilename, ".pdf", sep=""); pdf(file =figfile, width=24, height=12)
@@ -56,7 +56,8 @@ ylim = c(min(alldata), max(alldata))
 adhocboxplot(matrix_condition[index,], group, batch, ylim=ylim, xlab="Without batch effects", figureletter="a")
 adhocboxplot(matrix_conditionbatch[index,], group, batch, ylim=ylim, xlab="With batch effects added", figureletter="b")
 adhocboxplot(matrix_meancenter[index,], group, batch, ylim=ylim, xlab="Zero-centered per batch", figureletter="c")
-adhocboxplot(matrix_combat[index,], group, batch, ylim=ylim, xlab="ANOVA centered values (for tiden combat)", figureletter="d")
+#adhocboxplot(matrix_combat[index,], group, batch, ylim=ylim, xlab="ComBat adjusted data", figureletter="d")
+adhocboxplot(matrix_limma[index,], group, batch, ylim=ylim, xlab="ANOVA centered values", figureletter="d")
 adhocboxplot2(means_lm, group, batch, ylim=ylim, xlab="ANOVA estimates", figureletter="e")
 adhoclegend(group,batch)
 dev.off()

### 2014-07-17 commonscripts/helperfunctions.r
Updated R scripts
Ensured that all functions are present and also runs under Windows.

@@ -320,3 +320,36 @@ adhocpvalueplot = function(realcombatp, reallimmap, randomp, main="P-values", xr
 }
 
 
+# Run F test and store results in an object
+Ftest = function (dat, mod, mod0,adjust=1)
+{
+	obj=new.env()
+	class(obj)="Ftest"
+	n <- dim(dat)[2]
+	m <- dim(dat)[1]
+	df1 <- dim(mod)[2]
+	df0 <- dim(mod0)[2]
+	p <- rep(0, m)
+	Id <- diag(n)
+	resid <- dat %*% (Id - mod %*% solve(t(mod) %*% mod) %*% t(mod))
+	rss1 <- rowSums(resid * resid)
+	rm(resid)
+	resid0 <- dat %*% (Id - mod0 %*% solve(t(mod0) %*% mod0) %*%  t(mod0))
+	rss0 <- rowSums(resid0 * resid0)
+	rm(resid0)
+	fstats <- ((rss0 - rss1)/(df1 - df0))/(rss1/(n - df1))
+	if (is.null(adjust)) {
+		adjust=mean(rss1/(n - df1))/mean((rss0 - rss1)/(df1 - df0))
+	}
+	fadj=fstats*adjust
+	p <- 1 - pf(fadj, df1 = (df1 - df0), df2 = (n - df1))
+	obj$n1=n; obj$n0=m; obj$df1=df1-df0; obj$df0=n-df1;
+	obj$F0=fstats; obj$F=fadj
+	obj$MS1=(rss0-rss1)/(df1-df0)
+	obj$MS0=rss1/(n-df1)
+	obj$P=p
+	obj$adjust=adjust
+	return(obj)
+}
+
+

### 2014-07-17 reanalysis/Towfic2014/figurescript_towfic.r

@@ -7,7 +7,7 @@ downloaddata=TRUE
 set.seed(100)
 
 
-includelibs = c("Biobase", "GEOquery", "sva", "limma")
+includelibs = c("Biobase", "sva", "limma")
 lapply(includelibs, require, character.only=T)
 source("../../commonscripts/helperfunctions.r")
 source("helperfunctions_towfic.r")

### 2014-07-17 reanalysis/Towfic2014/helperfunctions_towfic.r

@@ -27,7 +27,7 @@ loadtowfic = function(downloaddata=TRUE)
     download.file(url="http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file",
                   destfile=temp, mode = "wb")
     tardirtemp = tempfile()
-    untar(temp, exdir = tardirtemp)
+    untar(temp, exdir = tardirtemp,tar="internal")
     rawannotation = read.table(paste(tardirtemp ,"/GPL6887_MouseWG-6_V2_0_R3_11278593_A.txt.gz", sep=""), 
                                sep="\t", header=TRUE, stringsAsFactors=FALSE, 
                                skip=8, comment.char="", quote="", fill=TRUE)

### 2014-07-17 commonscripts/helperfunctions-vegard.r
Returned to original helperfunctions

@@ -1,116 +1,0 @@
-
-# Creates a sampleannotation data.frame based on a list of batches with a vector of treatmentcounts.
-# Used when creating artificial data.
-createsampleannotation = function( treatment_in_batches)
-{
-	batches = vector()
-	treatments = vector()
-	for(i in 1:length(treatment_in_batches))
-	{
-		thisbatch = i
-		for(s in 1:length(treatment_in_batches[[i]]))  
-		{
-			batches = c(batches, rep(thisbatch, treatment_in_batches[[i]][s]))
-			treatments = c(treatments, rep( s, treatment_in_batches[[i]][s]))
-		}
-	}
-	sampleannotation = data.frame(id=paste("sample", 1:length(treatments), sep=""),    treatment = treatments, batch=batches)
-	return(sampleannotation)
-}
-
-
-# plots 3 histograms from p-value distibutions on top on each other.
-# draws a line instead of bars for clarity.
-# binsize is 0.01
-oldadhocpvalueplot = function(realcombatp, reallimmap, randomp, main="P-values", xrange=1:25)
-{
-  thiscolors = c("red", "blue", "black")
-  
-  a = hist(realcombatp, breaks=100, plot=F)$counts[xrange]
-  b = hist(reallimmap, breaks=100, plot=F)$counts[xrange]
-  c = hist(randomp, breaks=100, plot=F)$counts[xrange]
-  ylim=c(0, max(c(a,b,c)))
-  # reproduced ComBat + limma
-  plot((xrange)/100, a , ylim=ylim,
-       main=main, xlab="p-value", ylab="frequency",
-       type="l", lwd=2, col=thiscolors[1])
-  
-  # batch handled in limma
-  lines((xrange)/100, b,
-        col=thiscolors[2], lwd=2)
-  
-  # random ComBat + limma
-  lines((xrange)/100, c,
-        col=thiscolors[3], lwd=2)
-  
-  legend("topright", text.col=thiscolors,
-         legend=c("Real data, ComBat adjusted",
-                  "Real data, batch handled by Limma",
-                  "Random data, ComBat adjusted")) 
-}
-
-
[…]

### 2014-07-17 commonscripts/helperfunctions.r

@@ -1,16 +1,16 @@
 
+# Creates a sampleannotation data.frame based on a list of batches with a vector of treatmentcounts.
+# Used when creating artificial data.
 createsampleannotation = function( treatment_in_batches)
 {
 	batches = vector()
 	treatments = vector()
 	for(i in 1:length(treatment_in_batches))
 	{
-		#thisbatch = paste("batch", i, sep="")
 		thisbatch = i
-		for(s in 1:length(treatment_in_batches[[i]]))
+		for(s in 1:length(treatment_in_batches[[i]]))  
 		{
 			batches = c(batches, rep(thisbatch, treatment_in_batches[[i]][s]))
-			#treatments = c(treatments, rep(paste("treatment", s, sep=""), treatment_in_batches[[i]][s]))
 			treatments = c(treatments, rep( s, treatment_in_batches[[i]][s]))
 		}
 	}
@@ -19,280 +19,10 @@ createsampleannotation = function( treatment_in_batches)
 }
 
 
-
-
-createrandomdata = function(ngenes, sa, mean=0, sd=1)
-{
-	ret = matrix( rnorm(dim(sa)[1]*ngenes, mean=mean, sd=sd), ncol=dim(sa)[1], nrow=ngenes)
-	dimnames(ret) = list( paste("gene", 1:ngenes, sep=""),  sa$id)
-	return(ret)
-}
-
-addbatcheffect = function(df, batches, thismean=1, thissd=1)
-{
-	ret = df
-	for(i in 1:length(unique(batches)))
-	{
-		for(s in 1:dim(df)[1])
-		{
-			thisgenesbatcheffect = rnorm(1,mean=thismean, sd=thissd)
-			a=batches==i
-			#ret[s, a] = ret[s, a] + rnorm(length(ret[s, a]), mean=thisgenesbatcheffect)	
-			ret[s, a] = ret[s, a] + thisgenesbatcheffect
-		}
-	}
-	return(ret)
-}
-
-# 
-addconditioneffect = function(df, labels, ndiffgenes, thismean=1, betweensd=1, insidesd=1, affectedconditions)
-{
-	ret = df
-	if(ndiffgenes>0)
-	{
-		for(i in 1:length(affectedconditions))
-		{
-			for(s in 1:ndiffgenes)
-			{
-				thisgeneseffect = rnorm(1, thismean, sd=betweensd)
-				a=labels==affectedconditions[i]
[…]

### 2014-07-17 overviewplot-vegard.r
rydding

@@ -1,172 +1,0 @@
-
-
-# Script to create Figure 1
-
-library(sva)
-library(lsmeans)
-
-adhocboxplot = function(y, grouplabels, batchlabels, addlegend=FALSE, ylim=NULL, xlab="x", figureletter=NA)
-{
-  #x = batchlabels + grouplabels/8
-  colpalette = c("blue", "red", "darkgreen", "magenta")
-  pchselection = c(1, 2, 3)
-  x = batchlabels
-  lwd=2
-  
-  # add extra x for better visual separation
-  for(b in unique(batchlabels))
-  {
-    
-    separation = 0.15
-    xoffset=0    
-    for(g in unique(grouplabels[batchlabels==b]))
-    {
-      x[batchlabels==b & grouplabels==g] = x[batchlabels==b & grouplabels==g] + xoffset
-      xoffset = xoffset+separation
-    }
-  }
-  
-  xlim = c(0, length(unique(batchlabels))+2)
-  plot(x, y, col=colpalette[grouplabels], pch=pchselection[batchlabels], 
-       xlim=xlim, lwd=lwd, cex=2, ylim=ylim, xlab=xlab, ylab="", axes=FALSE, cex.lab =2)
-  
-  #add boxplot
-  groupnames = unique(grouplabels)
-  groupnames = groupnames[order(groupnames)]
-  xoffset=length(unique(batchlabels))+1
-  boxwidth=0.35
-  boxseparation = 0.25
-  for(g in groupnames)
-  { 
-
-    m = mean(y[grouplabels==g])
-    ci = t.test(y[grouplabels==g])[["conf.int"]]
-    xleft = xoffset
-    ybottom = ci[1]
-    xright = xleft + boxwidth
-    ytop = ci[2]
-    rect(xleft, ybottom, xright, ytop, border=colpalette[g], lwd=lwd)
-    lines(c(xoffset,xright), c(m,m), col=colpalette[g], lwd=lwd)
-    xoffset = xoffset+boxseparation
-  }
-  legendcex=2
-  if(addlegend)
-  {
-    
-    legend("topright", legend=paste("Group ", groupnames, sep=""), 
-           text.col=colpalette[groupnames], bty="n", cex=legendcex)
-    legend("bottomright", legend=paste("Batch ", unique(batchlabels), sep=""),
-           pch=pchselection[unique(batchlabels)], bty="n", cex=legendcex)
-  }
-  if(!is.na(figureletter))
-  {
[…]

### 2014-07-18 tex/manuscript.tex
rydding
justeringer

@@
 Extraneous variables, if left unaccounted for, have the potential to lead an investigator into drawing wrong conclusions.
-A common example is ``batch effects'' caused by reagents, microarray chips, and other equipment are made in batches, often have systematic effects on the measurements.
+A common example is ``batch effects'' caused by reagents, microarray chips, and other equipment made in batches, which often have systematic effects on the measurements.
 A similar example is ``centre effects'' when samples or data come from multiple sources.
@@
 
-The supporting information of the original ComBat article (\citealp{Johnson2007}) demonstrates the method on cells inhibited for the expression of the TAL1 gene were compared to controls on a microarray platform (denoted ``Data set 2'').
+The supporting information of the original ComBat article (\citealp{Johnson2007}) demonstrates the method on cells inhibited for the expression of the TAL1 gene compared to controls on a microarray platform (denoted ``Data set 2'').
 The experiment was conducted on three different time points (used as batches) with a total of 30 samples and a fairly balanced batch--treatment set up (6-2, 3-4 and 9-6).
@@
 
+Be aware that the different versions of the r-packages could produce different results. In particular, the newest version of limma (3.20.8) was needed to produce the plot in Fig. 1d.
+
 \NOTE{Vi bør klare aa tilfredstille kravene i \href{http://biostatistics.oxfordjournals.org/content/10/3/405.full}{Biostatistics}.

### 2014-07-19 tex/manuscript.tex
Update manuscript.tex

@@
 \\[2pt]
-\textit{\NOTE{Andre adresser}}
+\textit{Institute of Cancer Genetics and Informatics, Oslo University Hospital HF - Radiumhospitalet,
+Montebello, 
+0310 Oslo,
+Norway}
+\textit{Department of Informatics, University of Oslo, 
+0316 OSLO
+Norway
+Norway}
 \\[2pt]
@@
 Many methods and tools exist for this purpose.
-One method, ComBat which is part of the R package sva, is particularly popular due to its ability to remove batch differences even when batches are small and heterogeneous.
+One method, ComBat, which is part of the R package sva, is particularly popular due to its ability to remove batch differences even when batches are small and heterogeneous.
 It also has the option of preserving the difference between study groups, estimated from a two-way ANOVA model, to avoid conflating batch effects and group differences during batch adjustments.
@@
 Extraneous variables, if left unaccounted for, have the potential to lead an investigator into drawing wrong conclusions.
-A common example is ``batch effects'' caused by reagents, microarray chips, and other equipment made in batches, which often have systematic effects on the measurements.
-A similar example is ``centre effects'' when samples or data come from multiple sources.
+A common example is ``batch effects'' caused by reagents, microarray chips, and other equipment made in batches that may vary in some way, which often have systematic effects on the measurements.
+A similar example is ``centre effects'', when samples or data come from multiple sources.
 See \citet{Luo2010} for more examples.
 
-For a typical experiment comparing differences between study groups, the presence of batch effects will decrease the statistical power since it adds variation to the data.
-If the batch--group design is unbalanced, i.e if the study groups are not equally represented in all batches, batch effects may also act as a confounder and induce false differences between groups (\citealp{Leek2010}).
+For a typical experiment of comparing differences between study groups, the presence of batch effects will decrease the statistical power, since it adds variation to the data.
+If the batch--group design is unbalanced, i.e.
+if the study groups are not equally represented in all batches, batch effects may also act as a confounder and induce false differences between groups (\citealp{Leek2010}).
 
 The standard way to handle an extraneous variable is to include it in the statistical model employed in the inquiry.
-However, many analysis tools for high throughput data do not cater for this option, and when available it could still be outside the competence of the investigator.
+However, many analysis tools for high throughput data do not cater for this option, and when available, it could still be outside the competence of the investigator.
 Therefore, an alternative two step procedure has emerged.
@@
 In very uneven group--batch designs with multiple groups, spurious group differences may even be induced in this way.
-Figure~\ref{fig:boxplots} illustrates both these effects.
+Figure~\ref{fig:boxplots} illustrates both of these effects.
 
@@
 In addition, it uses an empirical Bayes approach to avoid over-correction for batch effects, which is critical for small batches.
-Increasingly, the inclusion of group difference as a covariate when removing batch effects has been recommended, both in the \texttt{sva} tutorial and user fora, and ComBat has thus help popularise this approach.
-Based on actual use of ComBat by the authors and others, ourselves included, we suspect thus adjusted data are commonly treated as ``batch effect free'' in subsequent analyses.
+Increasingly, the inclusion of group difference as a covariate when removing batch effects has been recommended, both in the \texttt{sva} tutorial and user fora, and ComBat has thus helped popularise this approach.
+Based on the actual use of ComBat by the authors and others, ourselves included, we suspect thus that adjusted data are commonly treated as ``batch effect free'' in subsequent analyses.
 And as a consequence, confidence in group effects has been overestimated and false results reported.
 
-Statisticians would most likely take extra precautions should the sample or batch sizes be very small.
+Statisticians would most likely take extra precautions, should the sample or batch sizes be very small.
 However, the effects of batch adjustment using two-way ANOVA or ComBat on unbalanced data sets remain even as sample and batch sizes increase.
-For example, group comparisons using one-way ANOVA on the batch adjusted data will essentially result in $F$ statistics that are inflated by a fixed factor which depends on the unevenness of the design rather than the size of the sample or batches.
+For example, group comparisons using one-way ANOVA on the batch adjusted data will essentially result in $F$ statistics that are inflated by a fixed factor which depends on the unevenness of the design, rather than the size of the sample or batches.
 The effect of this may be further exacerbated by running these analyses a large number of times, e.g.

### 2014-07-19 tex/manuscript.tex
Update manuscript.tex

@@
 When comparing groups, the common value $\bar{Y}$ has no effect, and so this is equilvalent to zero-centring each batch.
-If the groups are unevenly represented in the different batches, the batch average $\bar{Y}_i$ will tend to capture, through $\bar\beta_i$, group differences as well as batch effects:
+If the groups are unevenly represented in the different batches, the batch average $\bar{Y}_i$ will tend to capture, through $\bar\beta_i$, group differences, as well as batch effects:
 \begin{equation}
@@
 By reducing the differences between some groups, i.e.
-those found together in the same batches, it may also induce false differences between other groups, as is demonstrated in Figure~\ref{fig:boxplots}.
+those found together in the same batches, one may also induce false differences between other groups, as is demonstrated in Figure~\ref{fig:boxplots}.
 
-Removing batch effects while retaining group differences can be done through a two-way ANOVA in which group effects, $\beta_j$, and batch effects, $\gamma_i$, are estimated simultaneously.
+Removing batch effects while retaining group differences can be achieved through a two-way ANOVA, in which group effects, $\beta_j$, and batch effects, $\gamma_i$, are estimated simultaneously.
 Batch adjusted values may then be obtained by subtracting the estimated batch effects, $\hat\gamma_i$:
@@
 \end{equation}
-This will yield batch adjusted values where any systematic bias induced by the batch differences has been removed, while the group differences are retained.
+This will yield batch adjusted values, where any systematic bias induced by the batch differences has been removed, while the group differences are retained.
 
@@
 The undesired consequences of preserving group effects when correcting for batch effect is readily illustrated with a sanity check using random numbers.
-The documentation accompanying the \texttt{sva} library has a runnable example demonstrating how to adjust a data set with ComBat followed by an F-test.
+The documentation accompanying the \texttt{sva} library has a runnable example demonstrating how to adjust a data set with ComBat, followed by an F-test.
 Replacing the real data with random numbers from a standard normal distribution (mean=0, sd=1), but otherwise following the instructions, will generate the p-value distribution shown in Figure~\ref{fig:sanity}.
@@
 As can be seen from the QQ plot, the main effect of the procedure is to inflate the F-statistic by a factor.
-The size of this factor depends on how unbalanced the group--batch design is rather than on the sample size in itself.
+The size of this factor depends on how unbalanced the group--batch design is, rather than on the sample size in itself.
 Thus, increasing the sample size will not reduce the problem.
@@
 
-As the amount of false positive results when trying to retain group differences depends on the batch--group balance, we will show two examples with varying degree of unbalancedness.
+As the amount of false positive results when trying to retain group differences depends on the batch--group balance, we will show two examples with varying degree of unbalance.
 
@@
 
-In an experiment by \citet{Towfic2014}, cells were treated with either glatiramer acetate (a medicine for multiple sclerosis) or a generic drug, and mRNA was measured using microarrays alongside control samples.
-A batch effect correlating to the chip (Illumina WG-6\_V2, six samples per chip) was observed and adjusted for with ComBat, whereafter the data was tested for differentially expressed genes, yielding a list of 1000 genes (Table S5, \citealp{Towfic2014}).
-Unfortunately the batch--treatment design was unbalanced with several batches having only one of the main treatments of interest.
-When we re-analyzed their data without using ComBat, but instead blocked for batch effect in limma, only 9 genes were found ($\text{FDR}<0.05$).
-The above outlined sanity check with random numbers was also carried out.
+In an experiment by \citet{Towfic2014}, cells were treated with either glatiramer acetate (a medicine for multiple sclerosis) or a generic drug, and mRNA was measured using microarrays alongside control sample measurements.
+A batch effect correlating to the chip (Illumina WG-6\_V2, six samples per chip) was observed and adjusted for with ComBat, whereafter the data were tested for differentially expressed genes, yielding a list of 1000 genes (Table S5, \citealp{Towfic2014}).
+Unfortunately, the batch--treatment design was unbalanced, with several batches having only one of the main treatments of interest.
+When we re-analyzed their data without the use of ComBat, but instead blocked for batch effect in limma, only 9 genes were found ($\text{FDR}<0.05$).
+The sanity check  outlined above with random numbers was also carried out.
 The distribution of p-values for different settings are shown in Figure~\ref{fig:comparison}a.
 Our conclusion is that most of the genes reported as differentially expressed in \citet{Towfic2014} are false positives.
-This example is a sort of ``worst case'' scenario for applying ComBat, since it both has a very unbalanced batch--group design and a a priori assumption of no difference.
+This example is a sort of ``worst case'' scenario for applying ComBat, since it both has a very unbalanced batch--group design and an a priori assumption of no difference.
 The R-code for our analysis and a more extensive report can be downloaded from \href{https://github.com/vegardny/combat_tests.git}{Github} \NOTE{Update reference}.
@@
 The experiment was conducted on three different time points (used as batches) with a total of 30 samples and a fairly balanced batch--treatment set up (6-2, 3-4 and 9-6).
-ComBat was applied followed by a t-test in order to identify differentially expressed genes.
-First, we reproduced their analysis including the adjustment by ComBat, but using limma instead of the t-test, resulting in 1003 probes ($q<0.05$).
+ComBat was applied followed by a t-test, in order to identify differentially expressed genes.
+First, we reproduced their analysis, including the adjustment by ComBat, but using limma instead of the t-test, resulting in 1003 probes ($q<0.05$).
 Then, we analysed their data without batch adjustment in ComBat, but blocking for batch in limma, resulting in 377 probes ($q<0.05$).
@@
 The distribution of P-values for different settings are shown in Figure~\ref{fig:comparison}b.
-In contrast to the results obtained for \citet{Towfic2014} (Figure~\ref{fig:comparison}a), the P-value distributions for the alternative analysis does not indicate a huge difference.
+In contrast to the results obtained for \citet{Towfic2014} (Figure~\ref{fig:comparison}a), the P-value distributions for the alternative analysis do not indicate a huge difference.
 Nevertheless, the P-values may still be somewhat deflated in the ComBat adjusted analysis.
@@
 The impact is also more likely to be notable when used to analyse a large number of features, e.g.
-a large set of gene, followed by multiple testing corrections such as false discovery rate as the effect is more pronounced for more extreme values.
+a large set of genes, followed by multiple testing corrections such as false discovery rate, as the effect is more pronounced for more extreme values.
 
@@
 for retaining group differences for unbalanced designs, was optional and appeared subordinate, only exemplified in the supplementary information.
-However, over the years this feature became more important judging from advice given on user fora.
+However, over the years, this feature became more important, judging from advice given on user fora.
 Later, when ComBat was incorporated into the \texttt{sva} package (\citealp{Leek2012}), specification of a covariate model became required: batch adjustment without covariates was no longer a default behaviour.
 Although specification of a null model was still possible, the help file specifies that covariates should contain ``outcome of interest and other covariates besides batch''.
-Thus, we expect this usage to have grown more common.
+Thus, we expect this usage to have grown more common over time.
 
@@
 Our knowledge of the problem discussed in this article came through a typical use case when trying to adjust for batch effects in an unbalanced data set using ComBat.
-Upon realizing that the confidence in the estimated group differences were exaggerated, we searched the literature for a better understanding of correct use and potential limitations of ComBat.
+Upon realizing that the confidence in the estimated group differences was exaggerated, we searched the literature for a better understanding of correct use and the potential limitations of ComBat.
 However, the authors of ComBat and the \texttt{sva} package recommended including study group as a covariate as we had done.
@@
 A brief inquiry into  some of the articles citing ComBat (574 on Google Scholar) revealed few reported problems, although their method descriptions regarding ComBat were mostly sparse, limited to one or two sentences.
-A further indication of the carefree use of the procedure was the frequent omittance of program parameters that were used, i.e batch labels or whether group labels were supplied as covariates.
-Often no effort was done in order to substantiate the existence of batch effects beyond stating the presence of batches.
-The incorporation of the method into analysis tools and pipelines, like \COMMENT{TCGA}{does not seem to offer ComBat with covariates} and GenePattern (\citealp{Reich2006}), and other packages like \texttt{sva} (\citealp{Leek2012}) and \COMMENT{inSilicoDb (\citealp{Taminau2011})}{doesn't seem to offer ComBat with covariates} could make it even more accessible, but its use less transparent.
+A further indication of the carefree use of the procedure was the frequent omittance of program parameters that were applied, i.e batch labels or whether group labels were supplied as covariates.
+Often, no effort was undertaken in order to substantiate the existence of batch effects, beyond stating the presence of batches.
+The incorporation of the method into analysis tools and pipelines, like \COMMENT{TCGA}{does not seem to offer ComBat with covariates} and GenePattern (\citealp{Reich2006}), and other packages like \texttt{sva} (\citealp{Leek2012}) and \COMMENT{inSilicoDb (\citealp{Taminau2011})}{does not seem to offer ComBat with covariates} could make it even more accessible, but its use less transparent.
 We are concerned, in light of our findings, that many published results from data adjusted by ComBat with study group as covariate may be unreliable.
@@
 
-For an investigator facing an unbalanced data set with batch effects, our primary advice would be to adjust for batch inside the statistical test and avoid the two step procedure outlined above.
-If this is not possible, batch correction using outcome as covariate should only be done with great caution and downstream confidence estimates must be treated with suspicion.
+For an investigator facing an unbalanced data set with batch effects, our primary advice would be to adjust for batch inside the statistical test, and avoid the two step procedure outlined above.
+If this is not possible, batch correction using outcome as covariate should only be performed with great caution, and downstream confidence estimates must be treated with suspicion.
 
-Knowing that adjusting for batch effects while preserving the group difference may lead to varying degree of false results, to what extent can an investigator trust published results which has applied such a method?
+Knowing that adjustment for batch effects while preserving the group difference may lead to varying degree of false results, to what extent can an investigator trust published results where such a method has been applied?
 Essentially, when the batch--group configuration is balanced, or group difference is ignored  (i.e.
@@
 
-A superficial assessment of the reliability of results based on batch corrected data with group differences retain can often be made.
+A superficial assessment of the reliability of results based on batch corrected data with group differences retained can often be made.
 The effective sample size, $\nu$, may be compared against the nominal sample size, $\nu_0$, as in section~\label{sec:twogroups} for the case of two groups: similar computations, although more complicated, can be made for multiple groups, but the two-group assessment should still give a fair idea of the reliability of pairwise group comparisons.
-A major concern would be batches were either of the groups of interest are missing or strongly under-represented, which would contribute to the nominal sample size, but not to the effective sample size.
+A major concern would be batches where either of the groups of interest are missing or strongly under-represented, which would contribute to the nominal sample size, but not to the effective sample size.
 
 Alternatively, one may treat the results more like an ordered list of candidates, with the most likely true positives on top, de-emphasizing the somewhat deflated p-values.
-This would, however, be hypothesis generating rather than hypothesis testing.
+This would, however, be hypothesis generating, rather than hypothesis testing.
 While investigators should always assess the extent to which findings make biological or clinical sense, this is particularly true when the statistical assessment may be unreliable.
@@
 Finally, we would like to emphasise the importance of proper description of how data has been prepared for analysis, and what corrections and adjustments have been made.
-In cases where data preparation is done prior to and separate from the data analysis, as is now often the case, this is of particular importance as artifacts may be introduced in the data preparation which could influence the reliability of downstream analyses.
+In cases where data preparation is performed prior to, and separate from, the data analysis, as is now often the case, this is of particular importance, as artifacts may be introduced in the data preparation which could influence the reliability of downstream analyses.
 
@@
 \href{https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-figures.git}{GitHub}.
-Additional analyses, performed but not included in the article, may be found at
+Additional analyses, performed, but not included in the article, may be found at
 \href{ https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-reports.git}{GitHub}.
@@
 Real data are substituted with random numbers from a normal distribution, but the batch--group design is retained.
-ComBat is applied followed by an F-test.
+ComBat is applied, followed by an F-test.
 a) P-value distribution
@@
 First, analysed as described on the real data using ComBat.
-Secondly, with ComBat but with random numbers instead of real data.
-Thirdly, instead of ComBat, analyses of the real data applied \texttt{limma} blocking by batch.
+Secondly, with ComBat, but with random numbers instead of real data.
+Thirdly, instead of ComBat, analyses of the real data with  \texttt{limma} blocking by batch.
 a) Re-analysis of \citet{Towfic2014}, glatiramer acetate vs. generic 

### 2014-07-21 figurescript.r
div

@@ -2,6 +2,9 @@
 # Main script.
 # Calls the individual scripts that creates figures 1,2,3a,3b.
 
+source("http://bioconductor.org/biocLite.R")
+biocLite( c("sva", "limma", "lsmeans", "bladderbatch"))
+
 
 startwd = getwd()
 source("overviewplot.r")

### 2014-07-21 tex/manuscript.tex
Typeset manuscript
The latest diff.pdf shows Eivind's changes.
Update manuscript.tex
Updated so references work for others
Previously, they only worked for me since I was the only one with the
correct BibTeX bibliography library.
div

@@
 Norway}
+\\[2pt]
 \textit{Department of Informatics, University of Oslo, 
 0316 OSLO
-Norway
 Norway}
@@
 However, the authors of ComBat and the \texttt{sva} package recommended including study group as a covariate as we had done.
-In addition, other studies investigating batch effects were mostly recommending ComBat without much concern (\citealp{Kupfer2012}, \citealp{Kitchen2011}).
+In addition, other studies investigating batch effects were mostly recommending ComBat without much concern (\citealp{Kupfer2012}, \citealp{Kitchen2011}, (PMID:21386892).
 A brief inquiry into  some of the articles citing ComBat (574 on Google Scholar) revealed few reported problems, although their method descriptions regarding ComBat were mostly sparse, limited to one or two sentences.
@@
 \bibliographystyle{biorefs}
-\bibliography{D:/Projects/_Appl_/_TeX_/library}
+\input{manuscript.bbl} % Include bibliography bbl-file
+%\bibliography{D:/Projects/_Appl_/_TeX_/library} % To use with BibTeX to generate bbl-file
 %\bibliography{refs}

### 2014-07-21 tex/manuscript.tex
div

@@
 The supporting information of the original ComBat article (\citealp{Johnson2007}) demonstrates the method on cells inhibited for the expression of the TAL1 gene compared to controls on a microarray platform (denoted ``Data set 2'').
-The experiment was conducted on three different time points (used as batches) with a total of 30 samples and a fairly balanced batch--treatment set up (6-2, 3-4 and 9-6).
+The experiment was conducted on three different time points (used as batches) with a total of 30 samples and a fairly balanced batch--treatment set up containing 6+2, 3+4 and 9+6 samples.
 ComBat was applied followed by a t-test, in order to identify differentially expressed genes.
@@
 However, the authors of ComBat and the \texttt{sva} package recommended including study group as a covariate as we had done.
-In addition, other studies investigating batch effects were mostly recommending ComBat without much concern (\citealp{Kupfer2012}, \citealp{Kitchen2011}, (PMID:21386892).
+In addition, other studies investigating batch effects were mostly recommending ComBat without much concern (\citealp{Kupfer2012}, \citealp{Kitchen2011}, \CITE{PMID:21386892}).
 A brief inquiry into  some of the articles citing ComBat (574 on Google Scholar) revealed few reported problems, although their method descriptions regarding ComBat were mostly sparse, limited to one or two sentences.
@@
 The data and scripts used to generate the results in this work, i.e figures, are available at
-\href{https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-figures.git}{GitHub}.
+\href{https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-figures.git}{https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-figures.git}.
 Additional analyses, performed, but not included in the article, may be found at
-\href{ https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-reports.git}{GitHub}.
+\href{https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-reports.git}{https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-reports.git}.
 
 Be aware that the different versions of the r-packages could produce different results. In particular, the newest version of limma (3.20.8) was needed to produce the plot in Fig. 1d.
-
-\NOTE{Vi bør klare aa tilfredstille kravene i \href{http://biostatistics.oxfordjournals.org/content/10/3/405.full}{Biostatistics}.
-For å tilfredstille kravene skal data og script vare available.
-Jeg mener vi klarer dette.
-På githuben er ikke data fra Towfic, men det blir lastet av scriptet fra GEO.
-Ikke beskrevet her men i scriptet.
-Videre i forklaringen til Biostatistics står det at script og data skal bli lagt på deres sider, som vi kan hvis de vil.}
 
@@
 %\centering\includegraphics[width=13cm]{Fig/boxplots.png}
-\caption{Simulated data from four study groups was generated where groups 1 and 2 have lower means than groups 3 and 4.
-These were placed in three different batches with batch effects added.
-Values and boxes showing group mean with 95\%~confidence intervals are displayed for data without batch effects, after adding batch effects, after batch centering, and after two-way ANOVA based batch centering (using \texttt{limma}).
-The last frame shows the least squares estimates of the group means from a two-way ANOVA.
-This case, design and effects, was selected to illustrate the spurious effects that may arise.
+\caption{We simulated expression of one gene from four study groups unevenly distributed in 3 batches containing 50+0+0+20, 0+50+0+20 and 0+0+50+20 samples from groups 1 to 4.
+This case, design and effects, was selected to illustrate the spurious effects that may arise from different batch adjustments.
+The Y-axis represent the expression, while the X-axis is used to better visually differenitate between the batches.
+Boxes showing group mean with 95\% confidence intervals are displayed to the right of the measurments.
+a) The "true" values where groups 1 and 2 have lower means than groups 3 and 4.
+b) After adding batch effects, all the groups seem different.
+c) After batch centering, group 3 and 4 seem to differ.
+d) After two-way ANOVA based batch centering (using \texttt{limma}), group 1 and 2 seem to differ.
+e) The least squares estimates of the group means from a two-way ANOVA which represent a more reliable way of estimating group differences than the two-step procedure in c) and d).
 }

### 2014-07-21 tex/manuscript.tex
div

@@
 In very uneven group--batch designs with multiple groups, spurious group differences may even be induced in this way.
-Figure~\ref{fig:boxplots} illustrates both of these effects.
+Figure~\ref{fig:boxplots}c illustrates both of these effects.
 
@@
 Subsequent analyses will then systematically underestimate the statistical uncertainties and exaggerate the confidence of group differences.
-Figure~\ref{fig:boxplots} illustrates how statistical uncertainties are deflated by this batch adjustment method by comparing them to the uncertainties from the original ANOVA.
+Figure~\ref{fig:boxplots}d illustrates how statistical uncertainties are deflated by this batch adjustment method by comparing them to the uncertainties from the original ANOVA in Figure~\ref{fig:boxplots}e.
 
@@
 By reducing the differences between some groups, i.e.
-those found together in the same batches, one may also induce false differences between other groups, as is demonstrated in Figure~\ref{fig:boxplots}.
+those found together in the same batches, one may also induce false differences between other groups, as is demonstrated in Figure~\ref{fig:boxplots}c.
 
@@
 In an unbalanced design, however, the estimation errors $\hat\gamma_i-\gamma_i$ will induce systematic differences between groups which, when ignored in downstream analyses, may lead to over-confidence in estimated group differences.
-In Figure~\ref{fig:boxplots}, the effect of batch correction using group as covariate is shown with confidence intervals of the corrected data.
-For comparison, least square means estimates (R package \texttt{lsmeans}) are used to illustrate more appropriate confidence intervals that incorporate the uncertainties of the estimation errors.
+In Figure~\ref{fig:boxplots}d, the effect of batch correction using group as covariate is shown with confidence intervals of the corrected data.
+For comparison, least square means estimates (R package \texttt{lsmeans}) are used in Figure~\ref{fig:boxplots}e to illustrate more appropriate confidence intervals that incorporate the uncertainties of the estimation errors.
 
@@
 The documentation accompanying the \texttt{sva} library has a runnable example demonstrating how to adjust a data set with ComBat, followed by an F-test.
-Replacing the real data with random numbers from a standard normal distribution (mean=0, sd=1), but otherwise following the instructions, will generate the p-value distribution shown in Figure~\ref{fig:sanity}.
+Replacing the real data with random numbers from a standard normal distribution (mean=0, sd=1), but otherwise following the instructions, will generate the p-value distribution shown in Figure~\ref{fig:sanity}a.
 
-As can be seen from the QQ plot, the main effect of the procedure is to inflate the F-statistic by a factor.
+As can be seen from the QQ plot in Figure~\ref{fig:sanity}b, the main effect of the procedure is to inflate the F-statistic by a factor.
 The size of this factor depends on how unbalanced the group--batch design is, rather than on the sample size in itself.
@@
 
-Be aware that the different versions of the r-packages could produce different results. In particular, the newest version of limma (3.20.8) was needed to produce the plot in Fig. 1d.
+Be aware that the different versions of the r-packages could produce different results.
+In particular, the newest version of limma (3.20.8) was needed to produce the plot in Figure~\ref{fig:boxplots}d.
 
@@
 Boxes showing group mean with 95\% confidence intervals are displayed to the right of the measurments.
-a) The "true" values where groups 1 and 2 have lower means than groups 3 and 4.
+a) The ``true'' values where groups 1 and 2 have lower means than groups 3 and 4.
 b) After adding batch effects, all the groups seem different.
-c) After batch centering, group 3 and 4 seem to differ.
+c) After batch centering, group 3 and 4 seem to differ while group 1 and 3 are more similar.
 d) After two-way ANOVA based batch centering (using \texttt{limma}), group 1 and 2 seem to differ.

### 2014-07-21 tex/manuscript.tex
div

@@
 This example is a sort of ``worst case'' scenario for applying ComBat, since it both has a very unbalanced batch--group design and an a priori assumption of no difference.
-The R-code for our analysis and a more extensive report can be downloaded from \href{https://github.com/vegardny/combat_tests.git}{Github} \NOTE{Update reference}.
+The R-code for our analysis can be downloaded from \href{https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-figures.git}{https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-figures.git}.
 
@@
 Nevertheless, the P-values may still be somewhat deflated in the ComBat adjusted analysis.
-The R-code for our analysis can be downloaded from \href{https://github.com/vegardny/combat_tests.git}{Github} \NOTE{Update reference}.
+The R-code for our analysis can be downloaded from \href{https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-figures.git}{https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-figures.git}
 
@@
 Often, no effort was undertaken in order to substantiate the existence of batch effects, beyond stating the presence of batches.
-The incorporation of the method into analysis tools and pipelines, like \COMMENT{TCGA}{does not seem to offer ComBat with covariates} and GenePattern (\citealp{Reich2006}), and other packages like \texttt{sva} (\citealp{Leek2012}) and \COMMENT{inSilicoDb (\citealp{Taminau2011})}{does not seem to offer ComBat with covariates} could make it even more accessible, but its use less transparent.
+The incorporation of ComBat into analysis tools, like \COMMENT{TCGA}{does not seem to offer ComBat with covariates} and GenePattern (\citealp{Reich2006}), and other packages like \texttt{sva} (\citealp{Leek2012}) and \COMMENT{inSilicoDb (\citealp{Taminau2011})}{does not seem to offer ComBat with covariates} could make it more accessible, but its use less transparent.
 We are concerned, in light of our findings, that many published results from data adjusted by ComBat with study group as covariate may be unreliable.

### 2014-07-22 tex/manuscript.tex
div

@@
 Often, no effort was undertaken in order to substantiate the existence of batch effects, beyond stating the presence of batches.
-The incorporation of ComBat into analysis tools, like \COMMENT{TCGA}{does not seem to offer ComBat with covariates} and GenePattern (\citealp{Reich2006}), and other packages like \texttt{sva} (\citealp{Leek2012}) and \COMMENT{inSilicoDb (\citealp{Taminau2011})}{does not seem to offer ComBat with covariates} could make it more accessible, but its use less transparent.
+The incorporation of ComBat into analysis tools, like \COMMENT{TCGA}{does not seem to offer ComBat with covariates} and GenePattern (\citealp{Reich2006}), and other packages like \COMMENT{inSilicoDb (\citealp{Taminau2011})}{does not seem to offer ComBat with covariates} could make it more accessible, but its use less transparent.
 We are concerned, in light of our findings, that many published results from data adjusted by ComBat with study group as covariate may be unreliable.

### 2014-07-22 tex/manuscript.tex
div

@@
 However, many analysis tools for high throughput data do not cater for this option, and when available, it could still be outside the competence of the investigator.
-Therefore, an alternative two step procedure has emerged.
+Therefore, an alternative two-step procedure has emerged.
 First the batch effects are estimated and removed, creating a ``batch effect free'' data set.
@@
 
-For an investigator facing an unbalanced data set with batch effects, our primary advice would be to adjust for batch inside the statistical test, and avoid the two step procedure outlined above.
+For an investigator facing an unbalanced data set with batch effects, our primary advice would be to adjust for batch inside the statistical test, and avoid the two-step procedure outlined above.
 If this is not possible, batch correction using outcome as covariate should only be performed with great caution, and downstream confidence estimates must be treated with suspicion.

### 2014-07-22 figurescript.r

@@ -2,9 +2,11 @@
 # Main script.
 # Calls the individual scripts that creates figures 1,2,3a,3b.
 
-source("http://bioconductor.org/biocLite.R")
-biocLite( c("sva", "limma", "lsmeans", "bladderbatch"))
-
+# In order to reproduce the overview plot, version 3.20.8 of limma was needed.
+# R version 3.1.1 was needed for installation of limma 3.20.8.
+# The below two commands should download and install the nescesary packages if needed.
+# source("http://bioconductor.org/biocLite.R")
+# biocLite( c("sva", "limma", "lsmeans", "bladderbatch"))
 
 startwd = getwd()
 source("overviewplot.r")

### 2014-07-23 tex/manuscript.tex
noe spellcheck forandring

@@
 $\widetilde Y_{ijr}=\Delta\widetilde Y_{ijr}+\bar{Y}$.
-When comparing groups, the common value $\bar{Y}$ has no effect, and so this is equilvalent to zero-centring each batch.
+When comparing groups, the common value $\bar{Y}$ has no effect, and so this is equivalent to zero-centring each batch.
 If the groups are unevenly represented in the different batches, the batch average $\bar{Y}_i$ will tend to capture, through $\bar\beta_i$, group differences, as well as batch effects:
@@
 
-In the original ComBat article (\citealp{Johnson2007}), the main objective was to employ an emprical Bayes method to allow better handling of small batches.
+In the original ComBat article (\citealp{Johnson2007}), the main objective was to employ an empirical Bayes method to allow better handling of small batches.
 The inclusion of covariates, e.g.
@@
 This case, design and effects, was selected to illustrate the spurious effects that may arise from different batch adjustments.
-The Y-axis represent the expression, while the X-axis is used to better visually differenitate between the batches.
-Boxes showing group mean with 95\% confidence intervals are displayed to the right of the measurments.
+The Y-axis represent the expression, while the X-axis is used to better visually differentiate between the batches.
+Boxes showing group mean with 95\% confidence intervals are displayed to the right of the measurements.
 a) The ``true'' values where groups 1 and 2 have lower means than groups 3 and 4.

### 2014-07-23 tex/manuscript.tex
div

@@
 Unfortunately, the batch--treatment design was unbalanced, with several batches having only one of the main treatments of interest.
-When we re-analyzed their data without the use of ComBat, but instead blocked for batch effect in limma, only 9 genes were found ($\text{FDR}<0.05$).
+When we re-analyzed their data without the use of ComBat, but instead blocked for batch effect in \texttt{limma} (\CITE{limma}), only 9 genes were found ($\text{FDR}<0.05$).
 The sanity check  outlined above with random numbers was also carried out.
@@
 ComBat was applied followed by a t-test, in order to identify differentially expressed genes.
-First, we reproduced their analysis, including the adjustment by ComBat, but using limma instead of the t-test, resulting in 1003 probes ($q<0.05$).
-Then, we analysed their data without batch adjustment in ComBat, but blocking for batch in limma, resulting in 377 probes ($q<0.05$).
+First, we reproduced their analysis, including the adjustment by ComBat, but using \texttt{limma} instead of the t-test, resulting in 1003 probes ($q<0.05$).
+Then, we analysed their data without batch adjustment in ComBat, but blocking for batch in \texttt{limma}, resulting in 377 probes ($q<0.05$).
 In addition, the sanity checks outlined above were performed.
@@
 Be aware that the different versions of the r-packages could produce different results.
-In particular, the newest version of limma (3.20.8) was needed to produce the plot in Figure~\ref{fig:boxplots}d.
+In particular, the newest version of \texttt{limma} (3.20.8) was needed to produce the plot in Figure~\ref{fig:boxplots}d.
 

### 2014-07-24 tex/manuscript.tex
forandret litt på figure1

@@
 The Y-axis represent the expression, while the X-axis is used to better visually differentiate between the batches.
-Boxes showing group mean with 95\% confidence intervals are displayed to the right of the measurements.
-a) The ``true'' values where groups 1 and 2 have lower means than groups 3 and 4.
-b) After adding batch effects, all the groups seem different.
-c) After batch centering, group 3 and 4 seem to differ while group 1 and 3 are more similar.
-d) After two-way ANOVA based batch centering (using \texttt{limma}), group 1 and 2 seem to differ.
-e) The least squares estimates of the group means from a two-way ANOVA which represent a more reliable way of estimating group differences than the two-step procedure in c) and d).
+Circles, triangles and crosses indicate the measurment of each sample and which batch it is part of.
+Boxes showing group mean with 95\% confidence intervals are displayed to the right of the measurements and are ment to imply if a statistical test would find significant group differences, true or false.
+a) The ``true'' values measured in a system without batch effects. Groups 1 and 2 have lower means than groups 3 and 4.
+b) The same samples as in a) but measured in a system with batch effects. All the groups seem different.
+c) The measurements in b) are adjusted with batch centering. Group 3 and 4 seem to differ while group 1 and 3 are more similar.
+d) The measurements in b) are adjusted with two-way ANOVA based batch centering (using \texttt{limma}). Group 1 and 2 seem to differ.
+e) The least squares estimates of the group means from a two-way ANOVA which represent a more reliable way of estimating group differences of the measurements in b) than the two-step procedure in c) and d).
 }

### 2014-07-24 commonscripts/boxplot_function.r

@@ -24,13 +24,13 @@ adhocboxplot = function(y, grouplabels, batchlabels, addlegend=FALSE, ylim=NULL,
     }
   }
   
-  xlim = c(0, length(unique(batchlabels))+2)
+  xlim = c(0, length(unique(batchlabels))+3)
   plot(x, y, col=colpalette[grouplabels], pch=pchselection[batchlabels], 
        xlim=xlim, lwd=lwd, cex=cex, ylim=ylim, xlab=xlab, ylab="", axes=FALSE, cex.lab =labelcex)
 
   #add boxplot
   groupnames = sort(unique(grouplabels))
-  xoffset=length(unique(batchlabels))+1
+  xoffset=length(unique(batchlabels))+2
   boxwidth=0.35
   boxseparation = 0.25
   for(g in groupnames) {
@@ -47,7 +47,7 @@ adhocboxplot = function(y, grouplabels, batchlabels, addlegend=FALSE, ylim=NULL,
   }
 	if(!is.na(figureletter)) {
 		figureletter=paste("(", figureletter, ")", sep="")
-		text(x=0.1, y=ylim[2], labels=figureletter, cex=legendcex)
+		text(x=1, y=ylim[2], labels=figureletter, cex=legendcex)
 	}
 }
 
@@ -83,7 +83,7 @@ adhocboxplot2 = function(lsmeans, grouplabels, batchlabels, addlegend=FALSE, yli
 	}
 	if(!is.na(figureletter)) {
 		figureletter=paste("(", figureletter, ")", sep="")
-		text(x=0.1, y=ylim[2], labels=figureletter, cex=legendcex)
+		text(x=1, y=ylim[2], labels=figureletter, cex=legendcex)
 	}
 }
 

### 2014-08-10 tex/manuscript.tex
Update manuscript.tex
Corrected some minor things as per pdf coloring

@@
 Statisticians would most likely take extra precautions, should the sample or batch sizes be very small.
-However, the effects of batch adjustment using two-way ANOVA or ComBat on unbalanced data sets remain even as sample and batch sizes increase.
+However, the effects of batch adjustment using two-way ANOVA or ComBat on unbalanced data sets remain, even as sample and batch sizes increase.
 For example, group comparisons using one-way ANOVA on the batch adjusted data will essentially result in $F$ statistics that are inflated by a fixed factor which depends on the unevenness of the design, rather than the size of the sample or batches.
@@
 \end{equation}
-In a balanced group--batch design, the effect is the same all groups and thus does not influence group comparisons.
+In a balanced group--batch design, the effect is the same for all groups and thus does not influence group comparisons.
 In an unbalanced design, however, the estimation errors $\hat\gamma_i-\gamma_i$ will induce systematic differences between groups which, when ignored in downstream analyses, may lead to over-confidence in estimated group differences.
@@
 The undesired consequences of preserving group effects when correcting for batch effect is readily illustrated with a sanity check using random numbers.
-The documentation accompanying the \texttt{sva} library has a runnable example demonstrating how to adjust a data set with ComBat, followed by an F-test.
+The documentation accompanying the \texttt{sva} library has a executable example demonstrating how to adjust a data set with ComBat, followed by an F-test.
 Replacing the real data with random numbers from a standard normal distribution (mean=0, sd=1), but otherwise following the instructions, will generate the p-value distribution shown in Figure~\ref{fig:sanity}a.
@@
 
-To explain more clearly what is happening, and quantify the size of the problem, we may consider the simple case of estimating the difference $\Delta\beta=\beta_A-\beta_B$ between two groups, $A$ and $B$, when there are $m$ batches with batch $i$ containing $n_{iA}$ and $n_{iB}$ samples from each of the two groups for $i=1,\ldots,m$.
+To explain more clearly what is happening, and to quantify the size of the problem, we may consider the simple case of estimating the difference $\Delta\beta=\beta_A-\beta_B$ between two groups, $A$ and $B$, when there are $m$ batches with batch $i$ containing $n_{iA}$ and $n_{iB}$ samples from each of the two groups for $i=1,\ldots,m$.
 
@@
 For other cases, a re-analysis without using this approach is the most rigorous path.
-However, such a re-analysis is not feasible if the downstream  analysis can not adjust for batch effects by itself.
+However, such a re-analysis is not feasible if the downstream  analysis can not independently adjust for batch effects.
 To reach a reliable result, batch effects need to be handled in some way or another.

### 2014-08-11 tex/manuscript.tex
A few minor corrections (and one big one)
Modified the formulaes a little, and corrected the inequality to
nu_0>=nu. Tried to incorporate Eivind's comments. Updated the diff.pdf
to contain only the latest changes.

@@
 
-The ComBat method described in \citet{Johnson2007}, and included in the \texttt{sva} package (\citealp{Leek2012}), can use either of the two above described approaches to estimate batch differences.
-In addition, it uses an empirical Bayes approach to avoid over-correction for batch effects, which is critical for small batches.
+Two popular tools for batch adjustment of gene expression data are the R bioconductor packages \texttt{limma} (\citealp{Smyth2003}) and ComBat (\citealp{Johnson2007}) which is now included in the \texttt{sva} package (\citealp{Leek2012}).
+Both these packages allow the inclusion of covariates, e.g.
+group difference, the effects of which should not be removed in the batch adjustment.
+In addition, ComBat uses an empirical Bayes approach to avoid over-correction for batch effects, which is critical for small batches.
 Increasingly, the inclusion of group difference as a covariate when removing batch effects has been recommended, both in the \texttt{sva} tutorial and user fora, and ComBat has thus helped popularise this approach.
-Based on the actual use of ComBat by the authors and others, ourselves included, we suspect thus that adjusted data are commonly treated as ``batch effect free'' in subsequent analyses.
-And as a consequence, confidence in group effects has been overestimated and false results reported.
+Based on the actual use of ComBat by the authors and others, ourselves included, we suspect that adjusted data are commonly treated as ``batch effect free'' in subsequent analyses, and as a consequence, confidence in group effects has been overestimated and false results reported.
 
-Statisticians would most likely take extra precautions, should the sample or batch sizes be very small.
-However, the effects of batch adjustment using two-way ANOVA or ComBat on unbalanced data sets remain, even as sample and batch sizes increase.
+When sample or batch sizes are small, statisticians would most likely take extra precautions.
+However, batch adjustment using two-way ANOVA or ComBat on unbalanced data sets may be just as harmful on large samples and batch sizes as for small.
 For example, group comparisons using one-way ANOVA on the batch adjusted data will essentially result in $F$ statistics that are inflated by a fixed factor which depends on the unevenness of the design, rather than the size of the sample or batches.
@@
 \begin{equation}
-\Delta\widetilde Y_{ijr}=Y_{ijr}-\bar{Y}_{i}
+\widetilde{Y}^0_{ijr}=Y_{ijr}-\bar{Y}_{i}
 \quad\text{where}\quad
-\bar{Y}_i=\frac{\sum_{j=1}^M\sum_{r=1}^{n_{ij}} Y_{ijr}}{\sum_{j=1}^M n_{ij}}.
+\bar{Y}_i=\frac{1}{n_{i\textrm{-}}} \sum_{j=1}^M\sum_{r=1}^{n_{ij}} Y_{ijr},\;
+n_{i\textrm{-}}=\sum_{j=1}^M n_{ij}.
 \end{equation}
 An alternative is to centre each batch to the common average by adding the average value $\bar{Y}$ across the entire data set: i.e.
-$\widetilde Y_{ijr}=\Delta\widetilde Y_{ijr}+\bar{Y}$.
+$\widetilde{Y}^{\textrm{avg}}_{ijr}=\widetilde{Y}^0_{ijr}+\bar{Y}$.
 When comparing groups, the common value $\bar{Y}$ has no effect, and so this is equivalent to zero-centring each batch.
@@
 \begin{equation}
-\Delta\widetilde{Y}_{ijr}=\beta_j-\bar{\beta}_i+\epsilon_{ijr}-\bar{\epsilon}_i
+\widetilde{Y}^0_{ijr}=\beta_j-\bar{\beta}_i+\epsilon_{ijr}-\bar{\epsilon}_i
 \quad\text{where}\quad
-\bar{\beta}_i=\frac{\sum_{j=1}^M n_{ij}\beta_j}{\sum_{j=1}^M n_{ij}},\;
-\bar{\epsilon}_i=\frac{\sum_{j=1}^M\sum_{r=1}^{n_{ij}} \epsilon_{ijr}}{\sum_{j=1}^M n_{ij}}.
+\bar{\beta}_i={\sum_{j=1}^M \frac{n_{ij}}{n_{i\textrm{-}}}\beta_j},\;
+\bar{\epsilon}_i=\frac{1}{n_{i\textrm{-}}} \sum_{j=1}^M\sum_{r=1}^{n_{ij}} \epsilon_{ijr}.
 \end{equation}
@@
 \begin{equation}
-\widetilde Y_{ijr}=Y_{ijr}-\hat\gamma_i=\alpha+\beta_j+(\gamma_i-\hat\gamma_i)+\epsilon_{ijr}.
+\widetilde{Y}^{\textrm{cov}}_{ijr}=Y_{ijr}-\hat\gamma_i=\alpha+\beta_j+(\gamma_i-\hat\gamma_i)+\epsilon_{ijr}.
 \end{equation}
@@
 \begin{equation}
-\widetilde{Y}_{\textrm{-}j}
-=\frac{\sum_{i=1}^m\sum_{r=1}^{n_{ij}}\widetilde{Y}_{ijr}}{n_{\textrm{-}j}}
+\widetilde{Y}^{\textrm{cov}}_{\textrm{-}j}
+=\frac{1}{n_{\textrm{-}j}} \sum_{i=1}^m\sum_{r=1}^{n_{ij}}\widetilde{Y}_{ijr}
 =\alpha+\beta_j+\bar{\epsilon}_{\textrm{-}j}
-  -\sum_{i=1}^m \frac{n_{ij}}{n_{\textrm{-}j}}(\hat{\gamma}_i-\gamma_i),
-\quad
-n_{\textrm{-}j}=\sum_{i=1}^m n_{ij}.
+  -\sum_{i=1}^m \frac{n_{ij}}{n_{\textrm{-}j}}(\hat{\gamma}_i-\gamma_i)
+\quad\text{where}\quad
+n_{\textrm{-}j}=\sum_{i=1}^m n_{ij}
 \end{equation}
-In a balanced group--batch design, the effect is the same for all groups and thus does not influence group comparisons.
-In an unbalanced design, however, the estimation errors $\hat\gamma_i-\gamma_i$ will induce systematic differences between groups which, when ignored in downstream analyses, may lead to over-confidence in estimated group differences.
+so that
+\begin{equation}
+\widetilde{Y}^{\textrm{cov}}_{\textrm{-}j}-\widetilde{Y}^{\textrm{cov}}_{\textrm{-}j'}
+=\beta_j-\beta_{j'}+\bar{\epsilon}_{\textrm{-}j}-\bar{\epsilon}_{\textrm{-}j'}
+ -\sum_{i=1}^m \left(\frac{n_{ij}}{n_{\textrm{-}j}}-\frac{n_{ij'}}{n_{\textrm{-}j'}}\right)(\hat{\gamma}_i-\gamma_i).
+\end{equation}
+In a balanced group--batch design, the estimation error $\hat\gamma_i-\gamma_i$ has the same effect for all groups, and thus does not influence group comparisons.
+In an unbalanced design, however, it will induce systematic differences between groups which, when ignored in downstream analyses, may lead to biases or over-confidence in estimated group differences.
 In Figure~\ref{fig:boxplots}d, the effect of batch correction using group as covariate is shown with confidence intervals of the corrected data.
@@
 
-To explain more clearly what is happening, and to quantify the size of the problem, we may consider the simple case of estimating the difference $\Delta\beta=\beta_A-\beta_B$ between two groups, $A$ and $B$, when there are $m$ batches with batch $i$ containing $n_{iA}$ and $n_{iB}$ samples from each of the two groups for $i=1,\ldots,m$.
+To explain more clearly what is happening, and quantify the size of the problem, we may consider the simple case of estimating the difference $\Delta\beta=\beta_A-\beta_B$ between two groups, $A$ and $B$, when there are $m$ batches with batch $i=1,\ldots,m$ containing $n_{iA}$ and $n_{iB}$ samples from each of the two groups.
 
@@
 \end{equation}
-Using Jensen's inequality, we may then derive
+Using Jensen's inequality, we may derive
 \begin{equation}
@@
 =\frac{1}{\frac{1}{n_A}+\frac{1}{n_B}}=n\cdot\frac{n_A}{n}\cdot\frac{n_B}{n}
-\le
+\ge
 \nu
@@
 In effect, $\nu$ represents the effective sample size in the unbalanced group--batch design, while $\nu_0$ represents the nominal sample size when batches are ignored.
-The ratio $\nu/\nu_0\ge1$ indicates to what extent the uncertainty of the estimated group difference is deflated by the two-step procedure.
+The ratio $\nu_0/\nu\ge1$ indicates to what extent the uncertainty of the estimated group difference is deflated by the two-step procedure.
 
@@
 
-As the amount of false positive results when trying to retain group differences depends on the batch--group balance, we will show two examples with varying degree of unbalance.
+The extent to which batch adjustment with group differences retained will confound subsequent analyses depends on the batch--group balance.
+We have reanalysed to cases with varying degree of unbalance.
 
@@
 Unfortunately, the batch--treatment design was unbalanced, with several batches having only one of the main treatments of interest.
-When we re-analyzed their data without the use of ComBat, but instead blocked for batch effect in \texttt{limma} (\CITE{limma}), only 9 genes were found ($\text{FDR}<0.05$).
+When we re-analyzed their data without the use of ComBat, but instead blocked for batch effect in \texttt{limma}, only 9 genes were found ($\text{FDR}<0.05$).
 The sanity check  outlined above with random numbers was also carried out.
@@
 This example is a sort of ``worst case'' scenario for applying ComBat, since it both has a very unbalanced batch--group design and an a priori assumption of no difference.
-The R-code for our analysis can be downloaded from \href{https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-figures.git}{https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-figures.git}.
 
@@
 The experiment was conducted on three different time points (used as batches) with a total of 30 samples and a fairly balanced batch--treatment set up containing 6+2, 3+4 and 9+6 samples.
-ComBat was applied followed by a t-test, in order to identify differentially expressed genes.
-First, we reproduced their analysis, including the adjustment by ComBat, but using \texttt{limma} instead of the t-test, resulting in 1003 probes ($q<0.05$).
+ComBat was applied followed by a T-test, in order to identify differentially expressed genes.
+First, we reproduced their analysis, including the adjustment by ComBat, but using \texttt{limma} instead of the T-test, resulting in 1003 probes ($q<0.05$).
 Then, we analysed their data without batch adjustment in ComBat, but blocking for batch in \texttt{limma}, resulting in 377 probes ($q<0.05$).
@@
 Nevertheless, the P-values may still be somewhat deflated in the ComBat adjusted analysis.
-The R-code for our analysis can be downloaded from \href{https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-figures.git}{https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-figures.git}
 
@@
 However, the authors of ComBat and the \texttt{sva} package recommended including study group as a covariate as we had done.
-In addition, other studies investigating batch effects were mostly recommending ComBat without much concern (\citealp{Kupfer2012}, \citealp{Kitchen2011}, \CITE{PMID:21386892}).
+In addition, other studies investigating batch effects were mostly recommending ComBat without much concern (\citealp{Kupfer2012}, \citealp{Kitchen2011}, \citealp{Chen2011}).
 A brief inquiry into  some of the articles citing ComBat (574 on Google Scholar) revealed few reported problems, although their method descriptions regarding ComBat were mostly sparse, limited to one or two sentences.
+
 A further indication of the carefree use of the procedure was the frequent omittance of program parameters that were applied, i.e batch labels or whether group labels were supplied as covariates.
 Often, no effort was undertaken in order to substantiate the existence of batch effects, beyond stating the presence of batches.
-The incorporation of ComBat into analysis tools, like \COMMENT{TCGA}{does not seem to offer ComBat with covariates} and GenePattern (\citealp{Reich2006}), and other packages like \COMMENT{inSilicoDb (\citealp{Taminau2011})}{does not seem to offer ComBat with covariates} could make it more accessible, but its use less transparent.
-We are concerned, in light of our findings, that many published results from data adjusted by ComBat with study group as covariate may be unreliable.
+The incorporation of ComBat into various analysis tools appears to make it more accessible, but its use less transparent.
+Such tools include GenePattern (\citealp{Reich2006}), which offers ComBat with covariates, as well as TCGA and  inSilicoDb ((\citealp{Taminau2011}), presently without covariates as an option.
+
+We are concerned, in light of our findings, that many published results from batch adjusted data using study group as covariate may be unreliable.
 Furthermore, the frequent lack of proper method description accompanying published results makes it hard to judge if they are affected or not.
@@
 A superficial assessment of the reliability of results based on batch corrected data with group differences retained can often be made.
-The effective sample size, $\nu$, may be compared against the nominal sample size, $\nu_0$, as in section~\label{sec:twogroups} for the case of two groups: similar computations, although more complicated, can be made for multiple groups, but the two-group assessment should still give a fair idea of the reliability of pairwise group comparisons.
+In section~\label{sec:twogroups}, the ratio $\nu_0/\nu\ge1$ indicates by how much the effective sample size is overestimated when comparing two groups.
+A similar computations, although more complicated, can be made for multiple groups.
+However, the two-group assessment should still give a fair idea of the reliability of pairwise group comparisons.
 A major concern would be batches where either of the groups of interest are missing or strongly under-represented, which would contribute to the nominal sample size, but not to the effective sample size.
@@
 
-Finally, we would like to emphasise the importance of proper description of how data has been prepared for analysis, and what corrections and adjustments have been made.
+Finally, we would like to emphasise the importance of proper description of how data has been prepared for analysis, and what corrections and adjustments have been made (\citealp{Sandve2013}).
 In cases where data preparation is performed prior to, and separate from, the data analysis, as is now often the case, this is of particular importance, as artifacts may be introduced in the data preparation which could influence the reliability of downstream analyses.
@@
 
-The data and scripts used to generate the results in this work, i.e figures, are available at
+The data and scripts used to generate the results in this work are available at
 \href{https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-figures.git}{https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-figures.git}.
@@
 %\centering\includegraphics[width=13cm]{Fig/boxplots.png}
-\caption{We simulated expression of one gene from four study groups unevenly distributed in 3 batches containing 50+0+0+20, 0+50+0+20 and 0+0+50+20 samples from groups 1 to 4.
+\caption{We simulated expression of one gene from four study groups unevenly distributed in three batches containing 50+0+0+20, 0+50+0+20 and 0+0+50+20 samples from groups 1 to 4.
 This case, design and effects, was selected to illustrate the spurious effects that may arise from different batch adjustments.
-The Y-axis represent the expression, while the X-axis is used to better visually differentiate between the batches.
-Circles, triangles and crosses indicate the measurment of each sample and which batch it is part of.
-Boxes showing group mean with 95\% confidence intervals are displayed to the right of the measurements and are ment to imply if a statistical test would find significant group differences, true or false.
+The Y-axis represent the expression values, while the X-axis is used to visually separate the batches.
+Circles, triangles, and crosses indicate values from each of the three batches, with colours indicating study groups.
+Correspondingly coloured boxes to the right of the measurements show group means with 95\% confidence intervals.
+These give some indication as to which group differences would be found significant.
 a) The ``true'' values measured in a system without batch effects. Groups 1 and 2 have lower means than groups 3 and 4.
@@
 d) The measurements in b) are adjusted with two-way ANOVA based batch centering (using \texttt{limma}). Group 1 and 2 seem to differ.
-e) The least squares estimates of the group means from a two-way ANOVA which represent a more reliable way of estimating group differences of the measurements in b) than the two-step procedure in c) and d).
+e) The least squares estimates of the group means from a two-way ANOVA have the same means as in d), but more appropriate confidence intervals.
 }

### 2014-08-12 tex/manuscript.tex
Noen siste små justeringer

@@
 
+\renewcommand{\thefootnote}{§}
 \footnotetext{\color{blue}Manuscript version \today.}
@@
 % List of authors, with corresponding author marked by asterisk
-\author{VEGARD NYGAARD, EINAR ANDREAS RØDLAND\\[4pt]
+\renewcommand{\thefootnote}{\textdagger}
+\footnotetext{These two authors contributed equally.}
+\author{VEGARD NYGAARD$^\dagger$, EINAR ANDREAS RØDLAND$^\dagger$\\[4pt]
 % Author addresses
@@
 
-Two popular tools for batch adjustment of gene expression data are the R bioconductor packages \texttt{limma} (\citealp{Smyth2003}) and ComBat (\citealp{Johnson2007}) which is now included in the \texttt{sva} package (\citealp{Leek2012}).
-Both these packages allow the inclusion of covariates, e.g.
+A popular tool for batch adjustment of gene expression data is ComBat (\citealp{Johnson2007}) which is now included in the R bioconductor package \texttt{sva} (\citealp{Leek2012}).
+This allows the inclusion of covariates, e.g.
 group difference, the effects of which should not be removed in the batch adjustment.
-In addition, ComBat uses an empirical Bayes approach to avoid over-correction for batch effects, which is critical for small batches.
-Increasingly, the inclusion of group difference as a covariate when removing batch effects has been recommended, both in the \texttt{sva} tutorial and user fora, and ComBat has thus helped popularise this approach.
+In addition, it uses an empirical Bayes approach to avoid over-correction for batch effects, which is critical for small batches.
+Increasingly, the inclusion of group difference as a covariate when removing batch effects has been recommended, both in the \texttt{sva} tutorial and user fora.
+ComBat has thus helped popularise this approach.
 Based on the actual use of ComBat by the authors and others, ourselves included, we suspect that adjusted data are commonly treated as ``batch effect free'' in subsequent analyses, and as a consequence, confidence in group effects has been overestimated and false results reported.
+The R package \texttt{limma} (\citealp{Smyth2003}) also offers batch adjustment with covariates, but notes that this is not intended used prior to linear modeling.
 
 When sample or batch sizes are small, statisticians would most likely take extra precautions.
-However, batch adjustment using two-way ANOVA or ComBat on unbalanced data sets may be just as harmful on large samples and batch sizes as for small.
+However, batch adjustment using two-way ANOVA or ComBat on unbalanced data sets may be just as harmful for large samples and batch sizes as for small.
 For example, group comparisons using one-way ANOVA on the batch adjusted data will essentially result in $F$ statistics that are inflated by a fixed factor which depends on the unevenness of the design, rather than the size of the sample or batches.
@@
 \widetilde{Y}^{\textrm{cov}}_{\textrm{-}j}-\widetilde{Y}^{\textrm{cov}}_{\textrm{-}j'}
-=\beta_j-\beta_{j'}+\bar{\epsilon}_{\textrm{-}j}-\bar{\epsilon}_{\textrm{-}j'}
+=(\beta_j-\beta_{j'})+(\bar{\epsilon}_{\textrm{-}j}-\bar{\epsilon}_{\textrm{-}j'})
  -\sum_{i=1}^m \left(\frac{n_{ij}}{n_{\textrm{-}j}}-\frac{n_{ij'}}{n_{\textrm{-}j'}}\right)(\hat{\gamma}_i-\gamma_i).
@@
 The undesired consequences of preserving group effects when correcting for batch effect is readily illustrated with a sanity check using random numbers.
-The documentation accompanying the \texttt{sva} library has a executable example demonstrating how to adjust a data set with ComBat, followed by an F-test.
+The documentation accompanying the \texttt{sva} library has an executable example demonstrating how to adjust a data set with ComBat, followed by an F-test.
 Replacing the real data with random numbers from a standard normal distribution (mean=0, sd=1), but otherwise following the instructions, will generate the p-value distribution shown in Figure~\ref{fig:sanity}a.
@@
 The extent to which batch adjustment with group differences retained will confound subsequent analyses depends on the batch--group balance.
-We have reanalysed to cases with varying degree of unbalance.
+We have reanalysed two cases with varying degree of unbalance.
 
@@
 In addition, other studies investigating batch effects were mostly recommending ComBat without much concern (\citealp{Kupfer2012}, \citealp{Kitchen2011}, \citealp{Chen2011}).
-A brief inquiry into  some of the articles citing ComBat (574 on Google Scholar) revealed few reported problems, although their method descriptions regarding ComBat were mostly sparse, limited to one or two sentences.
+A brief inquiry into some of the articles citing ComBat (466 in Web of Science) revealed few reported problems, although their method descriptions regarding ComBat were mostly sparse, limited to one or two sentences.
+Some used ComBat with covariates, some without, but we did not find any that addressed the potential problems related to use of covariates.
 
@@
 The incorporation of ComBat into various analysis tools appears to make it more accessible, but its use less transparent.
-Such tools include GenePattern (\citealp{Reich2006}), which offers ComBat with covariates, as well as TCGA and  inSilicoDb ((\citealp{Taminau2011}), presently without covariates as an option.
+Such tools include GenePattern (\citealp{Reich2006}), which offers ComBat with covariates, as well as TCGA and  inSilicoDb (\citealp{Taminau2011}), presently without covariates as an option.
 
@@
 A superficial assessment of the reliability of results based on batch corrected data with group differences retained can often be made.
-In section~\label{sec:twogroups}, the ratio $\nu_0/\nu\ge1$ indicates by how much the effective sample size is overestimated when comparing two groups.
-A similar computations, although more complicated, can be made for multiple groups.
+In section~\ref{sec:twogroups}, the ratio $\nu_0/\nu\ge1$ indicates by how much the effective sample size is overestimated when comparing two groups.
+A similar computation, although more complicated, can be made for multiple groups.
 However, the two-group assessment should still give a fair idea of the reliability of pairwise group comparisons.
@@
 
-[...Acknowledgements...]
+We are grateful to Geir Kjetil Sandve for useful suggestions and discussions.
+
+This project was in part financed through the EUROCAN platform (VN) and the MetAction project (EAR).
+
 {\it Conflict of Interest}: None declared.

### 2014-08-13 tex/manuscript.tex
Siste rettelse

@@
 
-To explain more clearly what is happening, and quantify the size of the problem, we may consider the simple case of estimating the difference $\Delta\beta=\beta_A-\beta_B$ between two groups, $A$ and $B$, when there are $m$ batches with batch $i=1,\ldots,m$ containing $n_{iA}$ and $n_{iB}$ samples from each of the two groups.
+To explain more clearly what is happening, and to quantify the size of the problem, we may consider the simple case of estimating the difference $\Delta\beta=\beta_A-\beta_B$ between two groups, $A$ and $B$, when there are $m$ batches with batch $i=1,\ldots,m$ containing $n_{iA}$ and $n_{iB}$ samples from each of the two groups.
 

### 2014-08-13 figurescript.r
Lagt marg rundt boxplot og typesatt på ny

@@ -8,6 +8,9 @@
 # source("http://bioconductor.org/biocLite.R")
 # biocLite( c("sva", "limma", "lsmeans", "bladderbatch"))
 
+# If R is not started with the appropriate folder as work directory:
+#setwd("...path to batch-adjust-warning-figures folder...")
+
 startwd = getwd()
 source("overviewplot.r")
 

### 2014-08-13 overviewplot.r

@@ -46,7 +46,7 @@ figfilename = file.path( getwd(), "boxplots")
 figfile = paste( figfilename, ".pdf", sep=""); 
 pdf(file =figfile, width=24, height=12)
 
-op=par(mfrow=c(1, 5), 	xaxt="n",yaxt="n", mar=c(5,0,4,0))
+op=par(mfrow=c(1, 5), 	xaxt="n",yaxt="n", mar=c(5,0,0,0),oma=c(0,.5,.5,.5))
 alldata = c(matrix_condition[index,],matrix_conditionbatch[index,],matrix_meancenter[index,],matrix_limma[index,])
 ylim = c(min(alldata), max(alldata))
 adhocboxplot(matrix_condition[index,], group, batch, ylim=ylim, xlab="Without batch effects", figureletter="a")

### 2014-08-13 commonscripts/boxplot_function.r
Update boxplot_function.r

@@ -60,7 +60,7 @@ adhocboxplot2 = function(lsmeans, grouplabels, batchlabels, addlegend=FALSE, yli
 	x = as.numeric(batchlabels)
 	lwd=2
 	
-	xlim = c(0, length(unique(batchlabels))+3)
+	xlim = c(0, length(unique(batchlabels))+4)
 	plot(1, 1, col=colpalette[grouplabels], pch=pchselection[batchlabels], 
 			 xlim=xlim, lwd=lwd, cex=cex, ylim=ylim, xlab=xlab, ylab="", cex.lab=adhoc.labcex, type="n")
 	

### 2014-08-13 tex/manuscript.tex
Fikset opp i henhold til journalens normer

@@
-\documentclass{bio}
-%\documentclass[oupdraft]{bio}
+%\documentclass{bio}
+\documentclass[oupdraft]{bio}
 \usepackage[colorlinks=true, urlcolor=citecolor, linkcolor=citecolor, citecolor=citecolor]{hyperref}
@@
 %    \CITE{missing/unspecified citation}
-\usepackage{color}
-\definecolor{NBcol}{rgb}{0.7,0,1}
-\definecolor{CITEcol}{rgb}{1,0,0.7}
-\definecolor{COMMENTcol}{rgb}{0.7,0.2,0}
-\newcommand\COMMENT[2]{\textcolor{COMMENTcol}{#1}\NOTE{#2}}
-\newcommand\NB[1]{\textcolor{NBcol}{\textit{#1}}}
-\newcommand\NOTE[1]{\NB{[#1]}}
-\newcommand\CITE[1]{\textcolor{CITEcol}{[#1]}}
+%\usepackage{color}
+%\definecolor{NBcol}{rgb}{0.7,0,1}
+%\definecolor{CITEcol}{rgb}{1,0,0.7}
+%\definecolor{COMMENTcol}{rgb}{0.7,0.2,0}
+%\newcommand\COMMENT[2]{\textcolor{COMMENTcol}{#1}\NOTE{#2}}
+%\newcommand\NB[1]{\textcolor{NBcol}{\textit{#1}}}
+%\newcommand\NOTE[1]{\NB{[#1]}}
+%\newcommand\CITE[1]{\textcolor{CITEcol}{[#1]}}
 
@@
 
-\renewcommand{\thefootnote}{§}
-\footnotetext{\color{blue}Manuscript version \today.}
+% Manuscript version (date) in separate footnote
+\renewcommand{\thefootnote}{§}\footnotetext{\color{blue}Manuscript version \today.}
 
@@
 
-% List of authors, with corresponding author marked by asterisk
+% Set other footnote mark for shared first author footnote
 \renewcommand{\thefootnote}{\textdagger}
 \footnotetext{These two authors contributed equally.}
+
+% List of authors, with corresponding author marked by asterisk
 \author{VEGARD NYGAARD$^\dagger$, EINAR ANDREAS RØDLAND$^\dagger$\\[4pt]
@@
 \begin{equation}
-\widetilde{Y}^0_{ijr}=Y_{ijr}-\bar{Y}_{i}
+\tilde{Y}^0_{ijr}=Y_{ijr}-\bar{Y}_{i}
 \quad\text{where}\quad
@@
 An alternative is to centre each batch to the common average by adding the average value $\bar{Y}$ across the entire data set: i.e.
-$\widetilde{Y}^{\textrm{avg}}_{ijr}=\widetilde{Y}^0_{ijr}+\bar{Y}$.
+$\tilde{Y}^{\textrm{avg}}_{ijr}=\tilde{Y}^0_{ijr}+\bar{Y}$.
 When comparing groups, the common value $\bar{Y}$ has no effect, and so this is equivalent to zero-centring each batch.
@@
 \begin{equation}
-\widetilde{Y}^0_{ijr}=\beta_j-\bar{\beta}_i+\epsilon_{ijr}-\bar{\epsilon}_i
+\tilde{Y}^0_{ijr}=\beta_j-\bar{\beta}_i+\epsilon_{ijr}-\bar{\epsilon}_i
 \quad\text{where}\quad
@@
 \begin{equation}
-\widetilde{Y}^{\textrm{cov}}_{ijr}=Y_{ijr}-\hat\gamma_i=\alpha+\beta_j+(\gamma_i-\hat\gamma_i)+\epsilon_{ijr}.
+\tilde{Y}^{\textrm{cov}}_{ijr}=Y_{ijr}-\hat\gamma_i=\alpha+\beta_j+(\gamma_i-\hat\gamma_i)+\epsilon_{ijr}.
 \end{equation}
@@
 \begin{equation}
-\widetilde{Y}^{\textrm{cov}}_{\textrm{-}j}
-=\frac{1}{n_{\textrm{-}j}} \sum_{i=1}^m\sum_{r=1}^{n_{ij}}\widetilde{Y}_{ijr}
+\tilde{Y}^{\textrm{cov}}_{\textrm{-}j}
+=\frac{1}{n_{\textrm{-}j}} \sum_{i=1}^m\sum_{r=1}^{n_{ij}}\tilde{Y}_{ijr}
 =\alpha+\beta_j+\bar{\epsilon}_{\textrm{-}j}
@@
 \begin{equation}
-\widetilde{Y}^{\textrm{cov}}_{\textrm{-}j}-\widetilde{Y}^{\textrm{cov}}_{\textrm{-}j'}
+\tilde{Y}^{\textrm{cov}}_{\textrm{-}j}-\tilde{Y}^{\textrm{cov}}_{\textrm{-}j'}
 =(\beta_j-\beta_{j'})+(\bar{\epsilon}_{\textrm{-}j}-\bar{\epsilon}_{\textrm{-}j'})
@@
 \begin{equation}
-\widehat{\Delta\beta}_i=\bar{Y}_{iA}-\bar{Y}_{iB}\sim\text{N}\left(\Delta\beta,\frac{\sigma^2}{\nu_i}\right)
+\Delta\hat{\beta}_i=\bar{Y}_{iA}-\bar{Y}_{iB}\sim\text{N}\left(\Delta\beta,\frac{\sigma^2}{\nu_i}\right)
 \quad\text{where}\quad
@@
 \begin{equation}
-\widehat{\Delta\beta}=\frac{\sum_{i=1}^m \nu_i\,\widehat{\Delta\beta}_i}{\nu}
+\Delta\hat{\beta}=\frac{\sum_{i=1}^m \nu_i\,\Delta\hat{\beta}_i}{\nu}
 \sim\text{N}\left(\Delta\beta,\frac{\sigma^2}{\nu}\right)
@@
 
-If batch and group effects are estimated using a two-way ANOVA, the estimate $\widehat{\Delta\beta}$ will be as stated above, and so the estimated group difference is unaffected.
-The batch effects are then removed, and the estimated group differences retained, leaving the estimated $\widehat{\Delta\beta}$ unchanged by the batch adjustment.
-However, if this batch adjusted data set is analysed without considering batch effects, the variance of $\widehat{\Delta\beta}$ will be computed under the assumption that it is derived from a comparison of $n_A=\sum_{i=1}^m n_{iA}$ versus $n_B=\sum_{i=1}^m n_{iB}$ samples, and thus satisfy
+If batch and group effects are estimated using a two-way ANOVA, the estimate $\Delta\hat{\beta}$ will be as stated above, and so the estimated group difference is unaffected.
+The batch effects are then removed, and the estimated group differences retained, leaving the estimated $\Delta\hat{\beta}$ unchanged by the batch adjustment.
+However, if this batch adjusted data set is analysed without considering batch effects, the variance of $\Delta\hat{\beta}$ will be computed under the assumption that it is derived from a comparison of $n_A=\sum_{i=1}^m n_{iA}$ versus $n_B=\sum_{i=1}^m n_{iB}$ samples, and thus satisfy
 \begin{equation}
-\widehat{\Delta\beta}
+\Delta\hat{\beta}
 \sim\text{N}\left(\Delta\beta,\frac{\sigma^2}{\nu_0}\right)
@@
 
-\section{Supplementary Material}
-\label{supp}
+%\section{Supplementary Material}\label{supp}
 
-Supplementary material is available online at
-\href{http://biostatistics.oxfordjournals.org}%
-{http://biostatistics.oxfordjournals.org}.
+%Supplementary material is available online at
+%\href{http://biostatistics.oxfordjournals.org}%
+%{http://biostatistics.oxfordjournals.org}.
 
-\subsection{Reproducible research}
+\section*{Reproducible research}
 
@@
 
-This project was in part financed through the EUROCAN platform (VN) and the MetAction project (EAR).
+This work was supported by the EUROCAN platform (VN) and the MetAction project (EAR).
 

### 2014-08-14 commonscripts/boxplot_function.r
Revert "Revert "Delete tmpvegard.txt""
This reverts commit 26b8cc411590518fa9f20253dd48b556b2a9caac.

Revert "Delete tmpvegard.txt"

This reverts commit
913aac7267f670243e973f24af934fe204c95530.

@@ -1,7 +1,10 @@
 # Functions for producing a boxplot
 
-adhoc.palette =  c("blue", "red", "darkgreen", "darkorange")
-adhoc.pch = c(1, 2, 3)
+#adhoc.palette =  c("blue", "red", "darkgreen", "darkorange")
+#adhoc.pch = c(1, 2, 3,4)
+#adhoc.palette =  c("blue", "red", "darkgreen")
+adhoc.palette =  c("black","darkgray","gray","lightgray")
+adhoc.pch = c(1, 2, 3,4)
 adhoc.cex=3
 adhoc.legendcex=adhoc.cex
 adhoc.labcex=adhoc.cex
@@ -16,7 +19,7 @@ adhocboxplot = function(y, grouplabels, batchlabels, addlegend=FALSE, ylim=NULL,
   
   # add extra x for better visual separation
   for(b in unique(batchlabels)) {
-    separation = 0.15
+    separation = 0.25
     xoffset=0
     for(g in unique(grouplabels[batchlabels==b])) {
       x[batchlabels==b & grouplabels==g] = x[batchlabels==b & grouplabels==g] + xoffset
@@ -25,14 +28,16 @@ adhocboxplot = function(y, grouplabels, batchlabels, addlegend=FALSE, ylim=NULL,
   }
   
   xlim = c(0, length(unique(batchlabels))+4)
-  plot(x, y, col=colpalette[grouplabels], pch=pchselection[batchlabels], 
+#  plot(x, y, col=colpalette[grouplabels], pch=pchselection[batchlabels], 
+#       xlim=xlim, lwd=lwd, cex=cex, ylim=ylim, xlab=xlab, ylab="", cex.lab =labelcex)
+  plot(x, y, col=colpalette[batchlabels], pch=pchselection[grouplabels], 
        xlim=xlim, lwd=lwd, cex=cex, ylim=ylim, xlab=xlab, ylab="", cex.lab =labelcex)
 
   #add boxplot
   groupnames = sort(unique(grouplabels))
   xoffset=length(unique(batchlabels))+2
   boxwidth=0.35
-  boxseparation = 0.25
+  boxseparation = 0.45
   for(g in groupnames) {
     g=factor(g,levels=levels(groupnames))
     m = mean(y[grouplabels==g])

### 2014-08-14 commonscripts/boxplot_function.r
Tried to make B&W version of figure...not good

@@ -1,10 +1,7 @@
 # Functions for producing a boxplot
 
-#adhoc.palette =  c("blue", "red", "darkgreen", "darkorange")
-#adhoc.pch = c(1, 2, 3,4)
-#adhoc.palette =  c("blue", "red", "darkgreen")
-adhoc.palette =  c("black","darkgray","gray","lightgray")
-adhoc.pch = c(1, 2, 3,4)
+adhoc.palette =  c("blue", "red", "darkgreen", "darkorange")
+adhoc.pch = c(1, 2, 3)
 adhoc.cex=3
 adhoc.legendcex=adhoc.cex
 adhoc.labcex=adhoc.cex
@@ -19,7 +16,7 @@ adhocboxplot = function(y, grouplabels, batchlabels, addlegend=FALSE, ylim=NULL,
   
   # add extra x for better visual separation
   for(b in unique(batchlabels)) {
-    separation = 0.25
+    separation = 0.15
     xoffset=0
     for(g in unique(grouplabels[batchlabels==b])) {
       x[batchlabels==b & grouplabels==g] = x[batchlabels==b & grouplabels==g] + xoffset
@@ -28,16 +25,14 @@ adhocboxplot = function(y, grouplabels, batchlabels, addlegend=FALSE, ylim=NULL,
   }
   
   xlim = c(0, length(unique(batchlabels))+4)
-#  plot(x, y, col=colpalette[grouplabels], pch=pchselection[batchlabels], 
-#       xlim=xlim, lwd=lwd, cex=cex, ylim=ylim, xlab=xlab, ylab="", cex.lab =labelcex)
-  plot(x, y, col=colpalette[batchlabels], pch=pchselection[grouplabels], 
+  plot(x, y, col=colpalette[grouplabels], pch=pchselection[batchlabels], 
        xlim=xlim, lwd=lwd, cex=cex, ylim=ylim, xlab=xlab, ylab="", cex.lab =labelcex)
 
   #add boxplot
   groupnames = sort(unique(grouplabels))
   xoffset=length(unique(batchlabels))+2
   boxwidth=0.35
-  boxseparation = 0.45
+  boxseparation = 0.25
   for(g in groupnames) {
     g=factor(g,levels=levels(groupnames))
     m = mean(y[grouplabels==g])

### 2014-10-28 samplesizescalingplot.r
samplesizescalingplot

@@ -1,0 +1,70 @@
+
+### Figure of box-plots and confidence intervals
+
+library(limma)
+library(sva)
+
+source("commonscripts/helperfunctions.r")
+
+rseed=1111
+set.seed(rseed)
+
+
+pvalcounts = list()
+
+for(n in c(1,10,100))
+{
+	sampleannotation = createsampleannotation(  list(c(n,n*5), c(n*5,n) )  ) 	
+	
+	x = table(sampleannotation[,2:3])	
+	runname = paste(paste(x[,1], collapse="_"),  paste(x[,2], collapse="_"), sep="_x_")
+	
+	ngenes=20000
+	batchaffectedgenefraction = 0.0001
+
+	matrix_random = matrix(rnorm(ngenes * nrow(sampleannotation), mean=0, sd=1), nrow=ngenes, ncol=nrow(sampleannotation))
+	for(batch in (unique(sampleannotation$batch)))
+	{
+		for(s in 1: (dim(matrix_random)[1]*batchaffectedgenefraction) )
+		{
+			thisgenesbatcheffect = rnorm(1)
+			a= sampleannotation$batch==batch
+			matrix_random[s, a] = matrix_random[s, a] + thisgenesbatcheffect
+		}
+	}
+	
+		mod0 = model.matrix(~1,data=sampleannotation)
+ 		batch = factor(sampleannotation$batch)
+ 		mod = model.matrix(~as.factor(treatment), data=sampleannotation)
+# 	combat_edata = ComBat(dat=matrix_random, batch=batch, 
+# 												mod=mod, numCovs=NULL, 
+# 												par.prior=TRUE, prior.plots=FALSE)
+	
+	combat_edata = removeBatchEffect(matrix_random, batch=batch, design=mod)
+	
+	pValuesComBat = f.pvalue(combat_edata,mod,mod0)
+	pvalcounts[[runname]]=hist(pValuesComBat, plot=FALSE, breaks=100)$counts
+#qValuesComBat = p.adjust(pValuesComBat,method="BH")
+}
+
+figfilename = file.path( getwd(), "samplesizescaling")
+figfile = paste( figfilename, ".pdf", sep=""); 
+pdf(file =figfile, width=24, height=12)
+
+#colorpal = c("blue", "red", "green", "black")
+colorpal = c("black", "black","black","black")
+plot(  (1:25)/100, 1:25, 
[…]

### 2014-10-29 commonscripts/boxplot_function_v2.r
prøver å commite til revision

@@ -1,0 +1,91 @@
+# Functions for producing a boxplot
+
+#adhoc.palette =  c("blue", "red", "darkgreen", "darkorange")
+adhoc.palette =  c("black","black","black","black")
+adhoc.pch = c(1, 2, 3, 4)
+adhoc.cex=3
+adhoc.legendcex=adhoc.cex
+adhoc.labcex=adhoc.cex
+
+# y=matrix_conditionbatch[index,]
+# grouplabels=group
+# batchlabels=batch
+# addlegend=FALSE
+# ylim=ylim
+# xlab="Without batch effects"
+# figureletter="a"
+# cex=adhoc.cex
+# labelcex=adhoc.labcex
+# legendcex=adhoc.legendcex
+# estimatemethod="lsmeans"
+
+# Boxplot component: values and intervals
+adhocboxplot = function(y, grouplabels, batchlabels, addlegend=FALSE, ylim=NULL, xlab="x",figureletter=NA,cex=adhoc.cex,labelcex=adhoc.labcex, legendcex=adhoc.legendcex, estimatemethod="CI")
+{
+  colpalette = adhoc.palette
+  pchselection = adhoc.pch
+
+  lwd=2
+  
+  xlim = c(1, length(y)+50)
+  plot(y, col=colpalette[grouplabels], main=xlab, cex.main=labelcex, pch=pchselection[grouplabels],
+       xlim=xlim, lwd=lwd, cex=cex, ylim=ylim, ylab="", xlab=NA, cex.lab =labelcex)
+
+  #add batch separators
+  for(thisbatch in unique(batchlabels))
+  {
+    sepx = length(batchlabels) - match(thisbatch, batchlabels[length(batchlabels):1])+1+0.5
+    abline(v=sepx, col="black", lwd=lwd)
+    text(x=sepx-length(batchlabels)/7, y=ylim[2]-0.5, labels=paste("Batch ", thisbatch, sep=""), cex=legendcex)    
+  }
+  
+  #add boxplot
+  groupnames = sort(unique(grouplabels))
+  xoffset=round(length(y)* 1.1)
+  text(x=xoffset+boxwidth, y=ylim[2]-0.5, labels=estimatemethod, cex=legendcex)
+  boxwidth=round(length(y)/20)
+  boxseparation = round(boxwidth/1.5)
+  for(g in groupnames) {
+    g=factor(g,levels=levels(groupnames))
+    
+    if(estimatemethod=="lsmeans")
+    {
[…]

### 2014-10-29 overviewplot_v2.r

@@ -1,0 +1,64 @@
+
+### Figure of box-plots and confidence intervals
+
+library(limma)
+library(lsmeans)
+
+source("commonscripts/boxplot_function_v2.r")
+source("commonscripts/helperfunctions.r")
+
+# 3 batches.
+# 4 groups
+#sampleannotation = createsampleannotation(  list(c(100,0,0,20), c(0,100,0,20), c(0,0,100,20))) 
+sampleannotation = createsampleannotation(  list(c(50,0,0,20), c(0,50,0,20), c(0,0,50,20))) 
+#sampleannotation = createsampleannotation(  list(c(25,0,0,5), c(0,25,0,5), c(0,0,25,5))) 
+
+table(sampleannotation[,2:3])
+group=factor(sampleannotation$treatment)
+batch=factor(sampleannotation$batch)
+
+ngenes=1000
+#matrix_random = createrandomdata(ngenes, sampleannotation, mean=0, sd=1)
+rseed=1111
+set.seed(rseed)
+
+
+matrix_random = matrix(rnorm(ngenes * nrow(sampleannotation), mean=0, sd=1), nrow=ngenes, ncol=nrow(sampleannotation))
+
+matrix_condition = matrix_random
+matrix_condition[, group %in% c(3,4)] = matrix_condition[, group %in% c(3,4)] +3
+matrix_conditionbatch=matrix_condition
+matrix_conditionbatch[, batch ==2] = matrix_condition[, batch ==2] -1
+matrix_conditionbatch[, batch ==3] = matrix_condition[, batch ==3] +1
+matrix_meancenter = matrix_conditionbatch
+matrix_meancenter[,batch ==1] =  matrix_meancenter[,batch ==1] - rowMeans( matrix_meancenter[,batch ==1])
+matrix_meancenter[,batch ==2] =  matrix_meancenter[,batch ==2] - rowMeans( matrix_meancenter[,batch ==2])
+matrix_meancenter[,batch ==3] =  matrix_meancenter[,batch ==3] - rowMeans( matrix_meancenter[,batch ==3])
+mod = model.matrix(~group)
+matrix_limma = removeBatchEffect(matrix_conditionbatch, batch=batch, design=mod)
+
+
+index=1
+#fit_lm=lm(matrix_condition[index,] ~ group+batch)
+#means_lm=lsmeans(fit_lm,~group)
+#summary(means_lm)
+#contrast(means_lm,"pairwise")
+
[…]

### 2014-10-29 commonscripts/boxplot_function_v2.r
mer plots

@@ -41,10 +41,10 @@ adhocboxplot = function(y, grouplabels, batchlabels, addlegend=FALSE, ylim=NULL,
   
   #add boxplot
   groupnames = sort(unique(grouplabels))
-  xoffset=round(length(y)* 1.1)
-  text(x=xoffset+boxwidth, y=ylim[2]-0.5, labels=estimatemethod, cex=legendcex)
+  xoffset=round(length(y)* 1.1)  
   boxwidth=round(length(y)/20)
   boxseparation = round(boxwidth/1.5)
+  text(x=xoffset+boxwidth, y=ylim[2]-0.5, labels=estimatemethod, cex=legendcex)
   for(g in groupnames) {
     g=factor(g,levels=levels(groupnames))
     

### 2014-10-29 samplesizescalingplot.r

@@ -10,61 +10,76 @@ rseed=1111
 set.seed(rseed)
 
 
-pvalcounts = list()
+pvalcounts = list(ComBat=list(), LIMMA=list())
+batchaffectedgenefraction = 0
+samplesizescaleing = c(1,10,100)
+ngenes=20000
 
-for(n in c(1,10,100))
+for(n in samplesizescaleing)
 {
 	sampleannotation = createsampleannotation(  list(c(n,n*5), c(n*5,n) )  ) 	
 	
 	x = table(sampleannotation[,2:3])	
 	runname = paste(paste(x[,1], collapse="_"),  paste(x[,2], collapse="_"), sep="_x_")
 	
-	ngenes=20000
-	batchaffectedgenefraction = 0.0001
+	
+	
 
 	matrix_random = matrix(rnorm(ngenes * nrow(sampleannotation), mean=0, sd=1), nrow=ngenes, ncol=nrow(sampleannotation))
 	for(batch in (unique(sampleannotation$batch)))
 	{
-		for(s in 1: (dim(matrix_random)[1]*batchaffectedgenefraction) )
-		{
-			thisgenesbatcheffect = rnorm(1)
-			a= sampleannotation$batch==batch
-			matrix_random[s, a] = matrix_random[s, a] + thisgenesbatcheffect
-		}
+    if((dim(matrix_random)[1]*batchaffectedgenefraction) >= 1)
+    {
+		  for(s in 1: (dim(matrix_random)[1]*batchaffectedgenefraction) )
+		  {
+		  	thisgenesbatcheffect = rnorm(1)
+		  	a= sampleannotation$batch==batch
+		  	matrix_random[s, a] = matrix_random[s, a] + thisgenesbatcheffect
+		  }
+    }
 	}
 	
 		mod0 = model.matrix(~1,data=sampleannotation)
  		batch = factor(sampleannotation$batch)
- 		mod = model.matrix(~as.factor(treatment), data=sampleannotation)
-# 	combat_edata = ComBat(dat=matrix_random, batch=batch, 
-# 												mod=mod, numCovs=NULL, 
-# 												par.prior=TRUE, prior.plots=FALSE)
+
+    mod = model.matrix(~as.factor(treatment), data=sampleannotation)
+	matrix_adjusted = ComBat(dat=matrix_random, batch=batch, 
+ 												mod=mod, numCovs=NULL, 
+ 												par.prior=TRUE, prior.plots=FALSE)
[…]

### 2014-10-31 commonscripts/boxplot_function_v2.r
overviewplot v3

@@ -27,7 +27,7 @@ adhocboxplot = function(y, grouplabels, batchlabels, addlegend=FALSE, ylim=NULL,
 
   lwd=2
   
-  xlim = c(1, length(y)+50)
+  xlim = c(1, round(length(y)*1.25))
   plot(y, col=colpalette[grouplabels], main=xlab, cex.main=labelcex, pch=pchselection[grouplabels],
        xlim=xlim, lwd=lwd, cex=cex, ylim=ylim, ylab="", xlab=NA, cex.lab =labelcex)
 

### 2014-10-31 commonscripts/boxplot_function_v3.r

@@ -1,0 +1,112 @@
+
+
+
+# y=matrix_true[1,]
+# group=sampleannotation$group
+# batch=sampleannotation$batch
+# #batch=NA
+# estimatemethod="lsmeans"
+# ylim=NULL
+
+plot_one_gene = function(y, group, batch=NULL, ylim=NULL, main="", estimatemethod="none")
+{
+  lwd=2
+  a = order(group)
+  if(!is.null(batch))
+  {
+    a = order(batch, group)
+    batch=batch[a]
+  }
+  group=group[a]
+  
+  y=y[a]
+  
+  thispal=c("red", "blue", "green", "orange", "brown", "cyan")
+  
+  if(is.null(ylim))
+  {
+    ymax = round(max(y)+2, 1)
+    ymin = round(min(y)-2, 1)
+    ylim=c(ymin,ymax)
+  }
+  xlim = c(1, round(length(y)*1.25))
+  
+  # measurements
+  plot(y, ylim=ylim, xlim=xlim, col=thispal[as.factor(group)], main=main,  ylab="Expression", lwd=2, 
+       xlab="samples")
+  
+  # batch separator
+  if(!is.null(batch))
+  {
+    batchnames=unique(batch)
+    x = match(batchnames, batch)
+    abline(v = x[-1]-0.5)#
+    batchlabelx=vector()
+    for(i in 1:(length(x)-1))
+    {
+      batchlabelx = c(batchlabelx, (x[i+1]+x[i])/2)
+    }
+    batchlabelx = c(batchlabelx, (x[i+1]+length(y))/2)	  
+    text(labels="Batch", y=ymin+0.5, x=2)
+    text(labels=paste(batchnames, sep=""), y=ymin+0.5, x=batchlabelx, cex=1.5)	  
+  }
+  
+  # group means line
+  groupnames=factor(unique(group))
+  groupmeans = vector()
+  for(thisgroup in groupnames)
+  {
+    groupmeans[thisgroup]=round(mean(y[group==thisgroup]),2)
+  }  
+  segments(  x0=(1:length(y))-0.5, y0=groupmeans[group], 
+             x1=(1:length(y))+0.5, y1=groupmeans[group], 
+             col=thispal[as.factor(group)], lwd=2)
+  #legend("topright", legend=paste(groupnames, groupmeans), text.col=thispal[as.factor(groupnames)])
+  
+  
+  # Boxplots for CI etc.  
[…]

### 2014-10-31 overviewplot.r

@@ -13,7 +13,7 @@ source("commonscripts/helperfunctions.r")
 sampleannotation = createsampleannotation(  list(c(50,0,0,20), c(0,50,0,20), c(0,0,50,20))) 
 
 table(sampleannotation[,2:3])
-group=factor(sampleannotation$treatment)
+group=factor(sampleannotation$group)
 batch=factor(sampleannotation$batch)
 
 ngenes=1000

### 2014-10-31 overviewplot_v2.r

@@ -14,7 +14,7 @@ sampleannotation = createsampleannotation(  list(c(50,0,0,20), c(0,50,0,20), c(0
 #sampleannotation = createsampleannotation(  list(c(25,0,0,5), c(0,25,0,5), c(0,0,25,5))) 
 
 table(sampleannotation[,2:3])
-group=factor(sampleannotation$treatment)
+group=factor(sampleannotation$group)
 batch=factor(sampleannotation$batch)
 
 ngenes=1000

### 2014-10-31 samplesizescalingplot.r

@@ -42,7 +42,7 @@ for(n in samplesizescaleing)
 		mod0 = model.matrix(~1,data=sampleannotation)
  		batch = factor(sampleannotation$batch)
 
-    mod = model.matrix(~as.factor(treatment), data=sampleannotation)
+    mod = model.matrix(~as.factor(group), data=sampleannotation)
 	matrix_adjusted = ComBat(dat=matrix_random, batch=batch, 
  												mod=mod, numCovs=NULL, 
  												par.prior=TRUE, prior.plots=FALSE)

### 2014-10-31 alternativefigure1.rmd
overviewplot v3

---
title: "Alternative Overview Figures"
author: "Vegard"
date: "Thursday, October 30, 2014"
output: html_document
---

Two scenarios

- Two groups, three batches. False Positives due to batch and ComBat.
- Three groups, two batches. False negatives due to batch, false positve due to mean adjust.

### False Positives due to batch and ComBat.

lage data set, 1000gener, 10+10
ingen gruppe effect eller batch effect.

1. plotte, målineg + meanstrek + CI + FC + p

2.
Dele i batcher 9+2+9, legge til batch effect.
plotte, målineg + meanstrek + CI + FC + p med batch separator i plottet.

3. Combat adjust
plotte, målineg + meanstrek + CI + FC + p uten batch separator i plottet.

4 plot med 2way anova og least square. Funker vel neppe med 1+1 bathen?

```{r}
library(sva)
source("commonscripts/helperfunctions.r")
source("commonscripts/boxplot_function_v3.r")
library(limma)
library(lsmeans)

set.seed(100)
ngenes = 1000
sampleannotation = createsampleannotation(  list(c(9,0), c(3,3), c(0,9)))
sampleannotation$batch=as.factor(sampleannotation$batch)
sampleannotation$group=as.factor(sampleannotation$group)
matrix_true = matrix(rnorm(ngenes * nrow(sampleannotation), mean=5, sd=1), nrow=ngenes, ncol=nrow(sampleannotation))

```

```{r, echo=FALSE}
plot_one_gene(matrix_true[1,], group=sampleannotation$group, main=paste("True values"),
              estimatemethod="CI")
```

Add batch effect.
```{r}
matrix_batcheffect = matrix_true
matrix_batcheffect[,sampleannotation$batch==1] = matrix_batcheffect[,sampleannotation$batch==1]+1
matrix_batcheffect[,sampleannotation$batch==3] = matrix_batcheffect[,sampleannotation$batch==3]-1
plot_one_gene(matrix_batcheffect[1,], group=sampleannotation$group, batch=sampleannotation$batch,
              main=paste("Batch affected values"), estimatemethod="CI")
```

Adjust for batch effect.
```{r}

mod0 = model.matrix(~1,data=sampleannotation)
mod = model.matrix(~as.factor(sampleannotation$group))

matrix_batchadjusted = ComBat(dat=matrix_batcheffect, batch=as.factor(sampleannotation$batch), 
                      mod=mod, numCovs=NULL, 
                      par.prior=TRUE, prior.plots=FALSE)
matrix_batchadjusted = removeBatchEffect(matrix_batcheffect, batch=as.factor(sampleannotation$batch),
                                         design=mod)

plot_one_gene(matrix_batchadjusted[1,], group=sampleannotation$group, batch=sampleannotation$batch,
              main=paste("Batch adjusted values"), estimatemethod="CI")
```

```{r}
plot_one_gene(matrix_true[1,], group=sampleannotation$group, batch=sampleannotation$batch,
              main=paste("True values, batch included in ANOVA"),estimatemethod="lsmeans")

```

```{r}
batch2FC = mean(matrix_true[1,sampleannotation$group==1 & sampleannotation$batch==2]) - 
          mean(matrix_true[1,sampleannotation$group==2 & sampleannotation$batch==2])

```

### 2014-10-31 reanalysis/Towfic2014/figurescript_towfic.r
div

@@ -51,7 +51,7 @@ limma_p_woc = limma_ret_woc$p.value[,1]
 print(paste("Limma adjusted real data, significant probes (fdr<0.05): ",  
             sum(p.adjust(limma_p_woc, method="fdr")<0.05)))
 
-# radnom, ComBat adjusted
+# random, ComBat adjusted
 set.seed(100)
 randdata = rawdata
 randdata[,] = matrix(rnorm(length(randdata), mean=0, sd=1))

### 2014-10-31 reanalysis/Towfic2014/helperfunctions_towfic.r

@@ -1,9 +1,85 @@
 
 
-
-# Towfic et al.
+# Towfic et al.  For the correct GEO accession GSE61901
 # load data and probe annotation and some formatting
 loadtowfic = function(downloaddata=TRUE)
+{
+	geoaccession="GSE61901"
+	
+	if(downloaddata)
+	{
+		geoseries=getGEO(geoaccession)
+		eset = geoseries[[1]]
+	}else{
+		eset=getGEO(filename="not_in_github/GSE61901_series_matrix.txt.gz")
+	}
+	# will only use sample annotation from this eset. The data will be taken from the non-normalized matrix file.
+	
+	# getting the sample annotation, some renaming
+	duplicatesampleannotation = pData(eset)[, c("characteristics_ch1", "characteristics_ch1.3", "characteristics_ch1.2")]
+	names(duplicatesampleannotation) = c("batch", "covariate", "array_strip_address")
+	duplicatesampleannotation[,1]=gsub("batch: ", "", duplicatesampleannotation[,1] )
+	duplicatesampleannotation[,2]=gsub("batchcovar: ", "", duplicatesampleannotation[,2] )
+	duplicatesampleannotation[,3]=gsub("array_address: ", "", duplicatesampleannotation[,3] )
+	duplicatesampleannotation[,"array_hyb_address"]=gsub("_[12]", "", duplicatesampleannotation[,3] )
+	
+	# some renaming for conveniance
+	duplicatesampleannotation$covariate = make.names(duplicatesampleannotation$covariate)
+	duplicatesampleannotation$covariate[duplicatesampleannotation$covariate=="GA.DP"] = "DP"
+	duplicatesampleannotation$covariate[duplicatesampleannotation$covariate=="GA.Q"] = "N"
+	duplicatesampleannotation$covariate[duplicatesampleannotation$covariate=="Medium"] = "M"
+	duplicatesampleannotation$covariate[duplicatesampleannotation$covariate=="GA.RS"] = "RS"
+	rownames(duplicatesampleannotation) = duplicatesampleannotation[, "array_strip_address"]
+	
+	# Creating a sample annotation for the data set where replicates are combined.
+	sampleannotation=duplicatesampleannotation[(1:(nrow(duplicatesampleannotation)/2))*2,c(1,2,4)]
+	rownames(sampleannotation) = sampleannotation[, "array_hyb_address"]
+	
[…]

### 2014-11-03 alternativefigure3.rmd
overviewplot v3

---
title: "Alternative Overview Figures"
author: "Vegard"
date: "Thursday, October 30, 2014"
output: html_document
---

Two scenarios

- Two groups, three batches. False Positives due to batch and ComBat.
- Three groups, two batches. False negatives due to batch, false positve due to mean adjust.

### False Positives due to batch and ComBat.

lage data set, 1000gener, 10+10
ingen gruppe effect eller batch effect.

1. plotte, målineg + meanstrek + CI + FC + p

2.
Dele i batcher 9+2+9, legge til batch effect.
plotte, målineg + meanstrek + CI + FC + p med batch separator i plottet.

3. Combat adjust
plotte, målineg + meanstrek + CI + FC + p uten batch separator i plottet.

4 plot med 2way anova og least square. Funker vel neppe med 1+1 bathen?

```{r}
library(sva)
source("commonscripts/helperfunctions.r")
source("commonscripts/boxplot_function_v3.r")
library(limma)
library(lsmeans)

set.seed(100)
ngenes = 1000
index=1
sa = createsampleannotation(  list(c(8,0), c(2,2), c(0,8)), as.factors=TRUE)
matrix_true = matrix(rnorm(ngenes * nrow(sa), mean=5, sd=1), nrow=ngenes, ncol=nrow(sa))

```

```{r, dev="png"}
plot_one_gene(matrix_true[index,], group=sa$group, main=paste("True values"),
              estimatemethod="CI")
```

Add batch effect.
```{r, dev="svg"}
matrix_batcheffect = matrix_true
matrix_batcheffect[,sa$batch==1] = matrix_batcheffect[,sa$batch==1]+1
matrix_batcheffect[,sa$batch==3] = matrix_batcheffect[,sa$batch==3]-1
plot_one_gene(matrix_batcheffect[index,], group=sa$group, batch=sa$batch,
              main=paste("Batch affected values"), estimatemethod="CI")
```

Adjust for batch effect.
```{r, dev="png"}

mod0 = model.matrix(~1,data=sa)
mod = model.matrix(~as.factor(sa$group))

matrix_batchadjusted = ComBat(dat=matrix_batcheffect, batch=as.factor(sa$batch), 
                      mod=mod, numCovs=NULL, 
                      par.prior=TRUE, prior.plots=FALSE)
matrix_batchadjusted = removeBatchEffect(matrix_batcheffect, batch=as.factor(sa$batch),
                                         design=mod)

plot_one_gene(matrix_batchadjusted[index,], group=sa$group, batch=sa$batch,
              main=paste("Batch effect adjusted values"), estimatemethod="CI")
```

```{r, dev="png"}
plot_one_gene(matrix_batcheffect[index,], group=sa$group, batch=sa$batch,
              main=paste("Batch affected values, batch included in ANOVA"),estimatemethod="lsmeans")

```

```{r}
#batch2FC = mean(matrix_true[index,sa$group==1 & sa$batch==2]) - 
#          mean(matrix_true[index,sa$group==2 & sa$batch==2])

figfilename = file.path( getwd(), "plots", paste("boxplots_v3_anovaadjusted_", index, sep=""))
#figfile = paste( figfilename, ".png", sep=""); png(file = figfile, width=1600, height=800)
figfile = paste( figfilename, ".pdf", sep=""); 
cairo_pdf(file =figfile, width=6, height=6)
op=par(mfrow=c(4, 1),   xaxt="n", mar=c(2,3,2,1)+0.1)
ylim=c(min(matrix_batcheffect[index,]), max(matrix_batcheffect[index,]))
plot_one_gene(matrix_true[index,], group=sa$group, main=paste("True values"),
              estimatemethod="CI", ylim=ylim)
plot_one_gene(matrix_batcheffect[index,], group=sa$group, batch=sa$batch,
              main=paste("Batch affected values"), estimatemethod="CI", ylim=ylim)
plot_one_gene(matrix_batchadjusted[index,], group=sa$group, batch=sa$batch,
              main=paste("Batch effect adjusted values"), estimatemethod="CI", ylim=ylim)
plot_one_gene(matrix_batcheffect[index,], group=sa$group, batch=sa$batch, ylim=ylim,
              main=paste("Batch affected values, batch included in ANOVA"),estimatemethod="lsmeans")

dev.off()

```

### False Positives due to batch and mean adjustment.

```{r , dev="png"}

#set.seed(100)
ngenes = 1000
index=3
sa = createsampleannotation(  list(c(8,3,0), c(0,7,8)), as.factors=TRUE)
matrix_true = matrix(rnorm(ngenes * nrow(sa), mean=5, sd=1), nrow=ngenes, ncol=nrow(sa))

# add a group effect for group 3
matrix_true[,sa$group==3] = matrix_true[,sa$group==3]+4
plot_one_gene(matrix_true[index,], group=sa$group, main=paste("True values"),
              estimatemethod="CI")

```

Add batch effect.
```{r , dev="png"}
matrix_batcheffect = matrix_true
matrix_batcheffect[,sa$batch==1] = matrix_batcheffect[,sa$batch==1]-2
matrix_batcheffect[,sa$batch==2] = matrix_batcheffect[,sa$batch==2]+2
plot_one_gene(matrix_batcheffect[index,], group=sa$group, batch=sa$batch,
              main=paste("Batch affected values"), estimatemethod="CI")
```

Batch adjust with mean center.

```{r , dev="png"}
matrix_batchadjusted = matrix_batcheffect
genemeans = rowMeans(matrix_batcheffect)
matrix_batchadjusted[,sa$batch ==1] =  genemeans +
  matrix_batchadjusted[,sa$batch ==1] - rowMeans( matrix_batchadjusted[,sa$batch ==1])
matrix_batchadjusted[,sa$batch ==2] =  genemeans + 
  matrix_batchadjusted[,sa$batch ==2] - rowMeans( matrix_batchadjusted[,sa$batch ==2])

#index=index+1
plot_one_gene(matrix_batchadjusted[index,], group=sa$group, batch=sa$batch,
              main=paste("Batch effect adjusted values"), estimatemethod="CI")
```

Include batch as factor in ANOVA.
```{r , dev="png"}
plot_one_gene(matrix_batcheffect[index,], group=sa$group, batch=sa$batch,
              main=paste("Batch affected values, batch included in ANOVA"),estimatemethod="lsmeans")

```

```{r}

figfilename = file.path( getwd(), "plots", paste("boxplots_v3_meanadjusted_", index, sep=""))
figfile = paste( figfilename, ".pdf", sep=""); 
pdf(file =figfile, width=6, height=6)
ylim=c(min(matrix_batcheffect[index,]), max(matrix_batcheffect[index,]))
op=par(mfrow=c(4, 1),   xaxt="n", mar=c(2,3,2,1)+0.1)
plot_one_gene(matrix_true[index,], group=sa$group, main=paste("True values"),
              estimatemethod="CI", ylim=ylim)
plot_one_gene(matrix_batcheffect[index,], group=sa$group, batch=sa$batch,
              main=paste("Batch affected values"), estimatemethod="CI", ylim=ylim)
plot_one_gene(matrix_batchadjusted[index,], group=sa$group, batch=sa$batch,
              main=paste("Batch adjusted values"), estimatemethod="CI", ylim=ylim)
plot_one_gene(matrix_batcheffect[index,], group=sa$group, batch=sa$batch, ylim=ylim,
              main=paste("Batch affected values, batch included in ANOVA"),estimatemethod="lsmeans")

dev.off()

```

```{r}
# debug
# rart?
#plot(matrix_true[,10]-matrix_true[,11], matrix_batchadjusted[,10]-matrix_batchadjusted[,11])

#plot(matrix_true[,10]-matrix_true[,11], rowMeans(matrix_batchadjusted[,1:10])-rowMeans(matrix_batchadjusted[,11:20]))

```

### 2014-11-03 commonscripts/boxplot_function_v3.r

@@ -1,16 +1,28 @@
 
 
+adhoc.cex=1
+adhoc.legendcex=adhoc.cex
 
-# y=matrix_true[1,]
+# y= matrix_true[index,]
 # group=sampleannotation$group
-# batch=sampleannotation$batch
-# #batch=NA
-# estimatemethod="lsmeans"
+# main=paste("True values")
+# estimatemethod="CI"
+# 
+# # y=matrix_true[1,]
+# # group=sampleannotation$group
+# # batch=sampleannotation$batch
+# batch=NULL
+# # estimatemethod="lsmeans"
 # ylim=NULL
 
-plot_one_gene = function(y, group, batch=NULL, ylim=NULL, main="", estimatemethod="none")
-{
-  lwd=2
+plot_one_gene = function(y, group, batch=NULL, ylim=NULL, main="", estimatemethod="none", lwd=1)
+{  
+  # Boxplots for CI etc.  
+  xboxplots=round(length(y)* 1.1)    
+  boxwidth=round(length(y)/20)
+  boxseparation = (boxwidth * 1.5)
+  thispal=c("red", "blue", "brown", "cyan")
+  
   a = order(group)
   if(!is.null(batch))
   {
@@ -21,7 +33,7 @@ plot_one_gene = function(y, group, batch=NULL, ylim=NULL, main="", estimatemetho
   
   y=y[a]
   
-  thispal=c("red", "blue", "green", "orange", "brown", "cyan")
+  
   
   if(is.null(ylim))
   {
@@ -29,11 +41,11 @@ plot_one_gene = function(y, group, batch=NULL, ylim=NULL, main="", estimatemetho
     ymin = round(min(y)-2, 1)
     ylim=c(ymin,ymax)
   }
-  xlim = c(1, round(length(y)*1.25))
+  xlim = c(1, round(xboxplots + (boxseparation) * length(unique(group)))  )
   
   # measurements
-  plot(y, ylim=ylim, xlim=xlim, col=thispal[as.factor(group)], main=main,  ylab="Expression", lwd=2, 
-       xlab="samples")
+  plot(y, ylim=ylim, xlim=xlim, col=thispal[as.factor(group)], 
+       main=main,  ylab="Expression", lwd=lwd, cex.main=adhoc.legendcex)
   
   # batch separator
   if(!is.null(batch))
[…]

### 2014-11-03 tex/manuscript.tex
Submitted version
mer fikling

@@
 % Manuscript version (date) in separate footnote
-\renewcommand{\thefootnote}{§}\footnotetext{\color{blue}Manuscript version \today.}
+%\renewcommand{\thefootnote}{§}\footnotetext{\color{blue}Manuscript version \today.}
 
@@
 Based on the actual use of ComBat by the authors and others, ourselves included, we suspect that adjusted data are commonly treated as ``batch effect free'' in subsequent analyses, and as a consequence, confidence in group effects has been overestimated and false results reported.
-The R package \texttt{limma} (\citealp{Smyth2003}) also offers batch adjustment with covariates, but notes that this is not intended used prior to linear modeling.
+The commercial software Partek (Partek® Genomics Suite®, version 6.6 Copyright ©; 2014 Partek Inc., St.
+Louis, MO, USA), and the R packages \texttt{limma} (\citealp{Smyth2003}) and \texttt{ber} (\citealp{http://link.springer.com/article/10.1007/s12561-013-9081-1}) also offer batch adjustment with covariates.
+Seeing that \texttt{ber} is sparsly used, and \texttt{Partek} and the relevant method ``removeBatchEffect'' in \texttt{limma} note that the this is not intended used prior to linear modeling, we choose to focus more on the popular and well documented ComBat.
 
@@
 
-In an experiment by \citet{Towfic2014}, cells were treated with either glatiramer acetate (a medicine for multiple sclerosis) or a generic drug, and mRNA was measured using microarrays alongside control sample measurements.
-A batch effect correlating to the chip (Illumina WG-6\_V2, six samples per chip) was observed and adjusted for with ComBat, whereafter the data were tested for differentially expressed genes, yielding a list of 1000 genes (Table S5, \citealp{Towfic2014}).
-Unfortunately, the batch--treatment design was unbalanced, with several batches having only one of the main treatments of interest.
-When we re-analyzed their data without the use of ComBat, but instead blocked for batch effect in \texttt{limma}, only 9 genes were found ($\text{FDR}<0.05$).
-The sanity check  outlined above with random numbers was also carried out.
-The distribution of p-values for different settings are shown in Figure~\ref{fig:comparison}a.
-Our conclusion is that most of the genes reported as differentially expressed in \citet{Towfic2014} are false positives.
-This example is a sort of ``worst case'' scenario for applying ComBat, since it both has a very unbalanced batch--group design and an a priori assumption of no difference.
+In an experiment described in \citet{Towfic2014}, the effect of glatiramer acetate (a medicine for multiple sclerosis) was compared to the effect of a generic drug. Cells were treated with
+glatiramer acetate (34 samples), the generic drug (11 samples) or 14 different types of other treatments(64 samples), and mRNA was measured using microarrays.
+A batch effect correlating to the chip (Illumina
+WG-6\_V2, six samples per chip, 17 chips in total) was observed and adjusted for with
+ComBat, whereafter the data were tested for differentially expressed
+genes, yielding hundreds of differentially expressed probes (Table S5, \citealp{Towfic2014}).
+Unfortunately, the batch--treatment design was unbalanced, with several
+batches having only one of the main treatments of interest. When we
+re-analyzed the data without the use of ComBat, but instead blocked for
+batch effect in \texttt{limma}, only 9 genes were found
+($\text{FDR}<0.05$). The sanity check  outlined above with random numbers
+was also carried out. The distribution of p-values for different settings
+are shown in Figure~\ref{fig:comparison}a. Our conclusion is that most of
+the genes found as differentially expressed in the analysis described in
+\citet{Towfic2014} are false positives. This example represents somewhat
+of a ``worst case'' scenario for applying ComBat, since it both has a
+very unbalanced batch--group design and an a priori assumption of no
+difference. Our re-analysis, although based on the "methods" section of
+Towfic et al, deviates in important ways from the analysis that was
+actually performed in the paper (personal communication, Ben Zeskind).
+More detailed information on this aspect can be found in the extended github repostiory
 
@@
 The supporting information of the original ComBat article (\citealp{Johnson2007}) demonstrates the method on cells inhibited for the expression of the TAL1 gene compared to controls on a microarray platform (denoted ``Data set 2'').
-The experiment was conducted on three different time points (used as batches) with a total of 30 samples and a fairly balanced batch--treatment set up containing 6+2, 3+4 and 9+6 samples.
+The experiment consist of 30 samples in 3 batches (batch1:6/2, batch2:3/4 and batch3:9/6 treatment/control samples).
 ComBat was applied followed by a T-test, in order to identify differentially expressed genes.
@@
 The incorporation of ComBat into various analysis tools appears to make it more accessible, but its use less transparent.
-Such tools include GenePattern (\citealp{Reich2006}), which offers ComBat with covariates, as well as TCGA and  inSilicoDb (\citealp{Taminau2011}), presently without covariates as an option.
+Such tools include GenePattern (v3.9.0) (\citealp{Reich2006}), AltAnalyze (v2.0.8 http://nar.oxfordjournals.org/content/38/suppl_2/W755.full) and SCAN.UPC (v2.6.3 http://www.pnas.org/content/110/44/17778.long), which offers ComBat with covariates, as well as TCGA and  inSilicoMerging (v1.8.7) (\citealp{ http://www.biomedcentral.com/1471-2105/13/335}), presently without covariates as an option.
+Especially troublesome, are the ComBat implementations in ChAMP (v1.2.8) (\citealp{http://bioinformatics.oxfordjournals.org/content/30/3/428.full}) and intCor (v1.03) (http://bioinformatics.oxfordjournals.org/content/early/2014/07/24/bioinformatics.btu423.full), where the use of covariates seems automatic and hidden from the user in the method call, while there is no mention of covariates in the documentation or articles.
 
@@
 \href{https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-figures.git}{https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-figures.git}.
-Additional analyses, performed, but not included in the article, may be found at
+Additional analyses, performed, but not included in the article, may be found at the extended repository 
 \href{https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-reports.git}{https://github.com/ous-uio-bioinfo-core/batch-adjust-warning-reports.git}.

### 2014-11-04 alternativefigure3.rmd

@@
 Add batch effect.
-```{r, dev="svg"}
+```{r, dev="png"}
 matrix_batcheffect = matrix_true
@@
 
-figfilename = file.path( getwd(), "plots", paste("boxplots_v3_anovaadjusted_", index, sep=""))
+figfilename = file.path( getwd(), "plots", paste("boxplots_v3_anovaadjusted", sep=""))
 #figfile = paste( figfilename, ".png", sep=""); png(file = figfile, width=1600, height=800)
 figfile = paste( figfilename, ".pdf", sep=""); 
-cairo_pdf(file =figfile, width=6, height=6)
+pdf(file =figfile, width=6, height=6)
 op=par(mfrow=c(4, 1),   xaxt="n", mar=c(2,3,2,1)+0.1)
@@
 
-figfilename = file.path( getwd(), "plots", paste("boxplots_v3_meanadjusted_", index, sep=""))
+figfilename = file.path( getwd(), "plots", paste("boxplots_v3_meanadjusted" , sep=""))
 figfile = paste( figfilename, ".pdf", sep=""); 

### 2014-11-04 alternativefigure3.rmd
fiks

@@
 ylim=c(min(matrix_batcheffect[index,]), max(matrix_batcheffect[index,]))
-plot_one_gene(matrix_true[index,], group=sa$group, main=paste("True values"),
+plot_one_gene(matrix_true[index,], group=sa$group, main=paste("(a) True values"),
               estimatemethod="CI", ylim=ylim)
 plot_one_gene(matrix_batcheffect[index,], group=sa$group, batch=sa$batch,
-              main=paste("Batch affected values"), estimatemethod="CI", ylim=ylim)
+              main=paste("(b) Batch affected values"), estimatemethod="CI", ylim=ylim)
 plot_one_gene(matrix_batchadjusted[index,], group=sa$group, batch=sa$batch,
-              main=paste("Batch effect adjusted values"), estimatemethod="CI", ylim=ylim)
+              main=paste("(c) Batch effect adjusted values"), estimatemethod="CI", ylim=ylim)
 plot_one_gene(matrix_batcheffect[index,], group=sa$group, batch=sa$batch, ylim=ylim,
-              main=paste("Batch affected values, batch included in ANOVA"),estimatemethod="lsmeans")
+              main=paste("(d) Batch affected values, batch included in ANOVA"),estimatemethod="lsmeans")
 
@@
 op=par(mfrow=c(4, 1),   xaxt="n", mar=c(2,3,2,1)+0.1)
-plot_one_gene(matrix_true[index,], group=sa$group, main=paste("True values"),
+plot_one_gene(matrix_true[index,], group=sa$group, main=paste("(a) True values"),
               estimatemethod="CI", ylim=ylim)
 plot_one_gene(matrix_batcheffect[index,], group=sa$group, batch=sa$batch,
-              main=paste("Batch affected values"), estimatemethod="CI", ylim=ylim)
+              main=paste("(b) Batch affected values"), estimatemethod="CI", ylim=ylim)
 plot_one_gene(matrix_batchadjusted[index,], group=sa$group, batch=sa$batch,
-              main=paste("Batch adjusted values"), estimatemethod="CI", ylim=ylim)
+              main=paste("(c) Batch adjusted values"), estimatemethod="CI", ylim=ylim)
 plot_one_gene(matrix_batcheffect[index,], group=sa$group, batch=sa$batch, ylim=ylim,
-              main=paste("Batch affected values, batch included in ANOVA"),estimatemethod="lsmeans")
+              main=paste("(d) Batch affected values, batch included in ANOVA"),estimatemethod="lsmeans")
 

### 2014-11-04 commonscripts/boxplot_function_v3.r

@@ -1,7 +1,6 @@
 
 
-adhoc.cex=1
-adhoc.legendcex=adhoc.cex
+
 
 # y= matrix_true[index,]
 # group=sampleannotation$group
@@ -17,11 +16,17 @@ adhoc.legendcex=adhoc.cex
 
 plot_one_gene = function(y, group, batch=NULL, ylim=NULL, main="", estimatemethod="none", lwd=1)
 {  
+	adhoc.cex=1
+	adhoc.legendcex=adhoc.cex
+	adhoc.palette =  c("black","black","black","black")
+	#adhoc.palette =c("red", "blue", "brown", "cyan")
+	adhoc.pch = c(1, 2, 3, 4)
+	
   # Boxplots for CI etc.  
   xboxplots=round(length(y)* 1.1)    
   boxwidth=round(length(y)/20)
   boxseparation = (boxwidth * 1.5)
-  thispal=c("red", "blue", "brown", "cyan")
+ 
   
   a = order(group)
   if(!is.null(batch))
@@ -44,7 +49,7 @@ plot_one_gene = function(y, group, batch=NULL, ylim=NULL, main="", estimatemetho
   xlim = c(1, round(xboxplots + (boxseparation) * length(unique(group)))  )
   
   # measurements
-  plot(y, ylim=ylim, xlim=xlim, col=thispal[as.factor(group)], 
+  plot(y, ylim=ylim, xlim=xlim, col=adhoc.palette[as.factor(group)],  pch=adhoc.pch[as.factor(group)]
        main=main,  ylab="Expression", lwd=lwd, cex.main=adhoc.legendcex)
   
   # batch separator
@@ -72,7 +77,7 @@ plot_one_gene = function(y, group, batch=NULL, ylim=NULL, main="", estimatemetho
   }  
   segments(  x0=(1:length(y))-0.5, y0=groupmeans[group], 
              x1=(1:length(y))+0.5, y1=groupmeans[group], 
-             col=thispal[as.factor(group)], lwd=lwd)
+             col=thispal[as.factor(group)], lwd=adhoc.pch[group])
   #legend("topright", legend=paste(groupnames, groupmeans), text.col=thispal[as.factor(groupnames)])
   
   
@@ -100,8 +105,8 @@ plot_one_gene = function(y, group, batch=NULL, ylim=NULL, main="", estimatemetho
     xleft = xoffset    
     #print(xleft)
[…]

### 2014-11-04 overviewplot_v2.r

@@ -44,7 +44,7 @@ index=1
 #summary(means_lm)
 #contrast(means_lm,"pairwise")
 
-figfilename = file.path( getwd(), "plots", paste("boxplots_v2" sep=""))
+figfilename = file.path( getwd(), "plots", paste("boxplots_v2", sep=""))
 #figfile = paste( figfilename, ".png", sep=""); png(file = figfile, width=1600, height=800)
 figfile = paste( figfilename, ".pdf", sep=""); 
 pdf(file =figfile, width=24, height=24)

### 2014-11-04 reanalysis/Towfic2014/figurescript_towfic.r

@@ -3,11 +3,11 @@
 
 starttime = Sys.time()
 debug = FALSE
-downloaddata=TRUE
+downloaddata=FALSE
 set.seed(100)
 
 
-includelibs = c("Biobase", "sva", "limma")
+includelibs = c("Biobase", "sva", "limma", "GEOquery")
 lapply(includelibs, require, character.only=T)
 source("../../commonscripts/helperfunctions.r")
 source("helperfunctions_towfic.r")
@@ -23,9 +23,16 @@ if(debug)
 
 qnormdata = normalizeBetweenArrays(rawdata, method="quantile") 
 
+# Collapsing the dataset based on the replicate sub-arrays.
+qnormdata = avearrays(qnormdata,  sampleannotation$array_hyb_address)
+sampleannotation=sampleannotation[(1:(nrow(sampleannotation)/2))*2,c(1,2,4)]
+rownames(sampleannotation) = sampleannotation[, "array_hyb_address"]
+
+
+
 # combat adjust
 combatdata= as.matrix(ComBat(dat=qnormdata,
-                             batch=sampleannotation$chip,
+                             batch=sampleannotation$batch,
                              mod=model.matrix(~as.factor(sampleannotation$covariate)),
                              numCovs=NULL, par.prior=TRUE, prior.plots=FALSE))
 
@@ -35,13 +42,13 @@ design = model.matrix(~0 + group)
 fit = lmFit(combatdata, design)
 cont.matrix = makeContrasts ( contrasts="groupDP-groupN", levels=design)  
 fit2 = contrasts.fit(fit, cont.matrix)
-limma_p_alt = eBayes(fit2)$p.value[,1]
+limma_p_combat = eBayes(fit2)$p.value[,1]
 print(paste("ComBat adjusted real data, significant probes (fdr<0.05): ",
-            sum(p.adjust(limma_p_alt, method="fdr")<0.05)))
+            sum(p.adjust(limma_p_combat, method="fdr")<0.05)))
 
 #Limma blocked batch and significance test
 group = factor(sampleannotation$covariate)
-block = factor(sampleannotation$chip)
+block = factor(sampleannotation$batch)
 design = model.matrix(~0+group+block)
 fit = lmFit(qnormdata, design)
 cont.matrix = makeContrasts ( contrasts="groupDP-groupN", levels=design)  
@@ -53,10 +60,10 @@ print(paste("Limma adjusted real data, significant probes (fdr<0.05): ",
 
 # random, ComBat adjusted
 set.seed(100)
[…]

### 2014-11-04 reanalysis/Towfic2014/helperfunctions_towfic.r

@@ -1,7 +1,7 @@
 
 
 # Towfic et al.  For the correct GEO accession GSE61901
-# load data and probe annotation and some formatting
+# load data and sample annotation and some formatting
 loadtowfic = function(downloaddata=TRUE)
 {
 	geoaccession="GSE61901"
@@ -16,26 +16,22 @@ loadtowfic = function(downloaddata=TRUE)
 	# will only use sample annotation from this eset. The data will be taken from the non-normalized matrix file.
 	
 	# getting the sample annotation, some renaming
-	duplicatesampleannotation = pData(eset)[, c("characteristics_ch1", "characteristics_ch1.3", "characteristics_ch1.2")]
-	names(duplicatesampleannotation) = c("batch", "covariate", "array_strip_address")
-	duplicatesampleannotation[,1]=gsub("batch: ", "", duplicatesampleannotation[,1] )
-	duplicatesampleannotation[,2]=gsub("batchcovar: ", "", duplicatesampleannotation[,2] )
-	duplicatesampleannotation[,3]=gsub("array_address: ", "", duplicatesampleannotation[,3] )
-	duplicatesampleannotation[,"array_hyb_address"]=gsub("_[12]", "", duplicatesampleannotation[,3] )
+	sampleannotation = pData(eset)[, c("characteristics_ch1", "characteristics_ch1.3", "characteristics_ch1.2")]
+	names(sampleannotation) = c("batch", "covariate", "array_strip_address")
+	sampleannotation[,1]=gsub("batch: ", "", sampleannotation[,1] )
+	sampleannotation[,2]=gsub("batchcovar: ", "", sampleannotation[,2] )
+	sampleannotation[,3]=gsub("array_address: ", "", sampleannotation[,3] )
+	sampleannotation[,"array_hyb_address"]=gsub("_[12]", "", sampleannotation[,3] )
 	
 	# some renaming for conveniance
-	duplicatesampleannotation$covariate = make.names(duplicatesampleannotation$covariate)
-	duplicatesampleannotation$covariate[duplicatesampleannotation$covariate=="GA.DP"] = "DP"
-	duplicatesampleannotation$covariate[duplicatesampleannotation$covariate=="GA.Q"] = "N"
-	duplicatesampleannotation$covariate[duplicatesampleannotation$covariate=="Medium"] = "M"
-	duplicatesampleannotation$covariate[duplicatesampleannotation$covariate=="GA.RS"] = "RS"
[…]

### 2014-11-05 alternativefigure3.rmd
revisjon

--- 2014-11-04 alternativefigure3.rmd
@@
 pdf(file =figfile, width=6, height=6)
-op=par(mfrow=c(4, 1),   xaxt="n", mar=c(2,3,2,1)+0.1)
-ylim=c(min(matrix_batcheffect[index,]), max(matrix_batcheffect[index,]))
-plot_one_gene(matrix_true[index,], group=sa$group, main=paste("True values"),
+op=par(mfrow=c(4, 1),   xaxt="n", mar=c(1,3,2,1)+0.1)
+ylim=c(min(matrix_batcheffect[index,])-1, max(matrix_batcheffect[index,])+1)
+plot_one_gene(matrix_true[index,], group=sa$group, main=paste("(e) True values"),
               estimatemethod="CI", ylim=ylim)
 plot_one_gene(matrix_batcheffect[index,], group=sa$group, batch=sa$batch,
-              main=paste("Batch affected values"), estimatemethod="CI", ylim=ylim)
+              main=paste("(f) Batch affected values"), estimatemethod="CI", ylim=ylim)
 plot_one_gene(matrix_batchadjusted[index,], group=sa$group, batch=sa$batch,
-              main=paste("Batch effect adjusted values"), estimatemethod="CI", ylim=ylim)
+              main=paste("(g) Batch effect adjusted values"), estimatemethod="CI", ylim=ylim)
 plot_one_gene(matrix_batcheffect[index,], group=sa$group, batch=sa$batch, ylim=ylim,
-              main=paste("Batch affected values, batch included in ANOVA"),estimatemethod="lsmeans")
+              main=paste("(h) Batch affected values, batch included in ANOVA"),estimatemethod="lsmeans")
 
@@
 sa = createsampleannotation(  list(c(8,3,0), c(0,7,8)), as.factors=TRUE)
-matrix_true = matrix(rnorm(ngenes * nrow(sa), mean=5, sd=1), nrow=ngenes, ncol=nrow(sa))
+matrix_true = matrix(rnorm(ngenes * nrow(sa), mean=5, sd=0.5), nrow=ngenes, ncol=nrow(sa))
 
 # add a group effect for group 3
-matrix_true[,sa$group==3] = matrix_true[,sa$group==3]+4
+matrix_true[,sa$group==3] = matrix_true[,sa$group==3]+2
 plot_one_gene(matrix_true[index,], group=sa$group, main=paste("True values"),
@@
 matrix_batcheffect = matrix_true
-matrix_batcheffect[,sa$batch==1] = matrix_batcheffect[,sa$batch==1]-2
-matrix_batcheffect[,sa$batch==2] = matrix_batcheffect[,sa$batch==2]+2
+matrix_batcheffect[,sa$batch==1] = matrix_batcheffect[,sa$batch==1]-1
+matrix_batcheffect[,sa$batch==2] = matrix_batcheffect[,sa$batch==2]+1
 plot_one_gene(matrix_batcheffect[index,], group=sa$group, batch=sa$batch,
@@
 pdf(file =figfile, width=6, height=6)
-ylim=c(min(matrix_batcheffect[index,]), max(matrix_batcheffect[index,]))
-op=par(mfrow=c(4, 1),   xaxt="n", mar=c(2,3,2,1)+0.1)
-plot_one_gene(matrix_true[index,], group=sa$group, main=paste("True values"),
+ylim=c(min(matrix_batcheffect[index,])-1, max(matrix_batcheffect[index,])+1)
+op=par(mfrow=c(4, 1),   xaxt="n", mar=c(1,3,2,1)+0.1)
+plot_one_gene(matrix_true[index,], group=sa$group, main=paste("(a) True values"),
               estimatemethod="CI", ylim=ylim)
 plot_one_gene(matrix_batcheffect[index,], group=sa$group, batch=sa$batch,
-              main=paste("Batch affected values"), estimatemethod="CI", ylim=ylim)
+              main=paste("(b) Batch affected values"), estimatemethod="CI", ylim=ylim)
 plot_one_gene(matrix_batchadjusted[index,], group=sa$group, batch=sa$batch,
-              main=paste("Batch adjusted values"), estimatemethod="CI", ylim=ylim)
+              main=paste("(c) Batch adjusted values"), estimatemethod="CI", ylim=ylim)
 plot_one_gene(matrix_batcheffect[index,], group=sa$group, batch=sa$batch, ylim=ylim,
-              main=paste("Batch affected values, batch included in ANOVA"),estimatemethod="lsmeans")
+              main=paste("(d) Batch affected values, batch included in ANOVA"),estimatemethod="lsmeans")
 

### 2014-11-05 tex/manuscript.tex
revisjon

@@
 ComBat, whereafter the data were tested for differentially expressed
-genes, yielding hundreds of differentially expressed probes (Table S5, \citealp{Towfic2014}).
+genes, yielding hundreds of differentially expressed probes (Table S5, \citealp{Towfic2014}) using a $\text{FDR}<0.05 threshold.
 Unfortunately, the batch--treatment design was unbalanced, with several
@@
 re-analyzed the data without the use of ComBat, but instead blocked for
-batch effect in \texttt{limma}, only 9 genes were found
-($\text{FDR}<0.05$). The sanity check  outlined above with random numbers
+batch effect in \texttt{limma}, only 9 genes were found ($\text{FDR}<0.05$)
+. The sanity check  outlined above with random numbers
 was also carried out. The distribution of p-values for different settings
@@
 \begin{array}{cc}
-\includegraphics[width=6.5cm]{Fig/leekrandomdatapvalues.pdf} &
-\includegraphics[width=6.5cm]{Fig/leekqqplot.png}
+\includegraphics[width=4cm]{Fig/leekrandomdatapvalues.pdf} &
+\includegraphics[width=4cm]{Fig/leekqqplot.png}\includegraphics[width=4cm]{Fig/samplesizescaling_ComBat.pdf}
 \end{array}$
@@
 b) QQ plot of the F-statistics
+c) P-value distributions for 3 equally unbalanced random number experiments with different sample sizes, 12,120 and 1200 samples from two study groups with a 1:5 and 5:1 distribution in two batches.
+A random batch effect is added for 10\% of the 20000 genes.
+This example is not from the \texttt{sva} package.
 }
@@
 
+\begin{figure}[!p]
+\begin{center}$
+\begin{array}{cc}
+\includegraphics[width=6.5cm]{Fig/boxplots_v3_meanadjusted.pdf} &
+\includegraphics[width=6.5cm]{Fig/boxplots_v3_anovaadjusted.pdf}
+\end{array}$
+\end{center}
+\caption
+{
+We simulated expression of one gene from two and three study groups.
+These two cases, designs and effects, were selected to illustrate the spurious effects that may arise from different batch adjustments.
+The Y-axis represents the log expression values.
+while the X-axis is used to separate the samples.
+Circles, triangles, and crosses indicate sample measurements from separate groups.
+Horizontal dashed lines indicates the mean of the expression values for that groups samples.
+Vertical dotted lines separates batches with the batch numbers on the lower half of the plot.
+To the far right of the plots, a typical differential expression test is performed with estimated fold change and a p-value.
+The mean expression for each group and 95\% confidence interval is estimated and depicted as boxes.
+a) Simulated expression of one gene in 26 samples from 3 groups (circles, triangles and crosses).
+The fold change and p-value presented are from the circles vs.
+triangles comparison and calculated with a one-way ANOVA.
+The confidence intervals indicate that the crosses are different from the other two groups.
+b) The values from (a) is divided into 2 batches with circle and triangles in batch 1 and triangles and crosses in batch 2.
+Batch effects were added, -1 for batch 1 and +1 for batch 2.
+The ANOVA test now results in a p-value less than 0.00 between the circles and triangles.
+This plot illustrates the problem of disregarding batch effects.
+c) The values in (b) were batch adjusted using the mean adjustment method.
+In order to make the means of the two batches equal, some of the group-effect for the crosses were transferred over to the triangels within the same batch.
+As a result the ANOVA test now indicate a difference between the circle-group and the triangle-group (p<0.00).
+d) The same values as in (b) are shown and used in a two-way ANOVA with group and batch included in the model.
+The estimated fold change and corresponding p-value is in effect calculated from the samples in batch 1.
+e) Simulated expression of one gene in 20 samples from two groups (circles and triangles). The t-test indicates no significant group differences(p=0.82).
+f) The values from (e) are divided into 3 batches in an unbalanced way, with only batch 2 containing samples from both groups.
+Batch effects were added, +1 for batch 1 and -1 for batch 3.
+The t-test now results in p<0.00.
+This plot illustrates the problem with disregarding batch effects.
+g) The values in (f) were batch effect adjusted while retaining the group difference (using removeBatchEffects), thus creating a ``batch effet free'' data set.
+Notice that all the values in batch 1 (which also belong to the circle group) are adjusted to make the group mean coincide with the mean of the two circle-samples in batch 2.
+The triangle-samples in batch 3 is adjusted similarly to match the mean of the two triangle-sampels in batch 2.The t-test now results in a significant group difference (p=0.01) and a fold change equal to a fold change calculated from the batch 2 samples only.
+Thus, the fold change is in effect estimated from only 2+2 samples, but the t-test assumes 10+10 samples with the low p-value as a result.
+h) The same values as in (f) are shown and used in a two-way ANOVA with group and batch included in the model.
+The estimated fold change is again in effect only based on samples from batch 2 and is exactly the same as for (g), but the p-value is much higher and the confidence intervals are more appropriate.
+}
+\label{fig:boxplots_v3}
+\end{figure}
+
 \end{document}

### 2014-11-05 commonscripts/boxplot_function_v3.r

@@ -49,7 +49,7 @@ plot_one_gene = function(y, group, batch=NULL, ylim=NULL, main="", estimatemetho
   xlim = c(1, round(xboxplots + (boxseparation) * length(unique(group)))  )
   
   # measurements
-  plot(y, ylim=ylim, xlim=xlim, col=adhoc.palette[as.factor(group)],  pch=adhoc.pch[as.factor(group)]
+  plot(y, ylim=ylim, xlim=xlim, col=adhoc.palette[as.factor(group)],  pch=adhoc.pch[as.factor(group)],
        main=main,  ylab="Expression", lwd=lwd, cex.main=adhoc.legendcex)
   
   # batch separator
@@ -57,7 +57,7 @@ plot_one_gene = function(y, group, batch=NULL, ylim=NULL, main="", estimatemetho
   {
     batchnames=unique(batch)
     x = match(batchnames, batch)
-    abline(v = x[-1]-0.5)#
+    abline(v = x[-1]-0.5, lty=3)#
     batchlabelx=vector()
     for(i in 1:(length(x)-1))
     {
@@ -77,8 +77,8 @@ plot_one_gene = function(y, group, batch=NULL, ylim=NULL, main="", estimatemetho
   }  
   segments(  x0=(1:length(y))-0.5, y0=groupmeans[group], 
              x1=(1:length(y))+0.5, y1=groupmeans[group], 
-             col=thispal[as.factor(group)], lwd=adhoc.pch[group])
-  #legend("topright", legend=paste(groupnames, groupmeans), text.col=thispal[as.factor(groupnames)])
+             col=adhoc.palette[as.factor(group)], lwd=lwd, lty=2)
+  #legend("topright", legend=paste(groupnames, groupmeans), text.col=adhoc.palette[as.factor(groupnames)])
   
   
   
@@ -107,7 +107,7 @@ plot_one_gene = function(y, group, batch=NULL, ylim=NULL, main="", estimatemetho
     xright = xleft + boxwidth    
     rect(xleft, ybottom, xright, ytop, border=adhoc.palette[g], lwd=lwd, lty="solid")
     lines(c(xoffset,xright), c(m,m), col=adhoc.palette[g], lwd=lwd)
-    #points(x=xleft+boxwidth/2, y=ylim[1]+0.5, pch=pchselection[g], cex=legendcex, lwd=lwd)
[…]

### 2014-11-05 samplesizescalingplot.r

@@ -11,7 +11,7 @@ set.seed(rseed)
 
 
 pvalcounts = list(ComBat=list(), LIMMA=list())
-batchaffectedgenefraction = 0
+batchaffectedgenefraction = 0.1
 samplesizescaleing = c(1,10,100)
 ngenes=20000
 
@@ -62,14 +62,14 @@ for(i in 1:length(pvalcounts))
 {
   figfilename = file.path( getwd(), "plots", paste("samplesizescaling",
                                                    names(pvalcounts)[i], 
-                                                   batchaffectedgenefraction,sep="_"))
+                                                   sep="_"))
   figfile = paste( figfilename, ".pdf", sep=""); 
-  pdf(file =figfile, width=24, height=12)
+  pdf(file =figfile)
   
   
   
   plot(  (1:25)/100, 1:25, 
-  		 main="Effect of sample size scaling on P-values", 
+  		 main="(c) Effect of sample size scaling on P-values", 
   		 xlab="p-value", ylab="Frequency", type="n", 
   		 ylim=c(0, max(unlist( pvalcounts))))
   for(t in 1:length(pvalcounts[[i]]))

### 2014-11-12 tex/manuscript.tex
mer
figurtekstretting

@@
 We simulated expression of one gene from two and three study groups.
-These two cases, designs and effects, were selected to illustrate the spurious effects that may arise from different batch adjustments.
+These two(a-d and e-h) cases, designs and effects, were selected to illustrate the spurious effects that may arise from different batch adjustments.
 The Y-axis represents the log expression values.
@@
 The confidence intervals indicate that the crosses are different from the other two groups.
-b) The values from (a) is divided into 2 batches with circle and triangles in batch 1 and triangles and crosses in batch 2.
+b) The values from (a) is divided into 2 batches with batch 1 containing circles and triangles, and batch 2 containing triangles and crosses.
 Batch effects were added, -1 for batch 1 and +1 for batch 2.
-The ANOVA test now results in a p-value less than 0.00 between the circles and triangles.
+The ANOVA test now results in a p<.01 between the circles and triangles.
 This plot illustrates the problem of disregarding batch effects.
@@
 In order to make the means of the two batches equal, some of the group-effect for the crosses were transferred over to the triangels within the same batch.
-As a result the ANOVA test now indicate a difference between the circle-group and the triangle-group (p<0.00).
+As a result the ANOVA test now indicate a difference between the circle-group and the triangle-group (p<.01).
 d) The same values as in (b) are shown and used in a two-way ANOVA with group and batch included in the model.
 The estimated fold change and corresponding p-value is in effect calculated from the samples in batch 1.
-e) Simulated expression of one gene in 20 samples from two groups (circles and triangles). The t-test indicates no significant group differences(p=0.82).
+e) Simulated expression of one gene in 20 samples from two groups (circles and triangles). The t-test indicates no significant group differences(p=.82).
 f) The values from (e) are divided into 3 batches in an unbalanced way, with only batch 2 containing samples from both groups.
 Batch effects were added, +1 for batch 1 and -1 for batch 3.
-The t-test now results in p<0.00.
+The t-test now results in p<.01.
 This plot illustrates the problem with disregarding batch effects.
-g) The values in (f) were batch effect adjusted while retaining the group difference (using removeBatchEffects), thus creating a ``batch effet free'' data set.
-Notice that all the values in batch 1 (which also belong to the circle group) are adjusted to make the group mean coincide with the mean of the two circle-samples in batch 2.
-The triangle-samples in batch 3 is adjusted similarly to match the mean of the two triangle-sampels in batch 2.The t-test now results in a significant group difference (p=0.01) and a fold change equal to a fold change calculated from the batch 2 samples only.
+g) The values in (f) were batch effect adjusted while retaining the group difference (using removeBatchEffects), thus creating a ``batch effect free'' data set.
+Notice that all the values in the "circle only" batch 1 are adjusted to make the group mean coincide with the mean of the two circle-samples in batch 2.
+The triangle-samples in batch 3 are adjusted similarly to match the mean of the two triangle-sampels in batch 2.The t-test now results in a significant group difference (p=.01) and a fold change equal to a fold change calculated from the batch 2 samples only.
 Thus, the fold change is in effect estimated from only 2+2 samples, but the t-test assumes 10+10 samples with the low p-value as a result.

### 2014-11-14 tex/manuscript.tex
Merged revisions

@@
-%\documentclass{bio}
-\documentclass[oupdraft]{bio}
+\documentclass{bio}
+%\documentclass[oupdraft]{bio}
 \usepackage[colorlinks=true, urlcolor=citecolor, linkcolor=citecolor, citecolor=citecolor]{hyperref}
@@
 %    \CITE{missing/unspecified citation}
-%\usepackage{color}
-%\definecolor{NBcol}{rgb}{0.7,0,1}
-%\definecolor{CITEcol}{rgb}{1,0,0.7}
-%\definecolor{COMMENTcol}{rgb}{0.7,0.2,0}
-%\newcommand\COMMENT[2]{\textcolor{COMMENTcol}{#1}\NOTE{#2}}
-%\newcommand\NB[1]{\textcolor{NBcol}{\textit{#1}}}
-%\newcommand\NOTE[1]{\NB{[#1]}}
-%\newcommand\CITE[1]{\textcolor{CITEcol}{[#1]}}
+\usepackage{color}
+\definecolor{NBcol}{rgb}{0.7,0,1}
+\definecolor{CITEcol}{rgb}{1,0,0.7}
+\definecolor{COMMENTcol}{rgb}{0.7,0.2,0}
+\newcommand\COMMENT[2]{\textcolor{COMMENTcol}{#1}\NOTE{#2}}
+\newcommand\NB[1]{\textcolor{NBcol}{\textit{#1}}}
+\newcommand\NOTE[1]{\NB{[#1]}}
+\newcommand\CITE[1]{\textcolor{CITEcol}{[#1]}}
+\newcommand\citeurl[1]{\textcolor{CITEcol}{[\url{#1}]}}
 
@@
 % Manuscript version (date) in separate footnote
-%\renewcommand{\thefootnote}{§}\footnotetext{\color{blue}Manuscript version \today.}
+\renewcommand{\thefootnote}{§}\footnotetext{\color{blue}Manuscript version: \today.}
 
@@
 It also has the option of preserving the difference between study groups, estimated from a two-way ANOVA model, to avoid conflating batch effects and group differences during batch adjustments.
+This is a natural approach when study groups are not balanced between batches as batch adjustments may otherwise bias, usually deflate, group differences.
 Unfortunately, this frequently used and recommended approach may systematically induce incorrect group differences in downstream analyses when groups are distributed between the batches in an unbalanced manner.
-The scientific community seems to be largely unaware of this problem, which most likely has contributed to false discoveries being presented in the published literature.
+\COMMENT{The scientific community seems to be largely unaware of how this approach may lead to false discoveries.}{Har moderert påstanden noe, jfr.
+reviewer~2, pt.~6.} We would expect any method for removing batch effects while retaining group differences to run into the same problem.
 }
@@
 In addition, it uses an empirical Bayes approach to avoid over-correction for batch effects, which is critical for small batches.
-Increasingly, the inclusion of group difference as a covariate when removing batch effects has been recommended, both in the \texttt{sva} tutorial and user fora.
+Increasingly, the inclusion of group difference as a covariate when removing batch effects has previously been recommended, both in the \texttt{sva} tutorial and user fora.
 ComBat has thus helped popularise this approach.
 Based on the actual use of ComBat by the authors and others, ourselves included, we suspect that adjusted data are commonly treated as ``batch effect free'' in subsequent analyses, and as a consequence, confidence in group effects has been overestimated and false results reported.
-The commercial software Partek (Partek® Genomics Suite®, version 6.6 Copyright ©; 2014 Partek Inc., St.
-Louis, MO, USA), and the R packages \texttt{limma} (\citealp{Smyth2003}) and \texttt{ber} (\citealp{http://link.springer.com/article/10.1007/s12561-013-9081-1}) also offer batch adjustment with covariates.
-Seeing that \texttt{ber} is sparsly used, and \texttt{Partek} and the relevant method ``removeBatchEffect'' in \texttt{limma} note that the this is not intended used prior to linear modeling, we choose to focus more on the popular and well documented ComBat.
+The commercial software Partek \CITE{citation or reference needed?}, and the R packages \texttt{limma} (\citealp{Smyth2003}) and \texttt{ber} (\citealp{Giordan2013}) also offer batch adjustment with covariates.
+Seeing that \texttt{ber} is sparsly used, and \texttt{Partek} and the relevant method ``removeBatchEffect'' in \texttt{limma} note that the this is not intended used prior to linear modeling, we choose to use the more popular and well documented ComBat to demonstrate the problem.
 
@@
 on thousands of genes, and use false discovery rate to determine significant cases: an approach that is particularly sensitive to inflated false positive rates.
+
+In conclusion, when study groups are unevenly distributed across batches, batch effects can have confounding effects if not properly adjusted for.
+This is difficult to do through a two-step procedure as batch adjustment will not be able to produce a ``batch effect free'' data set, as has also been observed in \citet{Buhule2014}.
+Estimating batch and group effects simultaneously through two-way ANOVA tries to overcome this, and does succeed at removing systematic biases in the group differences induced by the batch effects, but at the cost of making the observations interdependent to an extent that may invalidate subsequent analyses.
+Proper statistical analysis of unbalanced designs require models that handle batch effects as part of the analysis.
+However, the best way to handle this will be to ensure a balanced study design from the start, both to avoid data analysis problems and the loss of statistical power the ensues when batch and group effects need to be disentangled.
 
@@
 
+\subsubsection{Zero-centering batches:}
+
 The most common method for removing batch effects is to zero-centre each batch:
@@
 those found together in the same batches, one may also induce false differences between other groups, as is demonstrated in Figure~\ref{fig:boxplots}c.
+
+\subsubsection{Batch adjustment using two-way ANOVA to estimate batch effects:}
 
@@
 
+\subsubsection{Group comparison from two-way ANOVA:}
+
 If we estimate the group difference within batch $i$, we get
@@
 \end{equation}
+This is the same as would be obtained from a two-way ANOVA analysis.
+
+\subsubsection{Group comparison from one-way ANOVA after batch adjustment:}
 
@@
 The batch effects are then removed, and the estimated group differences retained, leaving the estimated $\Delta\hat{\beta}$ unchanged by the batch adjustment.
+
 However, if this batch adjusted data set is analysed without considering batch effects, the variance of $\Delta\hat{\beta}$ will be computed under the assumption that it is derived from a comparison of $n_A=\sum_{i=1}^m n_{iA}$ versus $n_B=\sum_{i=1}^m n_{iB}$ samples, and thus satisfy
@@
 We have reanalysed two cases with varying degree of unbalance.
+The first case represents a ``worst case'' scenario since it both has a very unbalanced batch--group design, and was in part selected since it provided a methods description which allowed us to assess the use of ComBat, and perform a full re-analysis of the data.
+The second was taken from the original publication of the ComBat method (\citealp{Johnson2007}).
 
@@
 
-In an experiment described in \citet{Towfic2014}, the effect of glatiramer acetate (a medicine for multiple sclerosis) was compared to the effect of a generic drug. Cells were treated with
-glatiramer acetate (34 samples), the generic drug (11 samples) or 14 different types of other treatments(64 samples), and mRNA was measured using microarrays.
-A batch effect correlating to the chip (Illumina
-WG-6\_V2, six samples per chip, 17 chips in total) was observed and adjusted for with
-ComBat, whereafter the data were tested for differentially expressed
-genes, yielding hundreds of differentially expressed probes (Table S5, \citealp{Towfic2014}) using a $\text{FDR}<0.05 threshold.
-Unfortunately, the batch--treatment design was unbalanced, with several
-batches having only one of the main treatments of interest. When we
-re-analyzed the data without the use of ComBat, but instead blocked for
-batch effect in \texttt{limma}, only 9 genes were found ($\text{FDR}<0.05$)
-. The sanity check  outlined above with random numbers
-was also carried out. The distribution of p-values for different settings
-are shown in Figure~\ref{fig:comparison}a. Our conclusion is that most of
-the genes found as differentially expressed in the analysis described in
-\citet{Towfic2014} are false positives. This example represents somewhat
-of a ``worst case'' scenario for applying ComBat, since it both has a
-very unbalanced batch--group design and an a priori assumption of no
-difference. Our re-analysis, although based on the "methods" section of
-Towfic et al, deviates in important ways from the analysis that was
-actually performed in the paper (personal communication, Ben Zeskind).
-More detailed information on this aspect can be found in the extended github repostiory
+In an experiment described in \citet{Towfic2014}, the effect of glatiramer acetate (a medicine for multiple sclerosis) was compared to the effect of a generic drug.
+Cells were treated with glatiramer acetate (34 samples), the generic drug (11 samples), or one of 14 other treatments (64 samples), and mRNA was measured using microarrays.
+A batch effect correlating to the chip (Illumina WG-6\_V2, six samples per chip, 17 chips in total) was observed and adjusted for with ComBat using treatment as a covariate.
+Batch adjusted data were then tested for differentially expressed genes, yielding hundreds of differentially expressed probes (Table S5, \citealp{Towfic2014}) using a $\text{FDR}<0.05$ threshold.
+Unfortunately, the batch--treatment design was highly unbalanced, with several batches having only one of the main treatments of interest.
+We re-analyzed the cited data (GEO: GSE40566) without using ComBat, but instead blocked for batch effect in \texttt{limma}, detecting only 9 differentially expressed genes at $\text{FDR}<0.05$.
+The sanity check  outlined above with random numbers was also carried out.
+The distribution of p-values for different settings are shown in Figure~\ref{fig:comparison}a.
+
+\NB{Einars tekst:} It should be noted that the original analyses by \citet{Towfic2014} differed from our re-analysis in one critical manner which was clarified upon contacting the authors.
+Their analyses used a different preprocessing of the microarray data (GEO: GSE61901) in which two technical replicates of each biological sample was included, corresponding to the two different slots of the bead-microarray.
+They refer to \citet{Bakshi2013} for analyses where the two slots have been combined to provide one set of expression values per sample.
+
+\NB{Vegards tekst:} Our conclusion is that most of the genes found as differentially expressed in the analysis described in \citet{Towfic2014} are false positives.
+This example represents somewhat of a ``worst case'' scenario for applying ComBat, since it both has a very unbalanced batch--group design and an a priori assumption of no difference.
+Our re-analysis, although based on the "methods" section of Towfic et al, deviates in important ways from the analysis that was actually performed in the paper (personal communication, Ben Zeskind).
+More detailed information on this aspect can be found in the extended github repostiory.
 
@@
 The incorporation of ComBat into various analysis tools appears to make it more accessible, but its use less transparent.
-Such tools include GenePattern (v3.9.0) (\citealp{Reich2006}), AltAnalyze (v2.0.8 http://nar.oxfordjournals.org/content/38/suppl_2/W755.full) and SCAN.UPC (v2.6.3 http://www.pnas.org/content/110/44/17778.long), which offers ComBat with covariates, as well as TCGA and  inSilicoMerging (v1.8.7) (\citealp{ http://www.biomedcentral.com/1471-2105/13/335}), presently without covariates as an option.
-Especially troublesome, are the ComBat implementations in ChAMP (v1.2.8) (\citealp{http://bioinformatics.oxfordjournals.org/content/30/3/428.full}) and intCor (v1.03) (http://bioinformatics.oxfordjournals.org/content/early/2014/07/24/bioinformatics.btu423.full), where the use of covariates seems automatic and hidden from the user in the method call, while there is no mention of covariates in the documentation or articles.
+Such tools include GenePattern (\citealp{Reich2006}, version 3.9.0), AltAnalyze (\citealp{Emig2010}, version 2.0.8) and SCAN.UPC (\citealp{Piccolo2013}, version 2.6.3), which offers ComBat with covariates, as well as TCGA and  inSilicoMerging (\citealp{ Taminau2012}, version 1.8.7), presently without covariates as an option.
+Especially troublesome, are the ComBat implementations in ChAMP (\citealp{Morris2014}, version 1.2.8) and intCor (\citealp{Fernandez-Albert2014}, version 1.03), where the use of covariates seems automatic and hidden from the user in the method call, while there is no mention of covariates in the documentation or articles.
 
@@
 \bibliographystyle{biorefs}
-\input{manuscript.bbl} % Include bibliography bbl-file
-%\bibliography{D:/Projects/_Appl_/_TeX_/library} % To use with BibTeX to generate bbl-file
+%\input{manuscript.bbl} % Include bibliography bbl-file
+\bibliography{D:/Projects/_Appl_/_TeX_/library} % To use with BibTeX to generate bbl-file
 %\bibliography{refs}
@@
 \begin{center}$
-\begin{array}{cc}
-\includegraphics[width=4cm]{Fig/leekrandomdatapvalues.pdf} &
-\includegraphics[width=4cm]{Fig/leekqqplot.png}\includegraphics[width=4cm]{Fig/samplesizescaling_ComBat.pdf}
+\begin{array}{ccc}
+\includegraphics[width=4.5cm]{Fig/leekrandomdatapvalues.pdf} &
+\includegraphics[width=4.5cm]{Fig/leekqqplot.png}&
+\includegraphics[width=4.5cm]{Fig/samplesizescaling_ComBat.pdf}
 \end{array}$

### 2014-11-17 commonscripts/boxplot_function_v4.r
plot v4

@@ -1,0 +1,166 @@
+
+adhoc.palette =c("red", "blue", "brown", "cyan")
+adhoc.cex=1
+adhoc.legendcex=adhoc.cex
+
+#  y=matrix_true[index,]
+#  group=sa$group
+#  main=paste("(a) True values")
+#  estimatemethod="CI"
+#  ylim=ylim
+#  batch=sa$batch
+#  lwd=1
+#  bbh = batchboxheight
+#  bblo = batchboxlowmeanoffsets
+
+
+plot_one_gene = function(y, group, batch=NULL, ylim=NULL, main="", estimatemethod="none", lwd=1, bbh=NA, bblo=NA, usecolor=TRUE)
+{  
+	
+  
+	
+  if(!usecolor)
+    adhoc.palette =  c("black","black","black","black")
+	  
+	adhoc.pch = c(1, 2, 3, 4)
+	
+  # Boxplots for CI etc.  
+  xboxplots=round(length(y)* 1.1)    
+  boxwidth=round(length(y)/20)
+  boxseparation = (boxwidth * 1.5)
+ 
+  
+  a = order(group)
+  if(!is.null(batch))
+  {
+    a = order(batch, group)
+    batch=batch[a]
+  }
+  group=group[a]
+  
+  y=y[a]  
+  
+  if(is.null(ylim))
+  {
+    ymax = round(max(y)+2, 1)
+    ymin = round(min(y)-2, 1)
+    ylim=c(ymin,ymax)
+  }
+  #xlim = c(1, round(xboxplots + (boxseparation) * length(unique(group)))  )
+	xlim = c(-1, length(y))
+  
+  # measurements
+  plot(y, ylim=ylim, xlim=xlim, col=adhoc.palette[as.factor(group)],  pch=adhoc.pch[as.factor(group)], main=main,  ylab="Expression", lwd=lwd, cex.main=adhoc.legendcex, xaxt="n", yaxt="n")
+  
+  # batch separator
+  if(!is.null(batch))    
+  {
+    for(b in unique(batch))
+    {
+      batchnames=unique(batch)
+      #x = match(batchnames, batch)
+      #abline(v = x[-1]-0.5, lty=3)#
+    
+      xleft = match(b, batch)
+      xright = length(batch) - match(b, batch[length(batch):1]) + 1
+      batchmean = mean(y[batch==b])
+      ybottom = batchmean - bblo[as.numeric(b)]
+      ytop = ybottom + bbh    
[…]

### 2014-11-17 overviewplot_v4.r

@@ -1,0 +1,108 @@
+
+
+
+library(sva)
+source("commonscripts/helperfunctions.r")
+source("commonscripts/boxplot_function_v4.r")
+library(limma)
+library(lsmeans)
+
+
+
+
+
+
+#set.seed(100)
+ngenes = 1000
+index=1
+#sa = createsampleannotation(  list(c(10,5,0,0), c(0,5,5,0), c(0,0,5,10)), as.factors=TRUE)
+
+
+sa = createsampleannotation(  list(c(10,3,0,0), c(0,10,30,0), c(0,0,3,20)), as.factors=TRUE)
+
+
+#sa = createsampleannotation(  list(c(20,2,0), c(0,20,50)), as.factors=TRUE) # god 3 grupper 2 batcher
+
+
+#sa = createsampleannotation(  list(c(15,0), c(2,2), c(0,15)), as.factors=TRUE)
+matrix_true = matrix(rnorm(ngenes * nrow(sa), mean=5, sd=2), nrow=ngenes, ncol=nrow(sa))
+
+# add a group effect for group 3
+ matrix_true[,sa$group==1] = matrix_true[,sa$group==1]+4
+ matrix_true[,sa$group==2] = matrix_true[,sa$group==2]+4
+ matrix_true[,sa$group==3] = matrix_true[,sa$group==3]+0
+ matrix_true[,sa$group==4] = matrix_true[,sa$group==4]+0
+
+bbextra=1.2
+batchboxheight=(max(matrix_true[index,]) - min(matrix_true[index,]) ) * bbextra
+batchboxlow=min(matrix_true[index,]) - batchboxheight * (bbextra-1)/2
+batchboxtop=max(matrix_true[index,]) + batchboxheight * (bbextra-1)/2
+batchmeans = sapply( unique(sa$batch), FUN=function(x){  mean( matrix_true[index,sa$batch==x] ) } )
+batchboxlowmeanoffsets=batchmeans - batchboxlow
+
+
+matrix_batcheffect = matrix_true
+matrix_batcheffect[,sa$batch==1] = matrix_batcheffect[,sa$batch==1]+2
+matrix_batcheffect[,sa$batch==2] = matrix_batcheffect[,sa$batch==2]-2
+
+
+#Batch adjust with mean center.
+
+matrix_batchadjusted = matrix_batcheffect
+genemeans = rowMeans(matrix_batcheffect)
+matrix_batchadjusted[,sa$batch ==1] =  genemeans +
[…]

### 2014-11-17 reanalysis/Towfic2014/oldhelperfunctions_towfic.r

@@ -1,18 +1,17 @@
 
-# old GSE40566
+
 # Towfic et al.  For the correct GEO accession GSE61901
 # load data and sample annotation and some formatting
 loadtowfic = function(downloaddata=TRUE)
 {
-	geoaccession="GSE40566"
-  rawfn = "http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file&file=GSE40566%5Fnon%5Fnormalized%2Etxt%2Egz"
+	geoaccession="GSE61901"
 	
 	if(downloaddata)
 	{
 		geoseries=getGEO(geoaccession)
 		eset = geoseries[[1]]
 	}else{
-		eset=getGEO(filename=paste("not_in_github/",geoaccession,"_series_matrix.txt.gz", sep=""))
+		eset=getGEO(filename="not_in_github/GSE61901_series_matrix.txt.gz")
 	}
 	# will only use sample annotation from this eset. The data will be taken from the non-normalized matrix file.
 	
@@ -31,20 +30,19 @@ loadtowfic = function(downloaddata=TRUE)
 	sampleannotation$covariate[sampleannotation$covariate=="Medium"] = "M"
 	sampleannotation$covariate[sampleannotation$covariate=="GA.RS"] = "RS"
 	rownames(sampleannotation) = sampleannotation[, "array_strip_address"]
-	
-	sampleannotation = read.table("sampleannotation.txt", sep="\t",
-                                header=TRUE, stringsAsFactors=FALSE)
+		
 	
 	if(downloaddata)
 	{
-		temp = tempfile()    
-		download.file(url=rawfn, destfile=temp, mode = "wb")
+		temp = tempfile()
+		download.file(url="http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE61901&format=file&file=GSE61901%5Fnon%2Dnormalized%2Etxt%2Egz",
+									destfile=temp, mode = "wb")
 		orgrawtable = read.table(temp, sep="\t", header=TRUE, 
-														 stringsAsFactors=FALSE, skip=0, strip.white=TRUE, fill=TRUE)
+														 stringsAsFactors=FALSE, skip=5, strip.white=TRUE, fill=TRUE)
 		unlink(temp)
 	}else{
-		orgrawtable = read.table(paste("not_in_github/",geoaccession,"_non-normalized.txt", sep=""), sep="\t", header=TRUE, 
[…]

### 2014-11-17 commonscripts/boxplot_function.r
opprydding

@@ -1,97 +1,144 @@
-# Functions for producing a boxplot
 
-adhoc.palette =  c("blue", "red", "darkgreen", "darkorange")
-adhoc.pch = c(1, 2, 3)
-adhoc.cex=3
+adhoc.palette =c("red", "blue", "brown", "cyan")
+adhoc.cex=1
 adhoc.legendcex=adhoc.cex
-adhoc.labcex=adhoc.cex
 
-# Boxplot component: values and intervals
-adhocboxplot = function(y, grouplabels, batchlabels, addlegend=FALSE, ylim=NULL, xlab="x",figureletter=NA,cex=adhoc.cex,labelcex=adhoc.labcex, legendcex=adhoc.legendcex)
-{
-  colpalette = adhoc.palette
-  pchselection = adhoc.pch
-  x = as.numeric(batchlabels)
-  lwd=2
+plot_one_gene = function(y, group, batch=NULL, ylim=NULL, main="", estimatemethod="none", lwd=1, bbh=NA, bblo=NA, usecolor=TRUE)
+{  
+	
+  if(!usecolor)
+    adhoc.palette =  c("black","black","black","black")
+	  
+	adhoc.pch = c(1, 2, 3, 4)
+	
+  # Boxplots for CI etc.  
+  xboxplots=round(length(y)* 1.1)    
+  boxwidth=round(length(y)/20)
+  boxseparation = (boxwidth * 1.5)
+ 
   
-  # add extra x for better visual separation
-  for(b in unique(batchlabels)) {
-    separation = 0.15
-    xoffset=0
-    for(g in unique(grouplabels[batchlabels==b])) {
-      x[batchlabels==b & grouplabels==g] = x[batchlabels==b & grouplabels==g] + xoffset
-      xoffset = xoffset+separation
+  a = order(group)
+  if(!is.null(batch))
+  {
+    a = order(batch, group)
+    batch=batch[a]
+  }
+  group=group[a]
+  
+  y=y[a]  
+  
+  if(is.null(ylim))
+  {
+    ymax = round(max(y)+2, 1)
+    ymin = round(min(y)-2, 1)
+    ylim=c(ymin,ymax)
+  }
+  #xlim = c(1, round(xboxplots + (boxseparation) * length(unique(group)))  )
+	xlim = c(-1, length(y))
+  
+  # measurements
[…]

### 2014-11-17 commonscripts/boxplot_function_v4.r

@@ -1,166 +1,0 @@
-
-adhoc.palette =c("red", "blue", "brown", "cyan")
-adhoc.cex=1
-adhoc.legendcex=adhoc.cex
-
-#  y=matrix_true[index,]
-#  group=sa$group
-#  main=paste("(a) True values")
-#  estimatemethod="CI"
-#  ylim=ylim
-#  batch=sa$batch
-#  lwd=1
-#  bbh = batchboxheight
-#  bblo = batchboxlowmeanoffsets
-
-
-plot_one_gene = function(y, group, batch=NULL, ylim=NULL, main="", estimatemethod="none", lwd=1, bbh=NA, bblo=NA, usecolor=TRUE)
-{  
-	
-  
-	
-  if(!usecolor)
-    adhoc.palette =  c("black","black","black","black")
-	  
-	adhoc.pch = c(1, 2, 3, 4)
-	
-  # Boxplots for CI etc.  
-  xboxplots=round(length(y)* 1.1)    
-  boxwidth=round(length(y)/20)
-  boxseparation = (boxwidth * 1.5)
- 
-  
-  a = order(group)
-  if(!is.null(batch))
-  {
-    a = order(batch, group)
-    batch=batch[a]
-  }
-  group=group[a]
-  
-  y=y[a]  
-  
-  if(is.null(ylim))
-  {
-    ymax = round(max(y)+2, 1)
-    ymin = round(min(y)-2, 1)
-    ylim=c(ymin,ymax)
-  }
-  #xlim = c(1, round(xboxplots + (boxseparation) * length(unique(group)))  )
-	xlim = c(-1, length(y))
-  
-  # measurements
-  plot(y, ylim=ylim, xlim=xlim, col=adhoc.palette[as.factor(group)],  pch=adhoc.pch[as.factor(group)], main=main,  ylab="Expression", lwd=lwd, cex.main=adhoc.legendcex, xaxt="n", yaxt="n")
-  
-  # batch separator
-  if(!is.null(batch))    
-  {
-    for(b in unique(batch))
-    {
-      batchnames=unique(batch)
-      #x = match(batchnames, batch)
-      #abline(v = x[-1]-0.5, lty=3)#
-    
-      xleft = match(b, batch)
-      xright = length(batch) - match(b, batch[length(batch):1]) + 1
-      batchmean = mean(y[batch==b])
-      ybottom = batchmean - bblo[as.numeric(b)]
-      ytop = ybottom + bbh    
[…]

### 2014-11-17 overviewplot.r

@@ -1,61 +1,108 @@
 
-### Figure of box-plots and confidence intervals
 
+
+library(sva)
+source("commonscripts/helperfunctions.r")
+source("commonscripts/boxplot_function_v4.r")
 library(limma)
 library(lsmeans)
 
-source("commonscripts/boxplot_function.r")
-source("commonscripts/helperfunctions.r")
-
-# 3 batches.
-# 4 groups
-#sampleannotation = createsampleannotation(  list(c(100,0,0,20), c(0,100,0,20), c(0,0,100,20))) 
-sampleannotation = createsampleannotation(  list(c(50,0,0,20), c(0,50,0,20), c(0,0,50,20))) 
-
-table(sampleannotation[,2:3])
-group=factor(sampleannotation$group)
-batch=factor(sampleannotation$batch)
-
-ngenes=1000
-#matrix_random = createrandomdata(ngenes, sampleannotation, mean=0, sd=1)
-rseed=1111
-set.seed(rseed)
-matrix_random = matrix(rnorm(ngenes * nrow(sampleannotation), mean=0, sd=1), nrow=ngenes, ncol=nrow(sampleannotation))
-
-matrix_condition = matrix_random
-matrix_condition[, group %in% c(3,4)] = matrix_condition[, group %in% c(3,4)] +2
-matrix_conditionbatch=matrix_condition
-matrix_conditionbatch[, batch ==2] = matrix_condition[, batch ==2] -1
-matrix_conditionbatch[, batch ==3] = matrix_condition[, batch ==3] +1
-matrix_meancenter = matrix_conditionbatch
-matrix_meancenter[,batch ==1] =  matrix_meancenter[,batch ==1] - rowMeans( matrix_meancenter[,batch ==1])
-matrix_meancenter[,batch ==2] =  matrix_meancenter[,batch ==2] - rowMeans( matrix_meancenter[,batch ==2])
-matrix_meancenter[,batch ==3] =  matrix_meancenter[,batch ==3] - rowMeans( matrix_meancenter[,batch ==3])
-mod = model.matrix(~group)
-matrix_limma = removeBatchEffect(matrix_conditionbatch, batch=batch, design=mod)
 
 
+
+
+
+#set.seed(100)
+ngenes = 1000
 index=1
-fit_lm=lm(matrix_condition[index,] ~ group+batch)
-means_lm=lsmeans(fit_lm,~group)
-summary(means_lm)
[…]

### 2014-11-17 overviewplot_v4.r

@@ -1,108 +1,0 @@
-
-
-
-library(sva)
-source("commonscripts/helperfunctions.r")
-source("commonscripts/boxplot_function_v4.r")
-library(limma)
-library(lsmeans)
-
-
-
-
-
-
-#set.seed(100)
-ngenes = 1000
-index=1
-#sa = createsampleannotation(  list(c(10,5,0,0), c(0,5,5,0), c(0,0,5,10)), as.factors=TRUE)
-
-
-sa = createsampleannotation(  list(c(10,3,0,0), c(0,10,30,0), c(0,0,3,20)), as.factors=TRUE)
-
-
-#sa = createsampleannotation(  list(c(20,2,0), c(0,20,50)), as.factors=TRUE) # god 3 grupper 2 batcher
-
-
-#sa = createsampleannotation(  list(c(15,0), c(2,2), c(0,15)), as.factors=TRUE)
-matrix_true = matrix(rnorm(ngenes * nrow(sa), mean=5, sd=2), nrow=ngenes, ncol=nrow(sa))
-
-# add a group effect for group 3
- matrix_true[,sa$group==1] = matrix_true[,sa$group==1]+4
- matrix_true[,sa$group==2] = matrix_true[,sa$group==2]+4
- matrix_true[,sa$group==3] = matrix_true[,sa$group==3]+0
- matrix_true[,sa$group==4] = matrix_true[,sa$group==4]+0
-
-bbextra=1.2
-batchboxheight=(max(matrix_true[index,]) - min(matrix_true[index,]) ) * bbextra
-batchboxlow=min(matrix_true[index,]) - batchboxheight * (bbextra-1)/2
-batchboxtop=max(matrix_true[index,]) + batchboxheight * (bbextra-1)/2
-batchmeans = sapply( unique(sa$batch), FUN=function(x){  mean( matrix_true[index,sa$batch==x] ) } )
-batchboxlowmeanoffsets=batchmeans - batchboxlow
-
-
-matrix_batcheffect = matrix_true
-matrix_batcheffect[,sa$batch==1] = matrix_batcheffect[,sa$batch==1]+2
-matrix_batcheffect[,sa$batch==2] = matrix_batcheffect[,sa$batch==2]-2
-
-
-#Batch adjust with mean center.
-
-matrix_batchadjusted = matrix_batcheffect
-genemeans = rowMeans(matrix_batcheffect)
-matrix_batchadjusted[,sa$batch ==1] =  genemeans +
[…]

### 2014-11-17 reanalysis/Towfic2014/figurescript_towfic.r

@@ -23,13 +23,6 @@ if(debug)
 
 qnormdata = normalizeBetweenArrays(rawdata, method="quantile") 
 
-# Collapsing the dataset based on the replicate sub-arrays.
-qnormdata = avearrays(qnormdata,  sampleannotation$array_hyb_address)
-sampleannotation=sampleannotation[(1:(nrow(sampleannotation)/2))*2,c(1,2,4)]
-rownames(sampleannotation) = sampleannotation[, "array_hyb_address"]
-
-
-
 # combat adjust
 combatdata= as.matrix(ComBat(dat=qnormdata,
                              batch=sampleannotation$batch,

### 2014-11-17 reanalysis/Towfic2014/helperfunctions_towfic.r

@@ -1,37 +1,12 @@
 
-# old GSE40566
-# Towfic et al.  For the correct GEO accession GSE61901
-# load data and sample annotation and some formatting
+# Downloads the non-normalized data from GSE40566
+# Using the covariate annotation from GSE61901
 loadtowfic = function(downloaddata=TRUE)
 {
 	geoaccession="GSE40566"
-  rawfn = "http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file&file=GSE40566%5Fnon%5Fnormalized%2Etxt%2Egz"
+  rawfn = "http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file&file=GSE40566%5Fnon%5Fnormalized%2Etxt%2Egz"	
 	
-	if(downloaddata)
-	{
-		geoseries=getGEO(geoaccession)
-		eset = geoseries[[1]]
-	}else{
-		eset=getGEO(filename=paste("not_in_github/",geoaccession,"_series_matrix.txt.gz", sep=""))
-	}
-	# will only use sample annotation from this eset. The data will be taken from the non-normalized matrix file.
-	
-	# getting the sample annotation, some renaming
-	sampleannotation = pData(eset)[, c("characteristics_ch1", "characteristics_ch1.3", "characteristics_ch1.2")]
-	names(sampleannotation) = c("batch", "covariate", "array_strip_address")
-	sampleannotation[,1]=gsub("batch: ", "", sampleannotation[,1] )
-	sampleannotation[,2]=gsub("batchcovar: ", "", sampleannotation[,2] )
-	sampleannotation[,3]=gsub("array_address: ", "", sampleannotation[,3] )
-	sampleannotation[,"array_hyb_address"]=gsub("_[12]", "", sampleannotation[,3] )
-	
-	# some renaming for conveniance
-	sampleannotation$covariate = make.names(sampleannotation$covariate)
-	sampleannotation$covariate[sampleannotation$covariate=="GA.DP"] = "DP"
-	sampleannotation$covariate[sampleannotation$covariate=="GA.Q"] = "N"
-	sampleannotation$covariate[sampleannotation$covariate=="Medium"] = "M"
-	sampleannotation$covariate[sampleannotation$covariate=="GA.RS"] = "RS"
[…]

### 2014-11-17 reanalysis/Towfic2014/oldfigurescript_towfic.r

@@ -1,90 +1,0 @@
-
-# script to create Figure 3a
-
-starttime = Sys.time()
-debug = FALSE
-downloaddata=TRUE
-set.seed(100)
-
-
-includelibs = c("Biobase", "sva", "limma", "GEOquery")
-lapply(includelibs, require, character.only=T)
-source("../../commonscripts/helperfunctions.r")
-source("helperfunctions_towfic.r")
-
-print("Re-analysis and figure generation for example taken from Towfic et al. 2014")
-
-ret = loadtowfic(downloaddata)
-sampleannotation = ret[["sampleannotation"]]
-rawdata = ret[["data"]]
-
-if(debug)
-  rawdata = rawdata[1:1000,]
-
-qnormdata = normalizeBetweenArrays(rawdata, method="quantile") 
-
-# Collapsing the dataset based on the replicate sub-arrays.
-qnormdata = avearrays(qnormdata,  sampleannotation$array_hyb_address)
-sampleannotation=sampleannotation[(1:(nrow(sampleannotation)/2))*2,c(1,2,4)]
-rownames(sampleannotation) = sampleannotation[, "array_hyb_address"]
-
-
-
-# combat adjust
-combatdata= as.matrix(ComBat(dat=qnormdata,
-                             batch=sampleannotation$batch,
-                             mod=model.matrix(~as.factor(sampleannotation$covariate)),
-                             numCovs=NULL, par.prior=TRUE, prior.plots=FALSE))
-
-# Significance test DP vs N
-group = factor(sampleannotation$covariate)
-design = model.matrix(~0 + group)
-fit = lmFit(combatdata, design)
-cont.matrix = makeContrasts ( contrasts="groupDP-groupN", levels=design)  
-fit2 = contrasts.fit(fit, cont.matrix)
-limma_p_combat = eBayes(fit2)$p.value[,1]
-print(paste("ComBat adjusted real data, significant probes (fdr<0.05): ",
-            sum(p.adjust(limma_p_combat, method="fdr")<0.05)))
-
-#Limma blocked batch and significance test
-group = factor(sampleannotation$covariate)
-block = factor(sampleannotation$batch)
-design = model.matrix(~0+group+block)
-fit = lmFit(qnormdata, design)
-cont.matrix = makeContrasts ( contrasts="groupDP-groupN", levels=design)  
-fit2 = contrasts.fit(fit, cont.matrix)
-limma_ret_woc = eBayes(fit2)
[…]

### 2014-11-17 reanalysis/Towfic2014/oldhelperfunctions_towfic.r

@@ -1,75 +1,0 @@
-
-
-# Towfic et al.  For the correct GEO accession GSE61901
-# load data and sample annotation and some formatting
-loadtowfic = function(downloaddata=TRUE)
-{
-	geoaccession="GSE61901"
-	
-	if(downloaddata)
-	{
-		geoseries=getGEO(geoaccession)
-		eset = geoseries[[1]]
-	}else{
-		eset=getGEO(filename="not_in_github/GSE61901_series_matrix.txt.gz")
-	}
-	# will only use sample annotation from this eset. The data will be taken from the non-normalized matrix file.
-	
-	# getting the sample annotation, some renaming
-	sampleannotation = pData(eset)[, c("characteristics_ch1", "characteristics_ch1.3", "characteristics_ch1.2")]
-	names(sampleannotation) = c("batch", "covariate", "array_strip_address")
-	sampleannotation[,1]=gsub("batch: ", "", sampleannotation[,1] )
-	sampleannotation[,2]=gsub("batchcovar: ", "", sampleannotation[,2] )
-	sampleannotation[,3]=gsub("array_address: ", "", sampleannotation[,3] )
-	sampleannotation[,"array_hyb_address"]=gsub("_[12]", "", sampleannotation[,3] )
-	
-	# some renaming for conveniance
-	sampleannotation$covariate = make.names(sampleannotation$covariate)
-	sampleannotation$covariate[sampleannotation$covariate=="GA.DP"] = "DP"
-	sampleannotation$covariate[sampleannotation$covariate=="GA.Q"] = "N"
-	sampleannotation$covariate[sampleannotation$covariate=="Medium"] = "M"
-	sampleannotation$covariate[sampleannotation$covariate=="GA.RS"] = "RS"
-	rownames(sampleannotation) = sampleannotation[, "array_strip_address"]
-		
-	
-	if(downloaddata)
-	{
-		temp = tempfile()
-		download.file(url="http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE61901&format=file&file=GSE61901%5Fnon%2Dnormalized%2Etxt%2Egz",
-									destfile=temp, mode = "wb")
-		orgrawtable = read.table(temp, sep="\t", header=TRUE, 
-														 stringsAsFactors=FALSE, skip=5, strip.white=TRUE, fill=TRUE)
-		unlink(temp)
-	}else{
[…]

### 2014-11-17 samplesizescalingplot.r

@@ -10,7 +10,7 @@ rseed=1111
 set.seed(rseed)
 
 
-pvalcounts = list(ComBat=list(), LIMMA=list())
+pvalcounts = list()
 batchaffectedgenefraction = 0.1
 samplesizescaleing = c(1,10,100)
 ngenes=20000
@@ -39,22 +39,21 @@ for(n in samplesizescaleing)
     }
 	}
 	
-		mod0 = model.matrix(~1,data=sampleannotation)
- 		batch = factor(sampleannotation$batch)
+	mod0 = model.matrix(~1,data=sampleannotation)
+ 	batch = factor(sampleannotation$batch)
 
-    mod = model.matrix(~as.factor(group), data=sampleannotation)
+  mod = model.matrix(~as.factor(group), data=sampleannotation)
 	matrix_adjusted = ComBat(dat=matrix_random, batch=batch, 
  												mod=mod, numCovs=NULL, 
  												par.prior=TRUE, prior.plots=FALSE)
 	pValues = f.pvalue(matrix_adjusted,mod,mod0)
 	pvalcounts[["ComBat"]][[runname]]=hist(pValues, plot=FALSE, breaks=100)$counts
 	
-  
-	matrix_adjusted = removeBatchEffect(matrix_random, batch=batch, design=mod)
-	pValues = f.pvalue(matrix_adjusted,mod,mod0)
-	pvalcounts[["LIMMA"]][[runname]]=hist(pValues, plot=FALSE, breaks=100)$counts
-	
-#qValuesComBat = p.adjust(pValuesComBat,method="BH")
+	# using removeBatchEffect
+	#matrix_adjusted = removeBatchEffect(matrix_random, batch=batch, design=mod)
+	#pValues = f.pvalue(matrix_adjusted,mod,mod0)
+	#pvalcounts[["LIMMA"]][[runname]]=hist(pValues, plot=FALSE, breaks=100)$counts
+
 }
 
 

### 2014-11-17 figurescript.r
mer rydding

@@ -18,6 +18,8 @@ setwd("reanalysis/Leek2012")
 source("figurescript_leek.r")
 setwd(startwd)
 
+source("samplesizescalingplot.r")
+
 setwd("reanalysis/Towfic2014")
 source("figurescript_towfic.r")
 setwd(startwd)

### 2014-11-17 overviewplot.r

@@ -3,36 +3,36 @@
 
 library(sva)
 source("commonscripts/helperfunctions.r")
-source("commonscripts/boxplot_function_v4.r")
+source("commonscripts/boxplot_function.r")
 library(limma)
 library(lsmeans)
 
 
 
 
+#setseed=100
+#setseed = setseed+1
 
-
-#set.seed(100)
+set.seed(139)
 ngenes = 1000
 index=1
 #sa = createsampleannotation(  list(c(10,5,0,0), c(0,5,5,0), c(0,0,5,10)), as.factors=TRUE)
 
 
-sa = createsampleannotation(  list(c(10,3,0,0), c(0,10,30,0), c(0,0,3,20)), as.factors=TRUE)
-
-
-#sa = createsampleannotation(  list(c(20,2,0), c(0,20,50)), as.factors=TRUE) # god 3 grupper 2 batcher
+#sa = createsampleannotation(  list(c(10,3,0,0), c(0,10,30,0), c(0,0,3,20)), as.factors=TRUE) #  4 groups,  3 batches
+sa = createsampleannotation(  list(c(25,5,0), c(0,20,50)), as.factors=TRUE) #  3 groups,  2 batches
 
 
 #sa = createsampleannotation(  list(c(15,0), c(2,2), c(0,15)), as.factors=TRUE)
 matrix_true = matrix(rnorm(ngenes * nrow(sa), mean=5, sd=2), nrow=ngenes, ncol=nrow(sa))
 
-# add a group effect for group 3
- matrix_true[,sa$group==1] = matrix_true[,sa$group==1]+4
- matrix_true[,sa$group==2] = matrix_true[,sa$group==2]+4
- matrix_true[,sa$group==3] = matrix_true[,sa$group==3]+0
- matrix_true[,sa$group==4] = matrix_true[,sa$group==4]+0
+# add a group effect
+matrix_true[,sa$group==1] = matrix_true[,sa$group==1]+4
+matrix_true[,sa$group==2] = matrix_true[,sa$group==2]+4
+matrix_true[,sa$group==3] = matrix_true[,sa$group==3]+0
+matrix_true[,sa$group==4] = matrix_true[,sa$group==4]+0
 
+# calculating batchbox limits.
 bbextra=1.2
 batchboxheight=(max(matrix_true[index,]) - min(matrix_true[index,]) ) * bbextra
 batchboxlow=min(matrix_true[index,]) - batchboxheight * (bbextra-1)/2
[…]

### 2014-11-17 commonscripts/boxplot_function.r
overviewplot

@@ -2,14 +2,14 @@
 adhoc.palette =c("red", "blue", "brown", "cyan")
 adhoc.cex=1
 adhoc.legendcex=adhoc.cex
+adhoc.pch = c(1, 2, 3, 4)
 
 plot_one_gene = function(y, group, batch=NULL, ylim=NULL, main="", estimatemethod="none", lwd=1, bbh=NA, bblo=NA, usecolor=TRUE)
 {  
 	
   if(!usecolor)
     adhoc.palette =  c("black","black","black","black")
-	  
-	adhoc.pch = c(1, 2, 3, 4)
+  
 	
   # Boxplots for CI etc.  
   xboxplots=round(length(y)* 1.1)    
@@ -138,7 +138,7 @@ estimatesboxesonly = function(y, group, batch, ylim=NULL, main="", lwd=1, usecol
     rect(xleft, ybottom, xright, ytop, border=adjustcolor(adhoc.palette[g], linealpha), lwd=lwd, 
          lty="solid", density=-1, col=adjustcolor(adhoc.palette[g],boxalpha))
     lines(c(xleft,xright), c(m,m), col=adjustcolor(adhoc.palette[g],linealpha), lwd=lwd)
-    #points(x=(xleft+xright)/2, y=ybottom-1, pch=adhoc.pch[g], cex=adhoc.legendcex, lwd=lwd)
+    points(x=(xleft+xright)/2, y=ytop+( (ylim[2]-ylim[1])/30), pch=adhoc.pch[g], cex=adhoc.legendcex, lwd=lwd, col=adhoc.palette[g])
 
   }
 }

### 2014-11-17 overviewplot.r

@@ -70,6 +70,7 @@ ymax = max(t(t(tmpmatrix) + (batchboxtop - matrix_true[index,])))
 ylim = c(ymin, ymax)
 
 #op=par(mfrow=c(3, 2),   xaxt="n", mar=c(1,3,2,1)+0.1)
+source("commonscripts/boxplot_function.r")
 layout(  matrix(c(1,2,0,3,4,5), 2, 3, byrow = TRUE) , widths=c(1,1,0.35) )
 op=par(oma = c(5,4,0,0) + 0.1, mar = c(0,0,2,1) + 0.1)
 plot_one_gene(matrix_true[index,], group=sa$group,  batch=sa$batch,, 

### 2014-11-17 tex/manuscript.tex

@@
 One method, ComBat, which is part of the R package sva, is particularly popular due to its ability to remove batch differences even when batches are small and heterogeneous.
-It also has the option of preserving the difference between study groups, estimated from a two-way ANOVA model, to avoid conflating batch effects and group differences during batch adjustments.
-This is a natural approach when study groups are not balanced between batches as batch adjustments may otherwise bias, usually deflate, group differences.
+It also has the option of preserving the difference between study groups, as batch adjustments may otherwise bias, usually deflate, group differences when study groups are not evenly balanced across batches.
+Using a two-way ANOVA model to simultaneously estimate both group an batch effects is a natural approach in such cases.
 Unfortunately, this frequently used and recommended approach may systematically induce incorrect group differences in downstream analyses when groups are distributed between the batches in an unbalanced manner.
 \COMMENT{The scientific community seems to be largely unaware of how this approach may lead to false discoveries.}{Har moderert påstanden noe, jfr.
-reviewer~2, pt.~6.} We would expect any method for removing batch effects while retaining group differences to run into the same problem.
+reviewer~2, pt.~6.}
 }
@@
 In a balanced group--batch design, estimates of group differences and batch effects are independent, and this approach becomes identical to the above described zero-centering per batch.
-However, if the group--batch design is heavily unbalanced, estimation of group differences and batch effects are interdependent, and when applying the estimated group differences across the entire data set, the uncertainties of these estimates are ignored.
-Subsequent analyses will then systematically underestimate the statistical uncertainties and exaggerate the confidence of group differences.
-Figure~\ref{fig:boxplots}d illustrates how statistical uncertainties are deflated by this batch adjustment method by comparing them to the uncertainties from the original ANOVA in Figure~\ref{fig:boxplots}e.
+However, if the group--batch design is heavily unbalanced, estimation of group differences and batch effects are interdependent, and when applying the estimated group differences across the entire data set, point estimates are used while the estimation errors are ignored.
+Subsequent analyses will then systematically underestimate the size of estimation errors and exaggerate the confidence in group differences.
+Figure~\ref{fig:boxplots}d illustrates how confidence interval are deflated by this batch adjustment method by comparing them to confidence intervals from the original ANOVA in Figure~\ref{fig:boxplots}e.
 
@@
 In addition, it uses an empirical Bayes approach to avoid over-correction for batch effects, which is critical for small batches.
-Increasingly, the inclusion of group difference as a covariate when removing batch effects has previously been recommended, both in the \texttt{sva} tutorial and user fora.
+Increasingly, the inclusion of group difference as a covariate when removing batch effects has been recommended, both in the \texttt{sva} tutorial and user fora.
 ComBat has thus helped popularise this approach.
-Based on the actual use of ComBat by the authors and others, ourselves included, we suspect that adjusted data are commonly treated as ``batch effect free'' in subsequent analyses, and as a consequence, confidence in group effects has been overestimated and false results reported.
+Based on the actual use of batch adjustment methods like ComBat by the authors and others, ourselves included, we suspect that adjusted data are commonly treated as ``batch effect free'' in subsequent analyses, and as a consequence, confidence in group effects has been overestimated and false results reported.
 The commercial software Partek \CITE{citation or reference needed?}, and the R packages \texttt{limma} (\citealp{Smyth2003}) and \texttt{ber} (\citealp{Giordan2013}) also offer batch adjustment with covariates.
 Seeing that \texttt{ber} is sparsly used, and \texttt{Partek} and the relevant method ``removeBatchEffect'' in \texttt{limma} note that the this is not intended used prior to linear modeling, we choose to use the more popular and well documented ComBat to demonstrate the problem.
+We note that the \texttt{sva} are now being updated to address these concerns.
 
@@
 However, batch adjustment using two-way ANOVA or ComBat on unbalanced data sets may be just as harmful for large samples and batch sizes as for small.
-For example, group comparisons using one-way ANOVA on the batch adjusted data will essentially result in $F$ statistics that are inflated by a fixed factor which depends on the unevenness of the design, rather than the size of the sample or batches.
+For example, group comparisons using one-way ANOVA on the batch adjusted data will essentially result in $F$-statistics that are inflated by a fixed factor which depends on the unevenness of the design, rather than the size of the sample or batches.
 The effect of this may be further exacerbated by running these analyses a large number of times, e.g.
@@
 
-In conclusion, when study groups are unevenly distributed across batches, batch effects can have confounding effects if not properly adjusted for.
+In conclusion, when study groups are unevenly distributed across batches, batch effects can have confounding effects if not properly handled.
 This is difficult to do through a two-step procedure as batch adjustment will not be able to produce a ``batch effect free'' data set, as has also been observed in \citet{Buhule2014}.
@@
 Proper statistical analysis of unbalanced designs require models that handle batch effects as part of the analysis.
-However, the best way to handle this will be to ensure a balanced study design from the start, both to avoid data analysis problems and the loss of statistical power the ensues when batch and group effects need to be disentangled.
+However, the best approach is to ensure a balanced study design from the start, to avoid data analysis problems as well as the loss of statistical power that ensues when batch and group effects need to be disentangled.
 
@@
 We will base our discussion on a simple model for data with batch effects:
-\begin{equation}
+\begin{equation}\label{eq:model}
 Y_{ijr}=\alpha+\beta_j+\gamma_i+\epsilon_{ijr}
@@
 
-When combining data from more diverse data sources, e.g. microarray data from different platforms, a more general model is required. One such model, used by \citet{Johnson2007}, is
+When combining data from more diverse data sources, e.g. microarray data from different platforms, a more general model is required. The model used by ComBat (\citealp{Johnson2007}) is
 \begin{equation}
@@
 In Figure~\ref{fig:boxplots}d, the effect of batch correction using group as covariate is shown with confidence intervals of the corrected data.
-For comparison, least square means estimates (R package \texttt{lsmeans}) are used in Figure~\ref{fig:boxplots}e to illustrate more appropriate confidence intervals that incorporate the uncertainties of the estimation errors.
+For comparison, least square means estimates (R package \texttt{lsmeans}) are used in Figure~\ref{fig:boxplots}e to illustrate more appropriate confidence intervals that incorporate the uncertainties of the batch effect estimates.
 
@@
 The documentation accompanying the \texttt{sva} library has an executable example demonstrating how to adjust a data set with ComBat, followed by an F-test.
-Replacing the real data with random numbers from a standard normal distribution (mean=0, sd=1), but otherwise following the instructions, will generate the p-value distribution shown in Figure~\ref{fig:sanity}a.
+Replacing the real data with random numbers from a standard normal distribution, $\text{N}(0,1)$, but otherwise following the instructions, will generate the p-value distribution shown in Figure~\ref{fig:sanity}a.
 
@@
 In effect, $\nu$ represents the effective sample size in the unbalanced group--batch design, while $\nu_0$ represents the nominal sample size when batches are ignored.
-The ratio $\nu_0/\nu\ge1$ indicates to what extent the uncertainty of the estimated group difference is deflated by the two-step procedure.
+The ratio $\nu/\nu_0\le1$ indicates to what extent the two-step procedure leads to underestimates of the random variability, and hence false or over-confidence in group differences.
+
+\subsection{Distribution of $F$-statistic in the general case}
+
+The detailed calculations for general linear models are presented in the appendix.
+Here, we summarise the results for two-way ANOVA based batch adjustments based on \eqref{eq:model} when there are $m$ batches, $M$ study groups, and $n_{ij}$ samples in batch $i$ and group $j$.
+
+A one-way ANOVA checking for group differences will assume that the $F$-statistic follows an $\text{F}_{q,n-q-1}$ distribution.
+However, if the data have been batch adjusted prior to this analysis, the approximate distribution of this $F$-statistic, with correct first two momenta, will be
+\begin{equation}
+F\approx\frac{\tilde q\tilde\sigma^2}{q\sigma^2}\cdot\left(1+\frac{r}{n-q-r-1}\right)\cdot\text{F}_{\tilde q,n-q-r-1}
+\end{equation}
+where $\tilde q$ and $\tilde\sigma^2$ are as computed in the appendix and satisfy $\tilde q\tilde\sigma^2\ge q\sigma^2$ and $\tilde q\le q$ with equalities if and only if the groups are perfectly balanced between the batches.
+
+The two factors both cause a bias in the $F$-statistic.
+The first factor, $\tilde q\tilde\sigma^2/q\sigma^2$, captures the unbalancedness of the design but is otherwise independent of the sample size.
+The second factor, $q-r/(n-q-r-1)$ captures loss of variance due to the centering of the data, but is primarily a problem when there are few samples in each batch.
+Finally, the degrees of freedom, $\tilde q$ and $n-q-r-1$, of the F-distribution are smaller than assumed by the $\text{F}_{q,n-q-1}$ distribution, and so the variance is larger.
+
+As the sample size increases, assuming the distribution between groups and batches remain unchanged, the effect of the second factor and the second degree of freedom will diminish.
+However, the factor $\tilde q\tilde\sigma^2/q\sigma^2$ will remain, as will the reduced degree of freedom $\tilde q$.
 
@@
 
-\NB{Einars tekst:} It should be noted that the original analyses by \citet{Towfic2014} differed from our re-analysis in one critical manner which was clarified upon contacting the authors.
+It should be noted that the original analyses by \citet{Towfic2014} differed from our re-analysis in one critical manner which was clarified upon contacting the authors.
 Their analyses used a different preprocessing of the microarray data (GEO: GSE61901) in which two technical replicates of each biological sample was included, corresponding to the two different slots of the bead-microarray.
-They refer to \citet{Bakshi2013} for analyses where the two slots have been combined to provide one set of expression values per sample.
-
-\NB{Vegards tekst:} Our conclusion is that most of the genes found as differentially expressed in the analysis described in \citet{Towfic2014} are false positives.
-This example represents somewhat of a ``worst case'' scenario for applying ComBat, since it both has a very unbalanced batch--group design and an a priori assumption of no difference.
-Our re-analysis, although based on the "methods" section of Towfic et al, deviates in important ways from the analysis that was actually performed in the paper (personal communication, Ben Zeskind).
-More detailed information on this aspect can be found in the extended github repostiory.
+For analyses where the two slots have been combined to provide one set of expression values per sample, they refer to \citet{Bakshi2013} in which Partek was used for batch adjustment.
 
@@
 When the group--batch distribution is unbalanced, i.e.
-where batches do not have the same composition of groups, this will lead to deflated uncertainty estimates and over-confidence in the results.
-The problem is essentially independent of sample size.
+where batches do not have the same composition of groups, this will lead to deflated estimates of the estimation errors, and over-confidence in the results.
+The problem is essentially independent of sample size as illustrated in Figure~\ref{fig:comparison}c.
 
@@
 \subsection{Increased emphasis on preserving group difference}
+
+Multiple batch adjustment tools exist, most of which have batch centering as the main approach, but also offering the inclusion of covariates.
+Some of these give qualified warnings against using batch adjusted data in subsequent analyses.
 
@@
 Upon realizing that the confidence in the estimated group differences was exaggerated, we searched the literature for a better understanding of correct use and the potential limitations of ComBat.
-However, the authors of ComBat and the \texttt{sva} package recommended including study group as a covariate as we had done.
+However, the authors of ComBat and the \texttt{sva} package recommended including study group as a covariate as we had done: this is now changing as the supporting documentation of \texttt{sva} is being updated to warn users of the potential dangers.
 In addition, other studies investigating batch effects were mostly recommending ComBat without much concern (\citealp{Kupfer2012}, \citealp{Kitchen2011}, \citealp{Chen2011}).
@@
 Some used ComBat with covariates, some without, but we did not find any that addressed the potential problems related to use of covariates.
+However, the inability of batch adjustment methods in dealing with unbalanced designs has been noted (\citealp{Buhule2014}).
 
@@
 
+These problems are not specific to ComBat, but are shared by other batch adjustment methods: indeed, the empirical Bayes approach used by ComBat  will dampen the batch adjustments and may thus reduce the problem slightly relative to a more direct two-way ANOVA approach.
+Although some of the tools warn against performing subsequent analyses on batch adjusted data, these warnings are often somewhat vague and seem to be primarily concerned about the degrees of freedom in the data set.
+Batch adjustment will reduce the degrees of freedom in the data which may influence the distribution of the test statistics in subsequent analyses, e.g.
+$F$-statistics from ANOVA.
+For small data sets or where there are many small batches this may be a problem, while it would be less of a concern when there are a fair number of samples in each batch.
+However, we find that the main problem is not related to the degrees of freedom, but estimation errors during the batch adjustments which get applied across the data, which can induce group differences in the adjusted data set even in the absence of batch effects.
+
 We are concerned, in light of our findings, that many published results from batch adjusted data using study group as covariate may be unreliable.
@@
 \subsection{Practical advice}
+
+The main advice, particularly when batch effects are significant, is to ensure a balanced design in which study groups are evenly distributed across batches.
 
@@
 A superficial assessment of the reliability of results based on batch corrected data with group differences retained can often be made.
+A major concern would be batches where either of the groups of interest are missing or strongly under-represented, which would contribute to the nominal sample size, but not to the effective sample size.
 In section~\ref{sec:twogroups}, the ratio $\nu_0/\nu\ge1$ indicates by how much the effective sample size is overestimated when comparing two groups.
-A similar computation, although more complicated, can be made for multiple groups.
+A similar computation, although more complicated, can be made for multiple groups \NB{refer to appendix for details}.
 However, the two-group assessment should still give a fair idea of the reliability of pairwise group comparisons.
-A major concern would be batches where either of the groups of interest are missing or strongly under-represented, which would contribute to the nominal sample size, but not to the effective sample size.
+If the $\nu_0/\nu$ ratio is close to 1, results should still be reliable, while a ratio much larger than 1 will indicate exaggerated confidence.
+If the behaviour of the test statistic is well understood, as the $F$-statistic in ANOVA, one may adjust the statistic according to the $\nu_0/\nu$ ratio and recompute the $p$-value.
+However, this will require deep insight into the statistical behaviour of the model.
+Down-sampling to the effective samples size, i.e.
+to a $\nu/\nu_0$ portion of the samples, may also be done to see if results persist.
+However, none of these solutions are ideal.
 
-Alternatively, one may treat the results more like an ordered list of candidates, with the most likely true positives on top, de-emphasizing the somewhat deflated p-values.
+We did consider if batch adjustment could be modified so as to balance the need for removing batch effects with avoidance of introducing false group differences which would be somewhere between batch centering and two-way ANOVA based adjustment.
+For two-group comparisons, the batch centering will remove some of the group effects and thus underestimate group differences, while two-way ANOVA based batch adjustment will add random, but unbiased, group effects.
+Somewhere in between is a balance for which these two balance out.
+However, the applicability of this approach is very limited and potentially risky since the degree of moderation of the batch adjustment will depend on the subsequent comparison to be made.
+
+Alternatively, one may treat the results more like an ordered list of candidates, with the most likely true positives on top, de-emphasizing the somewhat deflated $p$-values.
 This would, however, be hypothesis generating, rather than hypothesis testing.
@@
 In cases where data preparation is performed prior to, and separate from, the data analysis, as is now often the case, this is of particular importance, as artifacts may be introduced in the data preparation which could influence the reliability of downstream analyses.
+When assessing published findings, we frequently found it hard to determine how batch adjustments had been done, and occasionally suspected that study group had been used as covariate without this being stated.
+Science not only depends on determining sound methods and finding reliable results, but also on communicating sufficient detail to allow readers to assess the extent to which findings can be trusted.
+Failing to do so may lead cause methodological problems to pass undetected, but also to cast doubt on sound results.
+An increased focus on making research reproducible is therefore of great importance.
+
+%% APPENDIX TO BE INCLUDED?
+
+\appendix
+\section{Distribution of error terms after two-way ANOVA batch adjustment}
+\label{sec:chiApprox}
+
+We assume a general linear model with intercept or more general covariates ($\alpha$), batch effects ($\beta$), and study groups or study parameters of interest ($\gamma$):
+\begin{equation}
+Y=\alpha A+\beta B+\gamma C+\epsilon
+\end{equation}
+where $Y=[Y_1,\ldots,Y_n]$ are the observable random variables, $\alpha$, $\beta$, and $\gamma$ is a $p$, $q$, and $r$ vectors with design matrices $A$, $B$, and $C$, and $\epsilon=[\epsilon_1,\ldots,\epsilon_n]\sim\text{N}(0,\sigma^2I_n)$ where $I_n$ is the $n\times n$ identity matrix.
+We also assume the full design matrix, combining $A$, $B$, and $C$, has full rank $p+q+r$: i.e., the covariates are linearly independent.
+
+In general, we wish to assess the explanatory power of $B$ including $C$ as batch effects and $A$ as additional covariates.
+Often $p=1$ with $\alpha$ the intercept term, but this formulation allows general covariates to be included in the analyses.
+The covariates of $B$ which are subject to testing are more commonly specified as contrasts on the full design matrix combining $A$, $B$, and $C$, but in our case this splitting is more convenient.
+
+We can eliminate $A$ from the model, simultaneously removing $p$ degrees of freedom.
+One way of doing this is to find an $n\times n$ rotation matrix $U$ so that $AU=[0\mid A']$ with $A'$ a $p\times p$ invertible matrix.
+Applying this rotation to all elements, we split the $n$ dimensions into $(n-p)+p$:  $YU=[Y_0\mid Y']$, $BU=[B_0\mid B']$, $BU=[C_0\mid C']$, and $\epsilon V=[\epsilon_0\mid\epsilon']\sim\text{N}(0,\sigma^2 I_n)$.
+We may then eliminate the degrees of freedom used to estimate $\alpha$ from the model yielding
+\begin{equation}
+Y_0=\beta_0 B_0+\gamma_0 C_0+\epsilon_0
+\end{equation}
+where $Y_0$ and $\epsilon_0$ are now $n-p$ vectors.
+If $\alpha$ is the intercept term, this corresponds to centering all variables and covariates, but the rotation also eliminates the corresponding degree of freedom.
+
+If there are no batches, i.e. $r=0$, we get the common linear model in which we get a decomposition $Y=\hat\beta_0B_0+\hat\epsilon_0$ with $\hat\gamma_0=Y_0B_0^t(B_0B_0^t)^{-1}$, where
+$|\hat\epsilon|^2\sim\sigma^2\chi_{n-p-q}$ and $|(\hat\beta_0-\beta_0)C_0|^2\sim\sigma^2\chi_q$, yielding the common
+$F$-statistic
+\begin{equation}
+F=\frac{|(\hat\beta_0-\beta_0)C_0|^2/q}{|\hat\epsilon_0|^2/(n-p-q)}\sim\text{F}_{q,n-p-q}.
+\end{equation}
+If we believe our data are  ``batch effect free'', this is the model we would use, and the distribution of the $F$-statistic we would assume.
+
+We define the batch adjusted data $\tilde Y_0=Y_0-\hat\gamma_0C_0$ using the estimated batch effect $\hat\gamma_0$ from the full model.
+If we let $X_0=[B_0^t\mid C_0^t]^t$ be the full design matrix after elimination of covariates $A$, the parameter estimates are $[\hat\beta_0\mid\hat\gamma_0]=Y_0X_0^t(X_0X_0^t)^{-1}$.
+The linear decomposition $Y_0=\hat\beta_0B_0+\hat\gamma_0C_0+\hat\epsilon_0$ has $|\hat\epsilon_0|^2\sim\sigma^2\chi^2_{n-p-q-r}$ and
+$|(\hat\beta_0-\beta_0)B_0+(\hat\gamma_0-\gamma_0)C_0|^2\sim\sigma^2\chi^2_{q+r}$, and so the variance of the error term is the same in the batch adjusted case $\tilde Y_0=\hat\beta_0C_0+\hat\epsilon_0$.
+However, the sum of squares $|(\hat\beta_0-\beta_0)B_0|^2$ explained by $B$ is less easily expressed.
+
+Solving $[\hat\beta_0\mid\hat\gamma_0]=Y_0X_0^t(X_0X_0^t)^{-1}$ and using
+\begin{equation}
+(X_0X_0^t)^{-1}
+\begin{pmatrix}B_0\\0\end{pmatrix}
+=
+\begin{pmatrix}P\\-(C_0C_0^t)^{-1}C_0B_0^tP\end{pmatrix} B_0
+\end{equation}
+with $P=(B_0\rho_{C_0}B_0^t)^{-1}$ and $\rho_{C_0}=I-C_0(C_0C_0^t)^{-1}C_0$
+yields
+\begin{equation}
+(\hat\beta_0-\beta_0)B_0=\epsilon_0\rho_{C_0}B_0^tPB_0\sim\text{N}(0,\sigma^2L_0)
+\quad\text{with}\quad
+L_0=B_0^tPB_0\rho_{C_0}B_0^tPB_0
+\end{equation}
+since $\epsilon_0\sim\text{N}(0,\sigma^2I)$.
+Let $\lambda_1,\ldots,\lambda_q\ge1$ be the non-zero eigenvalues of $L_0$: this holds since $L_0$ is a symmetric rank $q$ matrix \NB{complete the proof of this}.
+Then,
+\begin{equation}
+|(\hat\beta_0-\beta_0)B_0|^2=\sigma^2\sum_{i=1}^q\lambda_iU_i
+\quad\text{where}\quad
+U_1,\ldots,U_q\sim\chi^2_1.
+\end{equation}
+Approximating this so that the first two momenta match gives
+\begin{equation}
+|(\hat\beta_0-\beta_0)B_0|^2\approx\tilde\sigma^2\chi^2_{\tilde q}
+\quad\text{where}\quad
+\frac{\tilde q\tilde\sigma^2}{\sigma^2}=\sum_{i=1}^q\lambda_i=\text{tr}(L_0),\,
+\frac{\tilde q\tilde\sigma^4}{\sigma^4}=\sum_{i=1}^q\lambda_i^2=\text{tr}(L_0^2),
+\end{equation}
+from which it follows that $\tilde\sigma^2\ge\sigma^2$, $\tilde q\tilde\sigma^2\ge q\sigma^2$, and $\tilde q\le q$, where equality holds if and only if all $\lambda_i=1$, which happens when $B_0$ and $C_0$ are linearly independent, i.e.
+$B_0C_0^t=0$.
+
+We would like to express $\tilde\sigma^2$ and $\tilde q$ terms of the original design matrices.
+The rotated design matrices were defined as $B_0=BU_0$, etc., using the rotation $U=[U_0\mid U']$ so that $AU_0=0$ while $AU'$ is non-singular.
+Hence, it follows that $V_0V_0^t=\rho_A=1-A^t(AA^t)^{-1}A$ which is the projection onto the $n-p$-hyperplane perpendicular to the $p$-hyperplane spanned by $A$.
+This allows us to express
+\begin{equation}
+L_0=U_0^tLU_0
+\quad\text{where}\quad
+L=
+\end{equation}
 

### 2014-11-17 tex/manuscript.tex
Minor updates

@@
 \end{equation}
-where $\tilde q$ and $\tilde\sigma^2$ are as computed in the appendix and satisfy $\tilde q\tilde\sigma^2\ge q\sigma^2$ and $\tilde q\le q$ with equalities if and only if the groups are perfectly balanced between the batches.
+where $\tilde q$ and $\tilde\sigma^2$ are as computed in the appendix and satisfy $\tilde q\tilde\sigma^2\ge q\sigma^2$ and $\tilde q\le q$ with equalities if and only if the groups are evenly represented in all batches.
 
@@
 \end{equation}
+\NB{Need to complete this part...}
 

### 2014-11-18 commonscripts/boxplot_function.r
figure1

@@ -37,7 +37,7 @@ plot_one_gene = function(y, group, batch=NULL, ylim=NULL, main="", estimatemetho
 	xlim = c(-1, length(y))
   
   # measurements
-  plot(y, ylim=ylim, xlim=xlim, col=adhoc.palette[as.factor(group)],  pch=adhoc.pch[as.factor(group)], main=main,  ylab="Expression", lwd=lwd, cex.main=adhoc.legendcex, xaxt="n", yaxt="n")
+  plot(y, ylim=ylim, xlim=xlim, col=adhoc.palette[as.factor(group)],  pch=adhoc.pch[as.factor(group)], main=main,  ylab=NA, xlab=NA, lwd=lwd, cex.main=adhoc.legendcex, xaxt="n", yaxt="n")
   
   # batch separator
   if(!is.null(batch))    
@@ -116,7 +116,7 @@ estimatesboxesonly = function(y, group, batch, ylim=NULL, main="", lwd=1, usecol
     adhoc.palette =c("red", "blue", "brown", "cyan")
   
   groupnames=factor(unique(group))
-  plot( as.numeric(unique(group)),  xlim=c(1, length(unique(group))+1) , ylim=ylim, type="n", xaxt="n", yaxt="n", main=main, cex.main=adhoc.legendcex)
+  plot( as.numeric(unique(group)),  xlim=c(1, length(unique(group))+1) , ylim=ylim, type="n", xaxt="n", yaxt="n", main=main, cex.main=adhoc.legendcex, xlab=NA, ylab=NA)
   for(g in groupnames)
   {
     g=factor(g,levels=levels(groupnames))

### 2014-11-18 overviewplot.r

@@ -60,8 +60,8 @@ matrix_batchadjusted2 = removeBatchEffect(matrix_batcheffect[,], batch=as.factor
 
 figfilename = file.path( getwd(), "plots", paste("boxplots" , sep=""))
 figfile = paste( figfilename, ".pdf", sep=""); 
-#pdf(file =figfile, width=6, height=6)
-pdf(file =figfile)
+pdf(file =figfile, width=8, height=4)
+#pdf(file =figfile)
 
 # find the extreme ylims to be used by all plots.. 
 tmpmatrix = rbind(matrix_true[index,], matrix_batcheffect[index,], matrix_batchadjusted[index,], matrix_batchadjusted2[index,])
@@ -71,8 +71,9 @@ ylim = c(ymin, ymax)
 
 #op=par(mfrow=c(3, 2),   xaxt="n", mar=c(1,3,2,1)+0.1)
 source("commonscripts/boxplot_function.r")
-layout(  matrix(c(1,2,0,3,4,5), 2, 3, byrow = TRUE) , widths=c(1,1,0.35) )
-op=par(oma = c(5,4,0,0) + 0.1, mar = c(0,0,2,1) + 0.1)
+#layout(  matrix(c(1,2,0,3,4,5), 2, 3, byrow = TRUE) , widths=c(1,1,0.35) )
+layout(  matrix(c(1,2,3,4,5), 1, 5, byrow = TRUE) , widths=c(1,1,1,1,0.35) )
+op=par(oma = c(5,4,0,0) + 0.1, mar = c(0,0,2,1) + 0.1, xpd=NA)
 plot_one_gene(matrix_true[index,], group=sa$group,  batch=sa$batch,, 
               main=paste("(a) True values"), estimatemethod="CI", ylim=ylim,
               bbh=batchboxheight, bblo = batchboxlowmeanoffsets)
@@ -83,13 +84,13 @@ plot_one_gene(matrix_batcheffect[index,], group=sa$group, batch=sa$batch,
 plot_one_gene(matrix_batchadjusted[index,], group=sa$group, batch=sa$batch,
               main=paste("(c) Mean adjusted values"), estimatemethod="CI", ylim=ylim,
               bbh=batchboxheight, bblo = batchboxlowmeanoffsets)
-axis(side = 2, labels =TRUE)
+#axis(side = 2, labels =TRUE)
 plot_one_gene(matrix_batchadjusted2[index,], group=sa$group, batch=sa$batch,
[…]

### 2014-11-18 tex/manuscript.tex
Appendiks fullført, ny figur, diff oppdatert

--- 2014-11-17 tex/manuscript.tex
@@
 A one-way ANOVA checking for group differences will assume that the $F$-statistic follows an $\text{F}_{q,n-q-1}$ distribution.
-However, if the data have been batch adjusted prior to this analysis, the approximate distribution of this $F$-statistic, with correct first two momenta, will be
+However, if the data have been batch adjusted prior to this analysis using two-way ANOVA, the approximate distribution of this $F$-statistic, with correct first two momenta, will be
 \begin{equation}
-F\approx\frac{\tilde q\tilde\sigma^2}{q\sigma^2}\cdot\left(1+\frac{r}{n-q-r-1}\right)\cdot\text{F}_{\tilde q,n-q-r-1}
+F\sim\frac{\sum_{i=1}^q\lambda_i\chi^2_{1}}{\chi^2_{n-q-r-1}}
+\approx\frac{\tilde q\tilde\sigma^2}{q\sigma^2}\cdot\left(1+\frac{r}{n-q-r-1}\right)\cdot\text{F}_{\tilde q,n-q-r-1}
 \end{equation}
-where $\tilde q$ and $\tilde\sigma^2$ are as computed in the appendix and satisfy $\tilde q\tilde\sigma^2\ge q\sigma^2$ and $\tilde q\le q$ with equalities if and only if the groups are perfectly balanced between the batches.
+where the eigenvalues $\lambda_i$, $\tilde q$, and $\tilde\sigma^2$ are as computed in the appendix and satisfy $\tilde q\tilde\sigma^2\ge q\sigma^2$ and $\tilde q\le q$ with equalities if and only if the groups are evenly represented in all batches.
 
@@
 In section~\ref{sec:twogroups}, the ratio $\nu_0/\nu\ge1$ indicates by how much the effective sample size is overestimated when comparing two groups.
-A similar computation, although more complicated, can be made for multiple groups \NB{refer to appendix for details}.
+A similar computation, although more complicated, can be made for multiple groups: see appendix for details covering general linear models.
 However, the two-group assessment should still give a fair idea of the reliability of pairwise group comparisons.
@@
 If there are no batches, i.e. $r=0$, we get the common linear model in which we get a decomposition $Y=\hat\beta_0B_0+\hat\epsilon_0$ with $\hat\gamma_0=Y_0B_0^t(B_0B_0^t)^{-1}$, where
-$|\hat\epsilon|^2\sim\sigma^2\chi_{n-p-q}$ and $|(\hat\beta_0-\beta_0)C_0|^2\sim\sigma^2\chi_q$, yielding the common
+$|\hat\epsilon|^2\sim\sigma^2\chi^2_{n-p-q}$ and $|(\hat\beta_0-\beta_0)C_0|^2\sim\sigma^2\chi^2_q$, yielding the common
 $F$-statistic
@@
 \end{equation}
-If we believe our data are  ``batch effect free'', this is the model we would use, and the distribution of the $F$-statistic we would assume.
+If we believe our data are  ``batch effect free'', and therefore do not include batch in the model, this is the distribution of the $F$-statistic we would assume.
 
 We define the batch adjusted data $\tilde Y_0=Y_0-\hat\gamma_0C_0$ using the estimated batch effect $\hat\gamma_0$ from the full model.
-If we let $X_0=[B_0^t\mid C_0^t]^t$ be the full design matrix after elimination of covariates $A$, the parameter estimates are $[\hat\beta_0\mid\hat\gamma_0]=Y_0X_0^t(X_0X_0^t)^{-1}$.
+If we let $X_0=[B_0^t| C_0^t]^t$ be the full design matrix after elimination of covariates $A$, the parameter estimates are $[\hat\beta_0|\hat\gamma_0]=Y_0X_0^t(X_0X_0^t)^{-1}$.
 The linear decomposition $Y_0=\hat\beta_0B_0+\hat\gamma_0C_0+\hat\epsilon_0$ has $|\hat\epsilon_0|^2\sim\sigma^2\chi^2_{n-p-q-r}$ and
@@
 
-Solving $[\hat\beta_0\mid\hat\gamma_0]=Y_0X_0^t(X_0X_0^t)^{-1}$ and using
+Solving $[\hat\beta_0|\hat\gamma_0]=Y_0X_0^t(X_0X_0^t)^{-1}$ and using
 \begin{equation}
-(X_0X_0^t)^{-1}
-\begin{pmatrix}B_0\\0\end{pmatrix}
-=
-\begin{pmatrix}P\\-(C_0C_0^t)^{-1}C_0B_0^tP\end{pmatrix} B_0
+X_0^t(X_0X_0^t)^{-1} \begin{bmatrix}B_0\\0\end{bmatrix}
+=[B_0^t|C_0^t] \begin{bmatrix}P\\-(C_0C_0^t)^{-1}C_0B_0^tP\end{bmatrix} B_0
+=\rho_{C_0}B_0^tPB_0
 \end{equation}
-with $P=(B_0\rho_{C_0}B_0^t)^{-1}$ and $\rho_{C_0}=I-C_0(C_0C_0^t)^{-1}C_0$
-yields
+with $P=(B_0\rho_{C_0}B_0^t)^{-1}$, and $\rho_{W}=I-W^t(WW^t)^{-1}W$ for any $k\times n$ matrix with $k\le n$ defining the projection to the $n-k$-hyperplane perpendicular to the $k$-hyperplane spanned by $W$, yields
 \begin{equation}
-(\hat\beta_0-\beta_0)B_0=\epsilon_0\rho_{C_0}B_0^tPB_0\sim\text{N}(0,\sigma^2L_0)
-\quad\text{with}\quad
-L_0=B_0^tPB_0\rho_{C_0}B_0^tPB_0
+(\hat\beta_0-\beta_0)B_0=\epsilon_0\rho_{C_0}B_0^tPB_0\sim\text{N}(0,\sigma^2L)
+\quad\text{where}\quad
+L=B_0^tPB_0
 \end{equation}
 since $\epsilon_0\sim\text{N}(0,\sigma^2I)$.
-Let $\lambda_1,\ldots,\lambda_q\ge1$ be the non-zero eigenvalues of $L_0$: this holds since $L_0$ is a symmetric rank $q$ matrix \NB{complete the proof of this}.
+Let $\lambda_1,\ldots,\lambda_q\ge1$ be the non-zero eigenvalues of $L$: this holds since $L$ is a symmetric rank $q$ matrix,  has the same eigenvalues as $PB_0B_0^t$ the inverse of which, $(B_0B_0^t)^{-1}B_0\rho_{C}B_0^t$, has eigenvalues in $(0,1)$ (not 0 since $B_0$ and $C_0$ are linearly independent).
 Then,
@@
 \quad\text{where}\quad
-\frac{\tilde q\tilde\sigma^2}{\sigma^2}=\sum_{i=1}^q\lambda_i=\text{tr}(L_0),\,
-\frac{\tilde q\tilde\sigma^4}{\sigma^4}=\sum_{i=1}^q\lambda_i^2=\text{tr}(L_0^2),
+\frac{\tilde q\tilde\sigma^2}{\sigma^2}=\sum_{i=1}^q\lambda_i=\text{tr}(L),\,
+\frac{\tilde q\tilde\sigma^4}{\sigma^4}=\sum_{i=1}^q\lambda_i^2=\text{tr}(L^2),
 \end{equation}
@@
 We would like to express $\tilde\sigma^2$ and $\tilde q$ terms of the original design matrices.
-The rotated design matrices were defined as $B_0=BU_0$, etc., using the rotation $U=[U_0\mid U']$ so that $AU_0=0$ while $AU'$ is non-singular.
-Hence, it follows that $V_0V_0^t=\rho_A=1-A^t(AA^t)^{-1}A$ which is the projection onto the $n-p$-hyperplane perpendicular to the $p$-hyperplane spanned by $A$.
+The rotated design matrices were defined as $B_0=BU_0$, etc., using the rotation $U=[U_0|U']$ so that $AU_0=0$ while $AU'$ is non-singular.
+Hence, it follows that $V_0V_0^t=\rho_A=1-A^t(AA^t)^{-1}A$.
 This allows us to express
 \begin{equation}
-L_0=U_0^tLU_0
+L=U_0^tB^tPBU_0
 \quad\text{where}\quad
-L=
+P=(BU_0\rho_{C_0}U_0^tB^t)^{-1}=(B\rho_{A,C}B^t)^{-1}
 \end{equation}
+and $\rho_{A,C}=\rho_{[A^t|C^t]^t}$ is the projection to the $n-p-r$-hyperplane perpendicular to the $p+r$-hyperplane spanned by $A$ and $C$. We can then compute
+\begin{equation}
+\text{tr}(L)=\text{tr}(M),\,
+\text{tr}(L^2)=\text{tr}(M^2)
+\quad\text{where}\quad
+M=BU_0U_0^tB^tP=B\rho_A B^t(B\rho_{A,C}B^t)^{-1}
+\end{equation}
+and where $M$ and $L$ have the same eigenvalues.
+Notice that if $B$ is orthogonal to $A$, which can be achieved by replacing $B$ with $B-VA$ for some matrix $V$, then $B\rho_A=B$.
+If $C$ is also made orthogonal to $A$, then $\rho_{A,C}$ can be replaced by $\rho_C$.
 
@@
 \begin{figure}[!p]
-\centering\includegraphics[width=13cm]{Fig/boxplots.pdf}
+\centering\includegraphics[width=8cm]{Fig/boxplots.pdf}
+%\centering\includegraphics[width=13cm]{Fig/boxplots_old.pdf}% For the old version
 %\centering\includegraphics[width=13cm]{Fig/boxplots.png}

### 2014-11-19 tex/manuscript.tex
Oppdatert manus (og diff)
Figur 1 og figurtekst må oppdateres, ellers tror jeg det meste skulle
være på plass nå.

@@
 Unfortunately, this frequently used and recommended approach may systematically induce incorrect group differences in downstream analyses when groups are distributed between the batches in an unbalanced manner.
-\COMMENT{The scientific community seems to be largely unaware of how this approach may lead to false discoveries.}{Har moderert påstanden noe, jfr.
-reviewer~2, pt.~6.}
+The scientific community seems to be largely unaware of how this approach may lead to false discoveries.
 }
+microarrays, data normalisation, batch effect, ANOVA, linear models
 \end{abstract}
@@
 The detailed calculations for general linear models are presented in the appendix.
-Here, we summarise the results for two-way ANOVA based batch adjustments based on \eqref{eq:model} when there are $m$ batches, $M$ study groups, and $n_{ij}$ samples in batch $i$ and group $j$.
+Here, we summarise the results for two-way ANOVA based batch adjustments based on \eqref{eq:model} when there are $m$ batches, $M$ study groups, and $n$ samples with $n_{ij}$ samples in batch $i$ and group $j$.
+Assuming the two-way ANOVA model for batch and group effects is non-degenerate, the degrees of freedom in the ANOVA model as described in the appendix will be $p=1$ for the intercept, $q=M-1$ for the groups, and $r=m-1$ for the batches.
 
-A one-way ANOVA checking for group differences will assume that the $F$-statistic follows an $\text{F}_{q,n-q-1}$ distribution.
-However, if the data have been batch adjusted prior to this analysis using two-way ANOVA, the approximate distribution of this $F$-statistic, with correct first two momenta, will be
+A one-way ANOVA checking for group differences will assume that the $F$-statistic follows an $\text{F}_{M-1,n-M}$ distribution as group differences have $M-1$ degrees of freedom, while there $n-M$ degrees of freedom used to estimate the variance within groups.
+
+However, if the data have been batches adjusted prior to this analysis using two-way ANOVA, the approximate distribution of this $F$-statistic, with correct first two momenta, will be
 \begin{equation}
-F\sim\frac{\sum_{i=1}^q\lambda_i\chi^2_{1}}{\chi^2_{n-q-r-1}}
-\approx\frac{\tilde q\tilde\sigma^2}{q\sigma^2}\cdot\left(1+\frac{r}{n-q-r-1}\right)\cdot\text{F}_{\tilde q,n-q-r-1}
+F\sim\frac{\sum_{i=1}^{M-1}\lambda_i\chi^2_{1}}{\chi^2_{n-M-m+1}}
+\approx\frac{\tilde q\tilde\sigma^2}{(M-1)\sigma^2}\cdot\left(1+\frac{m-1}{n-M-m+1}\right)\cdot\text{F}_{\tilde q,n-M-m+1}
 \end{equation}
-where the eigenvalues $\lambda_i$, $\tilde q$, and $\tilde\sigma^2$ are as computed in the appendix and satisfy $\tilde q\tilde\sigma^2\ge q\sigma^2$ and $\tilde q\le q$ with equalities if and only if the groups are evenly represented in all batches.
+where the eigenvalues $\lambda_i$, $\tilde q$, and $\tilde\sigma^2$ are as computed in the appendix and satisfy $\tilde q\tilde\sigma^2\ge (M-1)\sigma^2$ and $\tilde q\le M-1$ with equalities if and only if the groups are evenly represented in all batches.
 
 The two factors both cause a bias in the $F$-statistic.
-The first factor, $\tilde q\tilde\sigma^2/q\sigma^2$, captures the unbalancedness of the design but is otherwise independent of the sample size.
-The second factor, $q-r/(n-q-r-1)$ captures loss of variance due to the centering of the data, but is primarily a problem when there are few samples in each batch.
-Finally, the degrees of freedom, $\tilde q$ and $n-q-r-1$, of the F-distribution are smaller than assumed by the $\text{F}_{q,n-q-1}$ distribution, and so the variance is larger.
+The first factor, $\tilde q\tilde\sigma^2/(M-1)\sigma^2$, captures the unbalancedness of the design, but is otherwise independent of the sample size.
+The second factor, $1-(m-1)/(n-M-m+1)$, is due to loss of variance from centering of the data, which can be influential when there are few samples in each batch.
+Finally, the degrees of freedom, $\tilde q$ and $n-M-m+1$, of the F-distribution are smaller than assumed by the $\text{F}_{M-1,n-M}$ distribution, and so the variance is larger.
 
 As the sample size increases, assuming the distribution between groups and batches remain unchanged, the effect of the second factor and the second degree of freedom will diminish.
-However, the factor $\tilde q\tilde\sigma^2/q\sigma^2$ will remain, as will the reduced degree of freedom $\tilde q$.
+However, the factor $\tilde q\tilde\sigma^2/(M-1)\sigma^2$ will remain, as will the reduced degree of freedom $\tilde q$.
 
@@
 Science not only depends on determining sound methods and finding reliable results, but also on communicating sufficient detail to allow readers to assess the extent to which findings can be trusted.
-Failing to do so may lead cause methodological problems to pass undetected, but also to cast doubt on sound results.
+Failing to do so may cause methodological problems to pass undetected, but also to cast doubt on sound results.
 An increased focus on making research reproducible is therefore of great importance.

### 2014-11-19 overviewplot.r
plot

@@ -24,7 +24,7 @@ sa = createsampleannotation(  list(c(25,5,0), c(0,20,50)), as.factors=TRUE) #  3
 
 
 #sa = createsampleannotation(  list(c(15,0), c(2,2), c(0,15)), as.factors=TRUE)
-matrix_true = matrix(rnorm(ngenes * nrow(sa), mean=5, sd=2), nrow=ngenes, ncol=nrow(sa))
+matrix_true = matrix(rnorm(ngenes * nrow(sa), mean=10, sd=2), nrow=ngenes, ncol=nrow(sa))
 
 # add a group effect
 matrix_true[,sa$group==1] = matrix_true[,sa$group==1]+4

### 2014-11-20 tex/manuscript.tex
Kun mindre rettelser

@@
 \end{equation}
-from which it follows that $\tilde\sigma^2\ge\sigma^2$, $\tilde q\tilde\sigma^2\ge q\sigma^2$, and $\tilde q\le q$, where equality holds if and only if all $\lambda_i=1$, which happens when $B_0$ and $C_0$ are linearly independent, i.e.
+from which it follows that $\tilde\sigma^2\ge\sigma^2$, $\tilde q\tilde\sigma^2\ge q\sigma^2$, and $\tilde q\le q$, and where equality holds if and only if all $\lambda_i=1$, which happens when $B_0$ and $C_0$ are orthogonal, i.e.
 $B_0C_0^t=0$.
@@
 \begin{equation}
-\text{tr}(L)=\text{tr}(M),\,
-\text{tr}(L^2)=\text{tr}(M^2)
-\quad\text{where}\quad
-M=BU_0U_0^tB^tP=B\rho_A B^t(B\rho_{A,C}B^t)^{-1}
+\frac{\tilde q\tilde\sigma^2}{\sigma^2}=\text{tr}(L)=\text{tr}(M),\,
+\frac{\tilde q\tilde\sigma^4}{\sigma^4}=\text{tr}(L^2)=\text{tr}(M^2)
+\quad\text{with}\quad
+M=B\rho_A B^t(B\rho_{A,C}B^t)^{-1}
 \end{equation}
-and where $M$ and $L$ have the same eigenvalues.
+since $M=BU_0U_0^tB^tP$ and $L=U_0^tB^tPBU_0$ have the same non-zero eigenvalues.
 Notice that if $B$ is orthogonal to $A$, which can be achieved by replacing $B$ with $B-VA$ for some matrix $V$, then $B\rho_A=B$.

### 2014-11-20 tex/manuscript.tex
rettinger

@@
 
-A simple removal of batch effects can be achieved by subtracting the mean of the measurements in one batch from all measurements in that batch, i.e mean adjustment or one-way ANOVA adjustment as implemented in the method \texttt{pamr.batchadjust} from the \texttt{pamr} package in R.
-When the batch--group design is balanced, mean-adjustment will remove most, but not necessarily all, variance attributed to batch and leave the between group variance, thus increasing the statistical power.
+A simple removal of batch effects can be achieved by subtracting the mean of the measurements in one batch from all measurements in that batch, i.e zero-centering or one-way ANOVA adjustment as implemented in the method \texttt{pamr.batchadjust} from the \texttt{pamr} package in R.
+When the batch--group design is balanced, zero-centering will remove most, but not necessarily all, variance attributed to batch and leave the between group variance, thus increasing the statistical power.
 However, when the batch--group design is unbalanced, batch differences will in part be influenced by group differences, and thus batch correction will reduce group differences and thereby reduce the statistical power.
@@
 Based on the actual use of batch adjustment methods like ComBat by the authors and others, ourselves included, we suspect that adjusted data are commonly treated as ``batch effect free'' in subsequent analyses, and as a consequence, confidence in group effects has been overestimated and false results reported.
-The commercial software Partek \CITE{citation or reference needed?}, and the R packages \texttt{limma} (\citealp{Smyth2003}) and \texttt{ber} (\citealp{Giordan2013}) also offer batch adjustment with covariates.
-Seeing that \texttt{ber} is sparsly used, and \texttt{Partek} and the relevant method ``removeBatchEffect'' in \texttt{limma} note that the this is not intended used prior to linear modeling, we choose to use the more popular and well documented ComBat to demonstrate the problem.
+The commercial software Partek Genomics Suite, and the R packages \texttt{limma} (\citealp{Smyth2003}) and \texttt{ber} (\citealp{Giordan2013}) also offer batch adjustment with covariates.
+Seeing that \texttt{ber} is sparsly used, and \texttt{Partek} and the relevant method ``removeBatchEffect'' in \texttt{limma} note that this is not intended used prior to linear modeling, we choose to use the more popular and well documented ComBat to demonstrate the problem.
 We note that the \texttt{sva} are now being updated to address these concerns.
@@
 $\tilde{Y}^{\textrm{avg}}_{ijr}=\tilde{Y}^0_{ijr}+\bar{Y}$.
-When comparing groups, the common value $\bar{Y}$ has no effect, and so this is equivalent to zero-centring each batch.
+When comparing groups, the common value $\bar{Y}$ has no effect, and so this mean-centring is equivalent to zero-centring each batch.
 If the groups are unevenly represented in the different batches, the batch average $\bar{Y}_i$ will tend to capture, through $\bar\beta_i$, group differences, as well as batch effects:
@@
 The undesired consequences of preserving group effects when correcting for batch effect is readily illustrated with a sanity check using random numbers.
-The documentation accompanying the \texttt{sva} library has an executable example demonstrating how to adjust a data set with ComBat, followed by an F-test.
+The documentation accompanying the \texttt{sva} library (v3.8.0) has an executable example demonstrating how to adjust a data set with ComBat, followed by an F-test.
 Replacing the real data with random numbers from a standard normal distribution, $\text{N}(0,1)$, but otherwise following the instructions, will generate the p-value distribution shown in Figure~\ref{fig:sanity}a.
@@
 The size of this factor depends on how unbalanced the group--batch design is, rather than on the sample size in itself.
-Thus, increasing the sample size will not reduce the problem.
+Thus, increasing the sample size will not reduce the problem as shown in Figure~\ref{fig:sanity}c.
 
@@
 In an experiment described in \citet{Towfic2014}, the effect of glatiramer acetate (a medicine for multiple sclerosis) was compared to the effect of a generic drug.
-Cells were treated with glatiramer acetate (34 samples), the generic drug (11 samples), or one of 14 other treatments (64 samples), and mRNA was measured using microarrays.
+Cells were treated with glatiramer acetate (34 samples), the generic drug (11 samples), or one of 14 other treatments (64 samples), and mRNA expression was measured using microarrays.
 A batch effect correlating to the chip (Illumina WG-6\_V2, six samples per chip, 17 chips in total) was observed and adjusted for with ComBat using treatment as a covariate.
@@
 Unfortunately, the batch--treatment design was highly unbalanced, with several batches having only one of the main treatments of interest.
-We re-analyzed the cited data (GEO: GSE40566) without using ComBat, but instead blocked for batch effect in \texttt{limma}, detecting only 9 differentially expressed genes at $\text{FDR}<0.05$.
-The sanity check  outlined above with random numbers was also carried out.
+We re-analyzed the cited data (GEO: GSE40566) without using ComBat, but instead blocked for batch effect in \texttt{limma}, detecting only 11 differentially expressed genes at $\text{FDR}<0.05$.
+The sanity check using random numbers, as described in section 3.1, was also carried out.
 The distribution of p-values for different settings are shown in Figure~\ref{fig:comparison}a.
@@
 Then, we analysed their data without batch adjustment in ComBat, but blocking for batch in \texttt{limma}, resulting in 377 probes ($q<0.05$).
-In addition, the sanity checks outlined above were performed.
+In addition, the sanity check using random numbers was performed.
 The distribution of P-values for different settings are shown in Figure~\ref{fig:comparison}b.
@@
 Later, when ComBat was incorporated into the \texttt{sva} package (\citealp{Leek2012}), specification of a covariate model became required: batch adjustment without covariates was no longer a default behaviour.
-Although specification of a null model was still possible, the help file specifies that covariates should contain ``outcome of interest and other covariates besides batch''.
+Although specification of an empty model was still possible, the help file specifies that covariates should contain ``outcome of interest and other covariates besides batch''.
 Thus, we expect this usage to have grown more common over time.
@@
 
-These problems are not specific to ComBat, but are shared by other batch adjustment methods: indeed, the empirical Bayes approach used by ComBat  will dampen the batch adjustments and may thus reduce the problem slightly relative to a more direct two-way ANOVA approach.
+Problems with batch adjusting unbalanced data sets are not specific to ComBat, but are shared by other batch adjustment methods: indeed, the empirical Bayes approach used by ComBat  will dampen the batch adjustments and may thus reduce the problem slightly relative to a more direct two-way ANOVA approach.
 Although some of the tools warn against performing subsequent analyses on batch adjusted data, these warnings are often somewhat vague and seem to be primarily concerned about the degrees of freedom in the data set.
@@
 If this is not possible, batch correction using outcome as covariate should only be performed with great caution, and downstream confidence estimates must be treated with suspicion.
+
+We did consider if batch adjustment could be modified so as to balance the need for removing batch effects with avoidance of introducing false group differences which would be somewhere between batch centering and two-way ANOVA based adjustment.
+For two-group comparisons, the batch centering will remove some of the group effects and thus underestimate group differences, while two-way ANOVA based batch adjustment will add random, but unbiased, group effects.
+Somewhere in between is a balance for which these two balance out.
+However, the applicability of this approach is very limited and potentially risky since the degree of moderation of the batch adjustment will depend on the subsequent comparison to be made.
+
+Alternatively, one may treat the results more like an ordered list of candidates, with the most likely true positives on top, de-emphasizing the somewhat deflated $p$-values.
+This would, however, be hypothesis generating, rather than hypothesis testing.
+While investigators should always assess the extent to which findings make biological or clinical sense, this is particularly true when the statistical assessment may be unreliable.
 
@@
 However, none of these solutions are ideal.
-
-We did consider if batch adjustment could be modified so as to balance the need for removing batch effects with avoidance of introducing false group differences which would be somewhere between batch centering and two-way ANOVA based adjustment.
-For two-group comparisons, the batch centering will remove some of the group effects and thus underestimate group differences, while two-way ANOVA based batch adjustment will add random, but unbiased, group effects.
-Somewhere in between is a balance for which these two balance out.
-However, the applicability of this approach is very limited and potentially risky since the degree of moderation of the batch adjustment will depend on the subsequent comparison to be made.
-
-Alternatively, one may treat the results more like an ordered list of candidates, with the most likely true positives on top, de-emphasizing the somewhat deflated $p$-values.
-This would, however, be hypothesis generating, rather than hypothesis testing.
-While investigators should always assess the extent to which findings make biological or clinical sense, this is particularly true when the statistical assessment may be unreliable.
 
@@
 
-We are grateful to Geir Kjetil Sandve for useful suggestions and discussions.
+We would like to thank Ben Zeskind and coauthors of \citet{Towfic2014} for their responsiveness and sharing scripts and data.
+We are also grateful to Geir Kjetil Sandve for useful suggestions and discussions.
 
-This work was supported by the EUROCAN platform (VN) and the MetAction project (EAR).
+This work was supported by the EUROCAN platform (VN, FP7/2007-2013, grant 260791) and the MetAction project (EAR).
 
@@
 
-\begin{figure}[!p]
-\begin{center}$
-\begin{array}{cc}
-\includegraphics[width=6.5cm]{Fig/boxplots_v3_meanadjusted.pdf} &
-\includegraphics[width=6.5cm]{Fig/boxplots_v3_anovaadjusted.pdf}
-\end{array}$
-\end{center}
-\caption
-{
-We simulated expression of one gene from two and three study groups.
-These two(a-d and e-h) cases, designs and effects, were selected to illustrate the spurious effects that may arise from different batch adjustments.
-The Y-axis represents the log expression values.
-while the X-axis is used to separate the samples.
-Circles, triangles, and crosses indicate sample measurements from separate groups.
-Horizontal dashed lines indicates the mean of the expression values for that groups samples.
-Vertical dotted lines separates batches with the batch numbers on the lower half of the plot.
-To the far right of the plots, a typical differential expression test is performed with estimated fold change and a p-value.
-The mean expression for each group and 95\% confidence interval is estimated and depicted as boxes.
-a) Simulated expression of one gene in 26 samples from 3 groups (circles, triangles and crosses).
-The fold change and p-value presented are from the circles vs.
-triangles comparison and calculated with a one-way ANOVA.
-The confidence intervals indicate that the crosses are different from the other two groups.
-b) The values from (a) is divided into 2 batches with batch 1 containing circles and triangles, and batch 2 containing triangles and crosses.
-Batch effects were added, -1 for batch 1 and +1 for batch 2.
-The ANOVA test now results in a p<.01 between the circles and triangles.
-This plot illustrates the problem of disregarding batch effects.
-c) The values in (b) were batch adjusted using the mean adjustment method.
-In order to make the means of the two batches equal, some of the group-effect for the crosses were transferred over to the triangels within the same batch.
-As a result the ANOVA test now indicate a difference between the circle-group and the triangle-group (p<.01).
-d) The same values as in (b) are shown and used in a two-way ANOVA with group and batch included in the model.
-The estimated fold change and corresponding p-value is in effect calculated from the samples in batch 1.
-e) Simulated expression of one gene in 20 samples from two groups (circles and triangles). The t-test indicates no significant group differences(p=.82).
-f) The values from (e) are divided into 3 batches in an unbalanced way, with only batch 2 containing samples from both groups.
-Batch effects were added, +1 for batch 1 and -1 for batch 3.
-The t-test now results in p<.01.
-This plot illustrates the problem with disregarding batch effects.
-g) The values in (f) were batch effect adjusted while retaining the group difference (using removeBatchEffects), thus creating a ``batch effect free'' data set.
-Notice that all the values in the "circle only" batch 1 are adjusted to make the group mean coincide with the mean of the two circle-samples in batch 2.
-The triangle-samples in batch 3 are adjusted similarly to match the mean of the two triangle-sampels in batch 2.The t-test now results in a significant group difference (p=.01) and a fold change equal to a fold change calculated from the batch 2 samples only.
-Thus, the fold change is in effect estimated from only 2+2 samples, but the t-test assumes 10+10 samples with the low p-value as a result.
-h) The same values as in (f) are shown and used in a two-way ANOVA with group and batch included in the model.
-The estimated fold change is again in effect only based on samples from batch 2 and is exactly the same as for (g), but the p-value is much higher and the confidence intervals are more appropriate.
-}
-\label{fig:boxplots_v3}
-\end{figure}
-
 \end{document}

### 2014-11-20 commonscripts/boxplot_function.r

@@ -138,7 +138,7 @@ estimatesboxesonly = function(y, group, batch, ylim=NULL, main="", lwd=1, usecol
     rect(xleft, ybottom, xright, ytop, border=adjustcolor(adhoc.palette[g], linealpha), lwd=lwd, 
          lty="solid", density=-1, col=adjustcolor(adhoc.palette[g],boxalpha))
     lines(c(xleft,xright), c(m,m), col=adjustcolor(adhoc.palette[g],linealpha), lwd=lwd)
-    points(x=(xleft+xright)/2, y=ytop+( (ylim[2]-ylim[1])/30), pch=adhoc.pch[g], cex=adhoc.legendcex, lwd=lwd, col=adhoc.palette[g])
+    #points(x=(xleft+xright)/2, y=ytop+( (ylim[2]-ylim[1])/30), pch=adhoc.pch[g], cex=adhoc.legendcex, lwd=lwd, col=adhoc.palette[g])
 
   }
 }

### 2014-11-20 overviewplot.r

@@ -12,7 +12,7 @@ library(lsmeans)
 
 #setseed=100
 #setseed = setseed+1
-
+#set.seed(setseed)
 set.seed(139)
 ngenes = 1000
 index=1
@@ -82,7 +82,7 @@ plot_one_gene(matrix_batcheffect[index,], group=sa$group, batch=sa$batch,
               main=paste("(b) Batch affected values"), estimatemethod="CI", ylim=ylim,
               bbh=batchboxheight, bblo = batchboxlowmeanoffsets)
 plot_one_gene(matrix_batchadjusted[index,], group=sa$group, batch=sa$batch,
-              main=paste("(c) Mean adjusted values"), estimatemethod="CI", ylim=ylim,
+              main=paste("(c) Mean centred values"), estimatemethod="CI", ylim=ylim,
               bbh=batchboxheight, bblo = batchboxlowmeanoffsets)
 #axis(side = 2, labels =TRUE)
 plot_one_gene(matrix_batchadjusted2[index,], group=sa$group, batch=sa$batch,

### 2014-11-20 tex/manuscript.tex
rettelser

@@
 The commercial software Partek Genomics Suite, and the R packages \texttt{limma} (\citealp{Smyth2003}) and \texttt{ber} (\citealp{Giordan2013}) also offer batch adjustment with covariates.
-Seeing that \texttt{ber} is sparsly used, and \texttt{Partek} and the relevant method ``removeBatchEffect'' in \texttt{limma} note that this is not intended used prior to linear modeling, we choose to use the more popular and well documented ComBat to demonstrate the problem.
+Seeing that \texttt{ber} is sparsely used, and \texttt{Partek} and the relevant method ``removeBatchEffect'' in \texttt{limma} note that this is not intended used prior to linear modeling, we choose to use the more popular and well documented ComBat to demonstrate the problem.
 We note that the \texttt{sva} are now being updated to address these concerns.

### 2014-11-21 tex/manuscript.tex
Oppdatert Figure 1 samt diskusjon
Diff mot versjon fra 17. november.

@@
 
-Problems with batch adjusting unbalanced data sets are not specific to ComBat, but are shared by other batch adjustment methods: indeed, the empirical Bayes approach used by ComBat  will dampen the batch adjustments and may thus reduce the problem slightly relative to a more direct two-way ANOVA approach.
+Problems with batch adjusting unbalanced data sets are not specific to ComBat, but are shared by other batch adjustment methods: indeed, the empirical Bayes approach used by ComBat  may dampen the batch adjustments and thereby reduce the problem slightly relative to a more direct two-way ANOVA approach.
 Although some of the tools warn against performing subsequent analyses on batch adjusted data, these warnings are often somewhat vague and seem to be primarily concerned about the degrees of freedom in the data set.
 Batch adjustment will reduce the degrees of freedom in the data which may influence the distribution of the test statistics in subsequent analyses, e.g.
-$F$-statistics from ANOVA.
+the variance of the $F$-statistics from ANOVA.
 For small data sets or where there are many small batches this may be a problem, while it would be less of a concern when there are a fair number of samples in each batch.
-However, we find that the main problem is not related to the degrees of freedom, but estimation errors during the batch adjustments which get applied across the data, which can induce group differences in the adjusted data set even in the absence of batch effects.
+However, we find that the main problem is a systematic bias, e.g.
+scaling the $F$-statistic by a factor, induced as estimation errors during the batch adjustments are applied across the data, which can induce group differences in the adjusted data set even in the absence of batch effects.
 
@@
 
-For an investigator facing an unbalanced data set with batch effects, our primary advice would be to adjust for batch inside the statistical test, and avoid the two-step procedure outlined above.
+For an investigator facing an unbalanced data set with batch effects, our primary advice would be to adjust for batch as part of the statistical analysis, and avoid the two-step procedure outlined above.
 If this is not possible, batch correction using outcome as covariate should only be performed with great caution, and downstream confidence estimates must be treated with suspicion.
+Comparing results from batch adjusted data with and without covariates may indicate to what extent the outcome depends on the mode of batch adjustment.
 
-We did consider if batch adjustment could be modified so as to balance the need for removing batch effects with avoidance of introducing false group differences which would be somewhere between batch centering and two-way ANOVA based adjustment.
+We did consider if a partial batch adjustment could be made to balance the need for removing batch effects against the risk of introducing false group differences which would be somewhere between batch centering and two-way ANOVA based adjustment.
 For two-group comparisons, the batch centering will remove some of the group effects and thus underestimate group differences, while two-way ANOVA based batch adjustment will add random, but unbiased, group effects.
-Somewhere in between is a balance for which these two balance out.
-However, the applicability of this approach is very limited and potentially risky since the degree of moderation of the batch adjustment will depend on the subsequent comparison to be made.
+Somewhere in between is a balance for which the effect of these two balance out in the statistical tests.
+However, the applicability of this approach is very limited and potentially risky since the appropriate degree of batch adjustment will depend on the subsequent comparison to be made.
 
@@
 Knowing that adjustment for batch effects while preserving the group difference may lead to varying degree of false results, to what extent can an investigator trust published results where such a method has been applied?
-Essentially, when the batch--group configuration is balanced, or group difference is ignored  (i.e.
-group labels not given as parameters to ComBat), problems related to preserving group differences will not occur.
-For other cases, a re-analysis without using this approach is the most rigorous path.
-However, such a re-analysis is not feasible if the downstream  analysis can not independently adjust for batch effects.
-To reach a reliable result, batch effects need to be handled in some way or another.
-To make matters worse, a re-analysis relies on the availability of the raw data and a description of processing and analysis steps taken in the original work.
-Even when this is available, the necessary statistics and bioinformatics skills and work hours could still be in short supply.
+Essentially, when the batch--group configuration is balanced, batch adjustment will not bias the group difference estimates.
+For unbalanced designs, however, batch effects may act as a confounder, and a re-analysis without relying on the two-step approach is the most rigorous path.
+However, such a re-analysis may be infeasible or impractical, depending on the availability of the raw data and proper method description, the ability of analysis tools to handle batch effects.
+To reach a reliable result, batch effects need to be handled in some way or another, and the necessary statistics and bioinformatics resources.
 
@@
 A similar computation, although more complicated, can be made for multiple groups: see appendix for details covering general linear models.
-However, the two-group assessment should still give a fair idea of the reliability of pairwise group comparisons.
 If the $\nu_0/\nu$ ratio is close to 1, results should still be reliable, while a ratio much larger than 1 will indicate exaggerated confidence.
@@
 \begin{figure}[!p]
-\centering\includegraphics[width=8cm]{Fig/boxplots.pdf}
-%\centering\includegraphics[width=13cm]{Fig/boxplots_old.pdf}% For the old version
-%\centering\includegraphics[width=13cm]{Fig/boxplots.png}
-\caption{We simulated expression of one gene from four study groups unevenly distributed in three batches containing 50+0+0+20, 0+50+0+20 and 0+0+50+20 samples from groups 1 to 4.
-This case, design and effects, was selected to illustrate the spurious effects that may arise from different batch adjustments.
-The Y-axis represent the expression values, while the X-axis is used to visually separate the batches.
-Circles, triangles, and crosses indicate values from each of the three batches, with colours indicating study groups.
-Correspondingly coloured boxes to the right of the measurements show group means with 95\% confidence intervals.
-These give some indication as to which group differences would be found significant.
-a) The ``true'' values measured in a system without batch effects. Groups 1 and 2 have lower means than groups 3 and 4.
-b) The same samples as in a) but measured in a system with batch effects. All the groups seem different.
-c) The measurements in b) are adjusted with batch centering. Group 3 and 4 seem to differ while group 1 and 3 are more similar.
-d) The measurements in b) are adjusted with two-way ANOVA based batch centering (using \texttt{limma}). Group 1 and 2 seem to differ.
+\centering\includegraphics[width=14.5cm]{Fig/boxplots.pdf}
+\caption{We simulated expression of one gene from three study groups unevenly distributed in three batches, designed to illustrate the spurious effects that may arise from different batch adjustments.
+The Y-axis represent the expression values, while the X-axis is used to visually separate the groups and batches.
+Circles, triangles, and crosses indicate values from each of the three groups.
+Surrounding boxes indicate batches.
+For each group, the coloured bar indicates mean and its 95\%~confidence interval, which give some indication as to which group differences would be found significant.
+a) The ``true values'' measured in a system without batch effects. The first two groups are drawn from normal distributions with the same mean and variance, while the third has a lower mean.
+b) Batch effects have been added to the ``true values'', as indicated by the shifting of the batch frames, leading to apparent differences between the first two groups.
+c) The measurements in b) are adjusted by mean-centering each batch. All groups appear significantly different.
+d) The measurements in b) are adjusted with two-way ANOVA based batch centering (using \texttt{limma}). Again, all groups appear to be significantly different.
 e) The least squares estimates of the group means from a two-way ANOVA have the same means as in d), but more appropriate confidence intervals.

### 2014-11-21 tex/manuscript.tex
Oppdatert svarbrev, og små justeringer i manus

@@
 It should be noted that the original analyses by \citet{Towfic2014} differed from our re-analysis in one critical manner which was clarified upon contacting the authors.
-Their analyses used a different preprocessing of the microarray data (GEO: GSE61901) in which two technical replicates of each biological sample was included, corresponding to the two different slots of the bead-microarray.
+Their analyses used a different preprocessing of the microarray data (GEO: GSE61901) which included two technical replicates per biological sample, corresponding to the two different slots of the bead-microarray.
 For analyses where the two slots have been combined to provide one set of expression values per sample, they refer to \citet{Bakshi2013} in which Partek was used for batch adjustment.
@@
 The supporting information of the original ComBat article (\citealp{Johnson2007}) demonstrates the method on cells inhibited for the expression of the TAL1 gene compared to controls on a microarray platform (denoted ``Data set 2'').
-The experiment consist of 30 samples in 3 batches (batch1:6/2, batch2:3/4 and batch3:9/6 treatment/control samples).
+The experiment consist of 30 samples in 3 batches: the number of treatment/control samples are batch1:~6/2, batch2:~3/4, and batch3:~9/6.
 ComBat was applied followed by a T-test, in order to identify differentially expressed genes.

### 2014-11-22 tex/manuscript.tex
feiltelling

@@
 In an experiment described in \citet{Towfic2014}, the effect of glatiramer acetate (a medicine for multiple sclerosis) was compared to the effect of a generic drug.
-Cells were treated with glatiramer acetate (34 samples), the generic drug (11 samples), or one of 14 other treatments (64 samples), and mRNA expression was measured using microarrays.
+Cells were treated with glatiramer acetate (34 samples), the generic drug (11 samples), or one of 14 other treatments (60 samples), and mRNA expression was measured using microarrays.
 A batch effect correlating to the chip (Illumina WG-6\_V2, six samples per chip, 17 chips in total) was observed and adjusted for with ComBat using treatment as a covariate.

### 2014-11-24 tex/manuscript.tex
Lagt inn Eivinds korreksjoner i manus

@@
 The two factors both cause a bias in the $F$-statistic.
-The first factor, $\tilde q\tilde\sigma^2/(M-1)\sigma^2$, captures the unbalancedness of the design, but is otherwise independent of the sample size.
+The first factor, $\tilde q\tilde\sigma^2/(M-1)\sigma^2$, captures the degree of unbalance of the design, but is otherwise independent of the sample size.
 The second factor, $1-(m-1)/(n-M-m+1)$, is due to loss of variance from centering of the data, which can be influential when there are few samples in each batch.
@@
 For unbalanced designs, however, batch effects may act as a confounder, and a re-analysis without relying on the two-step approach is the most rigorous path.
-However, such a re-analysis may be infeasible or impractical, depending on the availability of the raw data and proper method description, the ability of analysis tools to handle batch effects.
+However, such a re-analysis may be infeasible or impractical, depending on the availability of the raw data and proper method description, and on the ability of analysis tools to handle batch effects.
 To reach a reliable result, batch effects need to be handled in some way or another, and the necessary statistics and bioinformatics resources.
